\vspace{24pt}\LARGE \textbf{הפטרות} \normalsize\vspace{-16pt}
\addcontentsline{toc}{chapter}{הפטרות}
\newhaftarah{בראשית}
\label{haft_1}
\haftnote{The Haftara is Isaiah 42:5\verserangechar 43:10. On Erev Rosh \d{H}odesh, some read the Haftara on page \pageref{haft_macharchodesh}, while some read the Parsha’s Haftara even on Erev Rosh \d{H}odesh.}
\setcounter{chap}{42}
\setcounter{verse}{5}
\twoverse{\Roman{chap}:\arabic{verse}}
{כֹּה\maqqaf אָמַ֞ר הָאֵ֣ל \pasek  יְהֹוָ֗ה בּוֹרֵ֤א הַשָּׁמַ֙יִם֙ וְנ֣וֹטֵיהֶ֔ם רֹקַ֥ע הָאָ֖רֶץ וְצֶאֱצָאֶ֑יהָ נֹתֵ֤ן נְשָׁמָה֙ לָעָ֣ם עָלֶ֔יהָ וְר֖וּחַ לַהֹלְכִ֥ים בָּֽהּ׃}
{Thus saith God the \lord, He that created the heavens, and stretched them forth, He that spread forth the earth and that which cometh out of it, He that giveth breath unto the people upon it, And spirit to them that walk therein:}
\twoverse{\arabic{verse}}
{אֲנִ֧י יְהֹוָ֛ה קְרָאתִ֥יךָֽ בְצֶ֖דֶק וְאַחְזֵ֣ק בְּיָדֶ֑ךָ וְאֶצׇּרְךָ֗ וְאֶתֶּנְךָ֛ לִבְרִ֥ית עָ֖ם לְא֥וֹר גּוֹיִֽם׃}
{I the \lord\space have called thee in righteousness, And have taken hold of thy hand, And kept thee, and set thee for a covenant of the people, For a light of the nations;}
\twoverse{\arabic{verse}}
{לִפְקֹ֖חַ עֵינַ֣יִם עִוְר֑וֹת לְהוֹצִ֤יא מִמַּסְגֵּר֙ אַסִּ֔יר מִבֵּ֥ית כֶּ֖לֶא יֹ֥שְׁבֵי חֹֽשֶׁךְ׃}
{To open the blind eyes, To bring out the prisoners from the dungeon, And them that sit in darkness out of the prison-house.}
\twoverse{\arabic{verse}}
{אֲנִ֥י יְהֹוָ֖ה ה֣וּא שְׁמִ֑י וּכְבוֹדִי֙ לְאַחֵ֣ר לֹֽא\maqqaf אֶתֵּ֔ן וּתְהִלָּתִ֖י לַפְּסִילִֽים׃}
{I am the \lord, that is My name; And My glory will I not give to another, Neither My praise to graven images.}
\twoverse{\arabic{verse}}
{הָרִאשֹׁנ֖וֹת הִנֵּה\maqqaf בָ֑אוּ וַֽחֲדָשׁוֹת֙ אֲנִ֣י מַגִּ֔יד בְּטֶ֥רֶם תִּצְמַ֖חְנָה אַשְׁמִ֥יעַ אֶתְכֶֽם׃ \petucha }
{Behold, the former things are come to pass, And new things do I declare; Before they spring forth I tell you of them.}
\twoverse{\arabic{verse}}
{שִׁ֤ירוּ לַֽיהֹוָה֙ שִׁ֣יר חָדָ֔שׁ תְּהִלָּת֖וֹ מִקְצֵ֣ה הָאָ֑רֶץ יוֹרְדֵ֤י הַיָּם֙ וּמְלֹא֔וֹ אִיִּ֖ים וְיֹשְׁבֵיהֶֽם׃}
{Sing unto the \lord\space a new song, And His praise from the end of the earth; Ye that go down to the sea, and all that is therein, The isles, and the inhabitants thereof.}
\twoverse{\arabic{verse}}
{יִשְׂא֤וּ מִדְבָּר֙ וְעָרָ֔יו חֲצֵרִ֖ים תֵּשֵׁ֣ב קֵדָ֑ר יָרֹ֙נּוּ֙ יֹ֣שְׁבֵי סֶ֔לַע מֵרֹ֥אשׁ הָרִ֖ים יִצְוָֽחוּ׃}
{Let the wilderness and the cities thereof lift up their voice, The villages that Kedar doth inhabit; Let the inhabitants of Sela exult, Let them shout from the top of the mountains.}
\twoverse{\arabic{verse}}
{יָשִׂ֥ימוּ לַיהֹוָ֖ה כָּב֑וֹד וּתְהִלָּת֖וֹ בָּאִיִּ֥ים יַגִּֽידוּ׃}
{Let them give glory unto the \lord, And declare His praise in the islands.}
\twoverse{\arabic{verse}}
{יְהֹוָה֙ כַּגִּבּ֣וֹר יֵצֵ֔א כְּאִ֥ישׁ מִלְחָמ֖וֹת יָעִ֣יר קִנְאָ֑ה יָרִ֙יעַ֙ אַף\maqqaf יַצְרִ֔יחַ עַל\maqqaf אֹיְבָ֖יו יִתְגַּבָּֽר׃ \setuma }
{The \lord\space will go forth as a mighty man, He will stir up jealousy like a man of war; He will cry, yea, He will shout aloud, He will prove Himself mighty against His enemies.}
\twoverse{\arabic{verse}}
{הֶחֱשֵׁ֙יתִי֙ מֵֽעוֹלָ֔ם אַחֲרִ֖ישׁ אֶתְאַפָּ֑ק כַּיּוֹלֵדָ֣ה אֶפְעֶ֔ה אֶשֹּׁ֥ם וְאֶשְׁאַ֖ף יָֽחַד׃}
{I have long time held My peace, I have been still, and refrained Myself; Now will I cry like a travailing woman, Gasping and panting at once.}
\twoverse{\arabic{verse}}
{אַחֲרִ֤יב הָרִים֙ וּגְבָע֔וֹת וְכׇל\maqqaf עֶשְׂבָּ֖ם אוֹבִ֑ישׁ וְשַׂמְתִּ֤י נְהָרוֹת֙ לָאִיִּ֔ים וַאֲגַמִּ֖ים אוֹבִֽישׁ׃}
{I will make waste mountains and hills, And dry up all their herbs; And I will make the rivers islands, And will dry up the pools.}
\twoverse{\arabic{verse}}
{וְהוֹלַכְתִּ֣י עִוְרִ֗ים בְּדֶ֙רֶךְ֙ לֹ֣א יָדָ֔עוּ בִּנְתִיב֥וֹת לֹא\maqqaf יָדְע֖וּ אַדְרִיכֵ֑ם אָשִׂים֩ מַחְשָׁ֨ךְ לִפְנֵיהֶ֜ם לָא֗וֹר וּמַֽעֲקַשִּׁים֙ לְמִישׁ֔וֹר אֵ֚לֶּה הַדְּבָרִ֔ים עֲשִׂיתִ֖ם וְלֹ֥א עֲזַבְתִּֽים׃}
{And I will bring the blind by a way that they knew not, In paths that they knew not will I lead them; I will make darkness light before them, and rugged places plain. These things will I do, And I will not leave them undone.}
\twoverse{\arabic{verse}}
{נָסֹ֤גוּ אָחוֹר֙ יֵבֹ֣שׁוּ בֹ֔שֶׁת הַבֹּטְחִ֖ים בַּפָּ֑סֶל הָאֹמְרִ֥ים לְמַסֵּכָ֖ה אַתֶּ֥ם אֱלֹהֵֽינוּ׃ \petucha }
{They shall be turned back, greatly ashamed, That trust in graven images, That say unto molten images: ‘Ye are our gods.’}
\twoverse{\arabic{verse}}
{הַחֵרְשִׁ֖ים שְׁמָ֑עוּ וְהַעִוְרִ֖ים הַבִּ֥יטוּ לִרְאֽוֹת׃}
{Hear, ye deaf, And look, ye blind, that ye may see.}
\twoverse{\arabic{verse}}
{מִ֤י עִוֵּר֙ כִּ֣י אִם\maqqaf עַבְדִּ֔י וְחֵרֵ֖שׁ כְּמַלְאָכִ֣י אֶשְׁלָ֑ח מִ֤י עִוֵּר֙ כִּמְשֻׁלָּ֔ם וְעִוֵּ֖ר כְּעֶ֥בֶד יְהֹוָֽה׃}
{Who is blind, but My servant? Or deaf, as My messenger that I send? Who is blind as he that is wholehearted, And blind as the \lord\textsc{’s}\space servant?}
\twoverse{\arabic{verse}}
{\qk{רָא֥וֹת}{ראית} רַבּ֖וֹת וְלֹ֣א תִשְׁמֹ֑ר פָּק֥וֹחַ אׇזְנַ֖יִם וְלֹ֥א יִשְׁמָֽע׃}
{Seeing many things, thou observest not; Opening the ears, he heareth not.}
\twoverse{\arabic{verse}}
{יְהֹוָ֥ה חָפֵ֖ץ לְמַ֣עַן צִדְק֑וֹ יַגְדִּ֥יל תּוֹרָ֖ה וְיַאְדִּֽיר׃}
{The \lord\space was pleased, for His righteousness’ sake, To make the teaching great and glorious.}
\twoverse{\arabic{verse}}
{וְהוּא֮ עַם\maqqaf בָּז֣וּז וְשָׁסוּי֒ הָפֵ֤חַ בַּחוּרִים֙ כֻּלָּ֔ם וּבְבָתֵּ֥י כְלָאִ֖ים הׇחְבָּ֑אוּ הָי֤וּ לָבַז֙ וְאֵ֣ין מַצִּ֔יל מְשִׁסָּ֖ה וְאֵין\maqqaf אֹמֵ֥ר הָשַֽׁב׃}
{But this is a people robbed and spoiled, They are all of them snared in holes, And they are hid in prison-houses; They are for a prey, and none delivereth, For a spoil, and none saith: ‘Restore.’}
\twoverse{\arabic{verse}}
{מִ֥י בָכֶ֖ם יַאֲזִ֣ין זֹ֑את יַקְשִׁ֥ב וְיִשְׁמַ֖ע לְאָחֽוֹר׃}
{Who among you will give ear to this? Who will hearken and hear for the time to come?}
\twoverse{\arabic{verse}}
{מִֽי\maqqaf נָתַ֨ן \qk{לִמְשִׁסָּ֧ה}{למשוסה} יַעֲקֹ֛ב וְיִשְׂרָאֵ֥ל לְבֹזְזִ֖ים הֲל֣וֹא יְהֹוָ֑ה ז֚וּ חָטָ֣אנוּ ל֔וֹ וְלֹֽא\maqqaf אָב֤וּ בִדְרָכָיו֙ הָל֔וֹךְ וְלֹ֥א שָׁמְע֖וּ בְּתוֹרָתֽוֹ׃}
{Who gave Jacob for a spoil, and Israel to the robbers? Did not the \lord? He against whom we have sinned, And in whose ways they would not walk, Neither were they obedient unto His law.}
\twoverse{\arabic{verse}}
{וַיִּשְׁפֹּ֤ךְ עָלָיו֙ חֵמָ֣ה אַפּ֔וֹ וֶעֱז֖וּז מִלְחָמָ֑ה וַתְּלַהֲטֵ֤הוּ מִסָּבִיב֙ וְלֹ֣א יָדָ֔ע וַתִּבְעַר\maqqaf בּ֖וֹ וְלֹא\maqqaf יָשִׂ֥ים עַל\maqqaf לֵֽב׃}
{Therefore He poured upon him the fury of His anger, And the strength of battle; And it set him on fire round about, yet he knew not, And it burned him, yet he laid it not to heart.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{וְעַתָּ֞ה כֹּֽה\maqqaf אָמַ֤ר יְהֹוָה֙ בֹּרַאֲךָ֣ יַעֲקֹ֔ב וְיֹצֶרְךָ֖ יִשְׂרָאֵ֑ל אַל\maqqaf תִּירָא֙ כִּ֣י גְאַלְתִּ֔יךָ קָרָ֥אתִי בְשִׁמְךָ֖ לִי\maqqaf אָֽתָּה׃}
{But now thus saith the \lord\space that created thee, O Jacob, And He that formed thee, O Israel: Fear not, for I have redeemed thee, I have called thee by thy name, thou art Mine.}
\twoverse{\arabic{verse}}
{כִּֽי\maqqaf תַעֲבֹ֤ר בַּמַּ֙יִם֙ אִתְּךָ\maqqaf אָ֔נִי וּבַנְּהָר֖וֹת לֹ֣א יִשְׁטְפ֑וּךָ כִּֽי\maqqaf תֵלֵ֤ךְ בְּמוֹ\maqqaf אֵשׁ֙ לֹ֣א תִכָּוֶ֔ה וְלֶהָבָ֖ה לֹ֥א תִבְעַר\maqqaf בָּֽךְ׃}
{When thou passest through the waters, I will be with thee, And through the rivers, they shall not overflow thee; When thou walkest through the fire, thou shalt not be burned, Neither shall the flame kindle upon thee.}
\twoverse{\arabic{verse}}
{כִּ֗י אֲנִי֙ יְהֹוָ֣ה אֱלֹהֶ֔יךָ קְד֥וֹשׁ יִשְׂרָאֵ֖ל מוֹשִׁיעֶ֑ךָ נָתַ֤תִּי כׇפְרְךָ֙ מִצְרַ֔יִם כּ֥וּשׁ וּסְבָ֖א תַּחְתֶּֽיךָ׃}
{For I am the \lord\space thy God, The Holy One of Israel, thy Saviour; I have given Egypt as thy ransom, Ethiopia and Seba for thee.}
\twoverse{\arabic{verse}}
{מֵאֲשֶׁ֨ר יָקַ֧רְתָּ בְעֵינַ֛י נִכְבַּ֖דְתָּ וַאֲנִ֣י אֲהַבְתִּ֑יךָ וְאֶתֵּ֤ן אָדָם֙ תַּחְתֶּ֔יךָ וּלְאֻמִּ֖ים תַּ֥חַת נַפְשֶֽׁךָ׃}
{Since thou art precious in My sight, and honourable, and I have loved thee; Therefore will I give men for thee, And peoples for thy life.}
\twoverse{\arabic{verse}}
{אַל\maqqaf תִּירָ֖א כִּ֣י אִתְּךָ\maqqaf אָ֑נִי מִמִּזְרָח֙ אָבִ֣יא זַרְעֶ֔ךָ וּמִֽמַּעֲרָ֖ב אֲקַבְּצֶֽךָּ׃}
{Fear not, for I am with thee; I will bring thy seed from the east, And gather thee from the west;}
\twoverse{\arabic{verse}}
{אֹמַ֤ר לַצָּפוֹן֙ תֵּ֔נִי וּלְתֵימָ֖ן אַל\maqqaf תִּכְלָ֑אִי הָבִ֤יאִי בָנַי֙ מֵרָח֔וֹק וּבְנוֹתַ֖י מִקְצֵ֥ה הָאָֽרֶץ׃}
{I will say to the north: ‘Give up’, And to the south: ‘Keep not back, bring My sons from far, and My daughters from the end of the earth;}
\twoverse{\arabic{verse}}
{כֹּ֚ל הַנִּקְרָ֣א בִשְׁמִ֔י וְלִכְבוֹדִ֖י בְּרָאתִ֑יו יְצַרְתִּ֖יו אַף\maqqaf עֲשִׂיתִֽיו׃}
{Every one that is called by My name, and whom I have created for My glory, I have formed him, yea, I have made him.’}
\twoverse{\arabic{verse}}
{הוֹצִ֥יא עַם\maqqaf עִוֵּ֖ר וְעֵינַ֣יִם יֵ֑שׁ וְחֵרְשִׁ֖ים וְאׇזְנַ֥יִם לָֽמוֹ׃}
{The blind people that have eyes shall be brought forth, and the deaf that have ears.}
\twoverse{\arabic{verse}}
{כׇּֽל\maqqaf הַגּוֹיִ֞ם נִקְבְּצ֣וּ יַחְדָּ֗ו וְיֵאָֽסְפוּ֙ לְאֻמִּ֔ים מִ֤י בָהֶם֙ יַגִּ֣יד זֹ֔את וְרִאשֹׁנ֖וֹת יַשְׁמִיעֻ֑נוּ יִתְּנ֤וּ עֵֽדֵיהֶם֙ וְיִצְדָּ֔קוּ וְיִשְׁמְע֖וּ וְיֹאמְר֥וּ אֱמֶֽת׃}
{All the nations are gathered together, and the peoples are assembled; who among them can declare this, And announce to us former things? Let them bring their witnesses, that they may be justified; And let them hear, and say: ‘It is truth.’}
\twoverse{\arabic{verse}}
{אַתֶּ֤ם עֵדַי֙ נְאֻם\maqqaf יְהֹוָ֔ה וְעַבְדִּ֖י אֲשֶׁ֣ר בָּחָ֑רְתִּי לְמַ֣עַן תֵּ֠דְע֠וּ וְתַאֲמִ֨ינוּ לִ֤י וְתָבִ֙ינוּ֙ כִּֽי\maqqaf אֲנִ֣י ה֔וּא לְפָנַי֙ לֹא\maqqaf נ֣וֹצַר אֵ֔ל וְאַחֲרַ֖י לֹ֥א יִהְיֶֽה׃ \setuma }
{Ye are My witnesses, saith the \lord, and My servant whom I have chosen; that ye may know and believe Me, and understand That I am He; before Me there was no God formed, neither shall any be after Me.}
\newperek
\newhaftarah{נח}
\label{haft_2}
\haftnote{The Haftara is Isaiah 54:1\verserangechar 55:5. On Rosh \d{H}odesh, read the Maftir and Haftara on page \pageref{maft_ShabRCh}.}
\setcounter{chap}{54}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}
{רׇנִּ֥י עֲקָרָ֖ה לֹ֣א יָלָ֑דָה פִּצְחִ֨י רִנָּ֤ה וְצַהֲלִי֙ לֹא\maqqaf חָ֔לָה כִּֽי\maqqaf רַבִּ֧ים בְּֽנֵי\maqqaf שׁוֹמֵמָ֛ה מִבְּנֵ֥י בְעוּלָ֖ה אָמַ֥ר יְהֹוָֽה׃}
{Sing, O barren, thou that didst not bear, Break forth into singing, and cry aloud, thou that didst not travail; For more are the children of the desolate Than the children of the married wife, saith the \lord.}
\twoverse{\arabic{verse}}
{הַרְחִ֣יבִי \legarmeh  מְק֣וֹם אׇהֳלֵ֗ךְ וִֽירִיע֧וֹת מִשְׁכְּנוֹתַ֛יִךְ יַטּ֖וּ אַל\maqqaf תַּחְשֹׂ֑כִי הַאֲרִ֙יכִי֙ מֵיתָרַ֔יִךְ וִיתֵדֹתַ֖יִךְ חַזֵּֽקִי׃}
{Enlarge the place of thy tent, And let them stretch forth the curtains of thy habitations, spare not; Lengthen thy cords, and strengthen thy stakes.}
\twoverse{\arabic{verse}}
{כִּֽי\maqqaf יָמִ֥ין וּשְׂמֹ֖אול תִּפְרֹ֑צִי וְזַרְעֵךְ֙ גּוֹיִ֣ם יִירָ֔שׁ וְעָרִ֥ים נְשַׁמּ֖וֹת יוֹשִֽׁיבוּ׃}
{For thou shalt spread abroad on the right hand and on the left; And thy seed shall possess the nations, And make the desolate cities to be inhabited.}
\twoverse{\arabic{verse}}
{אַל\maqqaf תִּֽירְאִי֙ כִּי\maqqaf לֹ֣א תֵב֔וֹשִׁי וְאַל\maqqaf תִּכָּֽלְמִ֖י כִּ֣י לֹ֣א תַחְפִּ֑ירִי כִּ֣י בֹ֤שֶׁת עֲלוּמַ֙יִךְ֙ תִּשְׁכָּ֔חִי וְחֶרְפַּ֥ת אַלְמְנוּתַ֖יִךְ לֹ֥א תִזְכְּרִי\maqqaf עֽוֹד׃}
{Fear not, for thou shalt not be ashamed. Neither be thou confounded, for thou shalt not be put to shame; For thou shalt forget the shame of thy youth, And the reproach of thy widowhood shalt thou remember no more.}
\twoverse{\arabic{verse}}
{כִּ֤י בֹעֲלַ֙יִךְ֙ עֹשַׂ֔יִךְ יְהֹוָ֥ה צְבָא֖וֹת שְׁמ֑וֹ וְגֹֽאֲלֵךְ֙ קְד֣וֹשׁ יִשְׂרָאֵ֔ל אֱלֹהֵ֥י כׇל\maqqaf הָאָ֖רֶץ יִקָּרֵֽא׃}
{For thy Maker is thy husband, The \lord\space of hosts is His name; And the Holy One of Israel is thy Redeemer, The God of the whole earth shall He be called.}
\twoverse{\arabic{verse}}
{כִּֽי\maqqaf כְאִשָּׁ֧ה עֲזוּבָ֛ה וַעֲצ֥וּבַת ר֖וּחַ קְרָאָ֣ךְ יְהֹוָ֑ה וְאֵ֧שֶׁת נְעוּרִ֛ים כִּ֥י תִמָּאֵ֖ס אָמַ֥ר אֱלֹהָֽיִךְ׃}
{For the \lord\space hath called thee As a wife forsaken and grieved in spirit; And a wife of youth, can she be rejected? Saith thy God.}
\twoverse{\arabic{verse}}
{בְּרֶ֥גַע קָטֹ֖ן עֲזַבְתִּ֑יךְ וּבְרַחֲמִ֥ים גְּדֹלִ֖ים אֲקַבְּצֵֽךְ׃}
{For a small moment have I forsaken thee; But with great compassion will I gather thee.}
\twoverse{\arabic{verse}}
{בְּשֶׁ֣צֶף קֶ֗צֶף הִסְתַּ֨רְתִּי פָנַ֥י רֶ֙גַע֙ מִמֵּ֔ךְ וּבְחֶ֥סֶד עוֹלָ֖ם רִחַמְתִּ֑יךְ אָמַ֥ר גֹּאֲלֵ֖ךְ יְהֹוָֽה׃ \setuma }
{In a little wrath I hid My face from thee for a moment; But with everlasting kindness will I have compassion on thee, Saith the \lord\space thy Redeemer.}
\twoverse{\arabic{verse}}
{כִּֽי\maqqaf מֵ֥י נֹ֙חַ֙ זֹ֣את לִ֔י אֲשֶׁ֣ר נִשְׁבַּ֗עְתִּי מֵעֲבֹ֥ר מֵי\maqqaf נֹ֛חַ ע֖וֹד עַל\maqqaf הָאָ֑רֶץ כֵּ֥ן נִשְׁבַּ֛עְתִּי מִקְּצֹ֥ף עָלַ֖יִךְ וּמִגְּעׇר\maqqaf בָּֽךְ׃}
{For this is as the waters of Noah unto Me; for as I have sworn that the waters of Noah should no more go over the earth, so have I sworn that I would not be wroth with thee, nor rebuke thee.}
\twoverse{\arabic{verse}}
{כִּ֤י הֶהָרִים֙ יָמ֔וּשׁוּ וְהַגְּבָע֖וֹת תְּמוּטֶ֑ינָה וְחַסְדִּ֞י מֵאִתֵּ֣ךְ לֹא\maqqaf יָמ֗וּשׁ וּבְרִ֤ית שְׁלוֹמִי֙ לֹ֣א תָמ֔וּט אָמַ֥ר מְרַחֲמֵ֖ךְ יְהֹוָֽה׃ \setuma }
{For the mountains may depart, and the hills be removed; but My kindness shall not depart from thee, neither shall My covenant of peace be removed, saith the \lord\space that hath compassion on thee.}
\twoverse{\arabic{verse}}
{עֲנִיָּ֥ה סֹעֲרָ֖ה לֹ֣א נֻחָ֑מָה הִנֵּ֨ה אָנֹכִ֜י מַרְבִּ֤יץ בַּפּוּךְ֙ אֲבָנַ֔יִךְ וִֽיסַדְתִּ֖יךְ בַּסַּפִּירִֽים׃}
{O thou afflicted, tossed with tempest, and not comforted, behold, I will set thy stones in fair colours, And lay thy foundations with sapphires.}
\twoverse{\arabic{verse}}
{וְשַׂמְתִּ֤י כַּֽדְכֹד֙ שִׁמְשֹׁתַ֔יִךְ וּשְׁעָרַ֖יִךְ לְאַבְנֵ֣י אֶקְדָּ֑ח וְכׇל\maqqaf גְּבוּלֵ֖ךְ לְאַבְנֵי\maqqaf חֵֽפֶץ׃}
{And I will make thy pinnacles of rubies, and thy gates of carbuncles, and all thy border of precious stones.}
\twoverse{\arabic{verse}}
{וְכׇל\maqqaf בָּנַ֖יִךְ לִמּוּדֵ֣י יְהֹוָ֑ה וְרַ֖ב שְׁל֥וֹם בָּנָֽיִךְ׃}
{And all thy children shall be taught of the \lord; and great shall be the peace of thy children.}
\twoverse{\arabic{verse}}
{בִּצְדָקָ֖ה תִּכּוֹנָ֑נִי רַחֲקִ֤י מֵעֹ֙שֶׁק֙ כִּי\maqqaf לֹ֣א תִירָ֔אִי וּמִ֨מְּחִתָּ֔ה כִּ֥י לֹֽא\maqqaf תִקְרַ֖ב אֵלָֽיִךְ׃}
{In righteousness shalt thou be established; be thou far from oppression, for thou shalt not fear, And from ruin, for it shall not come near thee.}
\twoverse{\arabic{verse}}
{הֵ֣ן גּ֥וֹר יָג֛וּר אֶ֖פֶס מֵאוֹתִ֑י מִי\maqqaf גָ֥ר אִתָּ֖ךְ עָלַ֥יִךְ יִפּֽוֹל׃}
{Behold, they may gather together, but not by Me; Whosoever shall gather together against thee shall fall because of thee.}
\twoverse{\arabic{verse}}
{\qk{הִנֵּ֤ה}{הן} אָנֹכִי֙ בָּרָ֣אתִי חָרָ֔שׁ נֹפֵ֙חַ֙ בְּאֵ֣שׁ פֶּחָ֔ם וּמוֹצִ֥יא כְלִ֖י לְמַעֲשֵׂ֑הוּ וְאָנֹכִ֛י בָּרָ֥אתִי מַשְׁחִ֖ית לְחַבֵּֽל׃}
{Behold, I have created the smith That bloweth the fire of coals, And bringeth forth a weapon for his work; And I have created the waster to destroy.}
\twoverse{\arabic{verse}}
{כׇּל\maqqaf כְּלִ֞י יוּצַ֤ר עָלַ֙יִךְ֙ לֹ֣א יִצְלָ֔ח וְכׇל\maqqaf לָשׁ֛וֹן תָּקוּם\maqqaf אִתָּ֥ךְ לַמִּשְׁפָּ֖ט תַּרְשִׁ֑יעִי זֹ֡את נַחֲלַת֩ עַבְדֵ֨י יְהֹוָ֧ה וְצִדְקָתָ֛ם מֵאִתִּ֖י נְאֻם\maqqaf יְהֹוָֽה׃ \setuma }
{No weapon that is formed against thee shall prosper; And every tongue that shall rise against thee in judgment thou shalt condemn. This is the heritage of the servants of the \lord, And their due reward from Me, saith the \lord.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{ה֤וֹי כׇּל\maqqaf צָמֵא֙ לְכ֣וּ לַמַּ֔יִם וַאֲשֶׁ֥ר אֵֽין\maqqaf ל֖וֹ כָּ֑סֶף לְכ֤וּ שִׁבְרוּ֙ וֶאֱכֹ֔לוּ וּלְכ֣וּ שִׁבְר֗וּ בְּלוֹא\maqqaf כֶ֛סֶף וּבְל֥וֹא מְחִ֖יר יַ֥יִן וְחָלָֽב׃}
{Ho, every one that thirsteth, come ye for water, And he that hath no money; Come ye, buy, and eat; Yea, come, buy wine and milk Without money and without price.}
\twoverse{\arabic{verse}}
{לָ֤מָּה תִשְׁקְלוּ\maqqaf כֶ֙סֶף֙ בְּלוֹא\maqqaf לֶ֔חֶם וִיגִֽיעֲכֶ֖ם בְּל֣וֹא לְשׇׂבְעָ֑ה שִׁמְע֨וּ שָׁמ֤וֹעַ אֵלַי֙ וְאִכְלוּ\maqqaf ט֔וֹב וְתִתְעַנַּ֥ג בַּדֶּ֖שֶׁן נַפְשְׁכֶֽם׃}
{Wherefore do ye spend money for that which is not bread? And your gain for that which satisfieth not? Hearken diligently unto Me, and eat ye that which is good, And let your soul delight itself in fatness.}
\twoverse{\arabic{verse}}
{הַטּ֤וּ אׇזְנְכֶם֙ וּלְכ֣וּ אֵלַ֔י שִׁמְע֖וּ וּתְחִ֣י נַפְשְׁכֶ֑ם וְאֶכְרְתָ֤ה לָכֶם֙ בְּרִ֣ית עוֹלָ֔ם חַֽסְדֵ֥י דָוִ֖ד הַנֶּֽאֱמָנִֽים׃}
{Incline your ear, and come unto Me; Hear, and your soul shall live; And I will make an everlasting covenant with you, Even the sure mercies of David.}
\twoverse{\arabic{verse}}
{הֵ֛ן עֵ֥ד לְאוּמִּ֖ים נְתַתִּ֑יו נָגִ֥יד וּמְצַוֵּ֖ה לְאֻמִּֽים׃}
{Behold, I have given him for a witness to the peoples, A prince and commander to the peoples.}
\twoverse{\arabic{verse}}
{הֵ֣ן גּ֤וֹי לֹֽא\maqqaf תֵדַע֙ תִּקְרָ֔א וְג֥וֹי לֹֽא\maqqaf יְדָע֖וּךָ אֵלֶ֣יךָ יָר֑וּצוּ לְמַ֙עַן֙ יְהֹוָ֣ה אֱלֹהֶ֔יךָ וְלִקְד֥וֹשׁ יִשְׂרָאֵ֖ל כִּ֥י פֵאֲרָֽךְ׃ \setuma }
{Behold, thou shalt call a nation that thou knowest not, And a nation that knew not thee shall run unto thee; Because of the \lord\space thy God, And for the Holy One of Israel, for He hath glorified thee.}
\newperek
\newhaftarah{לך לך}
\label{haft_3}
\haftnote{The Haftara is Isaiah 40:27\verserangechar 41:16. }
\setcounter{chap}{40}
\setcounter{verse}{27}
\twoverse{\Roman{chap}:\arabic{verse}}
{לָ֤מָּה תֹאמַר֙ יַֽעֲקֹ֔ב וּתְדַבֵּ֖ר יִשְׂרָאֵ֑ל נִסְתְּרָ֤ה דַרְכִּי֙ מֵיְהֹוָ֔ה וּמֵאֱלֹהַ֖י מִשְׁפָּטִ֥י יַעֲבֽוֹר׃}
{Why sayest thou, O Jacob, and speakest, O Israel: ‘My way is hid from the \lord, and my right is passed over from my God’?}
\twoverse{\arabic{verse}}
{הֲל֨וֹא יָדַ֜עְתָּ אִם\maqqaf לֹ֣א שָׁמַ֗עְתָּ אֱלֹהֵ֨י עוֹלָ֤ם \pasek  יְהֹוָה֙ בּוֹרֵא֙ קְצ֣וֹת הָאָ֔רֶץ לֹ֥א יִיעַ֖ף וְלֹ֣א יִיגָ֑ע אֵ֥ין חֵ֖קֶר לִתְבוּנָתֽוֹ׃}
{Hast thou not known? hast thou not heard That the everlasting God, the \lord, The Creator of the ends of the earth, Fainteth not, neither is weary? His discernment is past searching out.}
\twoverse{\arabic{verse}}
{נֹתֵ֥ן לַיָּעֵ֖ף כֹּ֑חַ וּלְאֵ֥ין אוֹנִ֖ים עׇצְמָ֥ה יַרְבֶּֽה׃}
{He giveth power to the faint; And to him that hath no might He increaseth strength.}
\twoverse{\arabic{verse}}
{וְיִעֲפ֥וּ נְעָרִ֖ים וְיִגָ֑עוּ וּבַחוּרִ֖ים כָּשׁ֥וֹל יִכָּשֵֽׁלוּ׃}
{Even the youths shall faint and be weary, And the young men shall utterly fall;}
\twoverse{\arabic{verse}}
{וְקוֹיֵ֤ יְהֹוָה֙ יַחֲלִ֣יפוּ כֹ֔חַ יַעֲל֥וּ אֵ֖בֶר כַּנְּשָׁרִ֑ים יָר֙וּצוּ֙ וְלֹ֣א יִיגָ֔עוּ יֵלְכ֖וּ וְלֹ֥א יִיעָֽפוּ׃ \setuma }
{But they that wait for the \lord\space shall renew their strength; They shall mount up with wings as eagles; They shall run, and not be weary; They shall walk, and not faint.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{הַחֲרִ֤ישׁוּ אֵלַי֙ אִיִּ֔ים וּלְאֻמִּ֖ים יַחֲלִ֣יפוּ כֹ֑חַ יִגְּשׁוּ֙ אָ֣ז יְדַבֵּ֔רוּ יַחְדָּ֖ו לַמִּשְׁפָּ֥ט נִקְרָֽבָה׃}
{Keep silence before Me, O islands, And let the peoples renew their strength; Let them draw near, then let them speak; Let us come near together to judgment.}
\twoverse{\arabic{verse}}
{מִ֤י הֵעִיר֙ מִמִּזְרָ֔ח צֶ֖דֶק יִקְרָאֵ֣הוּ לְרַגְל֑וֹ יִתֵּ֨ן לְפָנָ֤יו גּוֹיִם֙ וּמְלָכִ֣ים יַ֔רְדְּ יִתֵּ֤ן כֶּֽעָפָר֙ חַרְבּ֔וֹ כְּקַ֥שׁ נִדָּ֖ף קַשְׁתּֽוֹ׃}
{Who hath raised up one from the east, At whose steps victory attendeth? He giveth nations before him, And maketh him rule over kings; His sword maketh them as the dust, His bow as the driven stubble.}
\twoverse{\arabic{verse}}
{יִרְדְּפֵ֖ם יַעֲב֣וֹר שָׁל֑וֹם אֹ֥רַח בְּרַגְלָ֖יו לֹ֥א יָבֽוֹא׃}
{He pursueth them, and passeth on safely; The way with his feet he treadeth not.}
\twoverse{\arabic{verse}}
{מִֽי\maqqaf פָעַ֣ל וְעָשָׂ֔ה קֹרֵ֥א הַדֹּר֖וֹת מֵרֹ֑אשׁ אֲנִ֤י יְהֹוָה֙ רִאשׁ֔וֹן וְאֶת\maqqaf אַחֲרֹנִ֖ים אֲנִי\maqqaf הֽוּא׃}
{Who hath wrought and done it? He that called the generations from the beginning. I, the \lord, who am the first, And with the last am the same.}
\twoverse{\arabic{verse}}
{רָא֤וּ אִיִּים֙ וְיִרָ֔אוּ קְצ֥וֹת הָאָ֖רֶץ יֶחֱרָ֑דוּ קָרְב֖וּ וַיֶּאֱתָיֽוּן׃}
{The isles saw, and feared; The ends of the earth trembled; They drew near, and came.}
\twoverse{\arabic{verse}}
{אִ֥ישׁ אֶת\maqqaf רֵעֵ֖הוּ יַעְזֹ֑רוּ וּלְאָחִ֖יו יֹאמַ֥ר חֲזָֽק׃}
{They helped every one his neighbour; And every one said to his brother: ‘Be of good courage.’}
\twoverse{\arabic{verse}}
{וַיְחַזֵּ֤ק חָרָשׁ֙ אֶת\maqqaf צֹרֵ֔ף מַחֲלִ֥יק פַּטִּ֖ישׁ אֶת\maqqaf ה֣וֹלֶם פָּ֑עַם אֹמֵ֤ר לַדֶּ֙בֶק֙ ט֣וֹב ה֔וּא וַיְחַזְּקֵ֥הוּ בְמַסְמְרִ֖ים לֹ֥א יִמּֽוֹט׃ \setuma }
{So the carpenter encouraged the goldsmith, And he that smootheth with the hammer him that smiteth the anvil, Saying of the soldering: ‘It is good’; And he fastened it with nails, that it should not be moved.}
\twoverse{\arabic{verse}}
{וְאַתָּה֙ יִשְׂרָאֵ֣ל עַבְדִּ֔י יַעֲקֹ֖ב אֲשֶׁ֣ר בְּחַרְתִּ֑יךָ זֶ֖רַע אַבְרָהָ֥ם אֹהֲבִֽי׃}
{But thou, Israel, My servant, Jacob whom I have chosen, The seed of Abraham My friend;}
\twoverse{\arabic{verse}}
{אֲשֶׁ֤ר הֶחֱזַקְתִּ֙יךָ֙ מִקְצ֣וֹת הָאָ֔רֶץ וּמֵאֲצִילֶ֖יהָ קְרָאתִ֑יךָ וָאֹ֤מַר לְךָ֙ עַבְדִּי\maqqaf אַ֔תָּה בְּחַרְתִּ֖יךָ וְלֹ֥א מְאַסְתִּֽיךָ׃}
{Thou whom I have taken hold of from the ends of the earth, And called thee from the uttermost parts thereof, And said unto thee: ‘Thou art My servant, I have chosen thee and not cast thee away’;}
\twoverse{\arabic{verse}}
{אַל\maqqaf תִּירָא֙ כִּ֣י עִמְּךָ\maqqaf אָ֔נִי אַל\maqqaf תִּשְׁתָּ֖ע כִּֽי\maqqaf אֲנִ֣י אֱלֹהֶ֑יךָ אִמַּצְתִּ֙יךָ֙ אַף\maqqaf עֲזַרְתִּ֔יךָ אַף\maqqaf תְּמַכְתִּ֖יךָ בִּימִ֥ין צִדְקִֽי׃}
{Fear thou not, for I am with thee, Be not dismayed, for I am thy God; I strengthen thee, yea, I help thee; Yea, I uphold thee with My victorious right hand.}
\twoverse{\arabic{verse}}
{הֵ֤ן יֵבֹ֙שׁוּ֙ וְיִכָּ֣לְמ֔וּ כֹּ֖ל הַנֶּחֱרִ֣ים בָּ֑ךְ יִהְי֥וּ כְאַ֛יִן וְיֹאבְד֖וּ אַנְשֵׁ֥י רִיבֶֽךָ׃}
{Behold, all they that were incensed against thee Shall be ashamed and confounded; They that strove with thee Shall be as nothing, and shall perish.}
\twoverse{\arabic{verse}}
{תְּבַקְשֵׁם֙ וְלֹ֣א תִמְצָאֵ֔ם אַנְשֵׁ֖י מַצֻּתֶ֑ךָ יִהְי֥וּ כְאַ֛יִן וּכְאֶ֖פֶס אַנְשֵׁ֥י מִלְחַמְתֶּֽךָ׃}
{Thou shalt seek them, and shalt not find them, Even them that contended with thee; They that warred against thee Shall be as nothing, and as a thing of nought.}
\twoverse{\arabic{verse}}
{כִּ֗י אֲנִ֛י יְהֹוָ֥ה אֱלֹהֶ֖יךָ מַחֲזִ֣יק יְמִינֶ֑ךָ הָאֹמֵ֥ר לְךָ֛ אַל\maqqaf תִּירָ֖א אֲנִ֥י עֲזַרְתִּֽיךָ׃ \setuma }
{For I the \lord\space thy God Hold thy right hand, Who say unto thee: ‘Fear not, I help thee.’}
\twoverse{\arabic{verse}}
{אַל\maqqaf תִּֽירְאִי֙ תּוֹלַ֣עַת יַעֲקֹ֔ב מְתֵ֖י יִשְׂרָאֵ֑ל אֲנִ֤י עֲזַרְתִּיךְ֙ נְאֻם\maqqaf יְהֹוָ֔ה וְגֹאֲלֵ֖ךְ קְד֥וֹשׁ יִשְׂרָאֵֽל׃}
{Fear not, thou worm Jacob, And ye men of Israel; I help thee, saith the \lord, And thy Redeemer, the Holy One of Israel.}
\twoverse{\arabic{verse}}
{הִנֵּ֣ה שַׂמְתִּ֗יךְ לְמוֹרַג֙ חָר֣וּץ חָדָ֔שׁ בַּ֖עַל פִּיפִיּ֑וֹת תָּד֤וּשׁ הָרִים֙ וְתָדֹ֔ק וּגְבָע֖וֹת כַּמֹּ֥ץ תָּשִֽׂים׃}
{Behold, I make thee a new threshing-sledge Having sharp teeth; Thou shalt thresh the mountains, and beat them small, And shalt make the hills as chaff.}
\twoverse{\arabic{verse}}
{תִּזְרֵם֙ וְר֣וּחַ תִּשָּׂאֵ֔ם וּסְעָרָ֖ה תָּפִ֣יץ אוֹתָ֑ם וְאַתָּה֙ תָּגִ֣יל בַּֽיהֹוָ֔ה בִּקְד֥וֹשׁ יִשְׂרָאֵ֖ל תִּתְהַלָּֽל׃ \setuma }
{Thou shalt fan them, and the wind shall carry them away, And the whirlwind shall scatter them; And thou shalt rejoice in the \lord, Thou shalt glory in the Holy One of Israel.}
\newperek
\newhaftarah{וירא}
\label{haft_4}
\haftnote{The Haftara is II Kings 4:1\verserangechar 4:37. }
\setcounter{chap}{4}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}
{וְאִשָּׁ֣ה אַחַ֣ת מִנְּשֵׁ֣י בְנֵֽי\maqqaf הַ֠נְּבִיאִ֠ים צָעֲקָ֨ה אֶל\maqqaf אֱלִישָׁ֜ע לֵאמֹ֗ר עַבְדְּךָ֤ אִישִׁי֙ מֵ֔ת וְאַתָּ֣ה יָדַ֔עְתָּ כִּ֣י עַבְדְּךָ֔ הָיָ֥ה יָרֵ֖א אֶת\maqqaf יְהֹוָ֑ה וְהַ֨נֹּשֶׁ֔ה בָּ֗א לָקַ֜חַת אֶת\maqqaf שְׁנֵ֧י יְלָדַ֛י ל֖וֹ לַעֲבָדִֽים׃}
{Now there cried a certain woman of the wives of the sons of the prophets unto Elisha, saying: ‘Thy servant my husband is dead; and thou knowest that thy servant did fear the \lord; and the creditor is come to take unto him my two children to be bondmen.’}
\twoverse{\arabic{verse}}
{וַיֹּ֨אמֶר אֵלֶ֤יהָ אֱלִישָׁע֙ מָ֣ה אֶֽעֱשֶׂה\maqqaf לָּ֔ךְ הַגִּ֣ידִי לִ֔י מַה\maqqaf יֶּשׁ\maqqaf \qk{לָ֖ךְ}{לכי} בַּבָּ֑יִת וַתֹּ֗אמֶר אֵ֣ין לְשִׁפְחָתְךָ֥ כֹל֙ בַּבַּ֔יִת כִּ֖י אִם\maqqaf אָס֥וּךְ שָֽׁמֶן׃}
{And Elisha said unto her: ‘What shall I do for thee? tell me; what hast thou in the house?’ And she said: ‘Thy handmaid hath not any thing in the house, save a pot of oil.’}
\twoverse{\arabic{verse}}
{וַיֹּ֗אמֶר לְכִ֨י שַׁאֲלִי\maqqaf לָ֤ךְ כֵּלִים֙ מִן\maqqaf הַח֔וּץ מֵאֵ֖ת כׇּל\maqqaf \qk{שְׁכֵנָ֑יִךְ}{שכנכי} כֵּלִ֥ים רֵקִ֖ים אַל\maqqaf תַּמְעִֽיטִי׃}
{Then he said: ‘Go, borrow thee vessels abroad of all thy neighbours, even empty vessels; borrow not a few.}
\twoverse{\arabic{verse}}
{וּבָ֗את וְסָגַ֤רְתְּ הַדֶּ֙לֶת֙ בַּעֲדֵ֣ךְ וּבְעַד\maqqaf בָּנַ֔יִךְ וְיָצַ֕קְתְּ עַ֥ל כׇּל\maqqaf הַכֵּלִ֖ים הָאֵ֑לֶּה וְהַמָּלֵ֖א תַּסִּֽיעִי׃}
{And thou shalt go in, and shut the door upon thee and upon thy sons, and pour out into all those vessels; and thou shalt set aside that which is full.’}
\twoverse{\arabic{verse}}
{וַתֵּ֙לֶךְ֙ מֵֽאִתּ֔וֹ וַתִּסְגֹּ֣ר הַדֶּ֔לֶת בַּעֲדָ֖הּ וּבְעַ֣ד בָּנֶ֑יהָ הֵ֛ם מַגִּישִׁ֥ים אֵלֶ֖יהָ וְהִ֥יא \qk{מוֹצָֽקֶת}{מיצקת}׃}
{So she went from him, and shut the door upon her and upon her sons; they brought the vessels to her, and she poured out.}
\twoverse{\arabic{verse}}
{וַיְהִ֣י \legarmeh  כִּמְלֹ֣את הַכֵּלִ֗ים וַתֹּ֤אמֶר אֶל\maqqaf בְּנָהּ֙ הַגִּ֨ישָׁה אֵלַ֥י עוֹד֙ כֶּ֔לִי וַיֹּ֣אמֶר אֵלֶ֔יהָ אֵ֥ין ע֖וֹד כֶּ֑לִי וַֽיַּעֲמֹ֖ד הַשָּֽׁמֶן׃}
{And it came to pass, when the vessels were full, that she said unto her son: ‘Bring me yet a vessel.’ And he said unto her: ‘There is not a vessel more.’ And the oil stayed.}
\twoverse{\arabic{verse}}
{וַתָּבֹ֗א וַתַּגֵּד֙ לְאִ֣ישׁ הָאֱלֹהִ֔ים וַיֹּ֗אמֶר לְכִי֙ מִכְרִ֣י אֶת\maqqaf הַשֶּׁ֔מֶן וְשַׁלְּמִ֖י אֶת\maqqaf \qk{נִשְׁיֵ֑ךְ}{נשיכי} וְאַ֣תְּ \qk{וּבָנַ֔יִךְ}{בניכי} תִּֽחְיִ֖י בַּנּוֹתָֽר׃ \petucha }
{Then she came and told the man of God. And he said: ‘Go, sell the oil, and pay thy debt, and live thou and thy sons of the rest.’}
\twoverse{\arabic{verse}}
{וַיְהִ֨י הַיּ֜וֹם וַיַּֽעֲבֹ֧ר אֱלִישָׁ֣ע אֶל\maqqaf שׁוּנֵ֗ם וְשָׁם֙ אִשָּׁ֣ה גְדוֹלָ֔ה וַתַּחֲזֶק\maqqaf בּ֖וֹ לֶאֱכׇל\maqqaf לָ֑חֶם וַֽיְהִי֙ מִדֵּ֣י עׇבְר֔וֹ יָסֻ֥ר שָׁ֖מָּה לֶאֱכׇל\maqqaf לָֽחֶם׃}
{And it fell on a day, that Elisha passed to Shunem, where was a great woman; and she constrained him to eat bread. And so it was, that as oft as he passed by, he turned in thither to eat bread.}
\twoverse{\arabic{verse}}
{וַתֹּ֙אמֶר֙ אֶל\maqqaf אִישָׁ֔הּ הִנֵּה\maqqaf נָ֣א יָדַ֔עְתִּי כִּ֛י אִ֥ישׁ אֱלֹהִ֖ים קָד֣וֹשׁ ה֑וּא עֹבֵ֥ר עָלֵ֖ינוּ תָּמִֽיד׃}
{And she said unto her husband: ‘Behold now, I perceive that this is a holy man of God, that passeth by us continually.}
\twoverse{\arabic{verse}}
{נַעֲשֶׂה\maqqaf נָּ֤א עֲלִיַּת\maqqaf קִיר֙ קְטַנָּ֔ה וְנָשִׂ֨ים ל֥וֹ שָׁ֛ם מִטָּ֥ה וְשֻׁלְחָ֖ן וְכִסֵּ֣א וּמְנוֹרָ֑ה וְהָיָ֛ה בְּבֹא֥וֹ אֵלֵ֖ינוּ יָס֥וּר שָֽׁמָּה׃}
{Let us make, I pray thee, a little chamber on the roof; and let us set for him there a bed, and a table, and a stool, and a candlestick; and it shall be, when he cometh to us, that he shall turn in thither.’}
\twoverse{\arabic{verse}}
{וַיְהִ֥י הַיּ֖וֹם וַיָּ֣בֹא שָׁ֑מָּה וַיָּ֥סַר אֶל\maqqaf הָעֲלִיָּ֖ה וַיִּשְׁכַּב\maqqaf שָֽׁמָּה׃}
{And it fell on a day, that he came thither, and he turned into the upper chamber and lay there.}
\twoverse{\arabic{verse}}
{וַיֹּ֙אמֶר֙ אֶל\maqqaf גֵּיחֲזִ֣י נַעֲר֔וֹ קְרָ֖א לַשּׁוּנַמִּ֣ית הַזֹּ֑את וַיִּ֨קְרָא\maqqaf לָ֔הּ וַֽתַּעֲמֹ֖ד לְפָנָֽיו׃}
{And he said to Gehazi his servant: ‘Call this Shunammite.’ And when he had called her, she stood before him.}
\twoverse{\arabic{verse}}
{וַיֹּ֣אמֶר ל֗וֹ אֱמׇר\maqqaf נָ֣א אֵלֶ֘יהָ֮ הִנֵּ֣ה חָרַ֣דְתְּ \pasek  אֵלֵ֘ינוּ֮ אֶת\maqqaf כׇּל\maqqaf הַחֲרָדָ֣ה הַזֹּאת֒ מֶ֚ה לַעֲשׂ֣וֹת לָ֔ךְ הֲיֵ֤שׁ לְדַבֶּר\maqqaf לָךְ֙ אֶל\maqqaf הַמֶּ֔לֶךְ א֖וֹ אֶל\maqqaf שַׂ֣ר הַצָּבָ֑א וַתֹּ֕אמֶר בְּת֥וֹךְ עַמִּ֖י אָנֹכִ֥י יֹשָֽׁבֶת׃}
{And he said unto him: ‘Say now unto her: Behold, thou hast been careful for us with all this care; what is to be done for thee? wouldest thou be spoken for to the king, or to the captain of the host?’ And she answered: ‘I dwell among mine own people.’}
\twoverse{\arabic{verse}}
{וַיֹּ֕אמֶר וּמֶ֖ה לַעֲשׂ֣וֹת לָ֑הּ וַיֹּ֣אמֶר גֵּיחֲזִ֗י אֲבָ֛ל בֵּ֥ן אֵֽין\maqqaf לָ֖הּ וְאִישָׁ֥הּ זָקֵֽן׃}
{And he said: ‘What then is to be done for her?’ And Gehazi answered: ‘Verily she hath no son, and her husband is old.’}
\twoverse{\arabic{verse}}
{וַיֹּ֖אמֶר קְרָא\maqqaf לָ֑הּ וַיִּ֨קְרָא\maqqaf לָ֔הּ וַֽתַּעֲמֹ֖ד בַּפָּֽתַח׃}
{And he said: ‘Call her.’ And when he had called her, she stood in the door.}
\twoverse{\arabic{verse}}
{וַיֹּ֗אמֶר לַמּוֹעֵ֤ד הַזֶּה֙ כָּעֵ֣ת חַיָּ֔ה \qk{אַ֖תְּ}{אתי} חֹבֶ֣קֶת בֵּ֑ן וַתֹּ֗אמֶר אַל\maqqaf אֲדֹנִי֙ אִ֣ישׁ הָאֱלֹהִ֔ים אַל\maqqaf תְּכַזֵּ֖ב בְּשִׁפְחָתֶֽךָ׃}
{And he said: ‘At this season, when the time cometh round, thou shalt embrace a son.’ And she said: ‘Nay, my lord, thou man of God, do not lie unto thy handmaid.’}
\twoverse{\arabic{verse}}
{וַתַּ֥הַר הָאִשָּׁ֖ה וַתֵּ֣לֶד בֵּ֑ן לַמּוֹעֵ֤ד הַזֶּה֙ כָּעֵ֣ת חַיָּ֔ה אֲשֶׁר\maqqaf דִּבֶּ֥ר אֵלֶ֖יהָ אֱלִישָֽׁע׃}
{And the woman conceived, and bore a son at that season, when the time came round, as Elisha had said unto her.}
\twoverse{\arabic{verse}}
{וַיִּגְדַּ֖ל הַיָּ֑לֶד וַיְהִ֣י הַיּ֔וֹם וַיֵּצֵ֥א אֶל\maqqaf אָבִ֖יו אֶל\maqqaf הַקֹּצְרִֽים׃}
{And when the child was grown, it fell on a day, that he went out to his father to the reapers.}
\twoverse{\arabic{verse}}
{וַיֹּ֥אמֶר אֶל\maqqaf אָבִ֖יו רֹאשִׁ֣י \pasek  רֹאשִׁ֑י וַיֹּ֙אמֶר֙ אֶל\maqqaf הַנַּ֔עַר שָׂאֵ֖הוּ אֶל\maqqaf אִמּֽוֹ׃}
{And he said unto his father: ‘My head, my head.’ And he said to his servant: ‘Carry him to his mother.’}
\twoverse{\arabic{verse}}
{וַיִּ֨שָּׂאֵ֔הוּ וַיְבִיאֵ֖הוּ אֶל\maqqaf אִמּ֑וֹ וַיֵּ֧שֶׁב עַל\maqqaf בִּרְכֶּ֛יהָ עַד\maqqaf הַֽצׇּהֳרַ֖יִם וַיָּמֹֽת׃}
{And when he had taken him, and brought him to his mother, he sat on her knees till noon, and then died.}
\twoverse{\arabic{verse}}
{וַתַּ֙עַל֙ וַתַּשְׁכִּבֵ֔הוּ עַל\maqqaf מִטַּ֖ת אִ֣ישׁ הָאֱלֹהִ֑ים וַתִּסְגֹּ֥ר בַּעֲד֖וֹ וַתֵּצֵֽא׃}
{And she went up, and laid him on the bed of the man of God, and shut the door upon him, and went out.}
\twoverse{\arabic{verse}}
{וַתִּקְרָא֮ אֶל\maqqaf אִישָׁהּ֒ וַתֹּ֗אמֶר שִׁלְחָ֨ה נָ֥א לִי֙ אֶחָ֣ד מִן\maqqaf הַנְּעָרִ֔ים וְאַחַ֖ת הָאֲתֹנ֑וֹת וְאָר֛וּצָה עַד\maqqaf אִ֥ישׁ הָאֱלֹהִ֖ים וְאָשֽׁוּבָה׃}
{And she called unto her husband, and said: ‘Send me, I pray thee, one of the servants, and one of the asses, that I may run to the man of God, and come back.’}
\twoverse{\arabic{verse}}
{וַיֹּ֗אמֶר מַ֠דּ֠וּעַ \qk{אַ֣תְּ}{אתי} \qk{הֹלֶ֤כֶת}{הלכתי} אֵלָיו֙ הַיּ֔וֹם לֹא\maqqaf חֹ֖דֶשׁ וְלֹ֣א שַׁבָּ֑ת וַתֹּ֖אמֶר שָׁלֽוֹם׃}
{And he said: Wherefore wilt thou go to him today? it is neither new moon nor sabbath.’ And she said: ‘It shall be well.’}
\twoverse{\arabic{verse}}
{וַֽתַּחֲבֹשׁ֙ הָאָת֔וֹן וַתֹּ֥אמֶר אֶֽל\maqqaf נַעֲרָ֖הּ נְהַ֣ג וָלֵ֑ךְ אַל\maqqaf תַּעֲצׇר\maqqaf לִ֣י לִרְכֹּ֔ב כִּ֖י אִם\maqqaf אָמַ֥רְתִּי לָֽךְ׃}
{Then she saddled an ass, and said to her servant: ‘Drive, and go forward; slacken me not the riding, except I bid thee.’}
\twoverse{\arabic{verse}}
{וַתֵּ֗לֶךְ וַתָּבֹ֛א אֶל\maqqaf אִ֥ישׁ הָאֱלֹהִ֖ים אֶל\maqqaf הַ֣ר הַכַּרְמֶ֑ל וַ֠יְהִ֠י כִּרְא֨וֹת אִישׁ\maqqaf הָאֱלֹהִ֤ים אֹתָהּ֙ מִנֶּ֔גֶד וַיֹּ֙אמֶר֙ אֶל\maqqaf גֵּיחֲזִ֣י נַעֲר֔וֹ הִנֵּ֖ה הַשּׁוּנַמִּ֥ית הַלָּֽז׃}
{So she went, and came unto the man of God to mount Carmel. And it came to pass, when the man of God saw her afar off, that he said to Gehazi his servant: ‘Behold, yonder is that Shunammite.}
\twoverse{\arabic{verse}}
{עַתָּה֮ רֽוּץ\maqqaf נָ֣א לִקְרָאתָהּ֒ וֶאֱמׇר\maqqaf לָ֗הּ הֲשָׁל֥וֹם לָ֛ךְ הֲשָׁל֥וֹם לְאִישֵׁ֖ךְ הֲשָׁל֣וֹם לַיָּ֑לֶד וַתֹּ֖אמֶר שָׁלֽוֹם׃}
{Run, I pray thee, now to meet her, and say unto her: Is it well with thee? is it well with thy husband? is it well with the child?’ And she answered: ‘It is well.’}
\twoverse{\arabic{verse}}
{וַתָּבֹ֞א אֶל\maqqaf אִ֤ישׁ הָאֱלֹהִים֙ אֶל\maqqaf הָהָ֔ר וַֽתַּחֲזֵ֖ק בְּרַגְלָ֑יו וַיִּגַּ֨שׁ גֵּיחֲזִ֜י לְהׇדְפָ֗הּ וַיֹּ֩אמֶר֩ אִ֨ישׁ הָאֱלֹהִ֤ים הַרְפֵּה\maqqaf לָהּ֙ כִּֽי\maqqaf נַפְשָׁ֣הּ מָֽרָה\maqqaf לָ֔הּ וַֽיהֹוָה֙ הֶעְלִ֣ים מִמֶּ֔נִּי וְלֹ֥א הִגִּ֖יד לִֽי׃}
{And when she came to the man of God to the hill, she caught hold of his feet. And Gehazi came near to thrust her away; but the man of God said: ‘Let her alone; for her soul is bitter within her; and the \lord\space hath hid it from me, and hath not told Me.’}
\twoverse{\arabic{verse}}
{וַתֹּ֕אמֶר הֲשָׁאַ֥לְתִּי בֵ֖ן מֵאֵ֣ת אֲדֹנִ֑י הֲלֹ֣א אָמַ֔רְתִּי לֹ֥א תַשְׁלֶ֖ה אֹתִֽי׃}
{Then she said: ‘Did I desire a son of my lord? did I not say: Do not deceive me?’}
\twoverse{\arabic{verse}}
{וַיֹּ֨אמֶר לְגֵיחֲזִ֜י חֲגֹ֣ר מׇתְנֶ֗יךָ וְקַ֨ח מִשְׁעַנְתִּ֣י בְיָדְךָ֮ וָלֵךְ֒ כִּי\maqqaf תִמְצָ֥א אִישׁ֙ לֹ֣א תְבָרְכֶ֔נּוּ וְכִֽי\maqqaf יְבָרֶכְךָ֥ אִ֖ישׁ לֹ֣א תַעֲנֶ֑נּוּ וְשַׂמְתָּ֥ מִשְׁעַנְתִּ֖י עַל\maqqaf פְּנֵ֥י הַנָּֽעַר׃}
{Then he said to Gehazi: ‘Gird up thy loins, and take my staff in thy hand, and go thy way; if thou meet any man, salute him not; and if any salute thee, answer him not; and lay my staff upon the face of the child.’}
\twoverse{\arabic{verse}}
{וַתֹּ֙אמֶר֙ אֵ֣ם הַנַּ֔עַר חַי\maqqaf יְהֹוָ֥ה וְחֵֽי\maqqaf נַפְשְׁךָ֖ אִם\maqqaf אֶעֶזְבֶ֑ךָּ וַיָּ֖קׇם וַיֵּ֥לֶךְ אַחֲרֶֽיהָ׃}
{And the mother of the child said: ‘As the \lord\space liveth, and as thy soul liveth, I will not leave thee.’ And he arose, and followed her.}
\twoverse{\arabic{verse}}
{וְגֵחֲזִ֞י עָבַ֣ר לִפְנֵיהֶ֗ם וַיָּ֤שֶׂם אֶת\maqqaf הַמִּשְׁעֶ֙נֶת֙ עַל\maqqaf פְּנֵ֣י הַנַּ֔עַר וְאֵ֥ין ק֖וֹל וְאֵ֣ין קָ֑שֶׁב וַיָּ֤שׇׁב לִקְרָאתוֹ֙ וַיַּגֶּד\maqqaf ל֣וֹ לֵאמֹ֔ר לֹ֥א הֵקִ֖יץ הַנָּֽעַר׃}
{And Gehazi passed on before them, and laid the staff upon the face of the child; but there was neither voice, nor hearing. Wherefore he returned to meet him, and told him, saying: ‘The child is not awaked.’}
\twoverse{\arabic{verse}}
{וַיָּבֹ֥א אֱלִישָׁ֖ע הַבָּ֑יְתָה וְהִנֵּ֤ה הַנַּ֙עַר֙ מֵ֔ת מֻשְׁכָּ֖ב עַל\maqqaf מִטָּתֽוֹ׃}
{And when Elisha was come into the house, behold, the child was dead, and laid upon his bed.}
\twoverse{\arabic{verse}}
{וַיָּבֹ֕א וַיִּסְגֹּ֥ר הַדֶּ֖לֶת בְּעַ֣ד שְׁנֵיהֶ֑ם וַיִּתְפַּלֵּ֖ל אֶל\maqqaf יְהֹוָֽה׃}
{He went in therefore, and shut the door upon them twain, and prayed unto the \lord.}
\twoverse{\arabic{verse}}
{וַיַּ֜עַל וַיִּשְׁכַּ֣ב עַל\maqqaf הַיֶּ֗לֶד וַיָּ֩שֶׂם֩ פִּ֨יו עַל\maqqaf פִּ֜יו וְעֵינָ֤יו עַל\maqqaf עֵינָיו֙ וְכַפָּ֣יו עַל\maqqaf כַּפָּ֔ו וַיִּגְהַ֖ר עָלָ֑יו וַיָּ֖חׇם בְּשַׂ֥ר הַיָּֽלֶד׃}
{And he went up, and lay upon the child, and put his mouth upon his mouth, and his eyes upon his eyes, and his hands upon his hands; and he stretched himself upon him; and the flesh of the child waxed warm.}
\twoverse{\arabic{verse}}
{וַיָּ֜שׇׁב וַיֵּ֣לֶךְ בַּבַּ֗יִת אַחַ֥ת הֵ֙נָּה֙ וְאַחַ֣ת הֵ֔נָּה וַיַּ֖עַל וַיִּגְהַ֣ר עָלָ֑יו וַיְזוֹרֵ֤ר הַנַּ֙עַר֙ עַד\maqqaf שֶׁ֣בַע פְּעָמִ֔ים וַיִּפְקַ֥ח הַנַּ֖עַר אֶת\maqqaf עֵינָֽיו׃}
{Then he returned, and walked in the house once to and fro; and went up, and stretched himself upon him; and the child sneezed seven times, and the child opened his eyes.}
\twoverse{\arabic{verse}}
{וַיִּקְרָ֣א אֶל\maqqaf גֵּיחֲזִ֗י וַיֹּ֙אמֶר֙ קְרָא֙ אֶל\maqqaf הַשֻּׁנַמִּ֣ית הַזֹּ֔את וַיִּקְרָאֶ֖הָ וַתָּבֹ֣א אֵלָ֑יו וַיֹּ֖אמֶר שְׂאִ֥י בְנֵֽךְ׃}
{And he called Gehazi, and said: ‘Call this Shunammite.’ So he called her. And when she was come in unto him, he said: ‘Take up thy son.’}
\twoverse{\arabic{verse}}
{וַתָּבֹא֙ וַתִּפֹּ֣ל עַל\maqqaf רַגְלָ֔יו וַתִּשְׁתַּ֖חוּ אָ֑רְצָה וַתִּשָּׂ֥א אֶת\maqqaf בְּנָ֖הּ וַתֵּצֵֽא׃ \petucha }
{Then she went in, and fell at his feet, and bowed down to the ground; and she took up her son, and went out.}
\newhaftarah{חיי שרה}
\label{haft_5}
\haftnote{The Haftara is I Kings 1:1\verserangechar 1:31. }
\setcounter{chap}{1}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}
{וְהַמֶּ֤לֶךְ דָּוִד֙ זָקֵ֔ן בָּ֖א בַּיָּמִ֑ים וַיְכַסֻּ֙הוּ֙ בַּבְּגָדִ֔ים וְלֹ֥א יִחַ֖ם לֽוֹ׃}
{Now King David was old and stricken in years; and they covered him with clothes, but he could get no heat.}
\twoverse{\arabic{verse}}
{וַיֹּ֧אמְרוּ ל֣וֹ עֲבָדָ֗יו יְבַקְשׁ֞וּ לַאדֹנִ֤י הַמֶּ֙לֶךְ֙ נַעֲרָ֣ה בְתוּלָ֔ה וְעָֽמְדָה֙ לִפְנֵ֣י הַמֶּ֔לֶךְ וּתְהִי\maqqaf ל֖וֹ סֹכֶ֑נֶת וְשָׁכְבָ֣ה בְחֵיקֶ֔ךָ וְחַ֖ם לַאדֹנִ֥י הַמֶּֽלֶךְ׃}
{Wherefore his servants said unto him: ‘Let there be sought for my lord the king a young virgin; and let her stand before the king, and be a companion unto him; and let her lie in thy bosom, that my lord the king may get heat.’}
\twoverse{\arabic{verse}}
{וַיְבַקְשׁוּ֙ נַעֲרָ֣ה יָפָ֔ה בְּכֹ֖ל גְּב֣וּל יִשְׂרָאֵ֑ל וַֽיִּמְצְא֗וּ אֶת\maqqaf אֲבִישַׁג֙ הַשּׁ֣וּנַמִּ֔ית וַיָּבִ֥אוּ אֹתָ֖הּ לַמֶּֽלֶךְ׃}
{So they sought for a fair damsel throughout all the borders of Israel, and found Abishag the Shunammite, and brought her to the king.}
\twoverse{\arabic{verse}}
{וְהַֽנַּעֲרָ֖ה יָפָ֣ה עַד\maqqaf מְאֹ֑ד וַתְּהִ֨י לַמֶּ֤לֶךְ סֹכֶ֙נֶת֙ וַתְּשָׁ֣רְתֵ֔הוּ וְהַמֶּ֖לֶךְ לֹ֥א יְדָעָֽהּ׃}
{And the damsel was very fair; and she became a companion unto the king, and ministered to him; but the king knew her not.}
\twoverse{\arabic{verse}}
{וַאֲדֹנִיָּ֧ה בֶן\maqqaf חַגִּ֛ית מִתְנַשֵּׂ֥א לֵאמֹ֖ר אֲנִ֣י אֶמְלֹ֑ךְ וַיַּ֣עַשׂ ל֗וֹ רֶ֚כֶב וּפָ֣רָשִׁ֔ים וַחֲמִשִּׁ֥ים אִ֖ישׁ רָצִ֥ים לְפָנָֽיו׃}
{Now Adonijah the son of Haggith exalted himself, saying: ‘I will be king’; and he prepared him chariots and horsemen, and fifty men to run before him.}
\twoverse{\arabic{verse}}
{וְלֹֽא\maqqaf עֲצָב֨וֹ אָבִ֤יו מִיָּמָיו֙ לֵאמֹ֔ר מַדּ֖וּעַ כָּ֣כָה עָשִׂ֑יתָ וְגַם\maqqaf ה֤וּא טֽוֹב\maqqaf תֹּ֙אַר֙ מְאֹ֔ד וְאֹת֥וֹ יָלְדָ֖ה אַחֲרֵ֥י אַבְשָׁלֽוֹם׃}
{And his father had not grieved him all his life in saying: ‘Why hast thou done so?’ and he was also a very goodly man; and he was born after Absalom.}
\twoverse{\arabic{verse}}
{וַיִּהְי֣וּ דְבָרָ֔יו עִ֚ם יוֹאָ֣ב בֶּן\maqqaf צְרוּיָ֔ה וְעִ֖ם אֶבְיָתָ֣ר הַכֹּהֵ֑ן וַֽיַּעְזְר֔וּ אַחֲרֵ֖י אֲדֹנִיָּֽה׃}
{And he conferred with Joab the son of Zeruiah, and with Abiathar the priest; and they following Adonijah helped him.}
\twoverse{\arabic{verse}}
{וְצָד֣וֹק הַ֠כֹּהֵ֠ן וּבְנָיָ֨הוּ בֶן\maqqaf יְהוֹיָדָ֜ע וְנָתָ֤ן הַנָּבִיא֙ וְשִׁמְעִ֣י וְרֵעִ֔י וְהַגִּבּוֹרִ֖ים אֲשֶׁ֣ר לְדָוִ֑ד לֹ֥א הָי֖וּ עִם\maqqaf אֲדֹנִיָּֽהוּ׃}
{But Zadok the priest, and Benaiah the son of Jehoiada, and Nathan the prophet, and Shimei, and Rei, and the mighty men that belonged to David, were not with Adonijah.}
\twoverse{\arabic{verse}}
{וַיִּזְבַּ֣ח אֲדֹנִיָּ֗הוּ צֹ֤אן וּבָקָר֙ וּמְרִ֔יא עִ֚ם אֶ֣בֶן הַזֹּחֶ֔לֶת אֲשֶׁר\maqqaf אֵ֖צֶל עֵ֣ין רֹגֵ֑ל וַיִּקְרָ֗א אֶת\maqqaf כׇּל\maqqaf אֶחָיו֙ בְּנֵ֣י הַמֶּ֔לֶךְ וּלְכׇל\maqqaf אַנְשֵׁ֥י יְהוּדָ֖ה עַבְדֵ֥י הַמֶּֽלֶךְ׃}
{And Adonijah slew sheep and oxen and fatlings by the stone of Zoheleth, which is beside En-rogel; and he called all his brethren the king’s sons, and all the men of Judah the king’s servants;}
\twoverse{\arabic{verse}}
{וְֽאֶת\maqqaf נָתָן֩ הַנָּבִ֨יא וּבְנָיָ֜הוּ וְֽאֶת\maqqaf הַגִּבּוֹרִ֛ים וְאֶת\maqqaf שְׁלֹמֹ֥ה אָחִ֖יו לֹ֥א קָרָֽא׃}
{but Nathan the prophet, and Benaiah, and the mighty men, and Solomon his brother, he called not.}
\twoverse{\arabic{verse}}
{וַיֹּ֣אמֶר נָתָ֗ן אֶל\maqqaf בַּת\maqqaf שֶׁ֤בַע אֵם\maqqaf שְׁלֹמֹה֙ לֵאמֹ֔ר הֲל֣וֹא שָׁמַ֔עַתְּ כִּ֥י מָלַ֖ךְ אֲדֹנִיָּ֣הוּ בֶן\maqqaf חַגִּ֑ית וַאֲדֹנֵ֥ינוּ דָוִ֖ד לֹ֥א יָדָֽע׃}
{Then Nathan spoke unto Bath-sheba the mother of Solomon, saying: ‘Hast thou not heard that Adonijah the son of Haggith doth reign, and David our lord knoweth it not?}
\twoverse{\arabic{verse}}
{וְעַתָּ֕ה לְכִ֛י אִיעָצֵ֥ךְ נָ֖א עֵצָ֑ה וּמַלְּטִי֙ אֶת\maqqaf נַפְשֵׁ֔ךְ וְאֶת\maqqaf נֶ֥פֶשׁ בְּנֵ֖ךְ שְׁלֹמֹֽה׃}
{Now therefore come, let me, I pray thee, give thee counsel, that thou mayest save thine own life, and the life of thy son Solomon.}
\twoverse{\arabic{verse}}
{לְכִ֞י וּבֹ֣אִי \legarmeh  אֶל\maqqaf הַמֶּ֣לֶךְ דָּוִ֗ד וְאָמַ֤רְתְּ אֵלָיו֙ הֲלֹֽא\maqqaf אַתָּ֞ה אֲדֹנִ֣י הַמֶּ֗לֶךְ נִשְׁבַּ֤עְתָּ לַאֲמָֽתְךָ֙ לֵאמֹ֔ר כִּֽי\maqqaf שְׁלֹמֹ֤ה בְנֵךְ֙ יִמְלֹ֣ךְ אַחֲרַ֔י וְה֖וּא יֵשֵׁ֣ב עַל\maqqaf כִּסְאִ֑י וּמַדּ֖וּעַ מָלַ֥ךְ אֲדֹנִיָּֽהוּ׃}
{Go and get thee in unto king David, and say unto him: Didst not thou, my lord, O king, swear unto thy handmaid, saying: Assuredly Solomon thy son shall reign after me, and he shall sit upon my throne? why then doth Adonijah reign?}
\twoverse{\arabic{verse}}
{הִנֵּ֗ה עוֹדָ֛ךְ מְדַבֶּ֥רֶת שָׁ֖ם עִם\maqqaf הַמֶּ֑לֶךְ וַֽאֲנִי֙ אָב֣וֹא אַחֲרַ֔יִךְ וּמִלֵּאתִ֖י אֶת\maqqaf דְּבָרָֽיִךְ׃}
{Behold, while thou yet talkest there with the king, I also will come in after thee, and confirm thy words.’}
\twoverse{\arabic{verse}}
{וַתָּבֹ֨א בַת\maqqaf שֶׁ֤בַע אֶל\maqqaf הַמֶּ֙לֶךְ֙ הַחַ֔דְרָה וְהַמֶּ֖לֶךְ זָקֵ֣ן מְאֹ֑ד וַאֲבִישַׁג֙ הַשּׁ֣וּנַמִּ֔ית מְשָׁרַ֖ת אֶת\maqqaf הַמֶּֽלֶךְ׃}
{And Bath-sheba went in unto the king into the chamber.—Now the king was very old; and Abishag the Shunammite ministered unto the king.—}
\twoverse{\arabic{verse}}
{וַתִּקֹּ֣ד בַּת\maqqaf שֶׁ֔בַע וַתִּשְׁתַּ֖חוּ לַמֶּ֑לֶךְ וַיֹּ֥אמֶר הַמֶּ֖לֶךְ מַה\maqqaf לָּֽךְ׃}
{And Bath-sheba bowed, and prostrated herself unto the king. And the king said: ‘What wouldest thou?’}
\twoverse{\arabic{verse}}
{וַתֹּ֣אמֶר ל֗וֹ אֲדֹנִי֙ אַתָּ֨ה נִשְׁבַּ֜עְתָּ בַּיהֹוָ֤ה אֱלֹהֶ֙יךָ֙ לַאֲמָתֶ֔ךָ כִּֽי\maqqaf שְׁלֹמֹ֥ה בְנֵ֖ךְ יִמְלֹ֣ךְ אַחֲרָ֑י וְה֖וּא יֵשֵׁ֥ב עַל\maqqaf כִּסְאִֽי׃}
{And she said unto him: ‘My lord, thou didst swear by the \lord\space thy God unto thy handmaid: Assuredly Solomon thy son shall reign after me, and he shall sit upon my throne.}
\twoverse{\arabic{verse}}
{וְעַתָּ֕ה הִנֵּ֥ה אֲדֹנִיָּ֖ה מָלָ֑ךְ וְעַתָּ֛ה אֲדֹנִ֥י הַמֶּ֖לֶךְ לֹ֥א יָדָֽעְתָּ׃}
{And now, behold, Adonijah reigneth; and thou, my lord the king, knowest it not.}
\twoverse{\arabic{verse}}
{וַ֠יִּזְבַּ֠ח שׁ֥וֹר וּֽמְרִיא\maqqaf וְצֹאן֮ לָרֹב֒ \petucha  וַיִּקְרָא֙ לְכׇל\maqqaf בְּנֵ֣י הַמֶּ֔לֶךְ וּלְאֶבְיָתָר֙ הַכֹּהֵ֔ן וּלְיֹאָ֖ב שַׂ֣ר הַצָּבָ֑א וְלִשְׁלֹמֹ֥ה עַבְדְּךָ֖ לֹ֥א קָרָֽא׃}
{And he hath slain oxen and fatlings and sheep in abundance, and hath called all the sons of the king, and Abiathar, the priest, and Joab the captain of the host; but Solomon thy servant hath he not called.}
\twoverse{\arabic{verse}}
{וְאַתָּה֙ אֲדֹנִ֣י הַמֶּ֔לֶךְ עֵינֵ֥י כׇל\maqqaf יִשְׂרָאֵ֖ל עָלֶ֑יךָ לְהַגִּ֣יד לָהֶ֔ם מִ֗י יֵשֵׁ֛ב עַל\maqqaf כִּסֵּ֥א אֲדֹנִֽי\maqqaf הַמֶּ֖לֶךְ אַחֲרָֽיו׃}
{And thou, my lord the king, the eyes of all Israel are upon thee, that thou shouldest tell them who shall sit on the throne of my lord the king after him.}
\twoverse{\arabic{verse}}
{וְהָיָ֕ה כִּשְׁכַ֥ב אֲדֹנִֽי\maqqaf הַמֶּ֖לֶךְ עִם\maqqaf אֲבֹתָ֑יו וְהָיִ֗יתִי אֲנִ֛י וּבְנִ֥י שְׁלֹמֹ֖ה חַטָּאִֽים׃}
{Otherwise it will come to pass, when my lord the king shall sleep with his fathers, that I and my son Solomon shall be counted offenders.’}
\twoverse{\arabic{verse}}
{וְהִנֵּ֛ה עוֹדֶ֥נָּה מְדַבֶּ֖רֶת עִם\maqqaf הַמֶּ֑לֶךְ וְנָתָ֥ן הַנָּבִ֖יא בָּֽא׃}
{And, lo, while she yet talked with the king, Nathan the prophet came in.}
\twoverse{\arabic{verse}}
{וַיַּגִּ֤ידוּ לַמֶּ֙לֶךְ֙ לֵאמֹ֔ר הִנֵּ֖ה נָתָ֣ן הַנָּבִ֑יא וַיָּבֹא֙ לִפְנֵ֣י הַמֶּ֔לֶךְ וַיִּשְׁתַּ֧חוּ לַמֶּ֛לֶךְ עַל\maqqaf אַפָּ֖יו אָֽרְצָה׃}
{And they told the king, saying: ‘Behold Nathan the prophet.’ And when he was come in before the king, he bowed down before the king with his face to the ground.}
\twoverse{\arabic{verse}}
{וַיֹּ֘אמֶר֮ נָתָן֒ אֲדֹנִ֣י הַמֶּ֔לֶךְ אַתָּ֣ה אָמַ֔רְתָּ אֲדֹנִיָּ֖הוּ יִמְלֹ֣ךְ אַחֲרָ֑י וְה֖וּא יֵשֵׁ֥ב עַל\maqqaf כִּסְאִֽי׃}
{And Nathan said: ‘My lord, O king, hast thou said: Adonijah shall reign after me, and he shall sit upon my throne?}
\twoverse{\arabic{verse}}
{כִּ֣י \legarmeh  יָרַ֣ד הַיּ֗וֹם וַ֠יִּזְבַּ֠ח שׁ֥וֹר וּֽמְרִיא\maqqaf וְצֹאן֮ לָרֹב֒ וַיִּקְרָא֩ לְכׇל\maqqaf בְּנֵ֨י הַמֶּ֜לֶךְ וּלְשָׂרֵ֤י הַצָּבָא֙ וּלְאֶבְיָתָ֣ר הַכֹּהֵ֔ן וְהִנָּ֛ם אֹכְלִ֥ים וְשֹׁתִ֖ים לְפָנָ֑יו וַיֹּ֣אמְר֔וּ יְחִ֖י הַמֶּ֥לֶךְ אֲדֹנִיָּֽהוּ׃}
{For he is gone down this day, and hath slain oxen and fatlings and sheep in abundance, and hath called all the king’s sons, and the captains of the host, and Abiathar the priest; and, behold, they eat and drink before him, and say: Long live king Adonijah.}
\twoverse{\arabic{verse}}
{וְלִ֣י אֲנִֽי\maqqaf עַ֠בְדֶּ֠ךָ וּלְצָדֹ֨ק הַכֹּהֵ֜ן וְלִבְנָיָ֧הוּ בֶן\maqqaf יְהוֹיָדָ֛ע וְלִשְׁלֹמֹ֥ה עַבְדְּךָ֖ לֹ֥א קָרָֽא׃}
{But me, even me thy servant, and Zadok the priest, and Benaiah the son of Jehoiada, and thy servant Solomon hath he not called.}
\twoverse{\arabic{verse}}
{אִ֗ם מֵאֵת֙ אֲדֹנִ֣י הַמֶּ֔לֶךְ נִֽהְיָ֖ה הַדָּבָ֣ר הַזֶּ֑ה וְלֹ֤א הוֹדַ֙עְתָּ֙ אֶֽת\maqqaf \qk{עַבְדְּךָ֔}{עבדיך} מִ֗י יֵשֵׁ֛ב עַל\maqqaf כִּסֵּ֥א אֲדֹנִֽי\maqqaf הַמֶּ֖לֶךְ אַחֲרָֽיו׃ \setuma }
{Is this thing done by my lord the king, and thou hast not declared unto thy servant who should sit on the throne of my lord the king after him?’}
\twoverse{\arabic{verse}}
{וַיַּ֨עַן הַמֶּ֤לֶךְ דָּוִד֙ וַיֹּ֔אמֶר קִרְאוּ\maqqaf לִ֖י לְבַת\maqqaf שָׁ֑בַע וַתָּבֹא֙ לִפְנֵ֣י הַמֶּ֔לֶךְ וַֽתַּעֲמֹ֖ד לִפְנֵ֥י הַמֶּֽלֶךְ׃}
{Then king David answered and said: ‘Call me Bath-sheba.’ And she came into the king’s presence, and stood before the king.}
\twoverse{\arabic{verse}}
{וַיִּשָּׁבַ֥ע הַמֶּ֖לֶךְ וַיֹּאמַ֑ר חַי\maqqaf יְהֹוָ֕ה אֲשֶׁר\maqqaf פָּדָ֥ה אֶת\maqqaf נַפְשִׁ֖י מִכׇּל\maqqaf צָרָֽה׃}
{And the king swore and said: ‘As the \lord\space liveth, who hath redeemed my soul out of all adversity,}
\twoverse{\arabic{verse}}
{כִּ֡י כַּאֲשֶׁר֩ נִשְׁבַּ֨עְתִּי לָ֜ךְ בַּיהֹוָ֨ה אֱלֹהֵ֤י יִשְׂרָאֵל֙ לֵאמֹ֔ר כִּֽי\maqqaf שְׁלֹמֹ֤ה בְנֵךְ֙ יִמְלֹ֣ךְ אַחֲרַ֔י וְה֛וּא יֵשֵׁ֥ב עַל\maqqaf כִּסְאִ֖י תַּחְתָּ֑י כִּ֛י כֵּ֥ן אֶעֱשֶׂ֖ה הַיּ֥וֹם הַזֶּֽה׃}
{verily as I swore unto thee by the \lord, the God of Israel, saying: Assuredly Solomon thy son shall reign after me, and he shall sit upon my throne in my stead; verily so will I do this day.’}
\twoverse{\arabic{verse}}
{וַתִּקֹּ֨ד בַּת\maqqaf שֶׁ֤בַע אַפַּ֙יִם֙ אֶ֔רֶץ וַתִּשְׁתַּ֖חוּ לַמֶּ֑לֶךְ וַתֹּ֕אמֶר יְחִ֗י אֲדֹנִ֛י הַמֶּ֥לֶךְ דָּוִ֖ד לְעֹלָֽם׃ \petucha }
{Then Bath-sheba bowed with her face to the earth, and prostrated herself to the king, and said: ‘Let my lord king David live for ever.’}
\newhaftarah{תולדות}
\label{haft_6}
\haftnote{The Haftara is Malachi 1:1\verserangechar 2:7. On Erev Rosh \d{H}odesh, read the Haftara on page \pageref{haft_macharchodesh}.}
\setcounter{chap}{1}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}
{מַשָּׂ֥א דְבַר\maqqaf יְהֹוָ֖ה אֶל\maqqaf יִשְׂרָאֵ֑ל בְּיַ֖ד מַלְאָכִֽי׃}
{The burden of the word of the \lord\space to Israel by Malachi.}
\twoverse{\arabic{verse}}
{אָהַ֤בְתִּי אֶתְכֶם֙ אָמַ֣ר יְהֹוָ֔ה וַֽאֲמַרְתֶּ֖ם בַּמָּ֣ה אֲהַבְתָּ֑נוּ הֲלוֹא\maqqaf אָ֨ח עֵשָׂ֤ו לְיַֽעֲקֹב֙ נְאֻם\maqqaf יְהֹוָ֔ה וָאֹהַ֖ב אֶֽת\maqqaf יַעֲקֹֽב׃}
{I have loved you, saith the \lord. Yet ye say: ‘Wherein hast Thou loved us?’ Was not Esau Jacob’s brother? Saith the \lord; Yet I loved Jacob;}
\twoverse{\arabic{verse}}
{וְאֶת\maqqaf עֵשָׂ֖ו שָׂנֵ֑אתִי וָאָשִׂ֤ים אֶת\maqqaf הָרָיו֙ שְׁמָמָ֔ה וְאֶת\maqqaf נַחֲלָת֖וֹ לְתַנּ֥וֹת מִדְבָּֽר׃}
{But Esau I hated, And made his mountains a desolation, And gave his heritage to the jackals of the wilderness.}
\twoverse{\arabic{verse}}
{כִּי\maqqaf תֹאמַ֨ר אֱד֜וֹם רֻשַּׁ֗שְׁנוּ וְנָשׁוּב֙ וְנִבְנֶ֣ה חֳרָב֔וֹת כֹּ֤ה אָמַר֙ יְהֹוָ֣ה צְבָא֔וֹת הֵ֥מָּה יִבְנ֖וּ וַאֲנִ֣י אֶהֱר֑וֹס וְקָרְא֤וּ לָהֶם֙ גְּב֣וּל רִשְׁעָ֔ה וְהָעָ֛ם אֲשֶׁר\maqqaf זָעַ֥ם יְהֹוָ֖ה עַד\maqqaf עוֹלָֽם׃}
{Whereas Edom saith: ‘We are beaten down, But we will return and build the waste places’ ; Thus saith the \lord\space of hosts: They shall build, but I will throw down; And they shall be called The border of wickedness, And The people whom the \lord\space execrateth for ever.}
\twoverse{\arabic{verse}}
{וְעֵינֵיכֶ֖ם תִּרְאֶ֑ינָה וְאַתֶּ֤ם תֹּֽאמְרוּ֙ יִגְדַּ֣ל יְהֹוָ֔ה מֵעַ֖ל לִגְב֥וּל יִשְׂרָאֵֽל׃}
{And your eyes shall see, And ye shall say: ‘The \lord\space is great beyond the border of Israel.’}
\twoverse{\arabic{verse}}
{בֵּ֛ן יְכַבֵּ֥ד אָ֖ב וְעֶ֣בֶד אֲדֹנָ֑יו וְאִם\maqqaf אָ֣ב אָ֣נִי אַיֵּ֣ה כְבוֹדִ֡י וְאִם\maqqaf אֲדוֹנִ֣ים אָ֩נִי֩ אַיֵּ֨ה מוֹרָאִ֜י אָמַ֣ר \legarmeh  יְהֹוָ֣ה צְבָא֗וֹת לָכֶם֙ הַכֹּֽהֲנִים֙ בּוֹזֵ֣י שְׁמִ֔י וַאֲמַרְתֶּ֕ם בַּמֶּ֥ה בָזִ֖ינוּ אֶת\maqqaf שְׁמֶֽךָ׃}
{A son honoureth his father, And a servant his master; If then I be a father, Where is My honour? And if I be a master, Where is My fear? Saith the \lord\space of hosts Unto you, O priests, that despise My name. And ye say: ‘Wherein have we despised Thy name?’}
\twoverse{\arabic{verse}}
{מַגִּישִׁ֤ים עַֽל\maqqaf מִזְבְּחִי֙ לֶ֣חֶם מְגֹאָ֔ל וַֽאֲמַרְתֶּ֖ם בַּמֶּ֣ה גֵאַלְנ֑וּךָ בֶּאֱמׇרְכֶ֕ם שֻׁלְחַ֥ן יְהֹוָ֖ה נִבְזֶ֥ה הֽוּא׃}
{Ye offer polluted bread upon Mine altar. And ye say: ‘Wherein have we polluted thee?’ In that ye say: ‘The table of the \lord\space is contemptible.’}
\twoverse{\arabic{verse}}
{וְכִֽי\maqqaf תַגִּישׁ֨וּן עִוֵּ֤ר לִזְבֹּ֙חַ֙ אֵ֣ין רָ֔ע וְכִ֥י תַגִּ֛ישׁוּ פִּסֵּ֥חַ וְחֹלֶ֖ה אֵ֣ין רָ֑ע הַקְרִיבֵ֨הוּ נָ֜א לְפֶחָתֶ֗ךָ הֲיִרְצְךָ֙ א֚וֹ הֲיִשָּׂ֣א פָנֶ֔יךָ אָמַ֖ר יְהֹוָ֥ה צְבָאֽוֹת׃}
{And when ye offer the blind for sacrifice, is it no evil! And when ye offer the lame and sick, is it no evil! Present it now unto thy governor; will he be pleased with thee? Or will he accept thy person? Saith the \lord\space of hosts.}
\twoverse{\arabic{verse}}
{וְעַתָּ֛ה חַלּוּ\maqqaf נָ֥א פְנֵי\maqqaf אֵ֖ל וִֽיחׇנֵּ֑נוּ מִיֶּדְכֶם֙ הָ֣יְתָה זֹּ֔את הֲיִשָּׂ֤א מִכֶּם֙ פָּנִ֔ים אָמַ֖ר יְהֹוָ֥ה צְבָאֽוֹת׃}
{And now, I pray you, entreat the favour of God That He may be gracious unto us!— This hath been of your doing.— Will He accept any of your persons? Saith the \lord\space of hosts.}
\twoverse{\arabic{verse}}
{מִ֤י גַם\maqqaf בָּכֶם֙ וְיִסְגֹּ֣ר דְּלָתַ֔יִם וְלֹא\maqqaf תָאִ֥ירוּ מִזְבְּחִ֖י חִנָּ֑ם אֵֽין\maqqaf לִ֨י חֵ֜פֶץ בָּכֶ֗ם אָמַר֙ יְהֹוָ֣ה צְבָא֔וֹת וּמִנְחָ֖ה לֹא\maqqaf אֶרְצֶ֥ה מִיֶּדְכֶֽם׃}
{Oh that there were even one among you that would shut the doors, That ye might not kindle fire on Mine altar in vain! I have no pleasure in you, Saith the \lord\space of hosts, Neither will I accept an offering at your hand.}
\twoverse{\arabic{verse}}
{כִּ֣י מִמִּזְרַח\maqqaf שֶׁ֜מֶשׁ וְעַד\maqqaf מְבוֹא֗וֹ גָּד֤וֹל שְׁמִי֙ בַּגּוֹיִ֔ם וּבְכׇל\maqqaf מָק֗וֹם מֻקְטָ֥ר מֻגָּ֛שׁ לִשְׁמִ֖י וּמִנְחָ֣ה טְהוֹרָ֑ה כִּֽי\maqqaf גָד֤וֹל שְׁמִי֙ בַּגּוֹיִ֔ם אָמַ֖ר יְהֹוָ֥ה צְבָאֽוֹת׃}
{For from the rising of the sun even unto the going down of the same My name is great among the nations; And in every place offerings are presented unto My name, Even pure oblations; For My name is great among the nations, Saith the \lord\space of hosts.}
\twoverse{\arabic{verse}}
{וְאַתֶּ֖ם מְחַלְּלִ֣ים אוֹת֑וֹ בֶּאֱמׇרְכֶ֗ם שֻׁלְחַ֤ן אֲדֹנָי֙ מְגֹאָ֣ל ה֔וּא וְנִיב֖וֹ נִבְזֶ֥ה אׇכְלֽוֹ׃}
{But ye profane it, In that ye say: ‘The table of the \lord\space is polluted, And the fruit thereof, even the food thereof, is contemptible.’}
\twoverse{\arabic{verse}}
{וַאֲמַרְתֶּם֩ הִנֵּ֨ה מַתְּלָאָ֜ה וְהִפַּחְתֶּ֣ם אוֹת֗וֹ אָמַר֙ יְהֹוָ֣ה צְבָא֔וֹת וַהֲבֵאתֶ֣ם גָּז֗וּל וְאֶת\maqqaf הַפִּסֵּ֙חַ֙ וְאֶת\maqqaf הַ֣חוֹלֶ֔ה וַהֲבֵאתֶ֖ם אֶת\maqqaf הַמִּנְחָ֑ה הַאֶרְצֶ֥ה אוֹתָ֛הּ מִיֶּדְכֶ֖ם אָמַ֥ר יְהֹוָֽה׃ \setuma }
{Ye say also: ‘Behold, what a weariness is it!’ And ye have snuffed at it, Saith the \lord\space of hosts; And ye have brought that which was taken by violence, And the lame, and the sick; Thus ye bring the offering; Should I accept this of your hand? Saith the \lord.}
\twoverse{\arabic{verse}}
{וְאָר֣וּר נוֹכֵ֗ל וְיֵ֤שׁ בְּעֶדְרוֹ֙ זָכָ֔ר וְנֹדֵ֛ר וְזֹבֵ֥חַ מׇשְׁחָ֖ת לַאדֹנָ֑י כִּי֩ מֶ֨לֶךְ גָּד֜וֹל אָ֗נִי אָמַר֙ יְהֹוָ֣ה צְבָא֔וֹת וּשְׁמִ֖י נוֹרָ֥א בַגּוֹיִֽם׃}
{But cursed be he that dealeth craftily, Whereas he hath in his flock a male, And voweth, and sacrificeth unto the Lord a blemished thing; For I am a great King, Saith the \lord\space of hosts, And My name is feared among the nations.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{וְעַתָּ֗ה אֲלֵיכֶ֛ם הַמִּצְוָ֥ה הַזֹּ֖את הַכֹּהֲנִֽים׃}
{And now, this commandment Is for you, O ye priests.}
\twoverse{\arabic{verse}}
{אִם\maqqaf לֹ֣א תִשְׁמְע֡וּ וְאִם\maqqaf לֹא֩ תָשִׂ֨ימוּ עַל\maqqaf לֵ֜ב לָתֵ֧ת כָּב֣וֹד לִשְׁמִ֗י אָמַר֙ יְהֹוָ֣ה צְבָא֔וֹת וְשִׁלַּחְתִּ֤י בָכֶם֙ אֶת\maqqaf הַמְּאֵרָ֔ה וְאָרוֹתִ֖י אֶת\maqqaf בִּרְכוֹתֵיכֶ֑ם וְגַם֙ אָרוֹתִ֔יהָ כִּ֥י אֵינְכֶ֖ם שָׂמִ֥ים עַל\maqqaf לֵֽב׃}
{If ye will not hearken, and if ye will not lay it to heart, To give glory unto My name, Saith the \lord\space of hosts, Then will I send the curse upon you, And I will curse your blessings; Yea, I curse them, Because ye do not lay it to heart.}
\twoverse{\arabic{verse}}
{הִנְנִ֨י גֹעֵ֤ר לָכֶם֙ אֶת\maqqaf הַזֶּ֔רַע וְזֵרִ֤יתִי פֶ֙רֶשׁ֙ עַל\maqqaf פְּנֵיכֶ֔ם פֶּ֖רֶשׁ חַגֵּיכֶ֑ם וְנָשָׂ֥א אֶתְכֶ֖ם אֵלָֽיו׃}
{Behold, I will rebuke the seed for your hurt, And will spread dung upon your faces, Even the dung of your sacrifices; And ye shall be taken away unto it.}
\twoverse{\arabic{verse}}
{וִידַעְתֶּ֕ם כִּ֚י שִׁלַּ֣חְתִּי אֲלֵיכֶ֔ם אֵ֖ת הַמִּצְוָ֣ה הַזֹּ֑את לִהְי֤וֹת בְּרִיתִי֙ אֶת\maqqaf לֵוִ֔י אָמַ֖ר יְהֹוָ֥ה צְבָאֽוֹת׃}
{Know then that I have sent This commandment unto you, That My covenant might be with Levi, Saith the \lord\space of hosts.}
\twoverse{\arabic{verse}}
{בְּרִיתִ֣י \legarmeh  הָיְתָ֣ה אִתּ֗וֹ הַֽחַיִּים֙ וְהַשָּׁל֔וֹם וָאֶתְּנֵֽם\maqqaf ל֥וֹ מוֹרָ֖א וַיִּֽירָאֵ֑נִי וּמִפְּנֵ֥י שְׁמִ֖י נִחַ֥ת הֽוּא׃}
{My covenant was with him Of life and peace, and I gave them to him, And of fear, and he feared Me, And was afraid of My name.}
\twoverse{\arabic{verse}}
{תּוֹרַ֤ת אֱמֶת֙ הָיְתָ֣ה בְּפִ֔יהוּ וְעַוְלָ֖ה לֹֽא\maqqaf נִמְצָ֣א בִשְׂפָתָ֑יו בְּשָׁל֤וֹם וּבְמִישׁוֹר֙ הָלַ֣ךְ אִתִּ֔י וְרַבִּ֖ים הֵשִׁ֥יב מֵעָוֺֽן׃}
{The law of truth was in his mouth, And unrighteousness was not found in his lips; He walked with Me in peace and uprightness, And did turn many away from iniquity.}
\twoverse{\arabic{verse}}
{כִּֽי\maqqaf שִׂפְתֵ֤י כֹהֵן֙ יִשְׁמְרוּ\maqqaf דַ֔עַת וְתוֹרָ֖ה יְבַקְשׁ֣וּ מִפִּ֑יהוּ כִּ֛י מַלְאַ֥ךְ יְהֹֽוָה\maqqaf צְבָא֖וֹת הֽוּא׃}
{For the priest’s lips should keep knowledge, And they should seek the law at his mouth; For he is the messenger of the \lord\space of hosts.}
\newperek
\newhaftarah{ויצא}
\label{haft_7}
\haftnote{The Haftara is Hosea 12:13\verserangechar 14:10. }
\setcounter{chap}{12}
\setcounter{verse}{13}
\twoverse{\Roman{chap}:\arabic{verse}}
{וַיִּבְרַ֥ח יַעֲקֹ֖ב שְׂדֵ֣ה אֲרָ֑ם וַיַּעֲבֹ֤ד יִשְׂרָאֵל֙ בְּאִשָּׁ֔ה וּבְאִשָּׁ֖ה שָׁמָֽר׃}
{And Jacob fled into the field of Aram, And Israel served for a wife, And for a wife he kept sheep.}
\twoverse{\arabic{verse}}
{וּבְנָבִ֕יא הֶעֱלָ֧ה יְהֹוָ֛ה אֶת\maqqaf יִשְׂרָאֵ֖ל מִמִּצְרָ֑יִם וּבְנָבִ֖יא נִשְׁמָֽר׃}
{And by a prophet the \lord\space brought Israel up out of Egypt, And by a prophet was he kept.}
\twoverse{\arabic{verse}}
{הִכְעִ֥יס אֶפְרַ֖יִם תַּמְרוּרִ֑ים וְדָמָיו֙ עָלָ֣יו יִטּ֔וֹשׁ וְחֶ֨רְפָּת֔וֹ יָשִׁ֥יב ל֖וֹ אֲדֹנָֽיו׃}
{Ephraim hath provoked most bitterly; Therefore shall his blood be cast upon him, And his reproach shall his Lord return unto him.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{כְּדַבֵּ֤ר אֶפְרַ֙יִם֙ רְתֵ֔ת נָשָׂ֥א ה֖וּא בְּיִשְׂרָאֵ֑ל וַיֶּאְשַׁ֥ם בַּבַּ֖עַל וַיָּמֹֽת׃}
{When Ephraim spoke, there was trembling, He exalted himself in Israel; But when he became guilty through Baal, he died.}
\twoverse{\arabic{verse}}
{וְעַתָּ֣ה \legarmeh  יוֹסִ֣פוּ לַחֲטֹ֗א וַיַּעֲשׂ֣וּ לָהֶם֩ מַסֵּכָ֨ה מִכַּסְפָּ֤ם כִּתְבוּנָם֙ עֲצַבִּ֔ים מַעֲשֵׂ֥ה חָרָשִׁ֖ים כֻּלֹּ֑ה לָהֶם֙ הֵ֣ם אֹמְרִ֔ים זֹבְחֵ֣י אָדָ֔ם עֲגָלִ֖ים יִשָּׁקֽוּן׃}
{And now they sin more and more, And have made them molten images of their silver, According to their own understanding, even idols, All of them the work of the craftsmen; Of them they say: ‘They that sacrifice men kiss calves.’}
\twoverse{\arabic{verse}}
{לָכֵ֗ן יִֽהְיוּ֙ כַּעֲנַן\maqqaf בֹּ֔קֶר וְכַטַּ֖ל מַשְׁכִּ֣ים הֹלֵ֑ךְ כְּמֹץ֙ יְסֹעֵ֣ר מִגֹּ֔רֶן וּכְעָשָׁ֖ן מֵאֲרֻבָּֽה׃}
{Therefore they shall be as the morning cloud, And as the dew that early passeth away, As the chaff that is driven with the wind out of the threshing-floor, And as the smoke out of the window.}
\twoverse{\arabic{verse}}
{וְאָנֹכִ֛י יְהֹוָ֥ה אֱלֹהֶ֖יךָ מֵאֶ֣רֶץ מִצְרָ֑יִם וֵאלֹהִ֤ים זוּלָתִי֙ לֹ֣א תֵדָ֔ע וּמוֹשִׁ֥יעַ אַ֖יִן בִּלְתִּֽי׃}
{Yet I am the \lord\space thy God From the land of Egypt; And thou knowest no God but Me, And beside Me there is no saviour.}
\twoverse{\arabic{verse}}
{אֲנִ֥י יְדַעְתִּ֖יךָ בַּמִּדְבָּ֑ר בְּאֶ֖רֶץ תַּלְאֻבֽוֹת׃}
{I did know thee in the wilderness, In the land of great drought.}
\twoverse{\arabic{verse}}
{כְּמַרְעִיתָם֙ וַיִּשְׂבָּ֔עוּ שָׂבְע֖וּ וַיָּ֣רׇם לִבָּ֑ם עַל\maqqaf כֵּ֖ן שְׁכֵחֽוּנִי׃}
{When they were fed, they became full, They were filled, and their heart was exalted; Therefore have they forgotten Me.}
\twoverse{\arabic{verse}}
{וָאֱהִ֥י לָהֶ֖ם כְּמוֹ\maqqaf שָׁ֑חַל כְּנָמֵ֖ר עַל\maqqaf דֶּ֥רֶךְ אָשֽׁוּר׃}
{Therefore am I become unto them as a lion; As a leopard will I watch by the way;}
\twoverse{\arabic{verse}}
{אֶפְגְּשֵׁם֙ כְּדֹ֣ב שַׁכּ֔וּל וְאֶקְרַ֖ע סְג֣וֹר לִבָּ֑ם וְאֹכְלֵ֥ם שָׁם֙ כְּלָבִ֔יא חַיַּ֥ת הַשָּׂדֶ֖ה תְּבַקְּעֵֽם׃}
{I will meet them as a bear that is bereaved of her whelps, And will rend the enclosure of their heart; And there will I devour them like a lioness; The wild beast shall tear them.}
\twoverse{\arabic{verse}}
{שִׁחֶתְךָ֥ יִשְׂרָאֵ֖ל כִּי\maqqaf בִ֥י בְעֶזְרֶֽךָ׃}
{It is thy destruction, O Israel, That thou art against Me, against thy help.}
\twoverse{\arabic{verse}}
{אֱהִ֤י מַלְכְּךָ֙ אֵפ֔וֹא וְיוֹשִׁיעֲךָ֖ בְּכׇל\maqqaf עָרֶ֑יךָ וְשֹׁ֣פְטֶ֔יךָ אֲשֶׁ֣ר אָמַ֔רְתָּ תְּנָה\maqqaf לִּ֖י מֶ֥לֶךְ וְשָׂרִֽים׃}
{Ho, now, thy king, That he may save thee in all thy cities! And thy judges, of whom thou saidst: ‘Give me a king and princes!’}
\twoverse{\arabic{verse}}
{אֶֽתֶּן\maqqaf לְךָ֥ מֶ֙לֶךְ֙ בְּאַפִּ֔י וְאֶקַּ֖ח בְּעֶבְרָתִֽי׃ \petucha }
{I give thee a king in Mine anger, And take him away in My wrath.}
\twoverse{\arabic{verse}}
{צָרוּר֙ עֲוֺ֣ן אֶפְרָ֔יִם צְפוּנָ֖ה חַטָּאתֽוֹ׃}
{The iniquity of Ephraim is bound up; His sin is laid up in store.}
\twoverse{\arabic{verse}}
{חֶבְלֵ֥י יוֹלֵדָ֖ה יָבֹ֣אוּ ל֑וֹ הוּא\maqqaf בֵן֙ לֹ֣א חָכָ֔ם כִּי\maqqaf עֵ֥ת לֹֽא\maqqaf יַעֲמֹ֖ד בְּמִשְׁבַּ֥ר בָּנִֽים׃}
{The throes of a travailing woman shall come upon him; He is an unwise son; For it is time he should not tarry In the place of the breaking forth of children.}
\twoverse{\arabic{verse}}
{מִיַּ֤ד שְׁאוֹל֙ אֶפְדֵּ֔ם מִמָּ֖וֶת אֶגְאָלֵ֑ם אֱהִ֨י דְבָרֶ֜יךָ מָ֗וֶת אֱהִ֤י קָֽטׇבְךָ֙ שְׁא֔וֹל נֹ֖חַם יִסָּתֵ֥ר מֵעֵינָֽי׃}
{Shall I ransom them from the power of the nether-world? Shall I redeem them from death? Ho, thy plagues, O death! Ho, thy destruction, O netherworld! Repentance be hid from Mine eyes!}
\twoverse{\arabic{verse}}
{כִּ֣י ה֔וּא בֵּ֥ין אַחִ֖ים יַפְרִ֑יא יָב֣וֹא קָדִים֩ ר֨וּחַ יְהֹוָ֜ה מִמִּדְבָּ֣ר עֹלֶ֗ה וְיֵב֤וֹשׁ מְקוֹרוֹ֙ וְיֶחֱרַ֣ב מַעְיָנ֔וֹ ה֣וּא יִשְׁסֶ֔ה אוֹצַ֖ר כׇּל\maqqaf כְּלִ֥י חֶמְדָּֽה׃}
{For though he be fruitful among the reed-plants, An east wind shall come, the wind of the \lord\space coming up from the wilderness, And his spring shall become dry, and his fountain shall be dried up; He shall spoil the treasure of all precious vessels.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{תֶּאְשַׁם֙ שֹׁמְר֔וֹן כִּ֥י מָרְתָ֖ה בֵּאלֹהֶ֑יהָ בַּחֶ֣רֶב יִפֹּ֔לוּ עֹלְלֵיהֶ֣ם יְרֻטָּ֔שׁוּ וְהָרִיּוֹתָ֖יו יְבֻקָּֽעוּ׃ \petucha }
{Samaria shall bear her guilt, For she hath rebelled against her God; They shall fall by the sword; Their infants shall be dashed in pieces, And their women with child shall be ripped up.}
\twoverse{\arabic{verse}}
{שׁ֚וּבָה יִשְׂרָאֵ֔ל עַ֖ד יְהֹוָ֣ה אֱלֹהֶ֑יךָ כִּ֥י כָשַׁ֖לְתָּ בַּעֲוֺנֶֽךָ׃}
{Return, O Israel, unto the \lord\space thy God; For thou hast stumbled in thine iniquity.}
\twoverse{\arabic{verse}}
{קְח֤וּ עִמָּכֶם֙ דְּבָרִ֔ים וְשׁ֖וּבוּ אֶל\maqqaf יְהֹוָ֑ה אִמְר֣וּ אֵלָ֗יו כׇּל\maqqaf תִּשָּׂ֤א עָוֺן֙ וְקַח\maqqaf ט֔וֹב וּֽנְשַׁלְּמָ֥ה פָרִ֖ים שְׂפָתֵֽינוּ׃}
{Take with you words, And return unto the \lord; Say unto Him: ‘Forgive all iniquity, And accept that which is good; So will we render for bullocks the offering of our lips.}
\twoverse{\arabic{verse}}
{אַשּׁ֣וּר \legarmeh  לֹ֣א יוֹשִׁיעֵ֗נוּ עַל\maqqaf סוּס֙ לֹ֣א נִרְכָּ֔ב וְלֹא\maqqaf נֹ֥אמַר ע֛וֹד אֱלֹהֵ֖ינוּ לְמַעֲשֵׂ֣ה יָדֵ֑ינוּ אֲשֶׁר\maqqaf בְּךָ֖ יְרֻחַ֥ם יָתֽוֹם׃}
{Asshur shall not save us; We will not ride upon horses; Neither will we call any more the work of our hands our gods; For in Thee the fatherless findeth mercy.’}
\twoverse{\arabic{verse}}
{אֶרְפָּא֙ מְשׁ֣וּבָתָ֔ם אֹהֲבֵ֖ם נְדָבָ֑ה כִּ֛י שָׁ֥ב אַפִּ֖י מִמֶּֽנּוּ׃}
{I will heal their backsliding, I will love them freely; For Mine anger is turned away from him.}
\twoverse{\arabic{verse}}
{אֶהְיֶ֤ה כַטַּל֙ לְיִשְׂרָאֵ֔ל יִפְרַ֖ח כַּשּׁוֹשַׁנָּ֑ה וְיַ֥ךְ שׇׁרָשָׁ֖יו כַּלְּבָנֽוֹן׃}
{I will be as the dew unto Israel; He shall blossom as the lily, And cast forth his roots as Lebanon.}
\twoverse{\arabic{verse}}
{יֵֽלְכוּ֙ יֹֽנְקוֹתָ֔יו וִיהִ֥י כַזַּ֖יִת הוֹד֑וֹ וְרֵ֥יחַֽ ל֖וֹ כַּלְּבָנֽוֹן׃}
{His branches shall spread, And his beauty shall be as the olive-tree, And his fragrance as Lebanon.}
\twoverse{\arabic{verse}}
{יָשֻׁ֙בוּ֙ יֹשְׁבֵ֣י בְצִלּ֔וֹ יְחַיּ֥וּ דָגָ֖ן וְיִפְרְח֣וּ כַגָּ֑פֶן זִכְר֖וֹ כְּיֵ֥ין לְבָנֽוֹן׃}
{They that dwell under his shadow shall again Make corn to grow, And shall blossom as the vine; The scent thereof shall be as the wine of Lebanon.}
\twoverse{\arabic{verse}}
{אֶפְרַ֕יִם מַה\maqqaf לִּ֥י ע֖וֹד לָעֲצַבִּ֑ים אֲנִ֧י עָנִ֣יתִי וַאֲשׁוּרֶ֗נּוּ אֲנִי֙ כִּבְר֣וֹשׁ רַעֲנָ֔ן מִמֶּ֖נִּי פֶּרְיְךָ֥ נִמְצָֽא׃}
{Ephraim [shall say]: ‘What have I to do any more with idols?’ As for Me, I respond and look on him; I am like a leafy cypress-tree; From Me is thy fruit found.}
\twoverse{\arabic{verse}}
{מִ֤י חָכָם֙ וְיָ֣בֵֽן אֵ֔לֶּה נָב֖וֹן וְיֵדָעֵ֑ם כִּֽי\maqqaf יְשָׁרִ֞ים דַּרְכֵ֣י יְהֹוָ֗ה וְצַדִּקִים֙ יֵ֣לְכוּ בָ֔ם וּפֹשְׁעִ֖ים יִכָּ֥שְׁלוּ בָֽם׃}
{Whoso is wise, let him understand these things, Whoso is prudent, let him know them. For the ways of the \lord\space are right, And the just do walk in them; But transgressors do stumble therein.}
\newperek
\newhaftarah{וישלח}
\label{haft_8}
\haftnote{The Haftara is Obadiah 1:1\verserangechar 1:21. }
\setcounter{chap}{1}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}
{חֲז֖וֹן עֹֽבַדְיָ֑ה כֹּֽה\maqqaf אָמַר֩ אֲדֹנָ֨י יֱהֹוִ֜ה לֶאֱד֗וֹם שְׁמוּעָ֨ה שָׁמַ֜עְנוּ מֵאֵ֤ת יְהֹוָה֙ וְצִיר֙ בַּגּוֹיִ֣ם שֻׁלָּ֔ח ק֛וּמוּ וְנָק֥וּמָה עָלֶ֖יהָ לַמִּלְחָמָֽה׃}
{The vision of Obadiah. Thus saith the Lord \textsc{God} concerning Edom: We have heard a message from the \lord, And an ambassador is sent among the nations: ‘Arise ye, and let us rise up against her in battle.’}
\twoverse{\arabic{verse}}
{הִנֵּ֥ה קָטֹ֛ן נְתַתִּ֖יךָ בַּגּוֹיִ֑ם בָּז֥וּי אַתָּ֖ה מְאֹֽד׃}
{Behold, I make thee small among the nations; Thou art greatly despised.}
\twoverse{\arabic{verse}}
{זְד֤וֹן לִבְּךָ֙ הִשִּׁיאֶ֔ךָ שֹׁכְנִ֥י בְחַגְוֵי\maqqaf סֶ֖לַע מְר֣וֹם שִׁבְתּ֑וֹ אֹמֵ֣ר בְּלִבּ֔וֹ מִ֥י יוֹרִדֵ֖נִי אָֽרֶץ׃}
{The pride of thy heart hath beguiled thee, O thou that dwellest in the clefts of the rock, Thy habitation on high; That sayest in thy heart: ‘Who shall bring me down to the ground?’}
\twoverse{\arabic{verse}}
{אִם\maqqaf תַּגְבִּ֣יהַּ כַּנֶּ֔שֶׁר וְאִם\maqqaf בֵּ֥ין כּֽוֹכָבִ֖ים שִׂ֣ים קִנֶּ֑ךָ מִשָּׁ֥ם אוֹרִֽידְךָ֖ נְאֻם\maqqaf יְהֹוָֽה׃}
{Though thou make thy nest as high as the eagle, And though thou set it among the stars, I will bring thee down from thence, saith the \lord.}
\twoverse{\arabic{verse}}
{אִם\maqqaf גַּנָּבִ֤ים בָּאֽוּ\maqqaf לְךָ֙ אִם\maqqaf שׁ֣וֹדְדֵי לַ֔יְלָה אֵ֣יךְ נִדְמֵ֔יתָה הֲל֥וֹא יִגְנְב֖וּ דַּיָּ֑ם אִם\maqqaf בֹּֽצְרִים֙ בָּ֣אוּ לָ֔ךְ הֲל֖וֹא יַשְׁאִ֥ירוּ עֹלֵלֽוֹת׃}
{If thieves came to thee, if robbers by night— How art thou cut off!— Would they not steal till they had enough? If grape-gatherers came to thee, Would they not leave some gleaning grapes?}
\twoverse{\arabic{verse}}
{אֵ֚יךְ נֶחְפְּשׂ֣וּ עֵשָׂ֔ו נִבְע֖וּ מַצְפֻּנָֽיו׃}
{How is Esau searched out! How are his hidden places sought out!}
\twoverse{\arabic{verse}}
{עַֽד\maqqaf הַגְּב֣וּל שִׁלְּח֗וּךָ כֹּ֚ל אַנְשֵׁ֣י בְרִיתֶ֔ךָ הִשִּׁיא֛וּךָ יָכְל֥וּ לְךָ֖ אַנְשֵׁ֣י שְׁלֹמֶ֑ךָ לַחְמְךָ֗ יָשִׂ֤ימוּ מָזוֹר֙ תַּחְתֶּ֔יךָ אֵ֥ין תְּבוּנָ֖ה בּֽוֹ׃}
{All the men of thy confederacy Have conducted thee to the border; The men that were at peace with thee Have beguiled thee, and prevailed against thee; They that eat thy bread lay a snare under thee, In whom there is no discernment.}
\twoverse{\arabic{verse}}
{הֲל֛וֹא בַּיּ֥וֹם הַה֖וּא נְאֻם\maqqaf יְהֹוָ֑ה וְהַאֲבַדְתִּ֤י חֲכָמִים֙ מֵֽאֱד֔וֹם וּתְבוּנָ֖ה מֵהַ֥ר עֵשָֽׂו׃}
{Shall I not in that day, saith the \lord, Destroy the wise men out of Edom, And discernment out of the mount of Esau?}
\twoverse{\arabic{verse}}
{וְחַתּ֥וּ גִבּוֹרֶ֖יךָ תֵּימָ֑ן לְמַ֧עַן יִכָּֽרֶת\maqqaf אִ֛ישׁ מֵהַ֥ר עֵשָׂ֖ו מִקָּֽטֶל׃}
{And thy mighty men, O Teman, shall be dismayed, To the end that every one may be cut off from the mount of Esau by slaughter.}
\twoverse{\arabic{verse}}
{מֵחֲמַ֛ס אָחִ֥יךָ יַעֲקֹ֖ב תְּכַסְּךָ֣ בוּשָׁ֑ה וְנִכְרַ֖תָּ לְעוֹלָֽם׃}
{For the violence done to thy brother Jacob shame shall cover thee, And thou shalt be cut off for ever.}
\twoverse{\arabic{verse}}
{בְּיוֹם֙ עֲמׇֽדְךָ֣ מִנֶּ֔גֶד בְּי֛וֹם שְׁב֥וֹת זָרִ֖ים חֵיל֑וֹ וְנׇכְרִ֞ים בָּ֣אוּ שְׁעָרָ֗ו וְעַל\maqqaf יְרוּשָׁלַ֙͏ִם֙ יַדּ֣וּ גוֹרָ֔ל גַּם\maqqaf אַתָּ֖ה כְּאַחַ֥ד מֵהֶֽם׃}
{In the day that thou didst stand aloof, In the day that strangers carried away his substance, And foreigners entered into his gates, And cast lots upon Jerusalem, Even thou wast as one of them.}
\twoverse{\arabic{verse}}
{וְאַל\maqqaf תֵּ֤רֶא בְיוֹם\maqqaf אָחִ֙יךָ֙ בְּי֣וֹם נׇכְר֔וֹ וְאַל\maqqaf תִּשְׂמַ֥ח לִבְנֵֽי\maqqaf יְהוּדָ֖ה בְּי֣וֹם אׇבְדָ֑ם וְאַל\maqqaf תַּגְדֵּ֥ל פִּ֖יךָ בְּי֥וֹם צָרָֽה׃}
{But thou shouldest not have gazed on the day of thy brother In the day of his disaster, Neither shouldest thou have rejoiced over the children of Judah In the day of their destruction; Neither shouldest thou have spoken proudly In the day of distress.}
\twoverse{\arabic{verse}}
{אַל\maqqaf תָּב֤וֹא בְשַֽׁעַר\maqqaf עַמִּי֙ בְּי֣וֹם אֵידָ֔ם אַל\maqqaf תֵּ֧רֶא גַם\maqqaf אַתָּ֛ה בְּרָעָת֖וֹ בְּי֣וֹם אֵיד֑וֹ וְאַל\maqqaf תִּשְׁלַ֥חְנָה בְחֵיל֖וֹ בְּי֥וֹם אֵידֽוֹ׃}
{Thou shouldest not have entered into the gate of My people In the day of their calamity; Yea, thou shouldest not have gazed on their affliction In the day of their calamity, Nor have laid hands on their substance In the day of their calamity.}
\twoverse{\arabic{verse}}
{וְאַֽל\maqqaf תַּעֲמֹד֙ עַל\maqqaf הַפֶּ֔רֶק לְהַכְרִ֖ית אֶת\maqqaf פְּלִיטָ֑יו וְאַל\maqqaf תַּסְגֵּ֥ר שְׂרִידָ֖יו בְּי֥וֹם צָרָֽה׃}
{Neither shouldest thou have stood in the crossway, To cut off those of his that escape; Neither shouldest thou have delivered up those of his That did remain in the day of distress.}
\twoverse{\arabic{verse}}
{כִּֽי\maqqaf קָר֥וֹב יוֹם\maqqaf יְהֹוָ֖ה עַל\maqqaf כׇּל\maqqaf הַגּוֹיִ֑ם כַּאֲשֶׁ֤ר עָשִׂ֙יתָ֙ יֵעָ֣שֶׂה לָּ֔ךְ גְּמֻלְךָ֖ יָשׁ֥וּב בְּרֹאשֶֽׁךָ׃}
{For the day of the \lord\space is near upon all the nations; As thou hast done, it shall be done unto thee; Thy dealing shall return upon thine own head.}
\twoverse{\arabic{verse}}
{כִּ֗י כַּֽאֲשֶׁ֤ר שְׁתִיתֶם֙ עַל\maqqaf הַ֣ר קׇדְשִׁ֔י יִשְׁתּ֥וּ כׇֽל\maqqaf הַגּוֹיִ֖ם תָּמִ֑יד וְשָׁת֣וּ וְלָע֔וּ וְהָי֖וּ כְּל֥וֹא הָיֽוּ׃}
{For as ye have drunk upon My holy mountain, So shall all the nations drink continually, Yea, they shall drink, and swallow down, And shall be as though they had not been.}
\twoverse{\arabic{verse}}
{וּבְהַ֥ר צִיּ֛וֹן תִּהְיֶ֥ה פְלֵיטָ֖ה וְהָ֣יָה קֹ֑דֶשׁ וְיָֽרְשׁוּ֙ בֵּ֣ית יַֽעֲקֹ֔ב אֵ֖ת מוֹרָֽשֵׁיהֶֽם׃}
{But in mount Zion there shall be those that escape, And it shall be holy; And the house of Jacob shall possess their possessions.}
\twoverse{\arabic{verse}}
{וְהָיָה֩ בֵית\maqqaf יַעֲקֹ֨ב אֵ֜שׁ וּבֵ֧ית יוֹסֵ֣ף לֶהָבָ֗ה וּבֵ֤ית עֵשָׂו֙ לְקַ֔שׁ וְדָלְק֥וּ בָהֶ֖ם וַאֲכָל֑וּם וְלֹֽא\maqqaf יִֽהְיֶ֤ה שָׂרִיד֙ לְבֵ֣ית עֵשָׂ֔ו כִּ֥י יְהֹוָ֖ה דִּבֵּֽר׃}
{And the house of Jacob shall be a fire, And the house of Joseph a flame, And the house of Esau for stubble, And they shall kindle in them, and devour them; And there shall not be any remaining of the house of Esau; For the \lord\space hath spoken.}
\twoverse{\arabic{verse}}
{וְיָרְשׁ֨וּ הַנֶּ֜גֶב אֶת\maqqaf הַ֣ר עֵשָׂ֗ו וְהַשְּׁפֵלָה֙ אֶת\maqqaf פְּלִשְׁתִּ֔ים וְיָרְשׁוּ֙ אֶת\maqqaf שְׂדֵ֣ה אֶפְרַ֔יִם וְאֵ֖ת שְׂדֵ֣ה שֹׁמְר֑וֹן וּבִנְיָמִ֖ן אֶת\maqqaf הַגִּלְעָֽד׃}
{And they of the South shall possess the mount of Esau, And they of the Lowland the Philistines; And they shall possess the field of Ephraim, And the field of Samaria; And Benjamin shall possess Gilead.}
\twoverse{\arabic{verse}}
{וְגָלֻ֣ת הַֽחֵל\maqqaf הַ֠זֶּ֠ה לִבְנֵ֨י יִשְׂרָאֵ֤ל אֲשֶֽׁר\maqqaf כְּנַעֲנִים֙ עַד\maqqaf צָ֣רְפַ֔ת וְגָלֻ֥ת יְרוּשָׁלַ֖͏ִם אֲשֶׁ֣ר בִּסְפָרַ֑ד יִֽרְשׁ֕וּ אֵ֖ת עָרֵ֥י הַנֶּֽגֶב׃}
{And the captivity of this host of the children of Israel, That are among the Canaanites, even unto Zarephath, And the captivity of Jerusalem, that is in Sepharad, Shall possess the cities of the South.}
\twoverse{\arabic{verse}}
{וְעָל֤וּ מֽוֹשִׁעִים֙ בְּהַ֣ר צִיּ֔וֹן לִשְׁפֹּ֖ט אֶת\maqqaf הַ֣ר עֵשָׂ֑ו וְהָיְתָ֥ה לַֽיהֹוָ֖ה הַמְּלוּכָֽה׃}
{And saviours shall come up on mount Zion To judge the mount of Esau; And the kingdom shall be the \lord\textsc{’s}.}
\newhaftarah{וישב}
\label{haft_9}
\haftnote{The Haftara is Amos 2:6\verserangechar 3:8. On \d{H}anukka, read the Maftir on page \pageref{maft_hanukka}. See there for notes on which Maftir to read.  The Haftara is on page \pageref{haft_hanukka1}.}
\setcounter{chap}{2}
\setcounter{verse}{6}
\twoverse{\Roman{chap}:\arabic{verse}}
{כֹּ֚ה אָמַ֣ר יְהֹוָ֔ה עַל\maqqaf שְׁלֹשָׁה֙ פִּשְׁעֵ֣י יִשְׂרָאֵ֔ל וְעַל\maqqaf אַרְבָּעָ֖ה לֹ֣א אֲשִׁיבֶ֑נּוּ עַל\maqqaf מִכְרָ֤ם בַּכֶּ֙סֶף֙ צַדִּ֔יק וְאֶבְי֖וֹן בַּֽעֲב֥וּר נַעֲלָֽיִם׃}
{Thus saith the \lord: For three transgressions of Israel, Yea, for four, I will not reverse it: Because they sell the righteous for silver, And the needy for a pair of shoes;}
\twoverse{\arabic{verse}}
{הַשֹּׁאֲפִ֤ים עַל\maqqaf עֲפַר\maqqaf אֶ֙רֶץ֙ בְּרֹ֣אשׁ דַּלִּ֔ים וְדֶ֥רֶךְ עֲנָוִ֖ים יַטּ֑וּ וְאִ֣ישׁ וְאָבִ֗יו יֵֽלְכוּ֙ אֶל\maqqaf הַֽנַּעֲרָ֔ה לְמַ֥עַן חַלֵּ֖ל אֶת\maqqaf שֵׁ֥ם קׇדְשִֽׁי׃}
{That pant after the dust of the earth on the head of the poor, And turn aside the way of the humble; And a man and his father go unto the same maid, To profane My holy name;}
\twoverse{\arabic{verse}}
{וְעַל\maqqaf בְּגָדִ֤ים חֲבֻלִים֙ יַטּ֔וּ אֵ֖צֶל כׇּל\maqqaf מִזְבֵּ֑חַ וְיֵ֤ין עֲנוּשִׁים֙ יִשְׁתּ֔וּ בֵּ֖ית אֱלֹהֵיהֶֽם׃}
{And they lay themselves down beside every altar Upon clothes taken in pledge, And in the house of their God they drink The wine of them that have been fined.}
\twoverse{\arabic{verse}}
{וְאָ֨נֹכִ֜י הִשְׁמַ֤דְתִּי אֶת\maqqaf הָאֱמֹרִי֙ מִפְּנֵיהֶ֔ם אֲשֶׁ֨ר כְּגֹ֤בַהּ אֲרָזִים֙ גׇּבְה֔וֹ וְחָסֹ֥ן ה֖וּא כָּאַלּוֹנִ֑ים וָאַשְׁמִ֤יד פִּרְיוֹ֙ מִמַּ֔עַל וְשׇׁרָשָׁ֖יו מִתָּֽחַת׃}
{Yet destroyed I the Amorite before them, Whose height was like the height of the cedars, And he was strong as the oaks; Yet I destroyed his fruit from above, And his roots from beneath.}
\twoverse{\arabic{verse}}
{וְאָנֹכִ֛י הֶעֱלֵ֥יתִי אֶתְכֶ֖ם מֵאֶ֣רֶץ מִצְרָ֑יִם וָאוֹלֵ֨ךְ אֶתְכֶ֤ם בַּמִּדְבָּר֙ אַרְבָּעִ֣ים שָׁנָ֔ה לָרֶ֖שֶׁת אֶת\maqqaf אֶ֥רֶץ הָאֱמֹרִֽי׃}
{Also I brought you up out of the land of Egypt, And led you forty years in the wilderness, To possess the land of the Amorites.}
\twoverse{\arabic{verse}}
{וָאָקִ֤ים מִבְּנֵיכֶם֙ לִנְבִיאִ֔ים וּמִבַּחוּרֵיכֶ֖ם לִנְזִרִ֑ים הַאַ֥ף אֵֽין\maqqaf זֹ֛את בְּנֵ֥י יִשְׂרָאֵ֖ל נְאֻם\maqqaf יְהֹוָֽה׃}
{And I raised up of your sons for prophets, And of your young men for Nazirites. Is it not even thus, O ye children of Israel? Saith the \lord.}
\twoverse{\arabic{verse}}
{וַתַּשְׁק֥וּ אֶת\maqqaf הַנְּזִרִ֖ים יָ֑יִן וְעַל\maqqaf הַנְּבִיאִים֙ צִוִּיתֶ֣ם לֵאמֹ֔ר לֹ֖א תִּנָּבְאֽוּ׃}
{But ye gave the Nazirites wine to drink; And commanded the prophets, saying: ‘Prophesy not.’}
\twoverse{\arabic{verse}}
{הִנֵּ֛ה אָנֹכִ֥י מֵעִ֖יק תַּחְתֵּיכֶ֑ם כַּאֲשֶׁ֤ר תָּעִיק֙ הָֽעֲגָלָ֔ה הַֽמְלֵאָ֥ה לָ֖הּ עָמִֽיר׃}
{Behold, I will make it creak under you, As a cart creaketh that is full of sheaves.}
\twoverse{\arabic{verse}}
{וְאָבַ֤ד מָנוֹס֙ מִקָּ֔ל וְחָזָ֖ק לֹא\maqqaf יְאַמֵּ֣ץ כֹּח֑וֹ וְגִבּ֖וֹר לֹא\maqqaf יְמַלֵּ֥ט נַפְשֽׁוֹ׃}
{And flight shall fail the swift, And the strong shall not exert his strength, Neither shall the mighty deliver himself;}
\twoverse{\arabic{verse}}
{וְתֹפֵ֤שׂ הַקֶּ֙שֶׁת֙ לֹ֣א יַעֲמֹ֔ד וְקַ֥ל בְּרַגְלָ֖יו לֹ֣א יְמַלֵּ֑ט וְרֹכֵ֣ב הַסּ֔וּס לֹ֥א יְמַלֵּ֖ט נַפְשֽׁוֹ׃}
{Neither shall he stand that handleth the bow; And he that is swift of foot shall not deliver himself; Neither shall he that rideth the horse deliver himself;}
\twoverse{\arabic{verse}}
{וְאַמִּ֥יץ לִבּ֖וֹ בַּגִּבּוֹרִ֑ים עָר֛וֹם יָנ֥וּס בַּיּוֹם\maqqaf הַה֖וּא נְאֻם\maqqaf יְהֹוָֽה׃ \petucha }
{And he that is courageous among the mighty Shall flee away naked in that day, Saith the \lord.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{שִׁמְע֞וּ אֶת\maqqaf הַדָּבָ֣ר הַזֶּ֗ה אֲשֶׁ֨ר דִּבֶּ֧ר יְהֹוָ֛ה עֲלֵיכֶ֖ם בְּנֵ֣י יִשְׂרָאֵ֑ל עַ֚ל כׇּל\maqqaf הַמִּשְׁפָּחָ֔ה אֲשֶׁ֧ר הֶעֱלֵ֛יתִי מֵאֶ֥רֶץ מִצְרַ֖יִם לֵאמֹֽר׃}
{Hear this word that the \lord\space hath spoken against you, O children of Israel, against the whole family which I brought up out of the land of Egypt, saying:}
\twoverse{\arabic{verse}}
{רַ֚ק אֶתְכֶ֣ם יָדַ֔עְתִּי מִכֹּ֖ל מִשְׁפְּח֣וֹת הָאֲדָמָ֑ה עַל\maqqaf כֵּן֙ אֶפְקֹ֣ד עֲלֵיכֶ֔ם אֵ֖ת כׇּל\maqqaf עֲוֺנֹתֵיכֶֽם׃}
{You only have I known of all the families of the earth; Therefore I will visit upon you all your iniquities.}
\twoverse{\arabic{verse}}
{הֲיֵלְכ֥וּ שְׁנַ֖יִם יַחְדָּ֑ו בִּלְתִּ֖י אִם\maqqaf נוֹעָֽדוּ׃}
{Will two walk together, Except they have agreed?}
\twoverse{\arabic{verse}}
{הֲיִשְׁאַ֤ג אַרְיֵה֙ בַּיַּ֔עַר וְטֶ֖רֶף אֵ֣ין ל֑וֹ הֲיִתֵּ֨ן כְּפִ֤יר קוֹלוֹ֙ מִמְּעֹ֣נָת֔וֹ בִּלְתִּ֖י אִם\maqqaf לָכָֽד׃}
{Will a lion roar in the forest, When he hath no prey? Will a young lion give forth his voice out of his den, If he have taken nothing?}
\twoverse{\arabic{verse}}
{הֲתִפֹּ֤ל צִפּוֹר֙ עַל\maqqaf פַּ֣ח הָאָ֔רֶץ וּמוֹקֵ֖שׁ אֵ֣ין לָ֑הּ הֲיַֽעֲלֶה\maqqaf פַּח֙ מִן\maqqaf הָ֣אֲדָמָ֔ה וְלָכ֖וֹד לֹ֥א יִלְכּֽוֹד׃}
{Will a bird fall in a snare upon the earth, Where there is no lure for it? Will a snare spring up from the ground, And have taken nothing at all?}
\twoverse{\arabic{verse}}
{אִם\maqqaf יִתָּקַ֤ע שׁוֹפָר֙ בְּעִ֔יר וְעָ֖ם לֹ֣א יֶחֱרָ֑דוּ אִם\maqqaf תִּֽהְיֶ֤ה רָעָה֙ בְּעִ֔יר וַיהֹוָ֖ה לֹ֥א עָשָֽׂה׃}
{Shall the horn be blown in a city, And the people not tremble? Shall evil befall a city, And the \lord\space hath not done it?}
\twoverse{\arabic{verse}}
{כִּ֣י לֹ֧א יַעֲשֶׂ֛ה אֲדֹנָ֥י יֱהֹוִ֖ה דָּבָ֑ר כִּ֚י אִם\maqqaf גָּלָ֣ה סוֹד֔וֹ אֶל\maqqaf עֲבָדָ֖יו הַנְּבִיאִֽים׃}
{For the Lord \textsc{God} will do nothing, But He revealeth His counsel unto His servants the prophets.}
\twoverse{\arabic{verse}}
{אַרְיֵ֥ה שָׁאָ֖ג מִ֣י לֹ֣א יִירָ֑א אֲדֹנָ֤י יֱהֹוִה֙ דִּבֶּ֔ר מִ֖י לֹ֥א יִנָּבֵֽא׃}
{The lion hath roared, Who will not fear? The Lord \textsc{God} hath spoken, Who can but prophesy?}
\newperek
\newhaftarah{מקץ}
\label{haft_10}
\haftnote{The Haftara is I Kings 3:15\verserangechar 4:1. On \d{H}anukka, read the Maftir on page \pageref{maft_hanukka}. See there for notes on which Maftir to read.  The Haftara for the first Shabbat on \d{H}anukka is on page \pageref{haft_hanukka1}, and for the second Shabbat on page \pageref{haft_hanukka2}. If Shabbat and \d{H}anukka coincides with Rosh \d{H}odesh, combine the 6th and 7th aliyot of the parsha, read the Maftir for Rosh \d{H}odesh as the 7th aliya on page \pageref{maft_ShabRCh}, and read the Maftir and Haftara for Shabbat \d{H}anukka.}
\setcounter{chap}{3}
\setcounter{verse}{15}
\twoverse{\Roman{chap}:\arabic{verse}}
{וַיִּקַ֥ץ שְׁלֹמֹ֖ה וְהִנֵּ֣ה חֲל֑וֹם וַיָּב֨וֹא יְרֽוּשָׁלַ֜͏ִם וַֽיַּעֲמֹ֣ד \legarmeh  לִפְנֵ֣י \legarmeh  אֲר֣וֹן בְּרִית\maqqaf אֲדֹנָ֗י וַיַּ֤עַל עֹלוֹת֙ וַיַּ֣עַשׂ שְׁלָמִ֔ים וַיַּ֥עַשׂ מִשְׁתֶּ֖ה לְכׇל\maqqaf עֲבָדָֽיו׃ \petucha }
{And Solomon awoke, and, behold, it was a dream; and he came to Jerusalem, and stood before the ark of the covenant of the \lord, and offered up burnt-offerings, and offered peace-offerings, and made a feast to all his servants.}
\twoverse{\arabic{verse}}
{אָ֣ז תָּבֹ֗אנָה שְׁתַּ֛יִם נָשִׁ֥ים זֹנ֖וֹת אֶל\maqqaf הַמֶּ֑לֶךְ וַֽתַּעֲמֹ֖דְנָה לְפָנָֽיו׃}
{Then came there two women, that were harlots, unto the king, and stood before him.}
\twoverse{\arabic{verse}}
{וַתֹּ֜אמֶר הָאִשָּׁ֤ה הָאַחַת֙ בִּ֣י אֲדֹנִ֔י אֲנִי֙ וְהָאִשָּׁ֣ה הַזֹּ֔את יֹשְׁבֹ֖ת בְּבַ֣יִת אֶחָ֑ד וָאֵלֵ֥ד עִמָּ֖הּ בַּבָּֽיִת׃}
{And the one woman said: ‘Oh, my lord, I and this woman dwell in one house; and I was delivered of a child with her in the house.}
\twoverse{\arabic{verse}}
{וַיְהִ֞י בַּיּ֤וֹם הַשְּׁלִישִׁי֙ לְלִדְתִּ֔י וַתֵּ֖לֶד גַּם\maqqaf הָאִשָּׁ֣ה הַזֹּ֑את וַאֲנַ֣חְנוּ יַחְדָּ֗ו אֵֽין\maqqaf זָ֤ר אִתָּ֙נוּ֙ בַּבַּ֔יִת זוּלָתִ֥י שְׁתַּֽיִם\maqqaf אֲנַ֖חְנוּ בַּבָּֽיִת׃}
{And it came to pass the third day after I was delivered, that this woman was delivered also; and we were together; there was no stranger with us in the house, save we two in the house.}
\twoverse{\arabic{verse}}
{וַיָּ֛מׇת בֶּן\maqqaf הָאִשָּׁ֥ה הַזֹּ֖את לָ֑יְלָה אֲשֶׁ֥ר שָׁכְבָ֖ה עָלָֽיו׃}
{And this woman’s child died in the night; because she overlay it.}
\twoverse{\arabic{verse}}
{וַתָּ֩קׇם֩ בְּת֨וֹךְ הַלַּ֜יְלָה וַתִּקַּ֧ח אֶת\maqqaf בְּנִ֣י מֵאֶצְלִ֗י וַאֲמָֽתְךָ֙ יְשֵׁנָ֔ה וַתַּשְׁכִּיבֵ֖הוּ בְּחֵיקָ֑הּ וְאֶת\maqqaf בְּנָ֥הּ הַמֵּ֖ת הִשְׁכִּ֥יבָה בְחֵיקִֽי׃}
{And she arose at midnight, and took my son from beside me, while thy handmaid slept, and laid it in her bosom, and laid her dead child in my bosom.}
\twoverse{\arabic{verse}}
{וָאָקֻ֥ם בַּבֹּ֛קֶר לְהֵינִ֥יק אֶת\maqqaf בְּנִ֖י וְהִנֵּה\maqqaf מֵ֑ת וָאֶתְבּוֹנֵ֤ן אֵלָיו֙ בַּבֹּ֔קֶר וְהִנֵּ֛ה לֹא\maqqaf הָיָ֥ה בְנִ֖י אֲשֶׁ֥ר יָלָֽדְתִּי׃}
{And when I rose in the morning to give my child suck, behold, it was dead; but when I had looked well at it in the morning, behold, it was not my son, whom I did bear.’}
\twoverse{\arabic{verse}}
{וַתֹּ֩אמֶר֩ הָאִשָּׁ֨ה הָאַחֶ֜רֶת לֹ֣א כִ֗י בְּנִ֤י הַחַי֙ וּבְנֵ֣ךְ הַמֵּ֔ת וְזֹ֤את אֹמֶ֙רֶת֙ לֹ֣א כִ֔י בְּנֵ֥ךְ הַמֵּ֖ת וּבְנִ֣י הֶחָ֑י וַתְּדַבֵּ֖רְנָה לִפְנֵ֥י הַמֶּֽלֶךְ׃}
{And the other woman said: ‘Nay; but the living is my son, and the dead is thy son.’ And this said: ‘No; but the dead is thy son, and the living is my son.’ Thus they spoke before the king.}
\twoverse{\arabic{verse}}
{וַיֹּ֣אמֶר הַמֶּ֔לֶךְ זֹ֣את אֹמֶ֔רֶת זֶה\maqqaf בְּנִ֥י הַחַ֖י וּבְנֵ֣ךְ הַמֵּ֑ת וְזֹ֤את אֹמֶ֙רֶת֙ לֹ֣א כִ֔י בְּנֵ֥ךְ הַמֵּ֖ת וּבְנִ֥י הֶחָֽי׃ \petucha }
{Then said the king: ‘The one saith: This is my son that liveth, and thy son is the dead; and the other saith: Nay; but thy son is the dead, and my son is the living.’}
\twoverse{\arabic{verse}}
{וַיֹּ֥אמֶר הַמֶּ֖לֶךְ קְח֣וּ לִי\maqqaf חָ֑רֶב וַיָּבִ֥אוּ הַחֶ֖רֶב לִפְנֵ֥י הַמֶּֽלֶךְ׃}
{And the king said: ‘Fetch me a sword.’ And they brought a sword before the king.}
\twoverse{\arabic{verse}}
{וַיֹּ֣אמֶר הַמֶּ֔לֶךְ גִּזְר֛וּ אֶת\maqqaf הַיֶּ֥לֶד הַחַ֖י לִשְׁנָ֑יִם וּתְנ֤וּ אֶֽת\maqqaf הַחֲצִי֙ לְאַחַ֔ת וְאֶֽת\maqqaf הַחֲצִ֖י לְאֶחָֽת׃}
{And the king said: ‘Divide the living child in two, and give half to the one, and half to the other.’}
\twoverse{\arabic{verse}}
{וַתֹּ֣אמֶר הָאִשָּׁה֩ אֲשֶׁר\maqqaf בְּנָ֨הּ הַחַ֜י אֶל\maqqaf הַמֶּ֗לֶךְ כִּֽי\maqqaf נִכְמְר֣וּ רַֽחֲמֶ֘יהָ֮ עַל\maqqaf בְּנָהּ֒ וַתֹּ֣אמֶר \legarmeh  בִּ֣י אֲדֹנִ֗י תְּנוּ\maqqaf לָהּ֙ אֶת\maqqaf הַיָּל֣וּד הַחַ֔י וְהָמֵ֖ת אַל\maqqaf תְּמִיתֻ֑הוּ וְזֹ֣את אֹמֶ֗רֶת גַּם\maqqaf לִ֥י גַם\maqqaf לָ֛ךְ לֹ֥א יִהְיֶ֖ה גְּזֹֽרוּ׃}
{Then spoke the woman whose the living child was unto the king, for her heart yearned upon her son, and she said: ‘Oh, my lord, give her the living child, and in no wise slay it.’ But the other said: ‘It shall be neither mine nor thine; divide it.’}
\twoverse{\arabic{verse}}
{וַיַּ֨עַן הַמֶּ֜לֶךְ וַיֹּ֗אמֶר תְּנוּ\maqqaf לָהּ֙ אֶת\maqqaf הַיָּל֣וּד הַחַ֔י וְהָמֵ֖ת לֹ֣א תְמִיתֻ֑הוּ הִ֖יא אִמּֽוֹ׃ \setuma }
{Then the king answered and said: ‘Give her the living child, and in no wise slay it: she is the mother thereof.’}
\twoverse{\arabic{verse}}
{וַיִּשְׁמְע֣וּ כׇל\maqqaf יִשְׂרָאֵ֗ל אֶת\maqqaf הַמִּשְׁפָּט֙ אֲשֶׁ֣ר שָׁפַ֣ט הַמֶּ֔לֶךְ וַיִּֽרְא֖וּ מִפְּנֵ֣י הַמֶּ֑לֶךְ כִּ֣י רָא֔וּ כִּֽי\maqqaf חׇכְמַ֧ת אֱלֹהִ֛ים בְּקִרְבּ֖וֹ לַעֲשׂ֥וֹת מִשְׁפָּֽט׃ \setuma }
{And all Israel heard of the judgment which the king had judged; and they feared the king; for they saw that the wisdom of God was in him, to do justice.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{וַֽיְהִי֙ הַמֶּ֣לֶךְ שְׁלֹמֹ֔ה מֶ֖לֶךְ עַל\maqqaf כׇּל\maqqaf יִשְׂרָאֵֽל׃ \setuma }
{And king Solomon was king over all Israel.}
\newperek
\newhaftarah{ויגש}
\label{haft_11}
\haftnote{The Haftara is Ezekiel 37:15\verserangechar 37:28. }
\setcounter{chap}{37}
\setcounter{verse}{15}
\twoverse{\Roman{chap}:\arabic{verse}}
{וַיְהִ֥י דְבַר\maqqaf יְהֹוָ֖ה אֵלַ֥י לֵאמֹֽר׃}
{And the word of the \lord\space came unto me, saying:}
\twoverse{\arabic{verse}}
{וְאַתָּ֣ה בֶן\maqqaf אָדָ֗ם קַח\maqqaf לְךָ֙ עֵ֣ץ אֶחָ֔ד וּכְתֹ֤ב עָלָיו֙ לִֽיהוּדָ֔ה וְלִבְנֵ֥י יִשְׂרָאֵ֖ל חֲבֵרָ֑ו וּלְקַח֙ עֵ֣ץ אֶחָ֔ד וּכְת֣וֹב עָלָ֗יו לְיוֹסֵף֙ עֵ֣ץ אֶפְרַ֔יִם וְכׇל\maqqaf בֵּ֥ית יִשְׂרָאֵ֖ל חֲבֵרָֽו׃}
{‘And thou, son of man, take thee one stick, and write upon it: For Judah, and for the children of Israel his companions; then take another stick, and write upon it: For Joseph, the stick of Ephraim, and of all the house of Israel his companions;}
\twoverse{\arabic{verse}}
{וְקָרַ֨ב אֹתָ֜ם אֶחָ֧ד אֶל\maqqaf אֶחָ֛ד לְךָ֖ לְעֵ֣ץ אֶחָ֑ד וְהָי֥וּ לַאֲחָדִ֖ים בְּיָדֶֽךָ׃}
{and join them for thee one to another into one stick, that they may become one in thy hand.}
\twoverse{\arabic{verse}}
{וְכַֽאֲשֶׁר֙ יֹאמְר֣וּ אֵלֶ֔יךָ בְּנֵ֥י עַמְּךָ֖ לֵאמֹ֑ר הֲלֽוֹא\maqqaf תַגִּ֥יד לָ֖נוּ מָה\maqqaf אֵ֥לֶּה לָּֽךְ׃}
{And when the children of thy people shall speak unto thee, saying: Wilt thou not tell us what thou meanest by these?}
\twoverse{\arabic{verse}}
{דַּבֵּ֣ר אֲלֵהֶ֗ם כֹּה\maqqaf אָמַר֮ אֲדֹנָ֣י יֱהֹוִה֒ הִנֵּה֩ אֲנִ֨י לֹקֵ֜חַ אֶת\maqqaf עֵ֤ץ יוֹסֵף֙ אֲשֶׁ֣ר בְּיַד\maqqaf אֶפְרַ֔יִם וְשִׁבְטֵ֥י יִשְׂרָאֵ֖ל חֲבֵרָ֑ו וְנָתַתִּי֩ אוֹתָ֨ם עָלָ֜יו אֶת\maqqaf עֵ֣ץ יְהוּדָ֗ה וַֽעֲשִׂיתִם֙ לְעֵ֣ץ אֶחָ֔ד וְהָי֥וּ אֶחָ֖ד בְּיָדִֽי׃}
{say into them: Thus saith the Lord \textsc{God}: Behold, I will take the stick of Joseph, which is in the hand of Ephraim, and the tribes of Israel his companions; and I will put them unto him together with the stick of Judah, and make them one stick, and they shall be one in My hand.}
\twoverse{\arabic{verse}}
{וְהָי֨וּ הָעֵצִ֜ים אֲֽשֶׁר\maqqaf תִּכְתֹּ֧ב עֲלֵיהֶ֛ם בְּיָדְךָ֖ לְעֵינֵיהֶֽם׃}
{And the sticks whereon thou writest shall be in thy hand before their eyes.}
\twoverse{\arabic{verse}}
{וְדַבֵּ֣ר אֲלֵיהֶ֗ם כֹּה\maqqaf אָמַר֮ אֲדֹנָ֣י יֱהֹוִה֒ הִנֵּ֨ה אֲנִ֤י לֹקֵ֙חַ֙ אֶת\maqqaf בְּנֵ֣י יִשְׂרָאֵ֔ל מִבֵּ֥ין הַגּוֹיִ֖ם אֲשֶׁ֣ר הָֽלְכוּ\maqqaf שָׁ֑ם וְקִבַּצְתִּ֤י אֹתָם֙ מִסָּבִ֔יב וְהֵבֵאתִ֥י אוֹתָ֖ם אֶל\maqqaf אַדְמָתָֽם׃}
{And say unto them: Thus saith the Lord \textsc{God}: Behold, I will take the children of Israel from among the nations, whither they are gone, and will gather them on every side, and bring them into their own land;}
\twoverse{\arabic{verse}}
{וְעָשִׂ֣יתִי אֹ֠תָ֠ם לְג֨וֹי אֶחָ֤ד בָּאָ֙רֶץ֙ בְּהָרֵ֣י יִשְׂרָאֵ֔ל וּמֶ֧לֶךְ אֶחָ֛ד יִהְיֶ֥ה לְכֻלָּ֖ם לְמֶ֑לֶךְ וְלֹ֤א \qk{יִֽהְיוּ\maqqaf }{יהיה} לִשְׁנֵ֣י גוֹיִ֔ם וְלֹ֨א יֵחָ֥צוּ ע֛וֹד לִשְׁתֵּ֥י מַמְלָכ֖וֹת עֽוֹד׃}
{and I will make them one nation in the land, upon the mountains of Israel, and one king shall be king to them all; and they shall be no more two nations, neither shall they be divided into two kingdoms any more at all;}
\twoverse{\arabic{verse}}
{וְלֹ֧א יִֽטַּמְּא֣וּ ע֗וֹד בְּגִלּֽוּלֵיהֶם֙ וּבְשִׁקּ֣וּצֵיהֶ֔ם וּבְכֹ֖ל פִּשְׁעֵיהֶ֑ם וְהוֹשַׁעְתִּ֣י אֹתָ֗ם מִכֹּ֤ל מוֹשְׁבֹֽתֵיהֶם֙ אֲשֶׁ֣ר חָטְא֣וּ בָהֶ֔ם וְטִהַרְתִּ֤י אוֹתָם֙ וְהָיוּ\maqqaf לִ֣י לְעָ֔ם וַאֲנִ֕י אֶהְיֶ֥ה לָהֶ֖ם לֵאלֹהִֽים׃}
{neither shall they defile themselves any more with their idols, nor with their detestable things, nor with any of their transgressions; but I will save them out of all their dwelling-places, wherein they have sinned, and will cleanse them; so shall they be My people, and I will be their God.}
\twoverse{\arabic{verse}}
{וְעַבְדִּ֤י דָוִד֙ מֶ֣לֶךְ עֲלֵיהֶ֔ם וְרוֹעֶ֥ה אֶחָ֖ד יִהְיֶ֣ה לְכֻלָּ֑ם וּבְמִשְׁפָּטַ֣י יֵלֵ֔כוּ וְחֻקּוֹתַ֥י יִשְׁמְר֖וּ וְעָשׂ֥וּ אוֹתָֽם׃}
{And My servant David shall be king over them, and they all shall have one shepherd; they shall also walk in Mine ordinances, and observe My statutes, and do them.}
\twoverse{\arabic{verse}}
{וְיָשְׁב֣וּ עַל\maqqaf הָאָ֗רֶץ אֲשֶׁ֤ר נָתַ֙תִּי֙ לְעַבְדִּ֣י לְיַֽעֲקֹ֔ב אֲשֶׁ֥ר יָֽשְׁבוּ\maqqaf בָ֖הּ אֲבֽוֹתֵיכֶ֑ם וְיָשְׁב֣וּ עָלֶ֡יהָ הֵ֠מָּה וּבְנֵיהֶ֞ם וּבְנֵ֤י בְנֵיהֶם֙ עַד\maqqaf עוֹלָ֔ם וְדָוִ֣ד עַבְדִּ֔י נָשִׂ֥יא לָהֶ֖ם לְעוֹלָֽם׃}
{And they shall dwell in the land that I have given unto Jacob My servant, wherein your fathers dwelt; and they shall dwell therein, they, and their children, and their children’s children, for ever; and David My servant shall be their prince for ever.}
\twoverse{\arabic{verse}}
{וְכָרַתִּ֤י לָהֶם֙ בְּרִ֣ית שָׁל֔וֹם בְּרִ֥ית עוֹלָ֖ם יִהְיֶ֣ה אוֹתָ֑ם וּנְתַתִּים֙ וְהִרְבֵּיתִ֣י אוֹתָ֔ם וְנָתַתִּ֧י אֶת\maqqaf מִקְדָּשִׁ֛י בְּתוֹכָ֖ם לְעוֹלָֽם׃}
{Moreover I will make a covenant of peace with them—it shall be an everlasting covenant with them; and I will establish them, and multiply them, and will set My sanctuary in the midst of them for ever.}
\twoverse{\arabic{verse}}
{וְהָיָ֤ה מִשְׁכָּנִי֙ עֲלֵיהֶ֔ם וְהָיִ֥יתִי לָהֶ֖ם לֵאלֹהִ֑ים וְהֵ֖מָּה יִהְיוּ\maqqaf לִ֥י לְעָֽם׃}
{My dwelling-place also shall be over them; and I will be their God, and they shall be My people.}
\twoverse{\arabic{verse}}
{וְיָֽדְעוּ֙ הַגּוֹיִ֔ם כִּ֚י אֲנִ֣י יְהֹוָ֔ה מְקַדֵּ֖שׁ אֶת\maqqaf יִשְׂרָאֵ֑ל בִּהְי֧וֹת מִקְדָּשִׁ֛י בְּתוֹכָ֖ם לְעוֹלָֽם׃ \petucha }
{And the nations shall know that I am the \lord\space that sanctify Israel, when My sanctuary shall be in the midst of them for ever.’}
\newhaftarah{ויחי}
\label{haft_12}
\haftnote{The Haftara is I Kings 2:1\verserangechar 2:12. }
\setcounter{chap}{2}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}
{וַיִּקְרְב֥וּ יְמֵֽי\maqqaf דָוִ֖ד לָמ֑וּת וַיְצַ֛ו אֶת\maqqaf שְׁלֹמֹ֥ה בְנ֖וֹ לֵאמֹֽר׃}
{Now the days of David drew nigh that he should die; and he charged Solomon his son, saying:}
\twoverse{\arabic{verse}}
{אָנֹכִ֣י הֹלֵ֔ךְ בְּדֶ֖רֶךְ כׇּל\maqqaf הָאָ֑רֶץ וְחָזַקְתָּ֖ וְהָיִ֥יתָֽ לְאִֽישׁ׃}
{‘I go the way of all the earth; be thou strong therefore, and show thyself a man;}
\twoverse{\arabic{verse}}
{וְשָׁמַרְתָּ֞ אֶת\maqqaf מִשְׁמֶ֣רֶת \legarmeh  יְהֹוָ֣ה אֱלֹהֶ֗יךָ לָלֶ֤כֶת בִּדְרָכָיו֙ לִשְׁמֹ֨ר חֻקֹּתָ֤יו מִצְוֺתָיו֙ וּמִשְׁפָּטָ֣יו וְעֵדְוֺתָ֔יו כַּכָּת֖וּב בְּתוֹרַ֣ת מֹשֶׁ֑ה לְמַ֣עַן תַּשְׂכִּ֗יל אֵ֚ת כׇּל\maqqaf אֲשֶׁ֣ר תַּעֲשֶׂ֔ה וְאֵ֛ת כׇּל\maqqaf אֲשֶׁ֥ר תִּפְנֶ֖ה שָֽׁם׃}
{and keep the charge of the \lord\space thy God, to walk in His ways, to keep His statutes, and His commandments, and His ordinances, and His testimonies, according to that which is written in the law of Moses, that thou mayest prosper in all that thou doest, and whithersoever thou turnest thyself;}
\twoverse{\arabic{verse}}
{לְמַ֩עַן֩ יָקִ֨ים יְהֹוָ֜ה אֶת\maqqaf דְּבָר֗וֹ אֲשֶׁ֨ר דִּבֶּ֣ר עָלַי֮ לֵאמֹר֒ אִם\maqqaf יִשְׁמְר֨וּ בָנֶ֜יךָ אֶת\maqqaf דַּרְכָּ֗ם לָלֶ֤כֶת לְפָנַי֙ בֶּאֱמֶ֔ת בְּכׇל\maqqaf לְבָבָ֖ם וּבְכׇל\maqqaf נַפְשָׁ֑ם לֵאמֹ֕ר לֹֽא\maqqaf יִכָּרֵ֤ת לְךָ֙ אִ֔ישׁ מֵעַ֖ל כִּסֵּ֥א יִשְׂרָאֵֽל׃}
{that the \lord\space may establish His word which He spoke concerning me, saying: If thy children take heed to their way, to walk before Me in truth with all their heart and with all their soul, there shall not fail thee, said He, a man on the throne of Israel.}
\twoverse{\arabic{verse}}
{וְגַ֣ם אַתָּ֣ה יָדַ֡עְתָּ אֵת֩ אֲשֶׁר\maqqaf עָ֨שָׂה לִ֜י יוֹאָ֣ב בֶּן\maqqaf צְרוּיָ֗ה אֲשֶׁ֣ר עָשָׂ֣ה לִשְׁנֵֽי\maqqaf שָׂרֵ֣י צִבְא֣וֹת יִ֠שְׂרָאֵ֠ל לְאַבְנֵ֨ר בֶּן\maqqaf נֵ֜ר וְלַעֲמָשָׂ֤א בֶן\maqqaf יֶ֙תֶר֙ וַיַּ֣הַרְגֵ֔ם וַיָּ֥שֶׂם דְּמֵֽי\maqqaf מִלְחָמָ֖ה בְּשָׁלֹ֑ם וַיִּתֵּ֞ן דְּמֵ֣י מִלְחָמָ֗ה בַּחֲגֹֽרָתוֹ֙ אֲשֶׁ֣ר בְּמׇתְנָ֔יו וּֽבְנַעֲל֖וֹ אֲשֶׁ֥ר בְּרַגְלָֽיו׃}
{Moreover thou knowest also what Joab the son of Zeruiah did unto me, even what he did to the two captains of the hosts of Israel, unto Abner the son of Ner and unto Amasa the son of Jether, whom he slew, and shed the blood of war in peace, and put the blood of war upon his girdle that was about his loins, and in his shoes that were on his feet.}
\twoverse{\arabic{verse}}
{וְעָשִׂ֖יתָ כְּחׇכְמָתֶ֑ךָ וְלֹֽא\maqqaf תוֹרֵ֧ד שֵׂיבָת֛וֹ בְּשָׁלֹ֖ם שְׁאֹֽל׃}
{Do therefore according to thy wisdom, and let not his hoar head go down to the grave in peace.}
\twoverse{\arabic{verse}}
{וְלִבְנֵ֨י בַרְזִלַּ֤י הַגִּלְעָדִי֙ תַּֽעֲשֶׂה\maqqaf חֶ֔סֶד וְהָי֖וּ בְּאֹכְלֵ֣י שֻׁלְחָנֶ֑ךָ כִּי\maqqaf כֵן֙ קָרְב֣וּ אֵלַ֔י בְּבׇרְחִ֕י מִפְּנֵ֖י אַבְשָׁל֥וֹם אָחִֽיךָ׃}
{But show kindness unto the sons of Barzillai the Gileadite, and let them be of those that eat at thy table; for so they drew nigh unto me when I fled from Absalom thy brother.}
\twoverse{\arabic{verse}}
{וְהִנֵּ֣ה עִ֠מְּךָ֠ שִֽׁמְעִ֨י בֶן\maqqaf גֵּרָ֥א בֶן\maqqaf הַיְמִינִי֮ מִבַּחֻרִים֒ וְה֤וּא קִֽלְלַ֙נִי֙ קְלָלָ֣ה נִמְרֶ֔צֶת בְּי֖וֹם לֶכְתִּ֣י מַחֲנָ֑יִם וְהֽוּא\maqqaf יָרַ֤ד לִקְרָאתִי֙ הַיַּרְדֵּ֔ן וָאֶשָּׁ֨בַֽע ל֤וֹ בַֽיהֹוָה֙ לֵאמֹ֔ר אִם\maqqaf אֲמִֽיתְךָ֖ בֶּחָֽרֶב׃}
{And, behold, there is with thee Shimei the son of Gera, the Benjamite, of Bahurim, who cursed me with a grievous curse in the day when I went to Mahanaim; but he came down to meet me at the Jordan, and I swore to him by the \lord, saying: I will not put thee to death with the sword.}
\twoverse{\arabic{verse}}
{וְעַתָּה֙ אַל\maqqaf תְּנַקֵּ֔הוּ כִּ֛י אִ֥ישׁ חָכָ֖ם אָ֑תָּה וְיָֽדַעְתָּ֙ אֵ֣ת אֲשֶׁ֣ר תַּֽעֲשֶׂה\maqqaf לּ֔וֹ וְהֽוֹרַדְתָּ֧ אֶת\maqqaf שֵׂיבָת֛וֹ בְּדָ֖ם שְׁאֽוֹל׃}
{Now therefore hold him not guiltless, for thou art a wise man; and thou wilt know what thou oughtest to do unto him, and thou shalt bring his hoar head down to the grave with blood.’}
\twoverse{\arabic{verse}}
{וַיִּשְׁכַּ֥ב דָּוִ֖ד עִם\maqqaf אֲבֹתָ֑יו וַיִּקָּבֵ֖ר בְּעִ֥יר דָּוִֽד׃ \petucha }
{And David slept with his fathers, and was buried in the city of David.}
\twoverse{\arabic{verse}}
{וְהַיָּמִ֗ים אֲשֶׁ֨ר מָלַ֤ךְ דָּוִד֙ עַל\maqqaf יִשְׂרָאֵ֔ל אַרְבָּעִ֖ים שָׁנָ֑ה בְּחֶבְר֤וֹן מָלַךְ֙ שֶׁ֣בַע שָׁנִ֔ים וּבִירֽוּשָׁלַ֣͏ִם מָלַ֔ךְ שְׁלֹשִׁ֥ים וְשָׁלֹ֖שׁ שָׁנִֽים׃}
{And the days that David reigned over Israel were forty years: seven years reigned he in Hebron, and thirty and three years reigned he in Jerusalem.}
\twoverse{\arabic{verse}}
{וּשְׁלֹמֹ֕ה יָשַׁ֕ב עַל\maqqaf כִּסֵּ֖א דָּוִ֣ד אָבִ֑יו וַתִּכֹּ֥ן מַלְכֻת֖וֹ מְאֹֽד׃ \setuma }
{And Solomon sat upon the throne of David his father; and his kingdom was established firmly.}
