\vspace{24pt}\LARGE \textbf{הפטרות} \normalsize\vspace{-16pt}
\addcontentsline{toc}{chapter}{הפטרות}
\newhaftarah{דברים}
\label{haft_44}
\haftnote{The Haftara is Isaiah 1:1\verserangechar 1:27. This is the Third Haftara of Rebuke.}
\setcounter{chap}{1}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}
{חֲזוֹן֙ יְשַֽׁעְיָ֣הוּ בֶן\maqqaf אָמ֔וֹץ אֲשֶׁ֣ר חָזָ֔ה עַל\maqqaf יְהוּדָ֖ה וִירֽוּשָׁלָ֑͏ִם בִּימֵ֨י עֻזִּיָּ֧הוּ יוֹתָ֛ם אָחָ֥ז יְחִזְקִיָּ֖הוּ מַלְכֵ֥י יְהוּדָֽה׃}
{The Vision of Isaiah the son of Amoz, which he saw concerning Judah and Jerusalem, in the days of Uzziah, Jotham, Ahaz, and Hezekiah, kings of Judah.}
\twoverse{\arabic{verse}}
{שִׁמְע֤וּ שָׁמַ֙יִם֙ וְהַאֲזִ֣ינִי אֶ֔רֶץ כִּ֥י יְהֹוָ֖ה דִּבֵּ֑ר בָּנִים֙ גִּדַּ֣לְתִּי וְרוֹמַ֔מְתִּי וְהֵ֖ם פָּ֥שְׁעוּ בִֽי׃}
{Hear, O heavens, and give ear, O earth, For the \lord\space hath spoken: Children I have reared, and brought up, And they have rebelled against Me.}
\twoverse{\arabic{verse}}
{יָדַ֥ע שׁוֹר֙ קֹנֵ֔הוּ וַחֲמ֖וֹר אֵב֣וּס בְּעָלָ֑יו יִשְׂרָאֵל֙ לֹ֣א יָדַ֔ע עַמִּ֖י לֹ֥א הִתְבּוֹנָֽן׃}
{The ox knoweth his owner, And the ass his master’s crib; But Israel doth not know, My people doth not consider.}
\twoverse{\arabic{verse}}
{ה֣וֹי \legarmeh  גּ֣וֹי חֹטֵ֗א עַ֚ם כֶּ֣בֶד עָוֺ֔ן זֶ֣רַע מְרֵעִ֔ים בָּנִ֖ים מַשְׁחִיתִ֑ים עָזְב֣וּ אֶת\maqqaf יְהֹוָ֗ה נִֽאֲצ֛וּ אֶת\maqqaf קְד֥וֹשׁ יִשְׂרָאֵ֖ל נָזֹ֥רוּ אָחֽוֹר׃}
{Ah sinful nation, A people laden with iniquity, A seed of evil-doers, Children that deal corruptly; They have forsaken the \lord, They have contemned the Holy One of Israel, They are turned away backward.}
\twoverse{\arabic{verse}}
{עַ֣ל מֶ֥ה תֻכּ֛וּ ע֖וֹד תּוֹסִ֣יפוּ סָרָ֑ה כׇּל\maqqaf רֹ֣אשׁ לׇֽחֳלִ֔י וְכׇל\maqqaf לֵבָ֖ב דַּוָּֽי׃}
{On what part will ye yet be stricken, Seeing ye stray away more and more? The whole head is sick, And the whole heart faint;}
\twoverse{\arabic{verse}}
{מִכַּף\maqqaf רֶ֤גֶל וְעַד\maqqaf רֹאשׁ֙ אֵֽין\maqqaf בּ֣וֹ מְתֹ֔ם פֶּ֥צַע וְחַבּוּרָ֖ה וּמַכָּ֣ה טְרִיָּ֑ה לֹא\maqqaf זֹ֙רוּ֙ וְלֹ֣א חֻבָּ֔שׁוּ וְלֹ֥א רֻכְּכָ֖ה בַּשָּֽׁמֶן׃}
{From the sole of the foot even unto the head There is no soundness in it; But wounds, and bruises, and festering sores: They have not been pressed, neither bound up, Neither mollified with oil.}
\twoverse{\arabic{verse}}
{אַרְצְכֶ֣ם שְׁמָמָ֔ה עָרֵיכֶ֖ם שְׂרֻפ֣וֹת אֵ֑שׁ אַדְמַתְכֶ֗ם לְנֶגְדְּכֶם֙ זָרִים֙ אֹכְלִ֣ים אֹתָ֔הּ וּשְׁמָמָ֖ה כְּמַהְפֵּכַ֥ת זָרִֽים׃}
{Your country is desolate; Your cities are burned with fire; Your land, strangers devour it in your presence, And it is desolate, as overthrown by floods.}
\twoverse{\arabic{verse}}
{וְנוֹתְרָ֥ה בַת\maqqaf צִיּ֖וֹן כְּסֻכָּ֣ה בְכָ֑רֶם כִּמְלוּנָ֥ה בְמִקְשָׁ֖ה כְּעִ֥יר נְצוּרָֽה׃}
{And the daughter of Zion is left As a booth in a vineyard, As a lodge in a garden of cucumbers, As a besieged city.}
\twoverse{\arabic{verse}}
{לוּלֵי֙ יְהֹוָ֣ה צְבָא֔וֹת הוֹתִ֥יר לָ֛נוּ שָׂרִ֖יד כִּמְעָ֑ט כִּסְדֹ֣ם הָיִ֔ינוּ לַעֲמֹרָ֖ה דָּמִֽינוּ׃ \petucha }
{Except the \lord\space of hosts Had left unto us a very small remnant, We should have been as Sodom, We should have been like unto Gomorrah.}
\twoverse{\arabic{verse}}
{שִׁמְע֥וּ דְבַר\maqqaf יְהֹוָ֖ה קְצִינֵ֣י סְדֹ֑ם הַאֲזִ֛ינוּ תּוֹרַ֥ת אֱלֹהֵ֖ינוּ עַ֥ם עֲמֹרָֽה׃}
{Hear the word of the \lord, Ye rulers of Sodom; Give ear unto the law of our God, Ye people of Gomorrah.}
\twoverse{\arabic{verse}}
{לָמָּה\maqqaf לִּ֤י רֹב\maqqaf זִבְחֵיכֶם֙ יֹאמַ֣ר יְהֹוָ֔ה שָׂבַ֛עְתִּי עֹל֥וֹת אֵילִ֖ים וְחֵ֣לֶב מְרִיאִ֑ים וְדַ֨ם פָּרִ֧ים וּכְבָשִׂ֛ים וְעַתּוּדִ֖ים לֹ֥א חָפָֽצְתִּי׃}
{To what purpose is the multitude of your sacrifices unto Me? Saith the \lord; I am full of the burnt-offerings of rams, And the fat of fed beasts; And I delight not in the blood Of bullocks, or of lambs, or of he-goats.}
\twoverse{\arabic{verse}}
{כִּ֣י תָבֹ֔אוּ לֵֽרָא֖וֹת פָּנָ֑י מִֽי\maqqaf בִקֵּ֥שׁ זֹ֛את מִיֶּדְכֶ֖ם רְמֹ֥ס חֲצֵרָֽי׃}
{When ye come to appear before Me, Who hath required this at your hand, To trample My courts?}
\twoverse{\arabic{verse}}
{לֹ֣א תוֹסִ֗יפוּ הָבִיא֙ מִנְחַת\maqqaf שָׁ֔וְא קְטֹ֧רֶת תּוֹעֵבָ֛ה הִ֖יא לִ֑י חֹ֤דֶשׁ וְשַׁבָּת֙ קְרֹ֣א מִקְרָ֔א לֹא\maqqaf אוּכַ֥ל אָ֖וֶן וַֽעֲצָרָֽה׃}
{Bring no more vain oblations; It is an offering of abomination unto Me; New moon and sabbath, the holding of convocations— I cannot endure iniquity along with the solemn assembly.}
\twoverse{\arabic{verse}}
{חׇדְשֵׁיכֶ֤ם וּמֽוֹעֲדֵיכֶם֙ שָֽׂנְאָ֣ה נַפְשִׁ֔י הָי֥וּ עָלַ֖י לָטֹ֑רַח נִלְאֵ֖יתִי נְשֹֽׂא׃}
{Your new moons and your appointed seasons My soul hateth; They are a burden unto Me; I am weary to bear them.}
\twoverse{\arabic{verse}}
{וּבְפָרִשְׂכֶ֣ם כַּפֵּיכֶ֗ם אַעְלִ֤ים עֵינַי֙ מִכֶּ֔ם גַּ֛ם כִּֽי\maqqaf תַרְבּ֥וּ תְפִלָּ֖ה אֵינֶ֣נִּי שֹׁמֵ֑עַ יְדֵיכֶ֖ם דָּמִ֥ים מָלֵֽאוּ׃}
{And when ye spread forth your hands, I will hide Mine eyes from you; Yea, when ye make many prayers, I will not hear; Your hands are full of blood.}
\twoverse{\arabic{verse}}
{רַֽחֲצוּ֙ הִזַּכּ֔וּ הָסִ֛ירוּ רֹ֥עַ מַעַלְלֵיכֶ֖ם מִנֶּ֣גֶד עֵינָ֑י חִדְל֖וּ הָרֵֽעַ׃}
{Wash you, make you clean, Put away the evil of your doings From before Mine eyes, Cease to do evil;}
\twoverse{\arabic{verse}}
{לִמְד֥וּ הֵיטֵ֛ב דִּרְשׁ֥וּ מִשְׁפָּ֖ט אַשְּׁר֣וּ חָמ֑וֹץ שִׁפְט֣וּ יָת֔וֹם רִ֖יבוּ אַלְמָנָֽה׃ \setuma }
{Learn to do well; Seek justice, relieve the oppressed, Judge the fatherless, plead for the widow.}
\twoverse{\arabic{verse}}
{לְכוּ\maqqaf נָ֛א וְנִוָּכְחָ֖ה יֹאמַ֣ר יְהֹוָ֑ה אִם\maqqaf יִהְי֨וּ חֲטָאֵיכֶ֤ם כַּשָּׁנִים֙ כַּשֶּׁ֣לֶג יַלְבִּ֔ינוּ אִם\maqqaf יַאְדִּ֥ימוּ כַתּוֹלָ֖ע כַּצֶּ֥מֶר יִֽהְיֽוּ׃}
{Come now, and let us reason together, Saith the \lord; Though your sins be as scarlet, They shall be as white as snow; Though they be red like crimson, They shall be as wool.}
\twoverse{\arabic{verse}}
{אִם\maqqaf תֹּאב֖וּ וּשְׁמַעְתֶּ֑ם ט֥וּב הָאָ֖רֶץ תֹּאכֵֽלוּ׃}
{If ye be willing and obedient, Ye shall eat the good of the land;}
\twoverse{\arabic{verse}}
{וְאִם\maqqaf תְּמָאֲנ֖וּ וּמְרִיתֶ֑ם חֶ֣רֶב תְּאֻכְּל֔וּ כִּ֛י פִּ֥י יְהֹוָ֖ה דִּבֵּֽר׃ \petucha }
{But if ye refuse and rebel, Ye shall be devoured with the sword; For the mouth of the \lord\space hath spoken.}
\twoverse{\arabic{verse}}
{אֵיכָה֙ הָיְתָ֣ה לְזוֹנָ֔ה קִרְיָ֖ה נֶאֱמָנָ֑ה מְלֵֽאֲתִ֣י מִשְׁפָּ֗ט צֶ֛דֶק יָלִ֥ין בָּ֖הּ וְעַתָּ֥ה מְרַצְּחִֽים׃}
{How is the faithful city Become a harlot! She that was full of justice, Righteousness lodged in her, But now murderers.}
\twoverse{\arabic{verse}}
{כַּסְפֵּ֖ךְ הָיָ֣ה לְסִיגִ֑ים סׇבְאֵ֖ךְ מָה֥וּל בַּמָּֽיִם׃}
{Thy silver is become dross, Thy wine mixed with water.}
\twoverse{\arabic{verse}}
{שָׂרַ֣יִךְ סוֹרְרִ֗ים וְחַבְרֵי֙ גַּנָּבִ֔ים כֻּלּוֹ֙ אֹהֵ֣ב שֹׁ֔חַד וְרֹדֵ֖ף שַׁלְמֹנִ֑ים יָתוֹם֙ לֹ֣א יִשְׁפֹּ֔טוּ וְרִ֥יב אַלְמָנָ֖ה לֹא\maqqaf יָב֥וֹא אֲלֵיהֶֽם׃ \setuma }
{Thy princes are rebellious, And companions of thieves; Every one loveth bribes, And followeth after rewards; They judge not the fatherless, Neither doth the cause of the widow come unto them.}
\twoverse{\arabic{verse}}
{לָכֵ֗ן נְאֻ֤ם הָאָדוֹן֙ יְהֹוָ֣ה צְבָא֔וֹת אֲבִ֖יר יִשְׂרָאֵ֑ל ה֚וֹי אֶנָּחֵ֣ם מִצָּרַ֔י וְאִנָּקְמָ֖ה מֵאוֹיְבָֽי׃}
{Therefore saith the Lord, the \lord\space of hosts, The Mighty One of Israel: Ah, I will ease Me of Mine adversaries, And avenge Me of Mine enemies;}
\twoverse{\arabic{verse}}
{וְאָשִׁ֤יבָה יָדִי֙ עָלַ֔יִךְ וְאֶצְרֹ֥ף כַּבֹּ֖ר סִיגָ֑יִךְ וְאָסִ֖ירָה כׇּל\maqqaf בְּדִילָֽיִךְ׃}
{And I will turn My hand upon thee, And purge away thy dross as with lye, And will take away all thine alloy;}
\twoverse{\arabic{verse}}
{וְאָשִׁ֤יבָה שֹׁפְטַ֙יִךְ֙ כְּבָרִ֣אשֹׁנָ֔ה וְיֹעֲצַ֖יִךְ כְּבַתְּחִלָּ֑ה אַֽחֲרֵי\maqqaf כֵ֗ן יִקָּ֤רֵא לָךְ֙ עִ֣יר הַצֶּ֔דֶק קִרְיָ֖ה נֶאֱמָנָֽה׃}
{And I will restore thy judges as at the first, And thy counsellors as at the beginning; Afterward thou shalt be called The city of righteousness, The faithful city.}
\twoverse{\arabic{verse}}
{צִיּ֖וֹן בְּמִשְׁפָּ֣ט תִּפָּדֶ֑ה וְשָׁבֶ֖יהָ בִּצְדָקָֽה׃}
{Zion shall be redeemed with justice, And they that return of her with righteousness.}
\newhaftarah{ואתחנן}
\label{haft_45}
\haftnote{The Haftara is Isaiah 40:1\verserangechar 40:26. This is the First Haftara of Consolation.}
\setcounter{chap}{40}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}
{נַחֲמ֥וּ נַחֲמ֖וּ עַמִּ֑י יֹאמַ֖ר אֱלֹֽהֵיכֶֽם׃}
{Comfort ye, comfort ye My people, saith your God.}
\twoverse{\arabic{verse}}
{דַּבְּר֞וּ עַל\maqqaf לֵ֤ב יְרוּשָׁלַ֙͏ִם֙ וְקִרְא֣וּ אֵלֶ֔יהָ כִּ֤י מָֽלְאָה֙ צְבָאָ֔הּ כִּ֥י נִרְצָ֖ה עֲוֺנָ֑הּ כִּ֤י לָֽקְחָה֙ מִיַּ֣ד יְהֹוָ֔ה כִּפְלַ֖יִם בְּכׇל\maqqaf חַטֹּאתֶֽיהָ׃ \setuma }
{Bid Jerusalem take heart, and proclaim unto her, that her time of service is accomplished, that her guilt is paid off; that she hath received of the \lord\textsc{’s}\space hand double for all her sins.}
\twoverse{\arabic{verse}}
{ק֣וֹל קוֹרֵ֔א בַּמִּדְבָּ֕ר פַּנּ֖וּ דֶּ֣רֶךְ יְהֹוָ֑ה יַשְּׁרוּ֙ בָּעֲרָבָ֔ה מְסִלָּ֖ה לֵאלֹהֵֽינוּ׃}
{Hark! one calleth: ‘Clear ye in the wilderness the way of the \lord, make plain in the desert a highway for our God.}
\twoverse{\arabic{verse}}
{כׇּל\maqqaf גֶּיא֙ יִנָּשֵׂ֔א וְכׇל\maqqaf הַ֥ר וְגִבְעָ֖ה יִשְׁפָּ֑לוּ וְהָיָ֤ה הֶֽעָקֹב֙ לְמִישׁ֔וֹר וְהָרְכָסִ֖ים לְבִקְעָֽה׃}
{Every valley shall be lifted up, and every mountain and hill shall be made low; and the rugged shall be made level, and the rough places a plain;}
\twoverse{\arabic{verse}}
{וְנִגְלָ֖ה כְּב֣וֹד יְהֹוָ֑ה וְרָא֤וּ כׇל\maqqaf בָּשָׂר֙ יַחְדָּ֔ו כִּ֛י פִּ֥י יְהֹוָ֖ה דִּבֵּֽר׃ \petucha }
{And the glory of the \lord\space shall be revealed, and all flesh shall see it together; for the mouth of the \lord\space hath spoken it.’}
\twoverse{\arabic{verse}}
{ק֚וֹל אֹמֵ֣ר קְרָ֔א וְאָמַ֖ר מָ֣ה אֶקְרָ֑א כׇּל\maqqaf הַבָּשָׂ֣ר חָצִ֔יר וְכׇל\maqqaf חַסְדּ֖וֹ כְּצִ֥יץ הַשָּׂדֶֽה׃}
{Hark! one saith: ‘Proclaim!’ And he saith: ‘What shall I proclaim?’ ’All flesh is grass, And all the goodliness thereof is as the flower of the field;}
\twoverse{\arabic{verse}}
{יָבֵ֤שׁ חָצִיר֙ נָ֣בֵֽל צִ֔יץ כִּ֛י ר֥וּחַ יְהֹוָ֖ה נָ֣שְׁבָה בּ֑וֹ אָכֵ֥ן חָצִ֖יר הָעָֽם׃}
{The grass withereth, the flower fadeth; Because the breath of the \lord\space bloweth upon it— Surely the people is grass.}
\twoverse{\arabic{verse}}
{יָבֵ֥שׁ חָצִ֖יר נָ֣בֵֽל צִ֑יץ וּדְבַר\maqqaf אֱלֹהֵ֖ינוּ יָק֥וּם לְעוֹלָֽם׃ \setuma }
{The grass withereth, the flower fadeth; But the word of our God shall stand for ever.’}
\twoverse{\arabic{verse}}
{עַ֣ל הַר\maqqaf גָּבֹ֤הַּ עֲלִי\maqqaf לָךְ֙ מְבַשֶּׂ֣רֶת צִיּ֔וֹן הָרִ֤ימִי בַכֹּ֙חַ֙ קוֹלֵ֔ךְ מְבַשֶּׂ֖רֶת יְרוּשָׁלָ֑͏ִם הָרִ֙ימִי֙ אַל\maqqaf תִּירָ֔אִי אִמְרִי֙ לְעָרֵ֣י יְהוּדָ֔ה הִנֵּ֖ה אֱלֹהֵיכֶֽם׃}
{O thou that tellest good tidings to Zion, Get thee up into the high mountain; O thou that tellest good tidings to Jerusalem, Lift up thy voice with strength; Lift it up, be not afraid; Say unto the cities of Judah: ‘Behold your God!’}
\twoverse{\arabic{verse}}
{הִנֵּ֨ה אֲדֹנָ֤י יֱהֹוִה֙ בְּחָזָ֣ק יָב֔וֹא וּזְרֹע֖וֹ מֹ֣שְׁלָה ל֑וֹ הִנֵּ֤ה שְׂכָרוֹ֙ אִתּ֔וֹ וּפְעֻלָּת֖וֹ לְפָנָֽיו׃}
{Behold, the Lord \textsc{God} will come as a Mighty One, And His arm will rule for Him; Behold, His reward is with Him, And His recompense before Him.}
\twoverse{\arabic{verse}}
{כְּרֹעֶה֙ עֶדְר֣וֹ יִרְעֶ֔ה בִּזְרֹעוֹ֙ יְקַבֵּ֣ץ טְלָאִ֔ים וּבְחֵיק֖וֹ יִשָּׂ֑א עָל֖וֹת יְנַהֵֽל׃ \setuma }
{Even as a shepherd that feedeth his flock, That gathereth the lambs in his arm, And carrieth them in his bosom, And gently leadeth those that give suck.}
\twoverse{\arabic{verse}}
{מִֽי\maqqaf מָדַ֨ד בְּשׇׁעֳל֜וֹ מַ֗יִם וְשָׁמַ֙יִם֙ בַּזֶּ֣רֶת תִּכֵּ֔ן וְכָ֥ל בַּשָּׁלִ֖שׁ עֲפַ֣ר הָאָ֑רֶץ וְשָׁקַ֤ל בַּפֶּ֙לֶס֙ הָרִ֔ים וּגְבָע֖וֹת בְּמֹאזְנָֽיִם׃}
{Who hath measured the waters in the hollow of his hand, And meted out heaven with the span, And comprehended the dust of the earth in a measure, And weighed the mountains in scales, And the hills in a balance?}
\twoverse{\arabic{verse}}
{מִֽי\maqqaf תִכֵּ֥ן אֶת\maqqaf ר֖וּחַ יְהֹוָ֑ה וְאִ֖ישׁ עֲצָת֥וֹ יוֹדִיעֶֽנּוּ׃}
{Who hath meted out the spirit of the \lord? Or who was His counsellor that he might instruct Him?}
\twoverse{\arabic{verse}}
{אֶת\maqqaf מִ֤י נוֹעָץ֙ וַיְבִינֵ֔הוּ וַֽיְלַמְּדֵ֖הוּ בְּאֹ֣רַח מִשְׁפָּ֑ט וַיְלַמְּדֵ֣הוּ דַ֔עַת וְדֶ֥רֶךְ תְּבוּנ֖וֹת יוֹדִיעֶֽנּוּ׃}
{With whom took He counsel, and who instructed Him, And taught Him in the path of right, And taught Him knowledge, And made Him to know the way of discernment?}
\twoverse{\arabic{verse}}
{הֵ֤ן גּוֹיִם֙ כְּמַ֣ר מִדְּלִ֔י וּכְשַׁ֥חַק מֹאזְנַ֖יִם נֶחְשָׁ֑בוּ הֵ֥ן אִיִּ֖ים כַּדַּ֥ק יִטּֽוֹל׃}
{Behold, the nations are as a drop of a bucket, And are counted as the small dust of the balance; Behold the isles are as a mote in weight.}
\twoverse{\arabic{verse}}
{וּלְבָנ֕וֹן אֵ֥ין דֵּ֖י בָּעֵ֑ר וְחַ֨יָּת֔וֹ אֵ֥ין דֵּ֖י עוֹלָֽה׃ \petucha }
{And Lebanon is not sufficient fuel, Nor the beasts thereof sufficient for burnt-offerings.}
\twoverse{\arabic{verse}}
{כׇּל\maqqaf הַגּוֹיִ֖ם כְּאַ֣יִן נֶגְדּ֑וֹ מֵאֶ֥פֶס וָתֹ֖הוּ נֶחְשְׁבוּ\maqqaf לֽוֹ׃}
{All the nations are as nothing before Him; They are accounted by Him as things of nought, and vanity.}
\twoverse{\arabic{verse}}
{וְאֶל\maqqaf מִ֖י תְּדַמְּי֣וּן אֵ֑ל וּמַה\maqqaf דְּמ֖וּת תַּ֥עַרְכוּ לֽוֹ׃}
{To whom then will ye liken God? Or what likeness will ye compare unto Him?}
\twoverse{\arabic{verse}}
{הַפֶּ֙סֶל֙ נָסַ֣ךְ חָרָ֔שׁ וְצֹרֵ֖ף בַּזָּהָ֣ב יְרַקְּעֶ֑נּוּ וּרְתֻק֥וֹת כֶּ֖סֶף צוֹרֵֽף׃}
{The image perchance, which the craftsman hath melted, And the goldsmith spread over with gold, The silversmith casting silver chains?}
\twoverse{\arabic{verse}}
{הַֽמְסֻכָּ֣ן תְּרוּמָ֔ה עֵ֥ץ לֹא\maqqaf יִרְקַ֖ב יִבְחָ֑ר חָרָ֤שׁ חָכָם֙ יְבַקֶּשׁ\maqqaf ל֔וֹ לְהָכִ֥ין פֶּ֖סֶל לֹ֥א יִמּֽוֹט׃ \setuma }
{A holm-oak is set apart, He chooseth a tree that will not rot; He seeketh unto him a cunning craftsman To set up an image, that shall not be moved.}
\twoverse{\arabic{verse}}
{הֲל֤וֹא תֵֽדְעוּ֙ הֲל֣וֹא תִשְׁמָ֔עוּ הֲל֛וֹא הֻגַּ֥ד מֵרֹ֖אשׁ לָכֶ֑ם הֲלוֹא֙ הֲבִ֣ינוֹתֶ֔ם מוֹסְד֖וֹת הָאָֽרֶץ׃}
{Know ye not? hear ye not? Hath it not been told you from the beginning? Have ye not understood the foundations of the earth?}
\twoverse{\arabic{verse}}
{הַיֹּשֵׁב֙ עַל\maqqaf ח֣וּג הָאָ֔רֶץ וְיֹשְׁבֶ֖יהָ כַּחֲגָבִ֑ים הַנּוֹטֶ֤ה כַדֹּק֙ שָׁמַ֔יִם וַיִּמְתָּחֵ֥ם כָּאֹ֖הֶל לָשָֽׁבֶת׃}
{It is He that sitteth above the circle of the earth, And the inhabitants thereof are as grasshoppers; That stretcheth out the heavens as a curtain, And spreadeth them out as a tent to dwell in;}
\twoverse{\arabic{verse}}
{הַנּוֹתֵ֥ן רוֹזְנִ֖ים לְאָ֑יִן שֹׁ֥פְטֵי אֶ֖רֶץ כַּתֹּ֥הוּ עָשָֽׂה׃}
{That bringeth princes to nothing; He maketh the judges of the earth as a thing of nought.}
\twoverse{\arabic{verse}}
{אַ֣ף בַּל\maqqaf נִטָּ֗עוּ אַ֚ף בַּל\maqqaf זֹרָ֔עוּ אַ֛ף בַּל\maqqaf שֹׁרֵ֥שׁ בָּאָ֖רֶץ גִּזְעָ֑ם וְגַם\maqqaf נָשַׁ֤ף בָּהֶם֙ וַיִּבָ֔שׁוּ וּסְעָרָ֖ה כַּקַּ֥שׁ תִּשָּׂאֵֽם׃ \setuma }
{Scarce are they planted, Scarce are they sown, Scarce hath their stock taken root in the earth; When He bloweth upon them, they wither, And the whirlwind taketh them away as stubble.}
\twoverse{\arabic{verse}}
{וְאֶל\maqqaf מִ֥י תְדַמְּי֖וּנִי וְאֶשְׁוֶ֑ה יֹאמַ֖ר קָדֽוֹשׁ׃}
{To whom then will ye liken Me, that I should be equal? Saith the Holy One.}
\twoverse{\arabic{verse}}
{שְׂאוּ\maqqaf מָר֨וֹם עֵינֵיכֶ֤ם וּרְאוּ֙ מִי\maqqaf בָרָ֣א אֵ֔לֶּה הַמּוֹצִ֥יא בְמִסְפָּ֖ר צְבָאָ֑ם לְכֻלָּם֙ בְּשֵׁ֣ם יִקְרָ֔א מֵרֹ֤ב אוֹנִים֙ וְאַמִּ֣יץ כֹּ֔חַ אִ֖ישׁ לֹ֥א נֶעְדָּֽר׃ \setuma }
{Lift up your eyes on high, And see: who hath created these? He that bringeth out their host by number, He calleth them all by name; By the greatness of His might, and for that He is strong in power, Not one faileth.}
\newhaftarah{עקב}
\label{haft_46}
\haftnote{The Haftara is Isaiah 49:14\verserangechar 51:3. This is the Second Haftara of Consolation.}
\setcounter{chap}{49}
\setcounter{verse}{14}
\twoverse{\Roman{chap}:\arabic{verse}}
{וַתֹּ֥אמֶר צִיּ֖וֹן עֲזָבַ֣נִי יְהֹוָ֑ה וַאדֹנָ֖י שְׁכֵחָֽנִי׃}
{But Zion said: ‘The \lord\space hath forsaken me, And the Lord hath forgotten me.’}
\twoverse{\arabic{verse}}
{הֲתִשְׁכַּ֤ח אִשָּׁה֙ עוּלָ֔הּ מֵרַחֵ֖ם בֶּן\maqqaf בִּטְנָ֑הּ גַּם\maqqaf אֵ֣לֶּה תִשְׁכַּ֔חְנָה וְאָנֹכִ֖י לֹ֥א אֶשְׁכָּחֵֽךְ׃}
{Can a woman forget her sucking child, That she should not have compassion on the son of her womb? Yea, these may forget, Yet will not I forget thee.}
\twoverse{\arabic{verse}}
{הֵ֥ן עַל\maqqaf כַּפַּ֖יִם חַקֹּתִ֑יךְ חוֹמֹתַ֥יִךְ נֶגְדִּ֖י תָּמִֽיד׃}
{Behold, I have graven thee upon the palms of My hands; Thy walls are continually before Me.}
\twoverse{\arabic{verse}}
{מִהֲר֖וּ בָּנָ֑יִךְ מְהָֽרְסַ֥יִךְ וּמַחֲרִיבַ֖יִךְ מִמֵּ֥ךְ יֵצֵֽאוּ׃}
{Thy children make haste; Thy destroyers and they that made thee waste shall go forth from thee.}
\twoverse{\arabic{verse}}
{שְׂאִֽי\maqqaf סָבִ֤יב עֵינַ֙יִךְ֙ וּרְאִ֔י כֻּלָּ֖ם נִקְבְּצ֣וּ בָֽאוּ\maqqaf לָ֑ךְ חַי\maqqaf אָ֣נִי נְאֻם\maqqaf יְהֹוָ֗ה כִּ֤י כֻלָּם֙ כָּעֲדִ֣י תִלְבָּ֔שִׁי וּֽתְקַשְּׁרִ֖ים כַּכַּלָּֽה׃}
{Lift up thine eyes round about, and behold: All these gather themselves together, and come to thee. As I live, saith the \lord, Thou shalt surely clothe thee with them all as with an ornament, And gird thyself with them, like a bride.}
\twoverse{\arabic{verse}}
{כִּ֤י חׇרְבֹתַ֙יִךְ֙ וְשֹׁ֣מְמֹתַ֔יִךְ וְאֶ֖רֶץ הֲרִסֻתֵ֑ךְ כִּ֤י עַתָּה֙ תֵּצְרִ֣י מִיּוֹשֵׁ֔ב וְרָחֲק֖וּ מְבַלְּעָֽיִךְ׃}
{For thy waste and thy desolate places And thy land that hath been destroyed— Surely now shalt thou be too strait for the inhabitants, And they that swallowed thee up shall be far away.}
\twoverse{\arabic{verse}}
{ע֚וֹד יֹאמְר֣וּ בְאׇזְנַ֔יִךְ בְּנֵ֖י שִׁכֻּלָ֑יִךְ צַר\maqqaf לִ֥י הַמָּק֖וֹם גְּשָׁה\maqqaf לִּ֥י וְאֵשֵֽׁבָה׃}
{The children of thy bereavement Shall yet say in thine ears: ‘The place is too strait for me; Give place to me that I may dwell.’}
\twoverse{\arabic{verse}}
{וְאָמַ֣רְתְּ בִּלְבָבֵ֗ךְ מִ֤י יָֽלַד\maqqaf לִי֙ אֶת\maqqaf אֵ֔לֶּה וַאֲנִ֥י שְׁכוּלָ֖ה וְגַלְמוּדָ֑ה גֹּלָ֣ה \legarmeh  וְסוּרָ֗ה וְאֵ֙לֶּה֙ מִ֣י גִדֵּ֔ל הֵ֤ן אֲנִי֙ נִשְׁאַ֣רְתִּי לְבַדִּ֔י אֵ֖לֶּה אֵיפֹ֥ה הֵֽם׃ \petucha }
{Then shalt thou say in thy heart: ‘Who hath begotten me these, Seeing I have been bereaved of my children, and am solitary, An exile, and wandering to and fro? And who hath brought up these? Behold, I was left alone; These, where were they?’}
\twoverse{\arabic{verse}}
{כֹּֽה\maqqaf אָמַ֞ר אֲדֹנָ֣י יֱהֹוִ֗ה הִנֵּ֨ה אֶשָּׂ֤א אֶל\maqqaf גּוֹיִם֙ יָדִ֔י וְאֶל\maqqaf עַמִּ֖ים אָרִ֣ים נִסִּ֑י וְהֵבִ֤יאוּ בָנַ֙יִךְ֙ בְּחֹ֔צֶן וּבְנֹתַ֖יִךְ עַל\maqqaf כָּתֵ֥ף תִּנָּשֶֽׂאנָה׃}
{Thus saith the Lord \textsc{God}: Behold, I will lift up My hand to the nations, And set up Mine ensign to the peoples, And they shall bring thy sons in their bosom, And thy daughters shall be carried upon their shoulders.}
\twoverse{\arabic{verse}}
{וְהָי֨וּ מְלָכִ֜ים אֹמְנַ֗יִךְ וְשָׂרֽוֹתֵיהֶם֙ מֵינִ֣יקֹתַ֔יִךְ אַפַּ֗יִם אֶ֚רֶץ יִשְׁתַּ֣חֲווּ לָ֔ךְ וַעֲפַ֥ר רַגְלַ֖יִךְ יְלַחֵ֑כוּ וְיָדַ֙עַתְּ֙ כִּי\maqqaf אֲנִ֣י יְהֹוָ֔ה אֲשֶׁ֥ר לֹא\maqqaf יֵבֹ֖שׁוּ קֹוָֽי׃ \setuma }
{And kings shall be thy foster-fathers, And their queens thy nursing mothers; They shall bow down to thee with their face to the earth, And lick the dust of thy feet; And thou shalt know that I am the \lord, For they shall not be ashamed that wait for Me.}
\twoverse{\arabic{verse}}
{הֲיֻקַּ֥ח מִגִּבּ֖וֹר מַלְק֑וֹחַ וְאִם\maqqaf שְׁבִ֥י צַדִּ֖יק יִמָּלֵֽט׃ \setuma }
{Shall the prey be taken from the mighty, Or the captives of the victorious be delivered?}
\twoverse{\arabic{verse}}
{כִּי\maqqaf כֹ֣ה \legarmeh  אָמַ֣ר יְהֹוָ֗ה גַּם\maqqaf שְׁבִ֤י גִבּוֹר֙ יֻקָּ֔ח וּמַלְק֥וֹחַ עָרִ֖יץ יִמָּלֵ֑ט וְאֶת\maqqaf יְרִיבֵךְ֙ אָנֹכִ֣י אָרִ֔יב וְאֶת\maqqaf בָּנַ֖יִךְ אָנֹכִ֥י אוֹשִֽׁיעַ׃}
{But thus saith the \lord: Even the captives of the mighty shall be taken away, And the prey of the terrible shall be delivered; And I will contend with him that contendeth with thee, And I will save thy children.}
\twoverse{\arabic{verse}}
{וְהַאֲכַלְתִּ֤י אֶת\maqqaf מוֹנַ֙יִךְ֙ אֶת\maqqaf בְּשָׂרָ֔ם וְכֶעָסִ֖יס דָּמָ֣ם יִשְׁכָּר֑וּן וְיָדְע֣וּ כׇל\maqqaf בָּשָׂ֗ר כִּ֣י אֲנִ֤י יְהֹוָה֙ מֽוֹשִׁיעֵ֔ךְ וְגֹאֲלֵ֖ךְ אֲבִ֥יר יַעֲקֹֽב׃ \setuma }
{And I will feed them that oppress thee with their own flesh; And they shall be drunken with their own blood, as with sweet wine; And all flesh shall know that I the \lord\space am thy Saviour, And thy Redeemer, the Mighty One of Jacob.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{כֹּ֣ה \legarmeh  אָמַ֣ר יְהֹוָ֗ה אֵ֣י זֶ֠ה סֵ֣פֶר כְּרִית֤וּת אִמְּכֶם֙ אֲשֶׁ֣ר שִׁלַּחְתִּ֔יהָ א֚וֹ מִ֣י מִנּוֹשַׁ֔י אֲשֶׁר\maqqaf מָכַ֥רְתִּי אֶתְכֶ֖ם ל֑וֹ הֵ֤ן בַּעֲוֺנֹֽתֵיכֶם֙ נִמְכַּרְתֶּ֔ם וּבְפִשְׁעֵיכֶ֖ם שֻׁלְּחָ֥ה אִמְּכֶֽם׃}
{Thus saith the \lord: Where is the bill of your mother’s divorcement, Wherewith I have put her away? Or which of My creditors is it To whom I have sold you? Behold, for your iniquities were ye sold, And for your transgressions was your mother put away.}
\twoverse{\arabic{verse}}
{מַדּ֨וּעַ בָּ֜אתִי וְאֵ֣ין אִ֗ישׁ קָרָ֘אתִי֮ וְאֵ֣ין עוֹנֶה֒ הֲקָצ֨וֹר קָצְרָ֤ה יָדִי֙ מִפְּד֔וּת וְאִם\maqqaf אֵֽין\maqqaf בִּ֥י כֹ֖חַ לְהַצִּ֑יל הֵ֣ן בְּגַעֲרָתִ֞י אַחֲרִ֣יב יָ֗ם אָשִׂ֤ים נְהָרוֹת֙ מִדְבָּ֔ר תִּבְאַ֤שׁ דְּגָתָם֙ מֵאֵ֣ין מַ֔יִם וְתָמֹ֖ת בַּצָּמָֽא׃}
{Wherefore, when I came, was there no man? When I called, was there none to answer? Is My hand shortened at all, that it cannot redeem? Or have I no power to deliver? Behold, at My rebuke I dry up the sea, I make the rivers a wilderness; Their fish become foul, because there is no water, And die for thirst.}
\twoverse{\arabic{verse}}
{אַלְבִּ֥ישׁ שָׁמַ֖יִם קַדְר֑וּת וְשַׂ֖ק אָשִׂ֥ים כְּסוּתָֽם׃ \petucha }
{I clothe the heavens with blackness, And I make sackcloth their covering.}
\twoverse{\arabic{verse}}
{אֲדֹנָ֣י יֱהֹוִ֗ה נָ֤תַן לִי֙ לְשׁ֣וֹן לִמּוּדִ֔ים לָדַ֛עַת לָע֥וּת אֶת\maqqaf יָעֵ֖ף דָּבָ֑ר יָעִ֣יר \legarmeh  בַּבֹּ֣קֶר בַּבֹּ֗קֶר יָעִ֥יר לִי֙ אֹ֔זֶן לִשְׁמֹ֖עַ כַּלִּמּוּדִֽים׃}
{The Lord \textsc{God} hath given me The tongue of them that are taught, That I should know how to sustain with words him that is weary; He wakeneth morning by morning, He wakeneth mine ear To hear as they that are taught.}
\twoverse{\arabic{verse}}
{אֲדֹנָ֤י יֱהֹוִה֙ פָּתַֽח\maqqaf לִ֣י אֹ֔זֶן וְאָנֹכִ֖י לֹ֣א מָרִ֑יתִי אָח֖וֹר לֹ֥א נְסוּגֹֽתִי׃}
{The Lord \textsc{God} hath opened mine ear, And I was not rebellious, Neither turned away backward.}
\twoverse{\arabic{verse}}
{גֵּוִי֙ נָתַ֣תִּי לְמַכִּ֔ים וּלְחָיַ֖י לְמֹֽרְטִ֑ים פָּנַי֙ לֹ֣א הִסְתַּ֔רְתִּי מִכְּלִמּ֖וֹת וָרֹֽק׃}
{I gave my back to the smiters, And my checks to them that plucked off the hair; I hid not my face from shame and spitting.}
\twoverse{\arabic{verse}}
{וַֽאדֹנָ֤י יֱהֹוִה֙ יַֽעֲזׇר\maqqaf לִ֔י עַל\maqqaf כֵּ֖ן לֹ֣א נִכְלָ֑מְתִּי עַל\maqqaf כֵּ֞ן שַׂ֤מְתִּי פָנַי֙ כַּחַלָּמִ֔ישׁ וָאֵדַ֖ע כִּי\maqqaf לֹ֥א אֵבֽוֹשׁ׃}
{For the Lord \textsc{God} will help me; Therefore have I not been confounded; Therefore have I set my face like a flint, And I know that I shall not be ashamed.}
\twoverse{\arabic{verse}}
{קָרוֹב֙ מַצְדִּיקִ֔י מִֽי\maqqaf יָרִ֥יב אִתִּ֖י נַ֣עַמְדָה יָּ֑חַד מִי\maqqaf בַ֥עַל מִשְׁפָּטִ֖י יִגַּ֥שׁ אֵלָֽי׃}
{He is near that justifieth me; Who will contend with me? let us stand up together; Who is mine adversary? let him come near to me.}
\twoverse{\arabic{verse}}
{הֵ֣ן אֲדֹנָ֤י יֱהֹוִה֙ יַעֲזׇר\maqqaf לִ֔י מִי\maqqaf ה֖וּא יַרְשִׁיעֵ֑נִי הֵ֤ן כֻּלָּם֙ כַּבֶּ֣גֶד יִבְל֔וּ עָ֖שׁ יֹאכְלֵֽם׃ \setuma }
{Behold, the Lord \textsc{God} will help me; Who is he that shall condemn me? Behold, they all shall wax old as a garment, The moth shall eat them up.}
\twoverse{\arabic{verse}}
{מִ֤י בָכֶם֙ יְרֵ֣א יְהֹוָ֔ה שֹׁמֵ֖עַ בְּק֣וֹל עַבְדּ֑וֹ אֲשֶׁ֣ר \legarmeh  הָלַ֣ךְ חֲשֵׁכִ֗ים וְאֵ֥ין נֹ֙גַהּ֙ ל֔וֹ יִבְטַח֙ בְּשֵׁ֣ם יְהֹוָ֔ה וְיִשָּׁעֵ֖ן בֵּאלֹהָֽיו׃ \setuma }
{Who is among you that feareth the \lord, That obeyeth the voice of His servant? Though he walketh in darkness, And hath no light, Let him trust in the name of the \lord, And stay upon his God.}
\twoverse{\arabic{verse}}
{הֵ֧ן כֻּלְּכֶ֛ם קֹ֥דְחֵי אֵ֖שׁ מְאַזְּרֵ֣י זִיק֑וֹת לְכ֣וּ \legarmeh  בְּא֣וּר אֶשְׁכֶ֗ם וּבְזִיקוֹת֙ בִּֽעַרְתֶּ֔ם מִיָּדִי֙ הָיְתָה\maqqaf זֹּ֣את לָכֶ֔ם לְמַעֲצֵבָ֖ה תִּשְׁכָּבֽוּן׃ \setuma }
{Behold, all ye that kindle a fire, That gird yourselves with firebrands, Begone in the flame of your fire, And among the brands that ye have kindled. This shall ye have of My hand; Ye shall lie down in sorrow.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{שִׁמְע֥וּ אֵלַ֛י רֹ֥דְפֵי צֶ֖דֶק מְבַקְשֵׁ֣י יְהֹוָ֑ה הַבִּ֙יטוּ֙ אֶל\maqqaf צ֣וּר חֻצַּבְתֶּ֔ם וְאֶל\maqqaf מַקֶּ֥בֶת בּ֖וֹר נֻקַּרְתֶּֽם׃}
{Hearken to Me, ye that follow after righteousness, Ye that seek the \lord; Look unto the rock whence ye were hewn, And to the hole of the pit whence ye were digged.}
\twoverse{\arabic{verse}}
{הַבִּ֙יטוּ֙ אֶל\maqqaf אַבְרָהָ֣ם אֲבִיכֶ֔ם וְאֶל\maqqaf שָׂרָ֖ה תְּחוֹלֶלְכֶ֑ם כִּֽי\maqqaf אֶחָ֣ד קְרָאתִ֔יו וַאֲבָרְכֵ֖הוּ וְאַרְבֵּֽהוּ׃}
{Look unto Abraham your father, And unto Sarah that bore you; For when he was but one I called him, And I blessed him, and made him many.}
\twoverse{\arabic{verse}}
{כִּֽי\maqqaf נִחַ֨ם יְהֹוָ֜ה צִיּ֗וֹן נִחַם֙ כׇּל\maqqaf חׇרְבֹתֶ֔יהָ וַיָּ֤שֶׂם מִדְבָּרָהּ֙ כְּעֵ֔דֶן וְעַרְבָתָ֖הּ כְּגַן\maqqaf יְהֹוָ֑ה שָׂשׂ֤וֹן וְשִׂמְחָה֙ יִמָּ֣צֵא בָ֔הּ תּוֹדָ֖ה וְק֥וֹל זִמְרָֽה׃ \setuma }
{For the \lord\space hath comforted Zion; He hath comforted all her waste places, And hath made her wilderness like Eden, And her desert like the garden of the \lord; Joy and gladness shall be found therein, Thanksgiving, and the voice of melody.}
\newperek
\newhaftarah{ראה}
\label{haft_47}
\haftnote{The Haftara is Isaiah 54:11\verserangechar 55:5. This is the Third Haftara of Consolation.  On Rosh \d{H}odesh, read the Maftir on page \pageref{maft_ShabRCh}. Most read the Haftara for Rosh \d{H}odesh when that occurs (and combine this Haftara with Ki Tetze), but the special Haftara for Erev Rosh \d{H}odesh is not read when this week is Erev Rosh \d{H}odesh.}
\setcounter{chap}{54}
\setcounter{verse}{11}
\twoverse{\Roman{chap}:\arabic{verse}}
{עֲנִיָּ֥ה סֹעֲרָ֖ה לֹ֣א נֻחָ֑מָה הִנֵּ֨ה אָנֹכִ֜י מַרְבִּ֤יץ בַּפּוּךְ֙ אֲבָנַ֔יִךְ וִֽיסַדְתִּ֖יךְ בַּסַּפִּירִֽים׃}
{O thou afflicted, tossed with tempest, and not comforted, behold, I will set thy stones in fair colours, And lay thy foundations with sapphires.}
\twoverse{\arabic{verse}}
{וְשַׂמְתִּ֤י כַּֽדְכֹד֙ שִׁמְשֹׁתַ֔יִךְ וּשְׁעָרַ֖יִךְ לְאַבְנֵ֣י אֶקְדָּ֑ח וְכׇל\maqqaf גְּבוּלֵ֖ךְ לְאַבְנֵי\maqqaf חֵֽפֶץ׃}
{And I will make thy pinnacles of rubies, and thy gates of carbuncles, and all thy border of precious stones.}
\twoverse{\arabic{verse}}
{וְכׇל\maqqaf בָּנַ֖יִךְ לִמּוּדֵ֣י יְהֹוָ֑ה וְרַ֖ב שְׁל֥וֹם בָּנָֽיִךְ׃}
{And all thy children shall be taught of the \lord; and great shall be the peace of thy children.}
\twoverse{\arabic{verse}}
{בִּצְדָקָ֖ה תִּכּוֹנָ֑נִי רַחֲקִ֤י מֵעֹ֙שֶׁק֙ כִּי\maqqaf לֹ֣א תִירָ֔אִי וּמִ֨מְּחִתָּ֔ה כִּ֥י לֹֽא\maqqaf תִקְרַ֖ב אֵלָֽיִךְ׃}
{In righteousness shalt thou be established; be thou far from oppression, for thou shalt not fear, And from ruin, for it shall not come near thee.}
\twoverse{\arabic{verse}}
{הֵ֣ן גּ֥וֹר יָג֛וּר אֶ֖פֶס מֵאוֹתִ֑י מִי\maqqaf גָ֥ר אִתָּ֖ךְ עָלַ֥יִךְ יִפּֽוֹל׃}
{Behold, they may gather together, but not by Me; Whosoever shall gather together against thee shall fall because of thee.}
\twoverse{\arabic{verse}}
{\qk{הִנֵּ֤ה}{הן} אָנֹכִי֙ בָּרָ֣אתִי חָרָ֔שׁ נֹפֵ֙חַ֙ בְּאֵ֣שׁ פֶּחָ֔ם וּמוֹצִ֥יא כְלִ֖י לְמַעֲשֵׂ֑הוּ וְאָנֹכִ֛י בָּרָ֥אתִי מַשְׁחִ֖ית לְחַבֵּֽל׃}
{Behold, I have created the smith That bloweth the fire of coals, And bringeth forth a weapon for his work; And I have created the waster to destroy.}
\twoverse{\arabic{verse}}
{כׇּל\maqqaf כְּלִ֞י יוּצַ֤ר עָלַ֙יִךְ֙ לֹ֣א יִצְלָ֔ח וְכׇל\maqqaf לָשׁ֛וֹן תָּקוּם\maqqaf אִתָּ֥ךְ לַמִּשְׁפָּ֖ט תַּרְשִׁ֑יעִי זֹ֡את נַחֲלַת֩ עַבְדֵ֨י יְהֹוָ֧ה וְצִדְקָתָ֛ם מֵאִתִּ֖י נְאֻם\maqqaf יְהֹוָֽה׃ \setuma }
{No weapon that is formed against thee shall prosper; And every tongue that shall rise against thee in judgment thou shalt condemn. This is the heritage of the servants of the \lord, And their due reward from Me, saith the \lord.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{ה֤וֹי כׇּל\maqqaf צָמֵא֙ לְכ֣וּ לַמַּ֔יִם וַאֲשֶׁ֥ר אֵֽין\maqqaf ל֖וֹ כָּ֑סֶף לְכ֤וּ שִׁבְרוּ֙ וֶאֱכֹ֔לוּ וּלְכ֣וּ שִׁבְר֗וּ בְּלוֹא\maqqaf כֶ֛סֶף וּבְל֥וֹא מְחִ֖יר יַ֥יִן וְחָלָֽב׃}
{Ho, every one that thirsteth, come ye for water, And he that hath no money; Come ye, buy, and eat; Yea, come, buy wine and milk Without money and without price.}
\twoverse{\arabic{verse}}
{לָ֤מָּה תִשְׁקְלוּ\maqqaf כֶ֙סֶף֙ בְּלוֹא\maqqaf לֶ֔חֶם וִיגִֽיעֲכֶ֖ם בְּל֣וֹא לְשׇׂבְעָ֑ה שִׁמְע֨וּ שָׁמ֤וֹעַ אֵלַי֙ וְאִכְלוּ\maqqaf ט֔וֹב וְתִתְעַנַּ֥ג בַּדֶּ֖שֶׁן נַפְשְׁכֶֽם׃}
{Wherefore do ye spend money for that which is not bread? And your gain for that which satisfieth not? Hearken diligently unto Me, and eat ye that which is good, And let your soul delight itself in fatness.}
\twoverse{\arabic{verse}}
{הַטּ֤וּ אׇזְנְכֶם֙ וּלְכ֣וּ אֵלַ֔י שִׁמְע֖וּ וּתְחִ֣י נַפְשְׁכֶ֑ם וְאֶכְרְתָ֤ה לָכֶם֙ בְּרִ֣ית עוֹלָ֔ם חַֽסְדֵ֥י דָוִ֖ד הַנֶּֽאֱמָנִֽים׃}
{Incline your ear, and come unto Me; Hear, and your soul shall live; And I will make an everlasting covenant with you, Even the sure mercies of David.}
\twoverse{\arabic{verse}}
{הֵ֛ן עֵ֥ד לְאוּמִּ֖ים נְתַתִּ֑יו נָגִ֥יד וּמְצַוֵּ֖ה לְאֻמִּֽים׃}
{Behold, I have given him for a witness to the peoples, A prince and commander to the peoples.}
\twoverse{\arabic{verse}}
{הֵ֣ן גּ֤וֹי לֹֽא\maqqaf תֵדַע֙ תִּקְרָ֔א וְג֥וֹי לֹֽא\maqqaf יְדָע֖וּךָ אֵלֶ֣יךָ יָר֑וּצוּ לְמַ֙עַן֙ יְהֹוָ֣ה אֱלֹהֶ֔יךָ וְלִקְד֥וֹשׁ יִשְׂרָאֵ֖ל כִּ֥י פֵאֲרָֽךְ׃ \setuma }
{Behold, thou shalt call a nation that thou knowest not, And a nation that knew not thee shall run unto thee; Because of the \lord\space thy God, And for the Holy One of Israel, for He hath glorified thee.}
\newperek
\newhaftarah{שופטים}
\label{haft_48}
\haftnote{The Haftara is Isaiah 51:12\verserangechar 52:12. This is the Fourth Haftara of Consolation.}
\setcounter{chap}{51}
\setcounter{verse}{12}
\twoverse{\Roman{chap}:\arabic{verse}}
{אָנֹכִ֧י אָנֹכִ֛י ה֖וּא מְנַחֶמְכֶ֑ם מִי\maqqaf אַ֤תְּ וַתִּֽירְאִי֙ מֵאֱנ֣וֹשׁ יָמ֔וּת וּמִבֶּן\maqqaf אָדָ֖ם חָצִ֥יר יִנָּתֵֽן׃}
{I, even I, am He that comforteth you: Who art thou, that thou art afraid of man that shall die, And of the son of man that shall be made as grass;}
\twoverse{\arabic{verse}}
{וַתִּשְׁכַּ֞ח יְהֹוָ֣ה עֹשֶׂ֗ךָ נוֹטֶ֣ה שָׁמַ֘יִם֮ וְיֹסֵ֣ד אָ֒רֶץ֒ וַתְּפַחֵ֨ד תָּמִ֜יד כׇּל\maqqaf הַיּ֗וֹם מִפְּנֵי֙ חֲמַ֣ת הַמֵּצִ֔יק כַּאֲשֶׁ֥ר כּוֹנֵ֖ן לְהַשְׁחִ֑ית וְאַיֵּ֖ה חֲמַ֥ת הַמֵּצִֽיק׃}
{And hast forgotten the \lord\space thy Maker, That stretched forth the heavens, And laid the foundations of the earth; And fearest continually all the day Because of the fury of the oppressor, As he maketh ready to destroy? And where is the fury of the oppressor?}
\twoverse{\arabic{verse}}
{מִהַ֥ר צֹעֶ֖ה לְהִפָּתֵ֑חַ וְלֹא\maqqaf יָמ֣וּת לַשַּׁ֔חַת וְלֹ֥א יֶחְסַ֖ר לַחְמֽוֹ׃}
{He that is bent down shall speedily be loosed; And he shall not go down dying into the pit, Neither shall his bread fail.}
\twoverse{\arabic{verse}}
{וְאָנֹכִי֙ יְהֹוָ֣ה אֱלֹהֶ֔יךָ רֹגַ֣ע הַיָּ֔ם וַיֶּהֱמ֖וּ גַּלָּ֑יו יְהֹוָ֥ה צְבָא֖וֹת שְׁמֽוֹ׃}
{For I am the \lord\space thy God, Who stirreth up the sea, that the waves thereof roar; The \lord\space of hosts is His name.}
\twoverse{\arabic{verse}}
{וָאָשִׂ֤ם דְּבָרַי֙ בְּפִ֔יךָ וּבְצֵ֥ל יָדִ֖י כִּסִּיתִ֑יךָ לִנְטֹ֤עַ שָׁמַ֙יִם֙ וְלִיסֹ֣ד אָ֔רֶץ וְלֵאמֹ֥ר לְצִיּ֖וֹן עַמִּי\maqqaf אָֽתָּה׃ \setuma }
{And I have put My words in thy mouth, And have covered thee in the shadow of My hand, That I may plant the heavens, And lay the foundations of the earth, And say unto Zion: ‘Thou art My people.’}
\twoverse{\arabic{verse}}
{הִתְעוֹרְרִ֣י הִֽתְעוֹרְרִ֗י ק֚וּמִי יְר֣וּשָׁלַ֔͏ִם אֲשֶׁ֥ר שָׁתִ֛ית מִיַּ֥ד יְהֹוָ֖ה אֶת\maqqaf כּ֣וֹס חֲמָת֑וֹ אֶת\maqqaf קֻבַּ֜עַת כּ֧וֹס הַתַּרְעֵלָ֛ה שָׁתִ֖ית מָצִֽית׃}
{Awake, awake, Stand up, O Jerusalem, That hast drunk at the hand of the \lord\space The cup of His fury; Thou hast drunken the beaker, even the cup of staggering, And drained it.}
\twoverse{\arabic{verse}}
{אֵין\maqqaf מְנַהֵ֣ל לָ֔הּ מִכׇּל\maqqaf בָּנִ֖ים יָלָ֑דָה וְאֵ֤ין מַחֲזִיק֙ בְּיָדָ֔הּ מִכׇּל\maqqaf בָּנִ֖ים גִּדֵּֽלָה׃}
{There is none to guide her Among all the sons whom she hath brought forth; Neither is there any that taketh her by the hand Of all the sons that she hath brought up.}
\twoverse{\arabic{verse}}
{שְׁתַּ֤יִם הֵ֙נָּה֙ קֹרְאֹתַ֔יִךְ מִ֖י יָנ֣וּד לָ֑ךְ הַשֹּׁ֧ד וְהַשֶּׁ֛בֶר וְהָרָעָ֥ב וְהַחֶ֖רֶב מִ֥י אֲנַחֲמֵֽךְ׃}
{These two things are befallen thee; Who shall bemoan thee? Desolation and destruction, And the famine and the sword; How shall I comfort thee?}
\twoverse{\arabic{verse}}
{בָּנַ֜יִךְ עֻלְּפ֥וּ שָׁכְב֛וּ בְּרֹ֥אשׁ כׇּל\maqqaf חוּצ֖וֹת כְּת֣וֹא מִכְמָ֑ר הַֽמְלֵאִ֥ים חֲמַת\maqqaf יְהֹוָ֖ה גַּעֲרַ֥ת אֱלֹהָֽיִךְ׃}
{Thy sons have fainted, they lie at the head of all the streets, As an antelope in a net; They are full of the fury of the \lord, The rebuke of thy God.}
\twoverse{\arabic{verse}}
{לָכֵ֛ן שִׁמְעִי\maqqaf נָ֥א זֹ֖את עֲנִיָּ֑ה וּשְׁכֻרַ֖ת וְלֹ֥א מִיָּֽיִן׃ \petucha }
{Therefore hear now this, thou afflicted, And drunken, but not with wine;}
\twoverse{\arabic{verse}}
{כֹּה\maqqaf אָמַ֞ר אֲדֹנַ֣יִךְ יְהֹוָ֗ה וֵאלֹהַ֙יִךְ֙ יָרִ֣יב עַמּ֔וֹ הִנֵּ֥ה לָקַ֛חְתִּי מִיָּדֵ֖ךְ אֶת\maqqaf כּ֣וֹס הַתַּרְעֵלָ֑ה אֶת\maqqaf קֻבַּ֙עַת֙ כּ֣וֹס חֲמָתִ֔י לֹא\maqqaf תוֹסִ֥יפִי לִשְׁתּוֹתָ֖הּ עֽוֹד׃}
{Thus saith thy Lord the \lord, And thy God that pleadeth the cause of His people: Behold, I have taken out of thy hand The cup of staggering; The beaker, even the cup of My fury, Thou shalt no more drink it again;}
\twoverse{\arabic{verse}}
{וְשַׂמְתִּ֙יהָ֙ בְּיַד\maqqaf מוֹגַ֔יִךְ אֲשֶׁר\maqqaf אָמְר֥וּ לְנַפְשֵׁ֖ךְ שְׁחִ֣י וְנַעֲבֹ֑רָה וַתָּשִׂ֤ימִי כָאָ֙רֶץ֙ גֵּוֵ֔ךְ וְכַח֖וּץ לַעֹבְרִֽים׃ \petucha }
{And I will put it into the hand of them that afflict thee; That have said to thy soul: ‘Bow down, that we may go over’; And thou hast laid thy back as the ground, And as the street, to them that go over.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{עוּרִ֥י עוּרִ֛י לִבְשִׁ֥י עֻזֵּ֖ךְ צִיּ֑וֹן לִבְשִׁ֣י \legarmeh  בִּגְדֵ֣י תִפְאַרְתֵּ֗ךְ יְרֽוּשָׁלַ֙͏ִם֙ עִ֣יר הַקֹּ֔דֶשׁ כִּ֣י לֹ֥א יוֹסִ֛יף יָבֹא\maqqaf בָ֥ךְ ע֖וֹד עָרֵ֥ל וְטָמֵֽא׃}
{Awake, awake, Put on thy strength, O Zion; Put on thy beautiful garments, O Jerusalem, the holy city; For henceforth there shall no more come into thee The uncircumcised and the unclean.}
\twoverse{\arabic{verse}}
{הִתְנַעֲרִ֧י מֵעָפָ֛ר ק֥וּמִי שְּׁבִ֖י יְרוּשָׁלָ֑͏ִם \qk{הִֽתְפַּתְּחִי֙}{התפתחו} מוֹסְרֵ֣י צַוָּארֵ֔ךְ שְׁבִיָּ֖ה בַּת\maqqaf צִיּֽוֹן׃ \setuma }
{Shake thyself from the dust; Arise, and sit down, O Jerusalem; Loose thyself from the bands of thy neck, O captive daughter of Zion.}
\twoverse{\arabic{verse}}
{כִּי\maqqaf כֹה֙ אָמַ֣ר יְהֹוָ֔ה חִנָּ֖ם נִמְכַּרְתֶּ֑ם וְלֹ֥א בְכֶ֖סֶף תִּגָּאֵֽלוּ׃ \setuma }
{For thus saith the \lord: Ye were sold for nought; And ye shall be redeemed without money.}
\twoverse{\arabic{verse}}
{כִּ֣י כֹ֤ה אָמַר֙ אֲדֹנָ֣י יֱהֹוִ֔ה מִצְרַ֛יִם יָֽרַד\maqqaf עַמִּ֥י בָרִֽאשֹׁנָ֖ה לָג֣וּר שָׁ֑ם וְאַשּׁ֖וּר בְּאֶ֥פֶס עֲשָׁקֽוֹ׃}
{For thus saith the Lord \textsc{God}: My people went down aforetime into Egypt to sojourn there; And the Assyrian oppressed them without cause.}
\twoverse{\arabic{verse}}
{וְעַתָּ֤ה מַה\maqqaf לִּי\maqqaf פֹה֙ נְאֻם\maqqaf יְהֹוָ֔ה כִּֽי\maqqaf לֻקַּ֥ח עַמִּ֖י חִנָּ֑ם מֹשְׁלָ֤ו יְהֵילִ֙ילוּ֙ נְאֻם\maqqaf יְהֹוָ֔ה וְתָמִ֥יד כׇּל\maqqaf הַיּ֖וֹם שְׁמִ֥י מִנֹּאָֽץ׃}
{Now therefore, what do I here, saith the \lord, Seeing that My people is taken away for nought? They that rule over them do howl, saith the \lord, And My name continually all the day is blasphemed.}
\twoverse{\arabic{verse}}
{לָכֵ֛ן יֵדַ֥ע עַמִּ֖י שְׁמִ֑י לָכֵן֙ בַּיּ֣וֹם הַה֔וּא כִּֽי\maqqaf אֲנִי\maqqaf ה֥וּא הַֽמְדַבֵּ֖ר הִנֵּֽנִי׃ \setuma }
{Therefore My people shall know My name; Therefore they shall know in that day That I, even He that spoke, behold, here I am.}
\twoverse{\arabic{verse}}
{מַה\maqqaf נָּאו֨וּ עַל\maqqaf הֶהָרִ֜ים רַגְלֵ֣י מְבַשֵּׂ֗ר מַשְׁמִ֧יעַ שָׁל֛וֹם מְבַשֵּׂ֥ר ט֖וֹב מַשְׁמִ֣יעַ יְשׁוּעָ֑ה אֹמֵ֥ר לְצִיּ֖וֹן מָלַ֥ךְ אֱלֹהָֽיִךְ׃}
{How beautiful upon the mountains Are the feet of the messenger of good tidings, That announceth peace, the harbinger of good tidings, That announceth salvation; That saith unto Zion: ‘Thy God reigneth!’}
\twoverse{\arabic{verse}}
{ק֥וֹל צֹפַ֛יִךְ נָ֥שְׂאוּ ק֖וֹל יַחְדָּ֣ו יְרַנֵּ֑נוּ כִּ֣י עַ֤יִן בְּעַ֙יִן֙ יִרְא֔וּ בְּשׁ֥וּב יְהֹוָ֖ה צִיּֽוֹן׃}
{Hark, thy watchmen! they lift up the voice, Together do they sing; For they shall see, eye to eye, The \lord\space returning to Zion.}
\twoverse{\arabic{verse}}
{פִּצְח֤וּ רַנְּנוּ֙ יַחְדָּ֔ו חׇרְב֖וֹת יְרוּשָׁלָ֑͏ִם כִּֽי\maqqaf נִחַ֤ם יְהֹוָה֙ עַמּ֔וֹ גָּאַ֖ל יְרוּשָׁלָֽ͏ִם׃}
{Break forth into joy, sing together, Ye waste places of Jerusalem; For the \lord\space hath comforted His people, He hath redeemed Jerusalem.}
\twoverse{\arabic{verse}}
{חָשַׂ֤ף יְהֹוָה֙ אֶת\maqqaf זְר֣וֹעַ קׇדְשׁ֔וֹ לְעֵינֵ֖י כׇּל\maqqaf הַגּוֹיִ֑ם וְרָאוּ֙ כׇּל\maqqaf אַפְסֵי\maqqaf אָ֔רֶץ אֵ֖ת יְשׁוּעַ֥ת אֱלֹהֵֽינוּ׃ \setuma }
{The \lord\space hath made bare His holy arm In the eyes of all the nations; And all the ends of the earth shall see The salvation of our God.}
\twoverse{\arabic{verse}}
{ס֤וּרוּ ס֙וּרוּ֙ צְא֣וּ מִשָּׁ֔ם טָמֵ֖א אַל\maqqaf תִּגָּ֑עוּ צְא֣וּ מִתּוֹכָ֔הּ הִבָּ֕רוּ נֹשְׂאֵ֖י כְּלֵ֥י יְהֹוָֽה׃}
{Depart ye, depart ye, go ye out from thence, Touch no unclean thing; Go ye out of the midst of her; be ye clean, Ye that bear the vessels of the \lord.}
\twoverse{\arabic{verse}}
{כִּ֣י לֹ֤א בְחִפָּזוֹן֙ תֵּצֵ֔אוּ וּבִמְנוּסָ֖ה לֹ֣א תֵלֵכ֑וּן כִּֽי\maqqaf הֹלֵ֤ךְ לִפְנֵיכֶם֙ יְהֹוָ֔ה וּמְאַסִּפְכֶ֖ם אֱלֹהֵ֥י יִשְׂרָאֵֽל׃ \setuma }
{For ye shall not go out in haste, Neither shall ye go by flight; For the \lord\space will go before you, And the God of Israel will be your rearward.}
\newperek
\newhaftarah{כי תצא}
\label{haft_49}
\haftnote{The Haftara is Isaiah 54:1\verserangechar 54:10. This is the Fifth Haftara of Consolation. If the Haftara from Re'eh was not read because it fell on Rosh \d{H}odesh, combine it with this Haftara. It is on page \pageref{haft_47}.}
\setcounter{chap}{54}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}
{רׇנִּ֥י עֲקָרָ֖ה לֹ֣א יָלָ֑דָה פִּצְחִ֨י רִנָּ֤ה וְצַהֲלִי֙ לֹא\maqqaf חָ֔לָה כִּֽי\maqqaf רַבִּ֧ים בְּֽנֵי\maqqaf שׁוֹמֵמָ֛ה מִבְּנֵ֥י בְעוּלָ֖ה אָמַ֥ר יְהֹוָֽה׃}
{Sing, O barren, thou that didst not bear, Break forth into singing, and cry aloud, thou that didst not travail; For more are the children of the desolate Than the children of the married wife, saith the \lord.}
\twoverse{\arabic{verse}}
{הַרְחִ֣יבִי \legarmeh  מְק֣וֹם אׇהֳלֵ֗ךְ וִֽירִיע֧וֹת מִשְׁכְּנוֹתַ֛יִךְ יַטּ֖וּ אַל\maqqaf תַּחְשֹׂ֑כִי הַאֲרִ֙יכִי֙ מֵיתָרַ֔יִךְ וִיתֵדֹתַ֖יִךְ חַזֵּֽקִי׃}
{Enlarge the place of thy tent, And let them stretch forth the curtains of thy habitations, spare not; Lengthen thy cords, and strengthen thy stakes.}
\twoverse{\arabic{verse}}
{כִּֽי\maqqaf יָמִ֥ין וּשְׂמֹ֖אול תִּפְרֹ֑צִי וְזַרְעֵךְ֙ גּוֹיִ֣ם יִירָ֔שׁ וְעָרִ֥ים נְשַׁמּ֖וֹת יוֹשִֽׁיבוּ׃}
{For thou shalt spread abroad on the right hand and on the left; And thy seed shall possess the nations, And make the desolate cities to be inhabited.}
\twoverse{\arabic{verse}}
{אַל\maqqaf תִּֽירְאִי֙ כִּי\maqqaf לֹ֣א תֵב֔וֹשִׁי וְאַל\maqqaf תִּכָּֽלְמִ֖י כִּ֣י לֹ֣א תַחְפִּ֑ירִי כִּ֣י בֹ֤שֶׁת עֲלוּמַ֙יִךְ֙ תִּשְׁכָּ֔חִי וְחֶרְפַּ֥ת אַלְמְנוּתַ֖יִךְ לֹ֥א תִזְכְּרִי\maqqaf עֽוֹד׃}
{Fear not, for thou shalt not be ashamed. Neither be thou confounded, for thou shalt not be put to shame; For thou shalt forget the shame of thy youth, And the reproach of thy widowhood shalt thou remember no more.}
\twoverse{\arabic{verse}}
{כִּ֤י בֹעֲלַ֙יִךְ֙ עֹשַׂ֔יִךְ יְהֹוָ֥ה צְבָא֖וֹת שְׁמ֑וֹ וְגֹֽאֲלֵךְ֙ קְד֣וֹשׁ יִשְׂרָאֵ֔ל אֱלֹהֵ֥י כׇל\maqqaf הָאָ֖רֶץ יִקָּרֵֽא׃}
{For thy Maker is thy husband, The \lord\space of hosts is His name; And the Holy One of Israel is thy Redeemer, The God of the whole earth shall He be called.}
\twoverse{\arabic{verse}}
{כִּֽי\maqqaf כְאִשָּׁ֧ה עֲזוּבָ֛ה וַעֲצ֥וּבַת ר֖וּחַ קְרָאָ֣ךְ יְהֹוָ֑ה וְאֵ֧שֶׁת נְעוּרִ֛ים כִּ֥י תִמָּאֵ֖ס אָמַ֥ר אֱלֹהָֽיִךְ׃}
{For the \lord\space hath called thee As a wife forsaken and grieved in spirit; And a wife of youth, can she be rejected? Saith thy God.}
\twoverse{\arabic{verse}}
{בְּרֶ֥גַע קָטֹ֖ן עֲזַבְתִּ֑יךְ וּבְרַחֲמִ֥ים גְּדֹלִ֖ים אֲקַבְּצֵֽךְ׃}
{For a small moment have I forsaken thee; But with great compassion will I gather thee.}
\twoverse{\arabic{verse}}
{בְּשֶׁ֣צֶף קֶ֗צֶף הִסְתַּ֨רְתִּי פָנַ֥י רֶ֙גַע֙ מִמֵּ֔ךְ וּבְחֶ֥סֶד עוֹלָ֖ם רִחַמְתִּ֑יךְ אָמַ֥ר גֹּאֲלֵ֖ךְ יְהֹוָֽה׃ \setuma }
{In a little wrath I hid My face from thee for a moment; But with everlasting kindness will I have compassion on thee, Saith the \lord\space thy Redeemer.}
\twoverse{\arabic{verse}}
{כִּֽי\maqqaf מֵ֥י נֹ֙חַ֙ זֹ֣את לִ֔י אֲשֶׁ֣ר נִשְׁבַּ֗עְתִּי מֵעֲבֹ֥ר מֵי\maqqaf נֹ֛חַ ע֖וֹד עַל\maqqaf הָאָ֑רֶץ כֵּ֥ן נִשְׁבַּ֛עְתִּי מִקְּצֹ֥ף עָלַ֖יִךְ וּמִגְּעׇר\maqqaf בָּֽךְ׃}
{For this is as the waters of Noah unto Me; for as I have sworn that the waters of Noah should no more go over the earth, so have I sworn that I would not be wroth with thee, nor rebuke thee.}
\twoverse{\arabic{verse}}
{כִּ֤י הֶהָרִים֙ יָמ֔וּשׁוּ וְהַגְּבָע֖וֹת תְּמוּטֶ֑ינָה וְחַסְדִּ֞י מֵאִתֵּ֣ךְ לֹא\maqqaf יָמ֗וּשׁ וּבְרִ֤ית שְׁלוֹמִי֙ לֹ֣א תָמ֔וּט אָמַ֥ר מְרַחֲמֵ֖ךְ יְהֹוָֽה׃ \setuma }
{For the mountains may depart, and the hills be removed; but My kindness shall not depart from thee, neither shall My covenant of peace be removed, saith the \lord\space that hath compassion on thee.}
\newhaftarah{כי תבוא}
\label{haft_50}
\haftnote{The Haftara is Isaiah 60:1\verserangechar 60:22. This is the Sixth Haftara of Consolation.}
\setcounter{chap}{60}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}
{ק֥וּמִי א֖וֹרִי כִּ֣י בָ֣א אוֹרֵ֑ךְ וּכְב֥וֹד יְהֹוָ֖ה עָלַ֥יִךְ זָרָֽח׃}
{Arise, shine, for thy light is come, And the glory of the \lord\space is risen upon thee.}
\twoverse{\arabic{verse}}
{כִּֽי\maqqaf הִנֵּ֤ה הַחֹ֙שֶׁךְ֙ יְכַסֶּה\maqqaf אֶ֔רֶץ וַעֲרָפֶ֖ל לְאֻמִּ֑ים וְעָלַ֙יִךְ֙ יִזְרַ֣ח יְהֹוָ֔ה וּכְבוֹד֖וֹ עָלַ֥יִךְ יֵרָאֶֽה׃}
{For, behold, darkness shall cover the earth, And gross darkness the peoples; But upon thee the \lord\space will arise, And His glory shall be seen upon thee.}
\twoverse{\arabic{verse}}
{וְהָלְכ֥וּ גוֹיִ֖ם לְאוֹרֵ֑ךְ וּמְלָכִ֖ים לְנֹ֥גַהּ זַרְחֵֽךְ׃}
{And nations shall walk at thy light, And kings at the brightness of thy rising.}
\twoverse{\arabic{verse}}
{שְׂאִֽי\maqqaf סָבִ֤יב עֵינַ֙יִךְ֙ וּרְאִ֔י כֻּלָּ֖ם נִקְבְּצ֣וּ בָֽאוּ\maqqaf לָ֑ךְ בָּנַ֙יִךְ֙ מֵרָח֣וֹק יָבֹ֔אוּ וּבְנוֹתַ֖יִךְ עַל\maqqaf צַ֥ד תֵּאָמַֽנָה׃}
{Lift Up thine eyes round about, and see: They all are gathered together, and come to thee; Thy sons come from far, And thy daughters are borne on the side.}
\twoverse{\arabic{verse}}
{אָ֤ז תִּרְאִי֙ וְנָהַ֔רְתְּ וּפָחַ֥ד וְרָחַ֖ב לְבָבֵ֑ךְ כִּֽי\maqqaf יֵהָפֵ֤ךְ עָלַ֙יִךְ֙ הֲמ֣וֹן יָ֔ם חֵ֥יל גּוֹיִ֖ם יָבֹ֥אוּ לָֽךְ׃}
{Then thou shalt see and be radiant, And thy heart shall throb and be enlarged; Because the abundance of the sea shall be turned unto thee, The wealth of the nations shall come unto thee.}
\twoverse{\arabic{verse}}
{שִֽׁפְעַ֨ת גְּמַלִּ֜ים תְּכַסֵּ֗ךְ בִּכְרֵ֤י מִדְיָן֙ וְעֵיפָ֔ה כֻּלָּ֖ם מִשְּׁבָ֣א יָבֹ֑אוּ זָהָ֤ב וּלְבוֹנָה֙ יִשָּׂ֔אוּ וּתְהִלֹּ֥ת יְהֹוָ֖ה יְבַשֵּֽׂרוּ׃}
{The caravan of camels shall cover thee, And of the young camels of Midian and Ephah, All coming from Sheba; They shall bring gold and incense, And shall proclaim the praises of the \lord.}
\twoverse{\arabic{verse}}
{כׇּל\maqqaf צֹ֤אן קֵדָר֙ יִקָּ֣בְצוּ לָ֔ךְ אֵילֵ֥י נְבָי֖וֹת יְשָֽׁרְת֑וּנֶךְ יַעֲל֤וּ עַל\maqqaf רָצוֹן֙ מִזְבְּחִ֔י וּבֵ֥ית תִּפְאַרְתִּ֖י אֲפָאֵֽר׃}
{All the flocks of Kedar shall be gathered together unto thee, The rams of Nebaioth shall minister unto thee; They shall come up with acceptance on Mine altar, And I will glorify My glorious house.}
\twoverse{\arabic{verse}}
{מִי\maqqaf אֵ֖לֶּה כָּעָ֣ב תְּעוּפֶ֑ינָה וְכַיּוֹנִ֖ים אֶל\maqqaf אֲרֻבֹּתֵיהֶֽם׃}
{Who are these that fly as a cloud, And as the doves to their cotes?}
\twoverse{\arabic{verse}}
{כִּי\maqqaf לִ֣י \legarmeh  אִיִּ֣ים יְקַוּ֗וּ וׇֽאֳנִיּ֤וֹת תַּרְשִׁישׁ֙ בָּרִ֣אשֹׁנָ֔ה לְהָבִ֤יא בָנַ֙יִךְ֙ מֵֽרָח֔וֹק כַּסְפָּ֥ם וּזְהָבָ֖ם אִתָּ֑ם לְשֵׁם֙ יְהֹוָ֣ה אֱלֹהַ֔יִךְ וְלִקְד֥וֹשׁ יִשְׂרָאֵ֖ל כִּ֥י פֵאֲרָֽךְ׃}
{Surely the isles shall wait for Me, And the ships of Tarshish first, To bring thy sons from far, Their silver and their gold with them, For the name of the \lord\space thy God, And for the Holy One of Israel, because He hath glorified thee.}
\twoverse{\arabic{verse}}
{וּבָנ֤וּ בְנֵֽי\maqqaf נֵכָר֙ חֹמֹתַ֔יִךְ וּמַלְכֵיהֶ֖ם יְשָֽׁרְת֑וּנֶךְ כִּ֤י בְקִצְפִּי֙ הִכִּיתִ֔יךְ וּבִרְצוֹנִ֖י רִחַמְתִּֽיךְ׃}
{And aliens shall build up thy walls, And their kings shall minister unto thee; For in My wrath I smote thee, But in My favour have I had compassion on thee.}
\twoverse{\arabic{verse}}
{וּפִתְּח֨וּ שְׁעָרַ֧יִךְ תָּמִ֛יד יוֹמָ֥ם וָלַ֖יְלָה לֹ֣א יִסָּגֵ֑רוּ לְהָבִ֤יא אֵלַ֙יִךְ֙ חֵ֣יל גּוֹיִ֔ם וּמַלְכֵיהֶ֖ם נְהוּגִֽים׃}
{Thy gates also shall be open continually, Day and night, they shall not be shut; That men may bring unto thee the wealth of the nations, And their kings in procession.}
\twoverse{\arabic{verse}}
{כִּֽי\maqqaf הַגּ֧וֹי וְהַמַּמְלָכָ֛ה אֲשֶׁ֥ר לֹא\maqqaf יַעַבְד֖וּךְ יֹאבֵ֑דוּ וְהַגּוֹיִ֖ם חָרֹ֥ב יֶחֱרָֽבוּ׃}
{For that nation and kingdom that will not serve thee shall perish; Yea, those nations shall be utterly wasted.}
\twoverse{\arabic{verse}}
{כְּב֤וֹד הַלְּבָנוֹן֙ אֵלַ֣יִךְ יָב֔וֹא בְּר֛וֹשׁ תִּדְהָ֥ר וּתְאַשּׁ֖וּר יַחְדָּ֑ו לְפָאֵר֙ מְק֣וֹם מִקְדָּשִׁ֔י וּמְק֥וֹם רַגְלַ֖י אֲכַבֵּֽד׃}
{The glory of Lebanon shall come unto thee, The cypress, the plane-tree and the larch together; To beautify the place of My sanctuary, And I will make the place of My feet glorious.}
\twoverse{\arabic{verse}}
{וְהָלְכ֨וּ אֵלַ֤יִךְ שְׁח֙וֹחַ֙ בְּנֵ֣י מְעַנַּ֔יִךְ וְהִֽשְׁתַּחֲו֛וּ עַל\maqqaf כַּפּ֥וֹת רַגְלַ֖יִךְ כׇּל\maqqaf מְנַֽאֲצָ֑יִךְ וְקָ֤רְאוּ לָךְ֙ עִ֣יר יְהֹוָ֔ה צִיּ֖וֹן קְד֥וֹשׁ יִשְׂרָאֵֽל׃}
{And the sons of them that afflicted thee Shall come bending unto thee, And all they that despised thee shall bow down At the soles of thy feet; And they shall call thee The city of the \lord, The Zion of the Holy One of Israel.}
\twoverse{\arabic{verse}}
{תַּ֧חַת הֱיוֹתֵ֛ךְ עֲזוּבָ֥ה וּשְׂנוּאָ֖ה וְאֵ֣ין עוֹבֵ֑ר וְשַׂמְתִּיךְ֙ לִגְא֣וֹן עוֹלָ֔ם מְשׂ֖וֹשׂ דּ֥וֹר וָדֽוֹר׃}
{Whereas thou hast been forsaken and hated, So that no man passed through thee, I will make thee an eternal excellency, A joy of many generations.}
\twoverse{\arabic{verse}}
{וְיָנַקְתְּ֙ חֲלֵ֣ב גּוֹיִ֔ם וְשֹׁ֥ד מְלָכִ֖ים תִּינָ֑קִי וְיָדַ֗עַתְּ כִּ֣י אֲנִ֤י יְהֹוָה֙ מֽוֹשִׁיעֵ֔ךְ וְגֹאֲלֵ֖ךְ אֲבִ֥יר יַעֲקֹֽב׃}
{Thou shalt also suck the milk of the nations, And shalt suck the breast of kings; And thou shalt know that I the \lord\space am thy Saviour, And I, the Mighty One of Jacob, thy Redeemer.}
\twoverse{\arabic{verse}}
{תַּ֣חַת הַנְּחֹ֜שֶׁת אָבִ֣יא זָהָ֗ב וְתַ֤חַת הַבַּרְזֶל֙ אָ֣בִיא כֶ֔סֶף וְתַ֤חַת הָעֵצִים֙ נְחֹ֔שֶׁת וְתַ֥חַת הָאֲבָנִ֖ים בַּרְזֶ֑ל וְשַׂמְתִּ֤י פְקֻדָּתֵךְ֙ שָׁל֔וֹם וְנֹגְשַׂ֖יִךְ צְדָקָֽה׃}
{For brass I will bring gold, And for iron I will bring silver, And for wood brass, And for stones iron; I will also make thy officers peace, And righteousness thy magistrates.}
\twoverse{\arabic{verse}}
{לֹא\maqqaf יִשָּׁמַ֨ע ע֤וֹד חָמָס֙ בְּאַרְצֵ֔ךְ שֹׁ֥ד וָשֶׁ֖בֶר בִּגְבוּלָ֑יִךְ וְקָרָ֤את יְשׁוּעָה֙ חוֹמֹתַ֔יִךְ וּשְׁעָרַ֖יִךְ תְּהִלָּֽה׃}
{Violence shall no more be heard in thy land, Desolation nor destruction within thy borders; But thou shalt call thy walls Salvation, And thy gates Praise.}
\twoverse{\arabic{verse}}
{לֹא\maqqaf יִֽהְיֶה\maqqaf לָּ֨ךְ ע֤וֹד הַשֶּׁ֙מֶשׁ֙ לְא֣וֹר יוֹמָ֔ם וּלְנֹ֕גַהּ הַיָּרֵ֖חַ לֹא\maqqaf יָאִ֣יר לָ֑ךְ וְהָֽיָה\maqqaf לָ֤ךְ יְהֹוָה֙ לְא֣וֹר עוֹלָ֔ם וֵאלֹהַ֖יִךְ לְתִפְאַרְתֵּֽךְ׃}
{The sun shall be no more thy light by day, Neither for brightness shall the moon give light unto thee; But the \lord\space shall be unto thee an everlasting light, And thy God thy glory.}
\twoverse{\arabic{verse}}
{לֹא\maqqaf יָב֥וֹא עוֹד֙ שִׁמְשֵׁ֔ךְ וִירֵחֵ֖ךְ לֹ֣א יֵֽאָסֵ֑ף כִּ֣י יְהֹוָ֗ה יִֽהְיֶה\maqqaf לָּךְ֙ לְא֣וֹר עוֹלָ֔ם וְשָׁלְמ֖וּ יְמֵ֥י אֶבְלֵֽךְ׃}
{Thy sun shall no more go down, Neither shall thy moon withdraw itself; For the \lord\space shall be thine everlasting light, And the days of thy mourning shall be ended.}
\twoverse{\arabic{verse}}
{וְעַמֵּךְ֙ כֻּלָּ֣ם צַדִּיקִ֔ים לְעוֹלָ֖ם יִ֣ירְשׁוּ אָ֑רֶץ נֵ֧צֶר \qk{מַטָּעַ֛י}{מטעו} מַעֲשֵׂ֥ה יָדַ֖י לְהִתְפָּאֵֽר׃}
{Thy people also shall be all righteous, They shall inherit the land for ever; The branch of My planting, the work of My hands, Wherein I glory.}
\twoverse{\arabic{verse}}
{הַקָּטֹן֙ יִהְיֶ֣ה לָאֶ֔לֶף וְהַצָּעִ֖יר לְג֣וֹי עָצ֑וּם אֲנִ֥י יְהֹוָ֖ה בְּעִתָּ֥הּ אֲחִישֶֽׁנָּה׃ \setuma }
{The smallest shall become a thousand, And the least a mighty nation; I the \lord\space will hasten it in its time.}
\newhaftarah{נצבים}
\label{haft_51}
\haftnote{The Haftara is Isaiah 61:10\verserangechar 63:9. This is the Seventh (and final) Haftara of Consolation, and is read on the Shabbat before Rosh HaShana.}
\setcounter{chap}{61}
\setcounter{verse}{10}
\twoverse{\Roman{chap}:\arabic{verse}}
{שׂ֧וֹשׂ אָשִׂ֣ישׂ בַּֽיהֹוָ֗ה תָּגֵ֤ל נַפְשִׁי֙ בֵּֽאלֹהַ֔י כִּ֤י הִלְבִּישַׁ֙נִי֙ בִּגְדֵי\maqqaf יֶ֔שַׁע מְעִ֥יל צְדָקָ֖ה יְעָטָ֑נִי כֶּֽחָתָן֙ יְכַהֵ֣ן פְּאֵ֔ר וְכַכַּלָּ֖ה תַּעְדֶּ֥ה כֵלֶֽיהָ׃}
{I will greatly rejoice in the \lord, My soul shall be joyful in my God; For He hath clothed me with the garments of salvation, He hath covered me with the robe of victory, As a bridegroom putteth on a priestly diadem, And as a bride adorneth herself with her jewels.}
\twoverse{\arabic{verse}}
{כִּ֤י כָאָ֙רֶץ֙ תּוֹצִ֣יא צִמְחָ֔הּ וּכְגַנָּ֖ה זֵרוּעֶ֣יהָ תַצְמִ֑יחַ כֵּ֣ן \legarmeh  אֲדֹנָ֣י יֱהֹוִ֗ה יַצְמִ֤יחַ צְדָקָה֙ וּתְהִלָּ֔ה נֶ֖גֶד כׇּל\maqqaf הַגּוֹיִֽם׃}
{For as the earth bringeth forth her growth, And as the garden causeth the things that are sown in it to spring forth; So the Lord \textsc{God} will cause victory and glory To spring forth before all the nations.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{לְמַ֤עַן צִיּוֹן֙ לֹ֣א אֶחֱשֶׁ֔ה וּלְמַ֥עַן יְרוּשָׁלַ֖͏ִם לֹ֣א אֶשְׁק֑וֹט עַד\maqqaf יֵצֵ֤א כַנֹּ֙גַהּ֙ צִדְקָ֔הּ וִישׁוּעָתָ֖הּ כְּלַפִּ֥יד יִבְעָֽר׃}
{For Zion’s sake will I not hold My peace, And for Jerusalem’s sake I will not rest, Until her triumph go forth as brightness, And her salvation as a torch that burneth.}
\twoverse{\arabic{verse}}
{וְרָא֤וּ גוֹיִם֙ צִדְקֵ֔ךְ וְכׇל\maqqaf מְלָכִ֖ים כְּבוֹדֵ֑ךְ וְקֹ֤רָא לָךְ֙ שֵׁ֣ם חָדָ֔שׁ אֲשֶׁ֛ר פִּ֥י יְהֹוָ֖ה יִקֳּבֶֽנּוּ׃}
{And the nations shall see thy triumph, And all kings thy glory; And thou shalt be called by a new name, Which the mouth of the \lord\space shall mark out.}
\twoverse{\arabic{verse}}
{וְהָיִ֛ית עֲטֶ֥רֶת תִּפְאֶ֖רֶת בְּיַד\maqqaf יְהֹוָ֑ה \qk{וּצְנִ֥יף}{וצנוף} מְלוּכָ֖ה בְּכַף\maqqaf אֱלֹהָֽיִךְ׃}
{Thou shalt also be a crown of beauty in the hand of the \lord, And a royal diadem in the open hand of thy God.}
\twoverse{\arabic{verse}}
{לֹא\maqqaf יֵאָמֵר֩ לָ֨ךְ ע֜וֹד עֲזוּבָ֗ה וּלְאַרְצֵךְ֙ לֹֽא\maqqaf יֵאָמֵ֥ר עוֹד֙ שְׁמָמָ֔ה כִּ֣י לָ֗ךְ יִקָּרֵא֙ חֶפְצִי\maqqaf בָ֔הּ וּלְאַרְצֵ֖ךְ בְּעוּלָ֑ה כִּֽי\maqqaf חָפֵ֤ץ יְהֹוָה֙ בָּ֔ךְ וְאַרְצֵ֖ךְ תִּבָּעֵֽל׃}
{Thou shalt no more be termed Forsaken, Neither shall thy land any more be termed Desolate; But thou shalt be called, My delight is in her, And thy land, Espoused; For the \lord\space delighteth in thee, And thy land shall be espoused.}
\twoverse{\arabic{verse}}
{כִּֽי\maqqaf יִבְעַ֤ל בָּחוּר֙ בְּתוּלָ֔ה יִבְעָל֖וּךְ בָּנָ֑יִךְ וּמְשׂ֤וֹשׂ חָתָן֙ עַל\maqqaf כַּלָּ֔ה יָשִׂ֥ישׂ עָלַ֖יִךְ אֱלֹהָֽיִךְ׃}
{For as a young man espouseth a virgin, So shall thy sons espouse thee; And as the bridegroom rejoiceth over the bride, So shall thy God rejoice over thee.}
\twoverse{\arabic{verse}}
{עַל\maqqaf חֽוֹמֹתַ֣יִךְ יְרוּשָׁלַ֗͏ִם הִפְקַ֙דְתִּי֙ שֹֽׁמְרִ֔ים כׇּל\maqqaf הַיּ֧וֹם וְכׇל\maqqaf הַלַּ֛יְלָה תָּמִ֖יד לֹ֣א יֶחֱשׁ֑וּ הַמַּזְכִּרִים֙ אֶת\maqqaf יְהֹוָ֔ה אַל\maqqaf דֳּמִ֖י לָכֶֽם׃}
{I have set watchmen Upon thy walls, O Jerusalem, They shall never hold their peace Day nor night: ‘Ye that are the \lord\textsc{’s}\space remembrancers, Take ye no rest,}
\twoverse{\arabic{verse}}
{וְאַֽל\maqqaf תִּתְּנ֥וּ דֳמִ֖י ל֑וֹ עַד\maqqaf יְכוֹנֵ֞ן וְעַד\maqqaf יָשִׂ֧ים אֶת\maqqaf יְרוּשָׁלַ֛͏ִם תְּהִלָּ֖ה בָּאָֽרֶץ׃}
{And give Him no rest, Till He establish, And till He make Jerusalem A praise in the earth.’}
\twoverse{\arabic{verse}}
{נִשְׁבַּ֧ע יְהֹוָ֛ה בִּימִינ֖וֹ וּבִזְר֣וֹעַ עֻזּ֑וֹ אִם\maqqaf אֶתֵּן֩ אֶת\maqqaf דְּגָנֵ֨ךְ ע֤וֹד מַֽאֲכָל֙ לְאֹ֣יְבַ֔יִךְ וְאִם\maqqaf יִשְׁתּ֤וּ בְנֵֽי\maqqaf נֵכָר֙ תִּֽירוֹשֵׁ֔ךְ אֲשֶׁ֥ר יָגַ֖עַתְּ בּֽוֹ׃}
{The \lord\space hath sworn by His right hand, And by the arm of His strength: Surely I will no more give thy corn To be food for thine enemies; And strangers shall not drink thy wine, For which thou hast laboured;}
\twoverse{\arabic{verse}}
{כִּ֤י מְאַסְפָיו֙ יֹאכְלֻ֔הוּ וְהִֽלְל֖וּ אֶת\maqqaf יְהֹוָ֑ה וּמְקַבְּצָ֥יו יִשְׁתֻּ֖הוּ בְּחַצְר֥וֹת קׇדְשִֽׁי׃ \setuma }
{But they that have garnered it shall eat it, And praise the \lord, And they that have gathered it shall drink it In the courts of My sanctuary.}
\twoverse{\arabic{verse}}
{עִבְר֤וּ עִבְרוּ֙ בַּשְּׁעָרִ֔ים פַּנּ֖וּ דֶּ֣רֶךְ הָעָ֑ם סֹ֣לּוּ סֹ֤לּוּ הַֽמְסִלָּה֙ סַקְּל֣וּ מֵאֶ֔בֶן הָרִ֥ימוּ נֵ֖ס עַל\maqqaf הָֽעַמִּֽים׃}
{Go through, go through the gates, Clear ye the way of the people; Cast up, cast up the highway, Gather out the stones; Lift up an ensign over the peoples.}
\twoverse{\arabic{verse}}
{הִנֵּ֣ה יְהֹוָ֗ה הִשְׁמִ֙יעַ֙ אֶל\maqqaf קְצֵ֣ה הָאָ֔רֶץ אִמְרוּ֙ לְבַת\maqqaf צִיּ֔וֹן הִנֵּ֥ה יִשְׁעֵ֖ךְ בָּ֑א הִנֵּ֤ה שְׂכָרוֹ֙ אִתּ֔וֹ וּפְעֻלָּת֖וֹ לְפָנָֽיו׃}
{Behold, the \lord\space hath proclaimed Unto the end of the earth: Say ye to the daughter of Zion: ‘Behold, thy salvation cometh; Behold, His reward is with Him, And His recompense before Him.’}
\twoverse{\arabic{verse}}
{וְקָרְא֥וּ לָהֶ֛ם עַם\maqqaf הַקֹּ֖דֶשׁ גְּאוּלֵ֣י יְהֹוָ֑ה וְלָךְ֙ יִקָּרֵ֣א דְרוּשָׁ֔ה עִ֖יר לֹ֥א נֶעֱזָֽבָה׃ \setuma }
{And they shall call them The holy people, The redeemed of the \lord; And thou shalt be called Sought out, A city not forsaken.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{מִי\maqqaf זֶ֣ה \legarmeh  בָּ֣א מֵאֱד֗וֹם חֲמ֤וּץ בְּגָדִים֙ מִבׇּצְרָ֔ה זֶ֚ה הָד֣וּר בִּלְבוּשׁ֔וֹ צֹעֶ֖ה בְּרֹ֣ב כֹּח֑וֹ אֲנִ֛י מְדַבֵּ֥ר בִּצְדָקָ֖ה רַ֥ב לְהוֹשִֽׁיעַ׃}
{‘Who is this that cometh from Edom, with crimsoned garments from Bozrah? This that is glorious in his apparel, stately in the greatness of his strength?’— ’I that speak in victory, mighty to save.’—}
\twoverse{\arabic{verse}}
{מַדּ֥וּעַ אָדֹ֖ם לִלְבוּשֶׁ֑ךָ וּבְגָדֶ֖יךָ כְּדֹרֵ֥ךְ בְּגַֽת׃}
{‘Wherefore is Thine apparel red, and Thy garments like his that treadeth in the winevat?’—}
\twoverse{\arabic{verse}}
{פּוּרָ֣ה \legarmeh  דָּרַ֣כְתִּי לְבַדִּ֗י וּמֵֽעַמִּים֙ אֵֽין\maqqaf אִ֣ישׁ אִתִּ֔י וְאֶדְרְכֵ֣ם בְּאַפִּ֔י וְאֶרְמְסֵ֖ם בַּחֲמָתִ֑י וְיֵ֤ז נִצְחָם֙ עַל\maqqaf בְּגָדַ֔י וְכׇל\maqqaf מַלְבּוּשַׁ֖י אֶגְאָֽלְתִּי׃}
{‘I have trodden the winepress alone, and of the peoples there was no man with Me; yea, I trod them in Mine anger, and trampled them in My fury; and their lifeblood is dashed against My garments, and I have stained all My raiment.}
\twoverse{\arabic{verse}}
{כִּ֛י י֥וֹם נָקָ֖ם בְּלִבִּ֑י וּשְׁנַ֥ת גְּאוּלַ֖י בָּֽאָה׃}
{For the day of vengeance that was in My heart, and My year of redemption are come.}
\twoverse{\arabic{verse}}
{וְאַבִּיט֙ וְאֵ֣ין עֹזֵ֔ר וְאֶשְׁתּוֹמֵ֖ם וְאֵ֣ין סוֹמֵ֑ךְ וַתּ֤וֹשַֽׁע לִי֙ זְרֹעִ֔י וַחֲמָתִ֖י הִ֥יא סְמָכָֽתְנִי׃}
{And I looked, and there was none to help, and I beheld in astonishment, and there was none to uphold; therefore Mine own arm brought salvation unto Me, And My fury, it upheld Me.}
\twoverse{\arabic{verse}}
{וְאָב֤וּס עַמִּים֙ בְּאַפִּ֔י וַאֲשַׁכְּרֵ֖ם בַּחֲמָתִ֑י וְאוֹרִ֥יד לָאָ֖רֶץ נִצְחָֽם׃ \setuma }
{And I trod down the peoples in Mine anger, and made them drunk with My fury, and I poured out their lifeblood on the earth.’}
\twoverse{\arabic{verse}}
{חַֽסְדֵ֨י יְהֹוָ֤ה \pasek  אַזְכִּיר֙ תְּהִלֹּ֣ת יְהֹוָ֔ה כְּעַ֕ל כֹּ֥ל אֲשֶׁר\maqqaf גְּמָלָ֖נוּ יְהֹוָ֑ה וְרַב\maqqaf טוּב֙ לְבֵ֣ית יִשְׂרָאֵ֔ל אֲשֶׁר\maqqaf גְּמָלָ֥ם כְּֽרַחֲמָ֖יו וּכְרֹ֥ב חֲסָדָֽיו׃}
{I will make mention of the mercies of the \lord, and the praises of the \lord, according to all that the \lord\space hath bestowed on us; and the great goodness toward the house of Israel, which He hath bestowed on them according to His compassions, and according to the multitude of His mercies.}
\twoverse{\arabic{verse}}
{וַיֹּ֙אמֶר֙ אַךְ\maqqaf עַמִּ֣י הֵ֔מָּה בָּנִ֖ים לֹ֣א יְשַׁקֵּ֑רוּ וַיְהִ֥י לָהֶ֖ם לְמוֹשִֽׁיעַ׃}
{For He said: ‘Surely, they are My people, children that will not deal falsely’; so He was their Saviour.}
\twoverse{\arabic{verse}}
{בְּֽכׇל\maqqaf צָרָתָ֣ם \legarmeh  \qk{ל֣וֹ}{לא} צָ֗ר וּמַלְאַ֤ךְ פָּנָיו֙ הוֹשִׁיעָ֔ם בְּאַהֲבָת֥וֹ וּבְחֶמְלָת֖וֹ ה֣וּא גְאָלָ֑ם וַֽיְנַטְּלֵ֥ם וַֽיְנַשְּׂאֵ֖ם כׇּל\maqqaf יְמֵ֥י עוֹלָֽם׃}
{In all their affliction He was afflicted, and the angel of His presence saved them; in His love and in His pity He redeemed them; And He bore them, and carried them all the days of old.}
\newperek
\newhaftarah{שובה}
\label{haft_52}
\haftnote{The Haftara is Hosea 14:2\verserangechar 14:10 \tworange Micah 7:18\verserangechar 7:20. This is the Haftara for the Shabbat between Rosh HaShana and Yom Kippur. Customs vary for what is read.  All begin with Hosea 14:2\verserangechar 10. Some add Micah 7:18–20 when this Shabbat coincides with Vayelech and Joel 2:15\verserangechar 27 when it coincides with Ha’azinu.  Others always read the passage from Micah, or the passage from Joel, or both.}
\setcounter{chap}{14}
\setcounter{verse}{2}
\twoverse{\Roman{chap}:\arabic{verse}}
{שׁ֚וּבָה יִשְׂרָאֵ֔ל עַ֖ד יְהֹוָ֣ה אֱלֹהֶ֑יךָ כִּ֥י כָשַׁ֖לְתָּ בַּעֲוֺנֶֽךָ׃}
{Return, O Israel, unto the \lord\space thy God; For thou hast stumbled in thine iniquity.}
\twoverse{\arabic{verse}}
{קְח֤וּ עִמָּכֶם֙ דְּבָרִ֔ים וְשׁ֖וּבוּ אֶל\maqqaf יְהֹוָ֑ה אִמְר֣וּ אֵלָ֗יו כׇּל\maqqaf תִּשָּׂ֤א עָוֺן֙ וְקַח\maqqaf ט֔וֹב וּֽנְשַׁלְּמָ֥ה פָרִ֖ים שְׂפָתֵֽינוּ׃}
{Take with you words, And return unto the \lord; Say unto Him: ‘Forgive all iniquity, And accept that which is good; So will we render for bullocks the offering of our lips.}
\twoverse{\arabic{verse}}
{אַשּׁ֣וּר \legarmeh  לֹ֣א יוֹשִׁיעֵ֗נוּ עַל\maqqaf סוּס֙ לֹ֣א נִרְכָּ֔ב וְלֹא\maqqaf נֹ֥אמַר ע֛וֹד אֱלֹהֵ֖ינוּ לְמַעֲשֵׂ֣ה יָדֵ֑ינוּ אֲשֶׁר\maqqaf בְּךָ֖ יְרֻחַ֥ם יָתֽוֹם׃}
{Asshur shall not save us; We will not ride upon horses; Neither will we call any more the work of our hands our gods; For in Thee the fatherless findeth mercy.’}
\twoverse{\arabic{verse}}
{אֶרְפָּא֙ מְשׁ֣וּבָתָ֔ם אֹהֲבֵ֖ם נְדָבָ֑ה כִּ֛י שָׁ֥ב אַפִּ֖י מִמֶּֽנּוּ׃}
{I will heal their backsliding, I will love them freely; For Mine anger is turned away from him.}
\twoverse{\arabic{verse}}
{אֶהְיֶ֤ה כַטַּל֙ לְיִשְׂרָאֵ֔ל יִפְרַ֖ח כַּשּׁוֹשַׁנָּ֑ה וְיַ֥ךְ שׇׁרָשָׁ֖יו כַּלְּבָנֽוֹן׃}
{I will be as the dew unto Israel; He shall blossom as the lily, And cast forth his roots as Lebanon.}
\twoverse{\arabic{verse}}
{יֵֽלְכוּ֙ יֹֽנְקוֹתָ֔יו וִיהִ֥י כַזַּ֖יִת הוֹד֑וֹ וְרֵ֥יחַֽ ל֖וֹ כַּלְּבָנֽוֹן׃}
{His branches shall spread, And his beauty shall be as the olive-tree, And his fragrance as Lebanon.}
\twoverse{\arabic{verse}}
{יָשֻׁ֙בוּ֙ יֹשְׁבֵ֣י בְצִלּ֔וֹ יְחַיּ֥וּ דָגָ֖ן וְיִפְרְח֣וּ כַגָּ֑פֶן זִכְר֖וֹ כְּיֵ֥ין לְבָנֽוֹן׃}
{They that dwell under his shadow shall again Make corn to grow, And shall blossom as the vine; The scent thereof shall be as the wine of Lebanon.}
\twoverse{\arabic{verse}}
{אֶפְרַ֕יִם מַה\maqqaf לִּ֥י ע֖וֹד לָעֲצַבִּ֑ים אֲנִ֧י עָנִ֣יתִי וַאֲשׁוּרֶ֗נּוּ אֲנִי֙ כִּבְר֣וֹשׁ רַעֲנָ֔ן מִמֶּ֖נִּי פֶּרְיְךָ֥ נִמְצָֽא׃}
{Ephraim [shall say]: ‘What have I to do any more with idols?’ As for Me, I respond and look on him; I am like a leafy cypress-tree; From Me is thy fruit found.}
\twoverse{\arabic{verse}}
{מִ֤י חָכָם֙ וְיָ֣בֵֽן אֵ֔לֶּה נָב֖וֹן וְיֵדָעֵ֑ם כִּֽי\maqqaf יְשָׁרִ֞ים דַּרְכֵ֣י יְהֹוָ֗ה וְצַדִּקִים֙ יֵ֣לְכוּ בָ֔ם וּפֹשְׁעִ֖ים יִכָּ֥שְׁלוּ בָֽם׃}
{Whoso is wise, let him understand these things, Whoso is prudent, let him know them. For the ways of the \lord\space are right, And the just do walk in them; But transgressors do stumble therein.}
\setcounter{chap}{7}
\setcounter{verse}{18}
\twoverse{\Roman{chap}:\arabic{verse}}
{מִי\maqqaf אֵ֣ל כָּמ֗וֹךָ נֹשֵׂ֤א עָוֺן֙ וְעֹבֵ֣ר עַל\maqqaf פֶּ֔שַׁע לִשְׁאֵרִ֖ית נַחֲלָת֑וֹ לֹֽא\maqqaf הֶחֱזִ֤יק לָעַד֙ אַפּ֔וֹ כִּֽי\maqqaf חָפֵ֥ץ חֶ֖סֶד הֽוּא׃}
{Who is a God like unto Thee, that pardoneth the iniquity, And passeth by the transgression of the remnant of His heritage? He retaineth not His anger for ever, Because He delighteth in mercy.}
\twoverse{\arabic{verse}}
{יָשׁ֣וּב יְרַחֲמֵ֔נוּ יִכְבֹּ֖שׁ עֲוֺנֹתֵ֑ינוּ וְתַשְׁלִ֛יךְ בִּמְצֻל֥וֹת יָ֖ם כׇּל\maqqaf חַטֹּאותָֽם׃}
{He will again have compassion upon us; He will subdue our iniquities; And Thou wilt cast all their sins into the depths of the sea.}
\twoverse{\arabic{verse}}
{תִּתֵּ֤ן אֱמֶת֙ לְיַֽעֲקֹ֔ב חֶ֖סֶד לְאַבְרָהָ֑ם אֲשֶׁר\maqqaf נִשְׁבַּ֥עְתָּ לַאֲבֹתֵ֖ינוּ מִ֥ימֵי קֶֽדֶם׃}
{Thou wilt show faithfulness to Jacob, mercy to Abraham, As Thou hast sworn unto our fathers from the days of old.}
\setcounter{chap}{2}
\setcounter{verse}{15}
\twoverse{\Roman{chap}:\arabic{verse}}
{תִּקְע֥וּ שׁוֹפָ֖ר בְּצִיּ֑וֹן קַדְּשׁוּ\maqqaf צ֖וֹם קִרְא֥וּ עֲצָרָֽה׃}
{Blow the horn in Zion, Sanctify a fast, call a solemn assembly;}
\twoverse{\arabic{verse}}
{אִסְפוּ\maqqaf עָ֞ם קַדְּשׁ֤וּ קָהָל֙ קִבְצ֣וּ זְקֵנִ֔ים אִסְפוּ֙ עֽוֹלָלִ֔ים וְיֹנְקֵ֖י שָׁדָ֑יִם יֵצֵ֤א חָתָן֙ מֵֽחֶדְר֔וֹ וְכַלָּ֖ה מֵחֻפָּתָֽהּ׃}
{Gather the people, Sanctify the congregation, Assemble the elders, Gather the children, And those that suck the breasts; Let the bridegroom go forth from his chamber, And the bride out of her pavilion.}
\twoverse{\arabic{verse}}
{בֵּ֤ין הָאוּלָם֙ וְלַמִּזְבֵּ֔חַ יִבְכּוּ֙ הַכֹּ֣הֲנִ֔ים מְשָׁרְתֵ֖י יְהֹוָ֑ה וְֽיֹאמְר֞וּ ח֧וּסָה יְהֹוָ֣ה עַל\maqqaf עַמֶּ֗ךָ וְאַל\maqqaf תִּתֵּ֨ן נַחֲלָתְךָ֤ לְחֶרְפָּה֙ לִמְשׇׁל\maqqaf בָּ֣ם גּוֹיִ֔ם לָ֚מָּה יֹאמְר֣וּ בָעַמִּ֔ים אַיֵּ֖ה אֱלֹהֵיהֶֽם׃}
{Let the priests, the ministers of the \lord, Weep between the porch and the altar, And let them say: ‘Spare thy people, O \lord, And give not Thy heritage to reproach, That the nations should make them a byword: Wherefore should they say among the peoples: Where is their God?’}
\twoverse{\arabic{verse}}
{וַיְקַנֵּ֥א יְהֹוָ֖ה לְאַרְצ֑וֹ וַיַּחְמֹ֖ל עַל\maqqaf עַמּֽוֹ׃}
{Then was the \lord\space jealous for His land, And had pity on His people.}
\twoverse{\arabic{verse}}
{וַיַּ֨עַן יְהֹוָ֜ה וַיֹּ֣אמֶר לְעַמּ֗וֹ הִנְנִ֨י שֹׁלֵ֤חַ לָכֶם֙ אֶת\maqqaf הַדָּגָן֙ וְהַתִּיר֣וֹשׁ וְהַיִּצְהָ֔ר וּשְׂבַעְתֶּ֖ם אֹת֑וֹ וְלֹא\maqqaf אֶתֵּ֨ן אֶתְכֶ֥ם ע֛וֹד חֶרְפָּ֖ה בַּגּוֹיִֽם׃}
{And the \lord\space answered and said unto His people: ‘Behold, I will send you corn, and wine, and oil, And ye shall be satisfied therewith; And I will no more make you a reproach among the nations;}
\twoverse{\arabic{verse}}
{וְֽאֶת\maqqaf הַצְּפוֹנִ֞י אַרְחִ֣יק מֵעֲלֵיכֶ֗ם וְהִדַּחְתִּיו֮ אֶל\maqqaf אֶ֣רֶץ צִיָּ֣ה וּשְׁמָמָה֒ אֶת\maqqaf פָּנָ֗יו אֶל\maqqaf הַיָּם֙ הַקַּדְמֹנִ֔י וְסֹפ֖וֹ אֶל\maqqaf הַיָּ֣ם הָאַחֲר֑וֹן וְעָלָ֣ה בָאְשׁ֗וֹ וְתַ֙עַל֙ צַחֲנָת֔וֹ כִּ֥י הִגְדִּ֖יל לַעֲשֽׂוֹת׃}
{But I will remove far off from you the northern one, And will drive him into a land barren and desolate, With his face toward the eastern sea, And his hinder part toward the western sea; that his foulness may come up, and his ill savour may come up, because he hath done great things.’}
\twoverse{\arabic{verse}}
{אַל\maqqaf תִּֽירְאִ֖י אֲדָמָ֑ה גִּ֣ילִי וּשְׂמָ֔חִי כִּֽי\maqqaf הִגְדִּ֥יל יְהֹוָ֖ה לַעֲשֽׂוֹת׃}
{Fear not, O land, be glad and rejoice; for the \lord\space hath done great things.}
\twoverse{\arabic{verse}}
{אַל\maqqaf תִּֽירְאוּ֙ בַּהֲמ֣וֹת שָׂדַ֔י כִּ֥י דָשְׁא֖וּ נְא֣וֹת מִדְבָּ֑ר כִּי\maqqaf עֵץ֙ נָשָׂ֣א פִרְי֔וֹ תְּאֵנָ֥ה וָגֶ֖פֶן נָתְנ֥וּ חֵילָֽם׃}
{Be not afraid, ye beasts of the field; for the pastures of the wilderness do spring, for the tree beareth its fruit, the fig-tree and the vine do yield their strength.}
\twoverse{\arabic{verse}}
{וּבְנֵ֣י צִיּ֗וֹן גִּ֤ילוּ וְשִׂמְחוּ֙ בַּיהֹוָ֣ה אֱלֹהֵיכֶ֔ם כִּֽי\maqqaf נָתַ֥ן לָכֶ֛ם אֶת\maqqaf הַמּוֹרֶ֖ה לִצְדָקָ֑ה וַיּ֣וֹרֶד לָכֶ֗ם גֶּ֛שֶׁם מוֹרֶ֥ה וּמַלְק֖וֹשׁ בָּרִאשֽׁוֹן׃}
{Be glad then, ye children of Zion, and rejoice in the \lord\space your God; For He giveth you the former rain in just measure, And He causeth to come down for you the rain, the former rain and the latter rain, at the first.}
\twoverse{\arabic{verse}}
{וּמָלְא֥וּ הַגֳּרָנ֖וֹת בָּ֑ר וְהֵשִׁ֥יקוּ הַיְקָבִ֖ים תִּיר֥וֹשׁ וְיִצְהָֽר׃}
{And the floors shall be full of corn, and the vats shall overflow with wine and oil.}
\twoverse{\arabic{verse}}
{וְשִׁלַּמְתִּ֤י לָכֶם֙ אֶת\maqqaf הַשָּׁנִ֔ים אֲשֶׁר֙ אָכַ֣ל הָאַרְבֶּ֔ה הַיֶּ֖לֶק וְהֶחָסִ֣יל וְהַגָּזָ֑ם חֵילִי֙ הַגָּד֔וֹל אֲשֶׁ֥ר שִׁלַּ֖חְתִּי בָּכֶֽם׃}
{And I will restore to you the years that the locust hath eaten, the canker-worm, and the caterpiller, and the palmer-worm, My great army which I sent among you.}
\twoverse{\arabic{verse}}
{וַאֲכַלְתֶּ֤ם אָכוֹל֙ וְשָׂב֔וֹעַ וְהִלַּלְתֶּ֗ם אֶת\maqqaf שֵׁ֤ם יְהֹוָה֙ אֱלֹ֣הֵיכֶ֔ם אֲשֶׁר\maqqaf עָשָׂ֥ה עִמָּכֶ֖ם לְהַפְלִ֑יא וְלֹא\maqqaf יֵבֹ֥שׁוּ עַמִּ֖י לְעוֹלָֽם׃}
{And ye shall eat in plenty and be satisfied, and shall praise the name of the \lord\space your God, that hath dealt wondrously with you; and My people shall never be ashamed.}
\twoverse{\arabic{verse}}
{וִידַעְתֶּ֗ם כִּ֣י בְקֶ֤רֶב יִשְׂרָאֵל֙ אָ֔נִי וַאֲנִ֛י יְהֹוָ֥ה אֱלֹהֵיכֶ֖ם וְאֵ֣ין ע֑וֹד וְלֹֽא\maqqaf יֵבֹ֥שׁוּ עַמִּ֖י לְעוֹלָֽם׃ \petucha }
{And ye shall know that I am in the midst of Israel, and that I am the \lord\space your God, and there is none else; and My people shall never be ashamed.}
\newhaftarah{האזינו}
\label{haft_53}
\haftnote{The Haftara is II Samuel 22:1\verserangechar 22:51. This Haftara is only read on the Shabbat between Yom Kippur and Sukkot, if there is one.  If it is the Shabbat between Rosh HaShana and Yom Kippur, read the Haftara for Shabbat Shuva on page \pageref{haft_52}.}
\setcounter{chap}{22}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}
{וַיְדַבֵּ֤ר דָּוִד֙ לַיהֹוָ֔ה אֶת\maqqaf דִּבְרֵ֖י הַשִּׁירָ֣ה הַזֹּ֑את בְּיוֹם֩ הִצִּ֨יל יְהֹוָ֥ה אֹת֛וֹ מִכַּ֥ף כׇּל\maqqaf אֹיְבָ֖יו וּמִכַּ֥ף שָׁאֽוּל׃        }
{And David spoke unto the \lord\space the words of this song in the day that the \lord\space delivered him out of the hand of all his enemies, and out of the hand of Saul;}
\twoverse{\arabic{verse}}
{וַיֹּאמַ֑ר יְהֹוָ֛ה סַֽלְעִ֥י וּמְצֻדָתִ֖י וּמְפַלְטִי\maqqaf לִֽי׃}
{and he said: The \lord\space is my rock, and my fortress, and my deliverer;}
\twoverse{\arabic{verse}}
{אֱלֹהֵ֥י צוּרִ֖י אֶחֱסֶה\maqqaf בּ֑וֹ מָגִנִּ֞י וְקֶ֣רֶן יִשְׁעִ֗י מִשְׂגַּבִּי֙ וּמְנוּסִ֔י מֹשִׁעִ֕י מֵחָמָ֖ס תֹּשִׁעֵֽנִי׃}
{The God who is my rock, in Him I take refuge; My shield, and my horn of salvation, my high tower, and my refuge; My saviour, Thou savest me from violence.}
\twoverse{\arabic{verse}}
{מְהֻלָּ֖ל אֶקְרָ֣א יְהֹוָ֑ה וּמֵאֹיְבַ֖י אִוָּשֵֽׁעַ׃}
{Praised, I cry, is the \lord, and I am saved from mine enemies.}
\twoverse{\arabic{verse}}
{כִּ֥י אֲפָפֻ֖נִי מִשְׁבְּרֵי\maqqaf מָ֑וֶת נַחֲלֵ֥י בְלִיַּ֖עַל יְבַעֲתֻֽנִי׃}
{For the waves of Death compassed me. The floods of Belial assailed me.}
\twoverse{\arabic{verse}}
{חֶבְלֵ֥י שְׁא֖וֹל סַבֻּ֑נִי קִדְּמֻ֖נִי מֹ֥קְשֵׁי מָֽוֶת׃}
{The cords of Sheol surrounded me; the snares of Death confronted me.}
\twoverse{\arabic{verse}}
{בַּצַּר\maqqaf לִי֙ אֶקְרָ֣א יְהֹוָ֔ה וְאֶל\maqqaf אֱלֹהַ֖י אֶקְרָ֑א וַיִּשְׁמַ֤ע מֵהֵֽיכָלוֹ֙ קוֹלִ֔י וְשַׁוְעָתִ֖י בְּאׇזְנָֽיו׃}
{In my distress I called upon the \lord, yea, I called unto my God; and out of His temple He heard my voice, and my cry did enter into His ears.}
\twoverse{\arabic{verse}}
{\qk{וַיִּתְגָּעַ֤שׁ}{ותגעש} וַתִּרְעַשׁ֙ הָאָ֔רֶץ מוֹסְד֥וֹת הַשָּׁמַ֖יִם יִרְגָּ֑זוּ וַיִּֽתְגָּעֲשׁ֖וּ כִּי\maqqaf חָ֥רָה לֽוֹ׃}
{Then the earth did shake and quake, the foundations of heaven did tremble; they were shaken, because He was wroth.}
\twoverse{\arabic{verse}}
{עָלָ֤ה עָשָׁן֙ בְּאַפּ֔וֹ וְאֵ֥שׁ מִפִּ֖יו תֹּאכֵ֑ל גֶּחָלִ֖ים בָּעֲר֥וּ מִמֶּֽנּוּ׃}
{Smoke arose up in His nostrils, and fire out of His mouth did devour; coals flamed forth from Him.}
\twoverse{\arabic{verse}}
{וַיֵּ֥ט שָׁמַ֖יִם וַיֵּרַ֑ד וַעֲרָפֶ֖ל תַּ֥חַת רַגְלָֽיו׃}
{He bowed the heavens also, and came down; and thick darkness was under His feet.}
\twoverse{\arabic{verse}}
{וַיִּרְכַּ֥ב עַל\maqqaf כְּר֖וּב וַיָּעֹ֑ף וַיֵּרָ֖א עַל\maqqaf כַּנְפֵי\maqqaf רֽוּחַ׃}
{And He rode upon a cherub, and did fly; Yea, He was seen upon the wings of the wind.}
\twoverse{\arabic{verse}}
{וַיָּ֥שֶׁת חֹ֛שֶׁךְ סְבִיבֹתָ֖יו סֻכּ֑וֹת חַֽשְׁרַת\maqqaf מַ֖יִם עָבֵ֥י שְׁחָקִֽים׃}
{And He made darkness pavilions round about Him, gathering of waters, thick clouds of the skies.}
\twoverse{\arabic{verse}}
{מִנֹּ֖גַהּ נֶגְדּ֑וֹ בָּעֲר֖וּ גַּחֲלֵי\maqqaf אֵֽשׁ׃        }
{At the brightness before Him coals of fire flamed forth.}
\twoverse{\arabic{verse}}
{יַרְעֵ֥ם מִן\maqqaf שָׁמַ֖יִם יְהֹוָ֑ה וְעֶלְי֖וֹן יִתֵּ֥ן קוֹלֽוֹ׃        }
{The \lord\space thundered from heaven, and the Most High gave forth His voice.}
\twoverse{\arabic{verse}}
{וַיִּשְׁלַ֥ח חִצִּ֖ים וַיְפִיצֵ֑ם בָּרָ֖ק \qk{וַיָּהֹֽם}{ויהמם}׃}
{And He sent out arrows, and scattered them; lightning, and discomfited them.}
\twoverse{\arabic{verse}}
{וַיֵּֽרָאוּ֙ אֲפִ֣קֵי יָ֔ם יִגָּל֖וּ מֹסְד֣וֹת תֵּבֵ֑ל בְּגַעֲרַ֣ת יְהֹוָ֔ה מִנִּשְׁמַ֖ת ר֥וּחַ אַפּֽוֹ׃}
{And the channels of the sea appeared, the foundations of the world were laid bare by the rebuke of the \lord, at the blast of the breath of His nostrils.}
\twoverse{\arabic{verse}}
{יִשְׁלַ֥ח מִמָּר֖וֹם יִקָּחֵ֑נִי יַֽמְשֵׁ֖נִי מִמַּ֥יִם רַבִּֽים׃}
{He sent from on high, He took me; He drew me out of many waters;}
\twoverse{\arabic{verse}}
{יַצִּילֵ֕נִי מֵאֹיְבִ֖י עָ֑ז מִשֹּׂ֣נְאַ֔י כִּ֥י אָמְצ֖וּ מִמֶּֽנִּי׃}
{He delivered me from mine enemy most strong, from them that hated me, for they were too mighty for me.}
\twoverse{\arabic{verse}}
{יְקַדְּמֻ֖נִי בְּי֣וֹם אֵידִ֑י וַיְהִ֧י יְהֹוָ֛ה מִשְׁעָ֖ן לִֽי׃        }
{They confronted me in the day of my calamity; but the \lord\space was a stay unto me.}
\twoverse{\arabic{verse}}
{וַיֹּצֵ֥א לַמֶּרְחָ֖ב אֹתִ֑י יְחַלְּצֵ֖נִי כִּי\maqqaf חָ֥פֵֽץ בִּֽי׃        }
{He brought me forth also into a large place; He delivered me, because He delighted in me.}
\twoverse{\arabic{verse}}
{יִגְמְלֵ֥נִי יְהֹוָ֖ה כְּצִדְקָתִ֑י כְּבֹ֥ר יָדַ֖י יָשִׁ֥יב לִֽי׃        }
{The \lord\space rewarded me according to my righteousness; According to the cleanness of my hands hath He recompensed me.}
\twoverse{\arabic{verse}}
{כִּ֥י שָׁמַ֖רְתִּי דַּרְכֵ֣י יְהֹוָ֑ה וְלֹ֥א רָשַׁ֖עְתִּי מֵאֱלֹהָֽי׃        }
{For I have kept the ways of the \lord, And have not wickedly departed from my God.}
\twoverse{\arabic{verse}}
{כִּ֥י כׇל\maqqaf מִשְׁפָּטָ֖ו לְנֶגְדִּ֑י וְחֻקֹּתָ֖יו לֹא\maqqaf אָס֥וּר מִמֶּֽנָּה׃}
{For all His ordinances were before me; And as for His statutes, I did not depart from them.}
\twoverse{\arabic{verse}}
{וָאֶהְיֶ֥ה תָמִ֖ים ל֑וֹ וָאֶשְׁתַּמְּרָ֖ה מֵעֲוֺנִֽי׃}
{And I was single-hearted toward Him, And I kept myself from mine iniquity.}
\twoverse{\arabic{verse}}
{וַיָּ֧שֶׁב יְהֹוָ֛ה לִ֖י כְּצִדְקָתִ֑י כְּבֹרִ֖י לְנֶ֥גֶד עֵינָֽיו׃}
{Therefore hath the \lord\space recompensed me according to my righteousness, According to my cleanness in His eyes.}
\twoverse{\arabic{verse}}
{עִם\maqqaf חָסִ֖יד תִּתְחַסָּ֑ד עִם\maqqaf גִּבּ֥וֹר תָּמִ֖ים תִּתַּמָּֽם׃}
{With the merciful Thou dost show Thyself merciful, With the upright man Thou dost show Thyself upright,}
\twoverse{\arabic{verse}}
{עִם\maqqaf נָבָ֖ר תִּתָּבָ֑ר וְעִם\maqqaf עִקֵּ֖שׁ תִּתַּפָּֽל׃}
{With the pure Thou dost show Thyself pure; And with the crooked Thou dost show Thyself subtle.}
\twoverse{\arabic{verse}}
{וְאֶת\maqqaf עַ֥ם עָנִ֖י תּוֹשִׁ֑יעַ וְעֵינֶ֥יךָ עַל\maqqaf רָמִ֖ים תַּשְׁפִּֽיל׃}
{And the afflicted people Thou dost save; But Thine eyes are upon the haughty, that Thou mayest humble them.}
\twoverse{\arabic{verse}}
{כִּֽי\maqqaf אַתָּ֥ה נֵירִ֖י יְהֹוָ֑ה וַיהֹוָ֖ה יַגִּ֥יהַּ חׇשְׁכִּֽי׃}
{For Thou art my lamp, O \lord; And the \lord\space doth lighten my darkness.}
\twoverse{\arabic{verse}}
{כִּ֥י בְכָ֖ה אָר֣וּץ גְּד֑וּד בֵּאלֹהַ֖י אֲדַלֶּג\maqqaf שֽׁוּר׃}
{For by Thee I run upon a troop; By my God do I scale a wall.}
\twoverse{\arabic{verse}}
{הָאֵ֖ל תָּמִ֣ים דַּרְכּ֑וֹ אִמְרַ֤ת יְהֹוָה֙ צְרוּפָ֔ה מָגֵ֣ן ה֔וּא לְכֹ֖ל הַֽחֹסִ֥ים בּֽוֹ׃}
{As for God, His way is perfect; The word of the \lord\space is tried; He is a shield unto all them that take refuge in Him.}
\twoverse{\arabic{verse}}
{כִּ֥י מִי\maqqaf אֵ֖ל מִבַּלְעֲדֵ֣י יְהֹוָ֑ה וּמִ֥י צ֖וּר מִֽבַּלְעֲדֵ֥י אֱלֹהֵֽינוּ׃}
{For who is God, save the \lord? And who is a Rock, save our God?}
\twoverse{\arabic{verse}}
{הָאֵ֥ל מָֽעוּזִּ֖י חָ֑יִל וַיַּתֵּ֥ר תָּמִ֖ים \qk{דַּרְכִּֽי}{דרכו}׃}
{The God who is my strong fortress, And who letteth my way go forth straight;}
\twoverse{\arabic{verse}}
{מְשַׁוֶּ֥ה \qk{רַגְלַ֖י}{רגליו} כָּאַיָּל֑וֹת וְעַ֥ל בָּמֹתַ֖י יַעֲמִדֵֽנִי׃}
{Who maketh my feet like hinds’ , And setteth me upon my high places;}
\twoverse{\arabic{verse}}
{מְלַמֵּ֥ד יָדַ֖י לַמִּלְחָמָ֑ה וְנִחַ֥ת קֶשֶׁת\maqqaf נְחוּשָׁ֖ה זְרֹעֹתָֽי׃}
{Who traineth my hands for war, So that mine arms do bend a bow of brass.}
\twoverse{\arabic{verse}}
{וַתִּתֶּן\maqqaf לִ֖י מָגֵ֣ן יִשְׁעֶ֑ךָ וַעֲנֹתְךָ֖ תַּרְבֵּֽנִי׃        }
{Thou hast also given me Thy shield of salvation; And Thy condescension hath made me great.}
\twoverse{\arabic{verse}}
{תַּרְחִ֥יב צַעֲדִ֖י תַּחְתֵּ֑נִי וְלֹ֥א מָעֲד֖וּ קַרְסֻלָּֽי׃}
{Thou hast enlarged my steps under me, And my feet have not slipped.}
\twoverse{\arabic{verse}}
{אֶרְדְּפָ֥ה אֹיְבַ֖י וָאַשְׁמִידֵ֑ם וְלֹ֥א אָשׁ֖וּב עַד\maqqaf כַּלּוֹתָֽם׃}
{I have pursued mine enemies, and destroyed them; Neither did I turn back till they were consumed.}
\twoverse{\arabic{verse}}
{וָאֲכַלֵּ֥ם וָאֶמְחָצֵ֖ם וְלֹ֣א יְקוּמ֑וּן וַֽיִּפְּל֖וּ תַּ֥חַת רַגְלָֽי׃}
{And I have consumed them, and smitten them through, that they cannot arise; Yea, they are fallen under my feet.}
\twoverse{\arabic{verse}}
{וַתַּזְרֵ֥נִי חַ֖יִל לַמִּלְחָמָ֑ה תַּכְרִ֥יעַ קָמַ֖י תַּחְתֵּֽנִי׃}
{For Thou hast girded me with strength unto the battle; Thou hast subdued under me those that rose up against me.}
\twoverse{\arabic{verse}}
{וְאֹ֣יְבַ֔י תַּ֥תָּה לִּ֖י עֹ֑רֶף מְשַׂנְאַ֖י וָאַצְמִיתֵֽם׃        }
{Thou hast also made mine enemies turn their backs unto me; Yea, them that hate me, that I might cut them off.}
\twoverse{\arabic{verse}}
{יִשְׁע֖וּ וְאֵ֣ין מֹשִׁ֑יעַ אֶל\maqqaf יְהֹוָ֖ה וְלֹ֥א עָנָֽם׃}
{They looked, but there was none to save; Even unto the \lord, but He answered them not.}
\twoverse{\arabic{verse}}
{וְאֶשְׁחָקֵ֖ם כַּעֲפַר\maqqaf אָ֑רֶץ כְּטִיט\maqqaf חוּצ֥וֹת אֲדִקֵּ֖ם אֶרְקָעֵֽם׃}
{Then did I beat them small as the dust of the earth, I did stamp them as the mire of the streets, and did tread them down.}
\twoverse{\arabic{verse}}
{וַֽתְּפַלְּטֵ֔נִי מֵרִיבֵ֖י עַמִּ֑י תִּשְׁמְרֵ֙נִי֙ לְרֹ֣אשׁ גּוֹיִ֔ם עַ֥ם לֹא\maqqaf יָדַ֖עְתִּי יַעַבְדֻֽנִי׃}
{Thou also hast delivered me from the contentions of my people; Thou hast kept me to be the head of the nations; A people whom I have not known serve me.}
\twoverse{\arabic{verse}}
{בְּנֵ֥י נֵכָ֖ר יִתְכַּחֲשׁוּ\maqqaf לִ֑י לִשְׁמ֥וֹעַ אֹ֖זֶן יִשָּׁ֥מְעוּ לִֽי׃        }
{The sons of the stranger dwindle away before me; As soon as they hear of me, they obey me.}
\twoverse{\arabic{verse}}
{בְּנֵ֥י נֵכָ֖ר יִבֹּ֑לוּ וְיַחְגְּר֖וּ מִמִּסְגְּרוֹתָֽם׃}
{The sons of the stranger fade away, And come halting out of their close places.}
\twoverse{\arabic{verse}}
{חַי\maqqaf יְהֹוָ֖ה וּבָר֣וּךְ צוּרִ֑י וְיָרֻ֕ם אֱלֹהֵ֖י צ֥וּר יִשְׁעִֽי׃}
{The \lord\space liveth, and blessed be my Rock; And exalted be the God, my Rock of salvation;}
\twoverse{\arabic{verse}}
{הָאֵ֕ל הַנֹּתֵ֥ן נְקָמֹ֖ת לִ֑י וּמֹרִ֥יד עַמִּ֖ים תַּחְתֵּֽנִי׃}
{Even the God that executeth vengeance for me, And bringeth down peoples under me,}
\twoverse{\arabic{verse}}
{וּמוֹצִיאִ֖י מֵאֹֽיְבָ֑י וּמִקָּמַי֙ תְּר֣וֹמְמֵ֔נִי מֵאִ֥ישׁ חֲמָסִ֖ים תַּצִּילֵֽנִי׃}
{And that bringeth me forth from mine enemies; Yea, Thou liftest me up above them that rise up against me; Thou deliverest me from the violent man.}
\twoverse{\arabic{verse}}
{עַל\maqqaf כֵּ֛ן אֽוֹדְךָ֥ יְהֹוָ֖ה בַּגּוֹיִ֑ם וּלְשִׁמְךָ֖ אֲזַמֵּֽר׃}
{Therefore I will give thanks unto Thee, O \lord, among the nations, And will sing praises unto Thy name.}
\twoverse{\arabic{verse}}
{\qk{מִגְדּ֖וֹל}{מגדיל} יְשׁוּע֣וֹת מַלְכּ֑וֹ וְעֹֽשֶׂה\maqqaf חֶ֧סֶד לִמְשִׁיח֛וֹ לְדָוִ֥ד וּלְזַרְע֖וֹ עַד\maqqaf עוֹלָֽם׃ \petucha }
{A tower of salvation is He to His king; And showeth mercy to His anointed, To David and to his seed, for evermore.}
\newhaftarah{וזאת הברכה}
\label{haft_54}
\haftnote{The Haftara is Joshua 1:1\verserangechar 1:18. }
\setcounter{chap}{1}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}
{וַֽיְהִ֗י אַחֲרֵ֛י מ֥וֹת מֹשֶׁ֖ה עֶ֣בֶד יְהֹוָ֑ה וַיֹּ֤אמֶר יְהֹוָה֙ אֶל\maqqaf יְהוֹשֻׁ֣עַ בִּן\maqqaf נ֔וּן מְשָׁרֵ֥ת מֹשֶׁ֖ה לֵאמֹֽר׃}
{Now it came to pass after the death of Moses the servant of the \lord, that the \lord\space spoke unto Joshua the son of Nun, Moses’ minister, saying:}
\twoverse{\arabic{verse}}
{מֹשֶׁ֥ה עַבְדִּ֖י מֵ֑ת וְעַתָּה֩ ק֨וּם עֲבֹ֜ר אֶת\maqqaf הַיַּרְדֵּ֣ן הַזֶּ֗ה אַתָּה֙ וְכׇל\maqqaf הָעָ֣ם הַזֶּ֔ה אֶל\maqqaf הָאָ֕רֶץ אֲשֶׁ֧ר אָנֹכִ֛י נֹתֵ֥ן לָהֶ֖ם לִבְנֵ֥י יִשְׂרָאֵֽל׃}
{‘Moses My servant is dead; now therefore arise, go over this Jordan, thou, and all this people, unto the land which I do give to them, even to the children of Israel.}
\twoverse{\arabic{verse}}
{כׇּל\maqqaf מָק֗וֹם אֲשֶׁ֨ר תִּדְרֹ֧ךְ כַּֽף\maqqaf רַגְלְכֶ֛ם בּ֖וֹ לָכֶ֣ם נְתַתִּ֑יו כַּאֲשֶׁ֥ר דִּבַּ֖רְתִּי אֶל\maqqaf מֹשֶֽׁה׃}
{Every place that the sole of your foot shall tread upon, to you have I given it, as I spoke unto Moses.}
\twoverse{\arabic{verse}}
{מֵהַמִּדְבָּר֩ וְהַלְּבָנ֨וֹן הַזֶּ֜ה וְֽעַד\maqqaf הַנָּהָ֧ר הַגָּד֣וֹל נְהַר\maqqaf פְּרָ֗ת כֹּ֚ל אֶ֣רֶץ הַֽחִתִּ֔ים וְעַד\maqqaf הַיָּ֥ם הַגָּד֖וֹל מְב֣וֹא הַשָּׁ֑מֶשׁ יִֽהְיֶ֖ה גְּבֽוּלְכֶֽם׃}
{From the wilderness, and this Lebanon, even unto the great river, the river Euphrates, all the land of the Hittites, and unto the Great Sea toward the going down of the sun, shall be your border.}
\twoverse{\arabic{verse}}
{לֹא\maqqaf יִתְיַצֵּ֥ב אִישׁ֙ לְפָנֶ֔יךָ כֹּ֖ל יְמֵ֣י חַיֶּ֑יךָ כַּאֲשֶׁ֨ר הָיִ֤יתִי עִם\maqqaf מֹשֶׁה֙ אֶהְיֶ֣ה עִמָּ֔ךְ לֹ֥א אַרְפְּךָ֖ וְלֹ֥א אֶעֶזְבֶֽךָּ׃}
{There shall not any man be able to stand before thee all the days of thy life; as I was with Moses, so I will be with thee; I will not fail thee, nor forsake thee.}
\twoverse{\arabic{verse}}
{חֲזַ֖ק וֶאֱמָ֑ץ כִּ֣י אַתָּ֗ה תַּנְחִיל֙ אֶת\maqqaf הָעָ֣ם הַזֶּ֔ה אֶת\maqqaf הָאָ֕רֶץ אֲשֶׁר\maqqaf נִשְׁבַּ֥עְתִּי לַאֲבוֹתָ֖ם לָתֵ֥ת לָהֶֽם׃}
{Be strong and of good courage; for thou shalt cause this people to inherit the land which I swore unto their fathers to give them.}
\twoverse{\arabic{verse}}
{רַק֩ חֲזַ֨ק וֶאֱמַ֜ץ מְאֹ֗ד לִשְׁמֹ֤ר לַֽעֲשׂוֹת֙ כְּכׇל\maqqaf הַתּוֹרָ֔ה אֲשֶׁ֤ר צִוְּךָ֙ מֹשֶׁ֣ה עַבְדִּ֔י אַל\maqqaf תָּס֥וּר מִמֶּ֖נּוּ יָמִ֣ין וּשְׂמֹ֑אול לְמַ֣עַן תַּשְׂכִּ֔יל בְּכֹ֖ל אֲשֶׁ֥ר תֵּלֵֽךְ׃}
{Only be strong and very courageous, to observe to do according to all the law, which Moses My servant commanded thee; turn not from it to the right hand or to the left, that thou mayest have good success whithersoever thou goest.}
\twoverse{\arabic{verse}}
{לֹֽא\maqqaf יָמ֡וּשׁ סֵ֩פֶר֩ הַתּוֹרָ֨ה הַזֶּ֜ה מִפִּ֗יךָ וְהָגִ֤יתָ בּוֹ֙ יוֹמָ֣ם וָלַ֔יְלָה לְמַ֙עַן֙ תִּשְׁמֹ֣ר לַעֲשׂ֔וֹת כְּכׇל\maqqaf הַכָּת֖וּב בּ֑וֹ כִּי\maqqaf אָ֛ז תַּצְלִ֥יחַ אֶת\maqqaf דְּרָכֶ֖ךָ וְאָ֥ז תַּשְׂכִּֽיל׃}
{This book of the law shall not depart out of thy mouth, but thou shalt meditate therein day and night, that thou mayest observe to do according to all that is written therein; for then thou shalt make thy ways prosperous, and then thou shalt have good success.}
\twoverse{\arabic{verse}}
{הֲל֤וֹא צִוִּיתִ֙יךָ֙ חֲזַ֣ק וֶאֱמָ֔ץ אַֽל\maqqaf תַּעֲרֹ֖ץ וְאַל\maqqaf תֵּחָ֑ת כִּ֤י עִמְּךָ֙ יְהֹוָ֣ה אֱלֹהֶ֔יךָ בְּכֹ֖ל אֲשֶׁ֥ר תֵּלֵֽךְ׃ \petucha }
{Have not I commanded thee? Be strong and of good courage; be not affrighted, neither be thou dismayed: for the \lord\space thy God is with thee whithersoever thou goest.’}
\twoverse{\arabic{verse}}
{וַיְצַ֣ו יְהוֹשֻׁ֔עַ אֶת\maqqaf שֹׁטְרֵ֥י הָעָ֖ם לֵאמֹֽר׃}
{Then Joshua commanded the officers of the people, saying:}
\twoverse{\arabic{verse}}
{עִבְר֣וּ \legarmeh  בְּקֶ֣רֶב הַֽמַּחֲנֶ֗ה וְצַוּ֤וּ אֶת\maqqaf הָעָם֙ לֵאמֹ֔ר הָכִ֥ינוּ לָכֶ֖ם צֵידָ֑ה כִּ֞י בְּע֣וֹד \legarmeh  שְׁלֹ֣שֶׁת יָמִ֗ים אַתֶּם֙ עֹֽבְרִים֙ אֶת\maqqaf הַיַּרְדֵּ֣ן הַזֶּ֔ה לָבוֹא֙ לָרֶ֣שֶׁת אֶת\maqqaf הָאָ֔רֶץ אֲשֶׁר֙ יְהֹוָ֣ה אֱלֹֽהֵיכֶ֔ם נֹתֵ֥ן לָכֶ֖ם לְרִשְׁתָּֽהּ׃ \petucha }
{‘Pass through the midst of the camp, and command the people, saying: Prepare you victuals; for within three days ye are to pass over this Jordan, to go in to possess the land, which the \lord\space your God giveth you to possess it.’}
\twoverse{\arabic{verse}}
{וְלָרֽאוּבֵנִי֙ וְלַגָּדִ֔י וְלַחֲצִ֖י שֵׁ֣בֶט הַֽמְנַשֶּׁ֑ה אָמַ֥ר יְהוֹשֻׁ֖עַ לֵאמֹֽר׃}
{And to the Reubenites, and to the Gadites, and to the half-tribe of Manasseh, spoke Joshua, saying:}
\twoverse{\arabic{verse}}
{זָכוֹר֙ אֶת\maqqaf הַדָּבָ֔ר אֲשֶׁ֨ר צִוָּ֥ה אֶתְכֶ֛ם מֹשֶׁ֥ה עֶבֶד\maqqaf יְהֹוָ֖ה לֵאמֹ֑ר יְהֹוָ֤ה אֱלֹֽהֵיכֶם֙ מֵנִ֣יחַ לָכֶ֔ם וְנָתַ֥ן לָכֶ֖ם אֶת\maqqaf הָאָ֥רֶץ הַזֹּֽאת׃}
{‘Remember the word which Moses the servant of the \lord\space commanded, you, saying: The \lord\space your God giveth you rest, and will give you this land.}
\twoverse{\arabic{verse}}
{נְשֵׁיכֶ֣ם טַפְּכֶם֮ וּמִקְנֵיכֶם֒ יֵשְׁב֕וּ בָּאָ֕רֶץ אֲשֶׁ֨ר נָתַ֥ן לָכֶ֛ם מֹשֶׁ֖ה בְּעֵ֣בֶר הַיַּרְדֵּ֑ן וְאַתֶּם֩ תַּעַבְר֨וּ חֲמֻשִׁ֜ים לִפְנֵ֣י אֲחֵיכֶ֗ם כֹּ֚ל גִּבּוֹרֵ֣י הַחַ֔יִל וַעֲזַרְתֶּ֖ם אוֹתָֽם׃}
{Your wives, your little ones, and your cattle, shall abide in the land which Moses gave you beyond the Jordan; but ye shall pass over before your brethren armed, all the mighty men of valour, and shall help them;}
\twoverse{\arabic{verse}}
{עַ֠ד אֲשֶׁר\maqqaf יָנִ֨יחַ יְהֹוָ֥ה \pasek  לַאֲחֵיכֶם֮ כָּכֶם֒ וְיָרְשׁ֣וּ גַם\maqqaf הֵ֔מָּה אֶת\maqqaf הָאָ֕רֶץ אֲשֶׁר\maqqaf יְהֹוָ֥ה אֱלֹהֵיכֶ֖ם נֹתֵ֣ן לָהֶ֑ם וְשַׁבְתֶּ֞ם לְאֶ֤רֶץ יְרֻשַּׁתְכֶם֙ וִירִשְׁתֶּ֣ם אוֹתָ֔הּ אֲשֶׁ֣ר \legarmeh  נָתַ֣ן לָכֶ֗ם מֹשֶׁה֙ עֶ֣בֶד יְהֹוָ֔ה בְּעֵ֥בֶר הַיַּרְדֵּ֖ן מִזְרַ֥ח הַשָּֽׁמֶשׁ׃}
{until the \lord\space have given your brethren rest, as unto you, and they also have possessed the land which the \lord\space your God giveth them; then ye shall return unto the land of your possession, and possess it, which Moses the servant of the \lord\space gave you beyond the Jordan toward the sunrising.’}
\twoverse{\arabic{verse}}
{וַֽיַּעֲנ֔וּ אֶת\maqqaf יְהוֹשֻׁ֖עַ לֵאמֹ֑ר כֹּ֤ל אֲשֶׁר\maqqaf צִוִּיתָ֙נוּ֙ נַעֲשֶׂ֔ה וְאֶֽל\maqqaf כׇּל\maqqaf אֲשֶׁ֥ר תִּשְׁלָחֵ֖נוּ נֵלֵֽךְ׃}
{And they answered Joshua, saying: ‘All that thou hast commanded us we will do, and whithersoever thou sendest us we will go.}
\twoverse{\arabic{verse}}
{כְּכֹ֤ל אֲשֶׁר\maqqaf שָׁמַ֙עְנוּ֙ אֶל\maqqaf מֹשֶׁ֔ה כֵּ֖ן נִשְׁמַ֣ע אֵלֶ֑יךָ רַ֠ק יִֽהְיֶ֞ה יְהֹוָ֤ה אֱלֹהֶ֙יךָ֙ עִמָּ֔ךְ כַּאֲשֶׁ֥ר הָיָ֖ה עִם\maqqaf מֹשֶֽׁה׃}
{According as we hearkened unto Moses in all things, so will we hearken unto thee; only the \lord\space thy God be with thee, as He was with Moses.}
\twoverse{\arabic{verse}}
{כׇּל\maqqaf אִ֞ישׁ אֲשֶׁר\maqqaf יַמְרֶ֣ה אֶת\maqqaf פִּ֗יךָ וְלֹֽא\maqqaf יִשְׁמַ֧ע אֶת\maqqaf דְּבָרֶ֛יךָ לְכֹ֥ל אֲשֶׁר\maqqaf תְּצַוֶּ֖נּוּ יוּמָ֑ת רַ֖ק חֲזַ֥ק וֶאֱמָֽץ׃ \petucha }
{Whosoever he be that shall rebel against thy commandment, and shall not hearken unto thy words in all that thou commandest him, he shall be put to death; only be strong and of good courage.’}
