\documentclass[11pt, openany]{book}
\usepackage[
paperheight=9.68in,
paperwidth=7.44in,
top=0.75in,bottom=0.75in,
inner=1in,
outer=.75in,
marginparsep=0.1in,
headsep=16pt
]{geometry}

\usepackage{chumash_format}
%\bibliography{Chumash_files/chumash_sources}
\newcommand{\texttitle}{ספר בראשית}
\newcommand{\engtitle}{Genesis}

\fancyhead[C]{\texttitle \space\textendash\space \englishfont Introduction}

\begin{document}

\include{chumash_files/0introduction}

\clearpage
\mainmatter

\pagenumbering{arabic}
\setmainheader

\include{chumash_files/1Genesis_chumash}
\fancyhead[C]{פרשת שמות למנחה בשבת ויחי}
\addcontentsline{toc}{chapter}{פרשת שמות למנחה בשבת ויחי}
\newparsha{שמות}
\setcounter{chap}{1}
\setcounter{verse}{1}
\twoverse{\textsf{שמות}}
{וְאֵ֗לֶּה שְׁמוֹת֙ בְּנֵ֣י יִשְׂרָאֵ֔ל הַבָּאִ֖ים מִצְרָ֑יְמָה אֵ֣ת יַעֲקֹ֔ב אִ֥ישׁ וּבֵית֖וֹ בָּֽאוּ׃}
{NOW THESE are the names of the sons of Israel, who came into Egypt with Jacob; every man came with his household:}
\twoverse{\arabic{verse}}
{רְאוּבֵ֣ן שִׁמְע֔וֹן לֵוִ֖י וִיהוּדָֽה׃}
{Reuben, Simeon, Levi, and Judah;}
\twoverse{\arabic{verse}}
{יִשָּׂשכָ֥ר זְבוּלֻ֖ן וּבִנְיָמִֽן׃}
{Issachar, Zebulun, and Benjamin;}
\twoverse{\arabic{verse}}
{דָּ֥ן וְנַפְתָּלִ֖י גָּ֥ד וְאָשֵֽׁר׃}
{Dan and Naphtali, Gad and Asher.}
\twoverse{\arabic{verse}}
{וַֽיְהִ֗י כׇּל\maqqaf נֶ֛פֶשׁ יֹצְאֵ֥י יֶֽרֶךְ\maqqaf יַעֲקֹ֖ב שִׁבְעִ֣ים נָ֑פֶשׁ וְיוֹסֵ֖ף הָיָ֥ה בְמִצְרָֽיִם׃}
{And all the souls that came out of the loins of Jacob were seventy souls; and Joseph was in Egypt already.}
\twoverse{\arabic{verse}}
{וַיָּ֤מׇת יוֹסֵף֙ וְכׇל\maqqaf אֶחָ֔יו וְכֹ֖ל הַדּ֥וֹר הַהֽוּא׃}
{And Joseph died, and all his brethren, and all that generation.}
\twoverse{\arabic{verse}}
{וּבְנֵ֣י יִשְׂרָאֵ֗ל פָּר֧וּ וַֽיִּשְׁרְצ֛וּ וַיִּרְבּ֥וּ וַיַּֽעַצְמ֖וּ בִּמְאֹ֣ד מְאֹ֑ד וַתִּמָּלֵ֥א הָאָ֖רֶץ אֹתָֽם׃ \petucha }
{And the children of Israel were fruitful, and increased abundantly, and multiplied, and waxed exceeding mighty; and the land was filled with them.}
\twoverse{\arabic{verse}}
{וַיָּ֥קׇם מֶֽלֶךְ\maqqaf חָדָ֖שׁ עַל\maqqaf מִצְרָ֑יִם אֲשֶׁ֥ר לֹֽא\maqqaf יָדַ֖ע אֶת\maqqaf יוֹסֵֽף׃}
{Now there arose a new king over Egypt, who knew not Joseph.}
\twoverse{\textsf{לוי}}
{וַיֹּ֖אמֶר אֶל\maqqaf עַמּ֑וֹ הִנֵּ֗ה עַ֚ם בְּנֵ֣י יִשְׂרָאֵ֔ל רַ֥ב וְעָצ֖וּם מִמֶּֽנּוּ׃}
{And he said unto his people: ‘Behold, the people of the children of Israel are too many and too mighty for us;}
\twoverse{\arabic{verse}}
{הָ֥בָה נִֽתְחַכְּמָ֖ה ל֑וֹ פֶּן\maqqaf יִרְבֶּ֗ה וְהָיָ֞ה כִּֽי\maqqaf תִקְרֶ֤אנָה מִלְחָמָה֙ וְנוֹסַ֤ף גַּם\maqqaf הוּא֙ עַל\maqqaf שֹׂ֣נְאֵ֔ינוּ וְנִלְחַם\maqqaf בָּ֖נוּ וְעָלָ֥ה מִן\maqqaf הָאָֽרֶץ׃}
{come, let us deal wisely with them, lest they multiply, and it come to pass, that, when there befalleth us any war, they also join themselves unto our enemies, and fight against us, and get them up out of the land.’}
\twoverse{\arabic{verse}}
{וַיָּשִׂ֤ימוּ עָלָיו֙ שָׂרֵ֣י מִסִּ֔ים לְמַ֥עַן עַנֹּת֖וֹ בְּסִבְלֹתָ֑ם וַיִּ֜בֶן עָרֵ֤י מִסְכְּנוֹת֙ לְפַרְעֹ֔ה אֶת\maqqaf פִּתֹ֖ם וְאֶת\maqqaf רַעַמְסֵֽס׃}
{Therefore they did set over them taskmasters to afflict them with their burdens. And they built for Pharaoh store-cities, Pithom and Raamses.}
\twoverse{\arabic{verse}}
{וְכַאֲשֶׁר֙ יְעַנּ֣וּ אֹת֔וֹ כֵּ֥ן יִרְבֶּ֖ה וְכֵ֣ן יִפְרֹ֑ץ וַיָּקֻ֕צוּ מִפְּנֵ֖י בְּנֵ֥י יִשְׂרָאֵֽל׃}
{But the more they afflicted them, the more they multiplied and the more they spread abroad. And they were adread because of the children of Israel.}
\twoverse{\textsf{ישראל}}
{וַיַּעֲבִ֧דוּ מִצְרַ֛יִם אֶת\maqqaf בְּנֵ֥י יִשְׂרָאֵ֖ל בְּפָֽרֶךְ׃}
{And the Egyptians made the children of Israel to serve with rigour.}
\twoverse{\arabic{verse}}
{וַיְמָרְר֨וּ אֶת\maqqaf חַיֵּיהֶ֜ם בַּעֲבֹדָ֣ה קָשָׁ֗ה בְּחֹ֙מֶר֙ וּבִלְבֵנִ֔ים וּבְכׇל\maqqaf עֲבֹדָ֖ה בַּשָּׂדֶ֑ה אֵ֚ת כׇּל\maqqaf עֲבֹ֣דָתָ֔ם אֲשֶׁר\maqqaf עָבְד֥וּ בָהֶ֖ם בְּפָֽרֶךְ׃}
{And they made their lives bitter with hard service, in mortar and in brick, and in all manner of service in the field; in all their service, wherein they made them serve with rigour.}
\twoverse{\arabic{verse}}
{וַיֹּ֙אמֶר֙ מֶ֣לֶךְ מִצְרַ֔יִם לַֽמְיַלְּדֹ֖ת הָֽעִבְרִיֹּ֑ת אֲשֶׁ֨ר שֵׁ֤ם הָֽאַחַת֙ שִׁפְרָ֔ה וְשֵׁ֥ם הַשֵּׁנִ֖ית פּוּעָֽה׃}
{And the king of Egypt spoke to the Hebrew midwives, of whom the name of the one was Shiphrah, and the name of the other Puah;}
\twoverse{\arabic{verse}}
{וַיֹּ֗אמֶר בְּיַלֶּדְכֶן֙ אֶת\maqqaf הָֽעִבְרִיּ֔וֹת וּרְאִיתֶ֖ן עַל\maqqaf הָאׇבְנָ֑יִם אִם\maqqaf בֵּ֥ן הוּא֙ וַהֲמִתֶּ֣ן אֹת֔וֹ וְאִם\maqqaf בַּ֥ת הִ֖וא וָחָֽיָה׃}
{and he said: ‘When ye do the office of a midwife to the Hebrew women, ye shall look upon the birthstool: if it be a son, then ye shall kill him; but if it be a daughter, then she shall live.’}
\twoverse{\arabic{verse}}
{וַתִּירֶ֤אןָ הַֽמְיַלְּדֹת֙ אֶת\maqqaf הָ֣אֱלֹהִ֔ים וְלֹ֣א עָשׂ֔וּ כַּאֲשֶׁ֛ר דִּבֶּ֥ר אֲלֵיהֶ֖ן מֶ֣לֶךְ מִצְרָ֑יִם וַתְּחַיֶּ֖יןָ אֶת\maqqaf הַיְלָדִֽים׃}
{But the midwives feared God, and did not as the king of Egypt commanded them, but saved the men-children alive.}

\clearpage
\fancyhead[C]{\haftarahname}
\include{chumash_files/1Genesis_haftarot}

\vspace{12pt}
\section*{מפטיר לשבת ראש חודש}
\addcontentsline{toc}{section}{מפטיר לשבת ראש חודש}
\renewcommand{\haftarahname}{מפטיר לשבת ראש חודש}
\vspace{-12pt}
\nopagebreak
\label{maft_ShabRCh}
\haftnote{The Maftir for Shabbat Rosh \d{H}odesh is Numbers 28:9\verserangechar 15.}
\setcounter{chap}{28}
\setcounter{verse}{9}
\twoverse{\Roman{chap}:\arabic{verse}}{וּבְיוֹם֙ הַשַּׁבָּ֔ת שְׁנֵֽי־כְבָשִׂ֥ים בְּנֵֽי־שָׁנָ֖ה תְּמִימִ֑ם וּשְׁנֵ֣י עֶשְׂרֹנִ֗ים סֹ֧לֶת מִנְחָ֛ה בְּלוּלָ֥ה בַשֶּׁ֖מֶן וְנִסְכּֽוֹ׃}{And on the sabbath day two he-lambs of the first year without blemish, and two tenth parts of an ephah of fine flour for a meal-offering, mingled with oil, and the drink-offering thereof.}
\twoverse{\arabic{verse}}{עֹלַ֥ת שַׁבַּ֖ת בְּשַׁבַּתּ֑וֹ עַל־עֹלַ֥ת הַתָּמִ֖יד וְנִסְכָּֽהּ׃ \petucha}{This is the burnt-offering of every sabbath, beside the continual burnt-offering, and the drink-offering thereof.}
\twoverse{\arabic{verse}}{וּבְרָאשֵׁי֙ חׇדְשֵׁיכֶ֔ם תַּקְרִ֥יבוּ עֹלָ֖ה לַיהֹוָ֑ה פָּרִ֨ים בְּנֵֽי־בָקָ֤ר שְׁנַ֙יִם֙ וְאַ֣יִל אֶחָ֔ד כְּבָשִׂ֧ים בְּנֵי־שָׁנָ֛ה שִׁבְעָ֖ה תְּמִימִֽם׃}{And in your new moons ye shall present a burnt-offering unto the \lord: two young bullocks, and one ram, seven he-lambs of the first year without blemish;}
\twoverse{\arabic{verse}}{וּשְׁלֹשָׁ֣ה עֶשְׂרֹנִ֗ים סֹ֤לֶת מִנְחָה֙ בְּלוּלָ֣ה בַשֶּׁ֔מֶן לַפָּ֖ר הָאֶחָ֑ד וּשְׁנֵ֣י עֶשְׂרֹנִ֗ים סֹ֤לֶת מִנְחָה֙ בְּלוּלָ֣ה בַשֶּׁ֔מֶן לָאַ֖יִל הָֽאֶחָֽד׃}{and three tenth parts of an ephah of fine flour for a meal-offering, mingled with oil, for each bullock; and two tenth parts of fine flour for a meal-offering, mingled with oil, for the one ram;}
\twoverse{\arabic{verse}}{וְעִשָּׂרֹ֣ן עִשָּׂר֗וֹן סֹ֤לֶת מִנְחָה֙ בְּלוּלָ֣ה בַשֶּׁ֔מֶן לַכֶּ֖בֶשׂ הָאֶחָ֑ד עֹלָה֙ רֵ֣יחַ נִיחֹ֔חַ אִשֶּׁ֖ה לַיהֹוָֽה׃}{and a several tenth part of fine flour mingled with oil for a meal-offering unto every lamb; for a burnt-offering of a sweet savour, an offering made by fire unto the \lord.}
\twoverse{\arabic{verse}}{וְנִסְכֵּיהֶ֗ם חֲצִ֣י הַהִין֩ יִהְיֶ֨ה לַפָּ֜ר וּשְׁלִישִׁ֧ת הַהִ֣ין לָאַ֗יִל וּרְבִיעִ֥ת הַהִ֛ין לַכֶּ֖בֶשׂ יָ֑יִן זֹ֣את עֹלַ֥ת חֹ֙דֶשׁ֙ בְּחׇדְשׁ֔וֹ לְחׇדְשֵׁ֖י הַשָּׁנָֽה׃}{And their drink-offerings shall be half a hin of wine for a bullock, and the third part of a hin for the ram, and the fourth part of a hin for a lamb. This is the burnt-offering of every new moon throughout the months of the year.}
\twoverse{\arabic{verse}}{וּשְׂעִ֨יר עִזִּ֥ים אֶחָ֛ד לְחַטָּ֖את לַיהֹוָ֑ה עַל־עֹלַ֧ת הַתָּמִ֛יד יֵעָשֶׂ֖ה וְנִסְכּֽוֹ׃ \setuma }{And one he-goat for a sin-offering unto the \lord; it shall be offered beside the continual burnt-offering, and the drink-offering thereof.}
\newhaftarah{שבת ראש חודש}
\label{haft_ShabRCh}
\haftnote{The Haftarah for Shabbat Rosh \d{H}odesh is Isaiah 66:1\verserangechar 24.}
\setcounter{chap}{66}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}{כֹּ֚ה אָמַ֣ר יְהֹוָ֔ה הַשָּׁמַ֣יִם כִּסְאִ֔י וְהָאָ֖רֶץ הֲדֹ֣ם רַגְלָ֑י אֵי־זֶ֥ה בַ֙יִת֙ אֲשֶׁ֣ר תִּבְנוּ־לִ֔י וְאֵי־זֶ֥ה מָק֖וֹם מְנוּחָתִֽי׃}{Thus saith the \lord: The heaven is My throne, and the earth is My footstool; where is the house that ye may build unto Me? And where is the place that may be My resting-place?}
\twoverse{\arabic{verse}}{וְאֶת־כׇּל־אֵ֙לֶּה֙ יָדִ֣י עָשָׂ֔תָה וַיִּהְי֥וּ כׇל־אֵ֖לֶּה נְאֻם־יְהֹוָ֑ה וְאֶל־זֶ֣ה אַבִּ֔יט אֶל־עָנִי֙ וּנְכֵה־ר֔וּחַ וְחָרֵ֖ד עַל־דְּבָרִֽי׃}{For all these things hath My hand made, and so all these things came to be, saith the \lord; but on this man will I look, even on him that is poor and of a contrite spirit, and trembleth at My word.}
\twoverse{\arabic{verse}}{שׁוֹחֵ֨ט הַשּׁ֜וֹר מַכֵּה־אִ֗ישׁ זוֹבֵ֤חַ הַשֶּׂה֙ עֹ֣רֵֽף כֶּ֔לֶב מַעֲלֵ֤ה מִנְחָה֙ דַּם־חֲזִ֔יר מַזְכִּ֥יר לְבֹנָ֖ה מְבָ֣רֵֽךְ אָ֑וֶן גַּם־הֵ֗מָּה בָּֽחֲרוּ֙ בְּדַרְכֵיהֶ֔ם וּבְשִׁקּוּצֵיהֶ֖ם נַפְשָׁ֥ם חָפֵֽצָה׃}{He that killeth an ox is as if he slew a man; he that sacrificeth a lamb, as if he broke a dog’s neck; He that offereth a meal-offering, as if he offered swine’s blood; he that maketh a memorial-offering of frankincense, as if he blessed an idol; according as they have chosen their own ways, and their soul delighteth in their abominations;}
\twoverse{\arabic{verse}}{גַּם־אֲנִ֞י אֶבְחַ֣ר בְּתַעֲלֻלֵיהֶ֗ם וּמְגֽוּרֹתָם֙ אָבִ֣יא לָהֶ֔ם יַ֤עַן קָרָ֙אתִי֙ וְאֵ֣ין עוֹנֶ֔ה דִּבַּ֖רְתִּי וְלֹ֣א שָׁמֵ֑עוּ וַיַּעֲשׂ֤וּ הָרַע֙ בְּעֵינַ֔י וּבַאֲשֶׁ֥ר לֹא־חָפַ֖צְתִּי בָּחָֽרוּ׃ \setuma }{Even so I will choose their mockings, And will bring their fears upon them; Because when I called, none did answer; When I spoke, they did not hear, But they did that which was evil in Mine eyes, And chose that in which I delighted not.}
\twoverse{\arabic{verse}}{שִׁמְעוּ֙ דְּבַר־יְהֹוָ֔ה הַחֲרֵדִ֖ים אֶל־דְּבָר֑וֹ אָמְרוּ֩ אֲחֵיכֶ֨ם שֹׂנְאֵיכֶ֜ם מְנַדֵּיכֶ֗ם לְמַ֤עַן שְׁמִי֙ יִכְבַּ֣ד יְהֹוָ֔ה וְנִרְאֶ֥ה בְשִׂמְחַתְכֶ֖ם וְהֵ֥ם יֵבֹֽשׁוּ׃}{Hear the word of the \lord, Ye that tremble at His word: Your brethren that hate you, that cast you out for My name’s sake, have said: ‘Let the \lord\space be glorified, That we may gaze upon your joy’, But they shall be ashamed.}
\twoverse{\arabic{verse}}{ק֤וֹל שָׁאוֹן֙ מֵעִ֔יר ק֖וֹל מֵהֵיכָ֑ל ק֣וֹל יְהֹוָ֔ה מְשַׁלֵּ֥ם גְּמ֖וּל לְאֹיְבָֽיו׃}{Hark! an uproar from the city, Hark! it cometh from the temple, Hark! the \lord\space rendereth recompense to His enemies.}
\twoverse{\arabic{verse}}{בְּטֶ֥רֶם תָּחִ֖יל יָלָ֑דָה בְּטֶ֨רֶם יָב֥וֹא חֵ֛בֶל לָ֖הּ וְהִמְלִ֥יטָה זָכָֽר׃}{Before she travailed, she brought forth; Before her pain came, She was delivered of a man-child.}
\twoverse{\arabic{verse}}{מִֽי־שָׁמַ֣ע כָּזֹ֗את מִ֤י רָאָה֙ כָּאֵ֔לֶּה הֲי֤וּחַל אֶ֙רֶץ֙ בְּי֣וֹם אֶחָ֔ד אִם־יִוָּ֥לֵֽד גּ֖וֹי פַּ֣עַם אֶחָ֑ת כִּי־חָ֛לָה גַּם־יָלְדָ֥ה צִיּ֖וֹן אֶת־בָּנֶֽיהָ׃}{Who hath heard such a thing? Who hath seen such things? Is a land born in one day? Is a nation brought forth at once? For as soon as Zion travailed, She brought forth her children.}
\twoverse{\arabic{verse}}{הַאֲנִ֥י אַשְׁבִּ֛יר וְלֹ֥א אוֹלִ֖יד יֹאמַ֣ר יְהֹוָ֑ה אִם־אֲנִ֧י הַמּוֹלִ֛יד וְעָצַ֖רְתִּי אָמַ֥ר אֱלֹהָֽיִךְ׃ \setuma }{Shall I bring to the birth, and not cause to bring forth? Saith the \lord; Shall I that cause to bring forth shut the womb? Saith thy God.}
\twoverse{\arabic{verse}}{שִׂמְח֧וּ אֶת־יְרוּשָׁלַ֛͏ִם וְגִ֥ילוּ בָ֖הּ כׇּל־אֹהֲבֶ֑יהָ שִׂ֤ישׂוּ אִתָּהּ֙ מָשׂ֔וֹשׂ כׇּל־הַמִּֽתְאַבְּלִ֖ים עָלֶֽיהָ׃}{Rejoice ye with Jerusalem, And be glad with her, all ye that love her; Rejoice for joy with her, All ye that mourn for her;}
\twoverse{\arabic{verse}}{לְמַ֤עַן תִּֽינְקוּ֙ וּשְׂבַעְתֶּ֔ם מִשֹּׁ֖ד תַּנְחֻמֶ֑יהָ לְמַ֧עַן תָּמֹ֛צּוּ וְהִתְעַנַּגְתֶּ֖ם מִזִּ֥יז כְּבוֹדָֽהּ׃ \setuma }{That ye may suck, and be satisfied With the breast of her consolations; That ye may drink deeply with delight Of the abundance of her glory.}
\twoverse{\arabic{verse}}{כִּי־כֹ֣ה \legarmeh  אָמַ֣ר יְהֹוָ֗ה הִנְנִ֣י נֹטֶֽה־אֵ֠לֶ֠יהָ כְּנָהָ֨ר שָׁל֜וֹם וּכְנַ֧חַל שׁוֹטֵ֛ף כְּב֥וֹד גּוֹיִ֖ם וִֽינַקְתֶּ֑ם עַל־צַד֙ תִּנָּשֵׂ֔אוּ וְעַל־בִּרְכַּ֖יִם תְּשׇׁעֳשָֽׁעוּ׃}{For thus saith the \lord: Behold, I will extend peace to her like a river. And the wealth of the nations like an overflowing stream, and ye shall suck thereof: Ye shall be borne upon the side, and shall be dandled upon the knees.}
\twoverse{\arabic{verse}}{כְּאִ֕ישׁ אֲשֶׁ֥ר אִמּ֖וֹ תְּנַחֲמֶ֑נּוּ כֵּ֤ן אָֽנֹכִי֙ אֲנַ֣חֶמְכֶ֔ם וּבִירֽוּשָׁלַ֖͏ִם תְּנֻחָֽמוּ׃}{As one whom his mother comforteth, So will I comfort you; and ye shall be comforted in Jerusalem.}
\twoverse{\arabic{verse}}{וּרְאִיתֶם֙ וְשָׂ֣שׂ לִבְּכֶ֔ם וְעַצְמוֹתֵיכֶ֖ם כַּדֶּ֣שֶׁא תִפְרַ֑חְנָה וְנוֹדְעָ֤ה יַד־יְהֹוָה֙ אֶת־עֲבָדָ֔יו וְזָעַ֖ם אֶת־אֹיְבָֽיו׃ \setuma }{And when ye see this, your heart shall rejoice, and your bones shall flourish like young grass; and the hand of the \lord\space shall be known toward His servants, and He will have indignation against His enemies.}
\twoverse{\arabic{verse}}{כִּי־הִנֵּ֤ה יְהֹוָה֙ בָּאֵ֣שׁ יָב֔וֹא וְכַסּוּפָ֖ה מַרְכְּבֹתָ֑יו לְהָשִׁ֤יב בְּחֵמָה֙ אַפּ֔וֹ וְגַעֲרָת֖וֹ בְּלַהֲבֵי־אֵֽשׁ׃}{For, behold, the \lord\space will come in fire, And His chariots shall be like the whirlwind; to render His anger with fury, And His rebuke with flames of fire. }
\twoverse{\arabic{verse}}{כִּ֤י בָאֵשׁ֙ יְהֹוָ֣ה נִשְׁפָּ֔ט וּבְחַרְבּ֖וֹ אֶת־כׇּל־בָּשָׂ֑ר וְרַבּ֖וּ חַֽלְלֵ֥י יְהֹוָֽה׃}{For by fire will the \lord\space contend, And by His sword with all flesh; And the slain of the \lord\space shall be many.}
\twoverse{\arabic{verse}}{הַמִּתְקַדְּשִׁ֨ים וְהַמִּֽטַּהֲרִ֜ים אֶל־הַגַּנּ֗וֹת אַחַ֤ר \qk{אַחַת֙}{אחד} בַּתָּ֔וֶךְ אֹֽכְלֵי֙ בְּשַׂ֣ר הַחֲזִ֔יר וְהַשֶּׁ֖קֶץ וְהָעַכְבָּ֑ר יַחְדָּ֥ו יָסֻ֖פוּ נְאֻם־יְהֹוָֽה׃}{They that sanctify themselves and purify themselves to go unto the gardens, behind one in the midst, Eating swine’s flesh, and the detestable thing, and the mouse, Shall be consumed together, saith the \lord.}
\twoverse{\arabic{verse}}{וְאָנֹכִ֗י מַֽעֲשֵׂיהֶם֙ וּמַחְשְׁבֹ֣תֵיהֶ֔ם בָּאָ֕ה לְקַבֵּ֥ץ אֶת־כׇּל־הַגּוֹיִ֖ם וְהַלְּשֹׁנ֑וֹת וּבָ֖אוּ וְרָא֥וּ אֶת־כְּבוֹדִֽי׃}{For I [know] their works and their thoughts; [the time] cometh, that I will gather all nations and tongues; and they shall come, and shall see My glory.}
\twoverse{\arabic{verse}}{וְשַׂמְתִּ֨י בָהֶ֜ם א֗וֹת וְשִׁלַּחְתִּ֣י מֵהֶ֣ם \small ׀ \normalsize פְּ֠לֵיטִ֠ים אֶֽל־הַגּוֹיִ֞ם תַּרְשִׁ֨ישׁ פּ֥וּל וְל֛וּד מֹ֥שְׁכֵי קֶ֖שֶׁת תֻּבַ֣ל וְיָוָ֑ן הָאִיִּ֣ים הָרְחֹקִ֗ים אֲשֶׁ֨ר לֹֽא־שָׁמְע֤וּ אֶת־שִׁמְעִי֙ וְלֹא־רָא֣וּ אֶת־כְּבוֹדִ֔י וְהִגִּ֥ידוּ אֶת־כְּבוֹדִ֖י בַּגּוֹיִֽם׃}{And I will work a sign among them, and I will send such as escape of them unto the nations, to Tarshish, Pul and Lud, that draw the bow, to Tubal and Javan, to the isles afar off, that have not heard My fame, neither have seen My glory; and they shall declare My glory among the nations.}
\twoverse{\arabic{verse}}{וְהֵבִ֣יאוּ אֶת־כׇּל־אֲחֵיכֶ֣ם מִכׇּל־הַגּוֹיִ֣ם \small ׀ \normalsize מִנְחָ֣ה \small ׀ \normalsize לַיהֹוָ֡ה בַּסּוּסִ֡ים וּ֠בָרֶ֠כֶב וּבַצַּבִּ֨ים וּבַפְּרָדִ֜ים וּבַכִּרְכָּר֗וֹת עַ֣ל הַ֥ר קׇדְשִׁ֛י יְרוּשָׁלַ֖͏ִם אָמַ֣ר יְהֹוָ֑ה כַּאֲשֶׁ֣ר יָבִ֩יאוּ֩ בְנֵ֨י יִשְׂרָאֵ֧ל אֶת־הַמִּנְחָ֛ה בִּכְלִ֥י טָה֖וֹר בֵּ֥ית יְהֹוָֽה׃}{And they shall bring all your brethren out of all the nations for an offering unto the \lord, upon horses, and in chariots, and in fitters, and upon mules, and upon swift beasts, to My holy mountain Jerusalem, saith the \lord, as the children of Israel bring their offering in a clean vessel into the house of the \lord.}
\twoverse{\arabic{verse}}{וְגַם־מֵהֶ֥ם אֶקַּ֛ח לַכֹּהֲנִ֥ים לַלְוִיִּ֖ם אָמַ֥ר יְהֹוָֽה׃}{And of them also will I take for the priests and for the Levites, saith the \lord.}
\twoverse{\arabic{verse}}{כִּ֣י כַאֲשֶׁ֣ר הַשָּׁמַ֣יִם הַ֠חֲדָשִׁ֠ים וְהָאָ֨רֶץ הַחֲדָשָׁ֜ה אֲשֶׁ֨ר אֲנִ֥י עֹשֶׂ֛ה עֹמְדִ֥ים לְפָנַ֖י נְאֻם־יְהֹוָ֑ה כֵּ֛ן יַעֲמֹ֥ד זַרְעֲכֶ֖ם וְשִׁמְכֶֽם׃}{For as the new heavens and the new earth, which I will make, shall remain before Me, saith the \lord, so shall your seed and your name remain.}
\twoverse{\arabic{verse}}{וְהָיָ֗ה מִֽדֵּי־חֹ֙דֶשׁ֙ בְּחׇדְשׁ֔וֹ וּמִדֵּ֥י שַׁבָּ֖ת בְּשַׁבַּתּ֑וֹ יָב֧וֹא כׇל־בָּשָׂ֛ר לְהִשְׁתַּחֲוֺ֥ת לְפָנַ֖י אָמַ֥ר יְהֹוָֽה׃}{And it shall come to pass, that from one new moon to another, and from one sabbath to another, shall all flesh come to worship before Me, Saith the \lord.}
\twoverse{\arabic{verse}}{וְיָצְא֣וּ וְרָא֔וּ בְּפִגְרֵי֙ הָאֲנָשִׁ֔ים הַפֹּשְׁעִ֖ים בִּ֑י כִּ֣י תוֹלַעְתָּ֞ם לֹ֣א תָמ֗וּת וְאִשָּׁם֙ לֹ֣א תִכְבֶּ֔ה וְהָי֥וּ דֵרָא֖וֹן לְכׇל־בָּשָֽׂר׃ \small והיה מדי חדש בחדשו ומדי שבת בשבתו יבוא כל בשר להשתחות לפני אמר יהוה\small}{And they shall go forth, and look Upon the carcasses of the men that have rebelled against Me; For their worm shall not die, Neither shall their fire be quenched; And they shall be an abhorring unto all flesh. }
\newhaftarah{שבת מחר חודש}
\label{haft_macharchodesh}
\haftnote{The Haftarah for Shabbat Ma\d{h}ar \d{H}odesh is I Samuel 20:18\verserangechar 42.}
\setcounter{chap}{20}
\setcounter{verse}{18}
\twoverse{\Roman{chap}:\arabic{verse}}{וַיֹּאמֶר־ל֥וֹ יְהוֹנָתָ֖ן מָחָ֣ר חֹ֑דֶשׁ וְנִפְקַ֕דְתָּ כִּ֥י יִפָּקֵ֖ד מוֹשָׁבֶֽךָ׃}{And Jonathan said unto him: ‘To-morrow is the new moon; and thou wilt be missed, thy seat will be empty.}
\twoverse{\arabic{verse}}{וְשִׁלַּשְׁתָּ֙ תֵּרֵ֣ד מְאֹ֔ד וּבָאתָ֙ אֶל־הַמָּק֔וֹם אֲשֶׁר־נִסְתַּ֥רְתָּ שָּׁ֖ם בְּי֣וֹם הַֽמַּעֲשֶׂ֑ה וְיָ֣שַׁבְתָּ֔ אֵ֖צֶל הָאֶ֥בֶן הָאָֽזֶל׃}{And in the third day thou shalt hide thyself well, and come to the place where thou didst hide thyself in the day of work, and shalt remain by the stone Ezel.}
\twoverse{\arabic{verse}}{וַאֲנִ֕י שְׁלֹ֥שֶׁת הַחִצִּ֖ים צִדָּ֣ה אוֹרֶ֑ה לְשַֽׁלַּֽח־לִ֖י לְמַטָּרָֽה׃}{And I will shoot three arrows to the side-ward, as though I shot at a mark.}
\twoverse{\arabic{verse}}{וְהִנֵּה֙ אֶשְׁלַ֣ח אֶת־הַנַּ֔עַר לֵ֖ךְ מְצָ֣א אֶת־הַחִצִּ֑ים אִם־אָמֹר֩ אֹמַ֨ר לַנַּ֜עַר הִנֵּ֥ה הַחִצִּ֣ים \legarmeh  מִמְּךָ֣ וָהֵ֗נָּה קָחֶ֧נּוּ \small׀\small וָבֹ֛אָה כִּי־שָׁל֥וֹם לְךָ֛ וְאֵ֥ין דָּבָ֖ר חַי־יְהֹוָֽה׃}{And, behold, I will send the lad: Go, find the arrows. If I say unto the lad: Behold, the arrows are on this side of thee; take them, and come; for there is peace to thee and no hurt, as the \lord\space liveth.}
\twoverse{\arabic{verse}}{וְאִם־כֹּ֤ה אֹמַר֙ לָעֶ֔לֶם הִנֵּ֥ה הַחִצִּ֖ים מִמְּךָ֣ וָהָ֑לְאָה לֵ֕ךְ כִּ֥י שִֽׁלַּחֲךָ֖ יְהֹוָֽה׃}{But if I say thus unto the boy: Behold, the arrows are beyond thee; go thy way; for the \lord\space hath sent thee away.}
\twoverse{\arabic{verse}}{וְהַ֨דָּבָ֔ר אֲשֶׁ֥ר דִּבַּ֖רְנוּ אֲנִ֣י וָאָ֑תָּה הִנֵּ֧ה יְהֹוָ֛ה בֵּינִ֥י וּבֵינְךָ֖ עַד־עוֹלָֽם׃ \setuma }{And as touching the matter which I and thou have spoken of, behold, the \lord\space is between me and thee for ever.’}
\twoverse{\arabic{verse}}{וַיִּסָּתֵ֥ר דָּוִ֖ד בַּשָּׂדֶ֑ה וַיְהִ֣י הַחֹ֔דֶשׁ וַיֵּ֧שֶׁב הַמֶּ֛לֶךְ \qk{אֶל־}{על} לֶאֱכֽוֹל׃}{So David hid himself in the field; and when the new moon was come, the king sat him down to the meal to eat.}
\twoverse{\arabic{verse}}{וַיֵּ֣שֶׁב הַ֠מֶּ֠לֶךְ עַל־מ֨וֹשָׁב֜וֹ כְּפַ֣עַם \legarmeh  בְּפַ֗עַם אֶל־מוֹשַׁב֙ הַקִּ֔יר וַיָּ֙קׇם֙ יְה֣וֹנָתָ֔ן וַיֵּ֥שֶׁב אַבְנֵ֖ר מִצַּ֣ד שָׁא֑וּל וַיִּפָּקֵ֖ד מְק֥וֹם דָּוִֽד׃}{And the king sat upon his seat, as at other times, even upon the seat by the wall; and Jonathan stood up, and Abner sat by Saul’s side; but David’s place was empty.}
\twoverse{\arabic{verse}}{וְלֹא־דִבֶּ֥ר שָׁא֛וּל מְא֖וּמָה בַּיּ֣וֹם הַה֑וּא כִּ֤י אָמַר֙ מִקְרֶ֣ה ה֔וּא בִּלְתִּ֥י טָה֛וֹר ה֖וּא כִּי־לֹ֥א טָהֽוֹר׃ \setuma }{Nevertheless Saul spoke not any thing that day; for he thought: ‘Something hath befallen him, he is unclean; surely he is not clean.’}
\twoverse{\arabic{verse}}{וַיְהִ֗י מִֽמׇּחֳרַ֤ת הַחֹ֙דֶשׁ֙ הַשֵּׁנִ֔י וַיִּפָּקֵ֖ד מְק֣וֹם דָּוִ֑ד \petucha וַיֹּ֤אמֶר שָׁאוּל֙ אֶל־יְהוֹנָתָ֣ן בְּנ֔וֹ מַדּ֜וּעַ לֹא־בָ֧א בֶן־יִשַׁ֛י גַּם־תְּמ֥וֹל גַּם־הַיּ֖וֹם אֶל־הַלָּֽחֶם׃}{And it came to pass on the morrow after the new moon, which was the second day, that David’s place was empty; and Saul said unto Jonathan his son: ‘Wherefore cometh not the son of Jesse to the meal, neither yesterday, nor to-day?’}
\twoverse{\arabic{verse}}{וַיַּ֥עַן יְהוֹנָתָ֖ן אֶת־שָׁא֑וּל נִשְׁאֹ֨ל נִשְׁאַ֥ל דָּוִ֛ד מֵעִמָּדִ֖י עַד־בֵּ֥ית לָֽחֶם׃}{And Jonathan answered Saul: ‘David earnestly asked leave of me to go to Beth-lehem;}
\twoverse{\arabic{verse}}{וַיֹּ֡אמֶר שַׁלְּחֵ֣נִי נָ֡א כִּ֣י זֶ֩בַח֩ מִשְׁפָּחָ֨ה לָ֜נוּ בָּעִ֗יר וְה֤וּא צִוָּה־לִי֙ אָחִ֔י וְעַתָּ֗ה אִם־מָצָ֤אתִי חֵן֙ בְּעֵינֶ֔יךָ אִמָּ֥לְטָה נָּ֖א וְאֶרְאֶ֣ה אֶת־אֶחָ֑י עַל־כֵּ֣ן לֹא־בָ֔א אֶל־שֻׁלְחַ֖ן הַמֶּֽלֶךְ׃ \setuma }{and he said: Let me go, I pray thee; for our family hath a sacrifice in the city; and my brother, he hath commanded me; and now, if I have found favour in thine eyes, let me get away, I pray thee, and see my brethren. Therefore he is not come unto the king’s table.’}
\twoverse{\arabic{verse}}{וַיִּֽחַר־אַ֤ף שָׁאוּל֙ בִּיה֣וֹנָתָ֔ן וַיֹּ֣אמֶר ל֔וֹ בֶּֽן־נַעֲוַ֖ת הַמַּרְדּ֑וּת הֲל֣וֹא יָדַ֗עְתִּי כִּֽי־בֹחֵ֤ר אַתָּה֙ לְבֶן־יִשַׁ֔י לְבׇ֨שְׁתְּךָ֔ וּלְבֹ֖שֶׁת עֶרְוַ֥ת אִמֶּֽךָ׃}{Then Saul’s anger was kindled against Jonathan, and he said unto him: ‘Thou son of perverse rebellion, do not I know that thou hast chosen the son of Jesse to thine own shame, and unto the shame of thy mother’s nakedness?}
\twoverse{\arabic{verse}}{כִּ֣י כׇל־הַיָּמִ֗ים אֲשֶׁ֤ר בֶּן־יִשַׁי֙ חַ֣י עַל־הָאֲדָמָ֔ה לֹ֥א תִכּ֖וֹן אַתָּ֣ה וּמַלְכוּתֶ֑ךָ וְעַתָּ֗ה שְׁלַ֨ח וְקַ֤ח אֹתוֹ֙ אֵלַ֔י כִּ֥י בֶן־מָ֖וֶת הֽוּא׃ \setuma }{For as long as the son of Jesse liveth upon the earth, thou shalt not be established, nor thy kingdom. Wherefore now send and fetch him unto me, for he deserveth to die.’}
\twoverse{\arabic{verse}}{וַיַּ֙עַן֙ יְה֣וֹנָתָ֔ן אֶת־שָׁא֖וּל אָבִ֑יו וַיֹּ֧אמֶר אֵלָ֛יו לָ֥מָּה יוּמַ֖ת מֶ֥ה עָשָֽׂה׃}{And Jonathan answered Saul his father, and said unto him: ‘Wherefore should he be put to death? what hath he done?’}
\twoverse{\arabic{verse}}{וַיָּ֨טֶל שָׁא֧וּל אֶֽת־הַחֲנִ֛ית עָלָ֖יו לְהַכֹּת֑וֹ וַיֵּ֙דַע֙ יְה֣וֹנָתָ֔ן כִּי־כָ֥לָה הִ֛יא מֵעִ֥ם אָבִ֖יו לְהָמִ֥ית אֶת־דָּוִֽד׃ \setuma }{And Saul cast his spear at him to smite him; whereby Jonathan knew that it was determined of his father to put David to death.}
\twoverse{\arabic{verse}}{וַיָּ֧קׇם יְהֽוֹנָתָ֛ן מֵעִ֥ם הַשֻּׁלְחָ֖ן בׇּחֳרִי־אָ֑ף וְלֹֽא־אָכַ֞ל בְּיוֹם־הַחֹ֤דֶשׁ הַשֵּׁנִי֙ לֶ֔חֶם כִּ֤י נֶעְצַב֙ אֶל־דָּוִ֔ד כִּ֥י הִכְלִמ֖וֹ אָבִֽיו׃ \setuma }{So Jonathan arose from the table in fierce anger, and did eat no food the second day of the month; for he was grieved for David, and because his father had put him to shame.}
\twoverse{\arabic{verse}}{וַיְהִ֣י בַבֹּ֔קֶר וַיֵּצֵ֧א יְהוֹנָתָ֛ן הַשָּׂדֶ֖ה לְמוֹעֵ֣ד דָּוִ֑ד וְנַ֥עַר קָטֹ֖ן עִמּֽוֹ׃}{And it came to pass in the morning, that Jonathan went out into the field at the time appointed with David, and a little lad with him.}
\twoverse{\arabic{verse}}{וַיֹּ֣אמֶר לְנַֽעֲר֔וֹ רֻ֗ץ מְצָ֥א נָא֙ אֶת־הַ֣חִצִּ֔ים אֲשֶׁ֥ר אָנֹכִ֖י מוֹרֶ֑ה הַנַּ֣עַר רָ֔ץ וְהוּא־יָרָ֥ה הַחֵ֖צִי לְהַעֲבִרֽוֹ׃}{And he said unto his lad: ‘Run, find now the arrows which I shoot.’ And as the lad ran, he shot an arrow beyond him.}
\twoverse{\arabic{verse}}{וַיָּבֹ֤א הַנַּ֙עַר֙ עַד־מְק֣וֹם הַחֵ֔צִי אֲשֶׁ֥ר יָרָ֖ה יְהוֹנָתָ֑ן וַיִּקְרָ֨א יְהוֹנָתָ֜ן אַחֲרֵ֤י הַנַּ֙עַר֙ וַיֹּ֔אמֶר הֲל֥וֹא הַחֵ֖צִי מִמְּךָ֥ וָהָֽלְאָה׃}{And when the lad was come to the place of the arrow which Jonathan had shot, Jonathan cried after the lad, and said: ‘Is not the arrow beyond thee?’}
\twoverse{\arabic{verse}}{וַיִּקְרָ֤א יְהֽוֹנָתָן֙ אַחֲרֵ֣י הַנַּ֔עַר מְהֵרָ֥ה ח֖וּשָׁה אַֽל־תַּעֲמֹ֑ד וַיְלַקֵּ֞ט נַ֤עַר יְהֽוֹנָתָן֙ אֶת־\qk{הַ֣חִצִּ֔ים}{החצי} וַיָּבֹ֖א אֶל־אֲדֹנָֽיו׃}{And Jonathan cried after the lad: ‘Make speed, hasten, stay not.’ And Jonathan’s lad gathered up the arrows, and came to his master.}
\twoverse{\arabic{verse}}{וְהַנַּ֖עַר לֹא־יָדַ֣ע מְא֑וּמָה אַ֤ךְ יְהֽוֹנָתָן֙ וְדָוִ֔ד יָדְע֖וּ אֶת־הַדָּבָֽר׃ \setuma }{But the lad knew not any thing; only Jonathan and David knew the matter.}
\twoverse{\arabic{verse}}{וַיִּתֵּ֤ן יְהֽוֹנָתָן֙ אֶת־כֵּלָ֔יו אֶל־הַנַּ֖עַר אֲשֶׁר־ל֑וֹ וַיֹּ֣אמֶר ל֔וֹ לֵ֖ךְ הָבֵ֥יא הָעִֽיר׃}{And Jonathan gave his weapons unto his lad, and said unto him: ‘Go, carry them to the city.’}
\twoverse{\arabic{verse}}{הַנַּ֘עַר֮ בָּא֒ וְדָוִ֗ד קָ֚ם מֵאֵ֣צֶל הַנֶּ֔גֶב וַיִּפֹּ֨ל לְאַפָּ֥יו אַ֛רְצָה וַיִּשְׁתַּ֖חוּ שָׁלֹ֣שׁ פְּעָמִ֑ים וַֽיִּשְּׁק֣וּ \legarmeh  אִ֣ישׁ אֶת־רֵעֵ֗הוּ וַיִּבְכּוּ֙ אִ֣ישׁ אֶת־רֵעֵ֔הוּ עַד־דָּוִ֖ד הִגְדִּֽיל׃}{And as soon as the lad was gone, David arose out of a place toward the South, and fell on his face to the ground, and bowed down three times; and they kissed one another, and wept one with another, until David exceeded.}
\twoverse{\arabic{verse}}{וַיֹּ֧אמֶר יְהוֹנָתָ֛ן לְדָוִ֖ד לֵ֣ךְ לְשָׁל֑וֹם אֲשֶׁר֩ נִשְׁבַּ֨עְנוּ שְׁנֵ֜ינוּ אֲנַ֗חְנוּ בְּשֵׁ֤ם יְהֹוָה֙ לֵאמֹ֔ר יְהֹוָ֞ה יִהְיֶ֣ה \legarmeh  בֵּינִ֣י וּבֵינֶ֗ךָ וּבֵ֥ין זַרְעִ֛י וּבֵ֥ין זַרְעֲךָ֖ עַד־עוֹלָֽם׃ \petucha}{And Jonathan said to David: ‘Go in peace, forasmuch as we have sworn both of us in the name of the \lord, saying: The \lord\space shall be between me and thee, and Between my seed and thy seed, for ever.’}
\vspace{12pt}
\section*{מפטיר לשבת חנכה}
\addcontentsline{toc}{section}{מפטיר לשבת חנכה}
\renewcommand{\haftarahname}{מפטיר לשבת חנכה}
\vspace{-12pt}
\nopagebreak
\label{maft_hanukka}
\haftnote{The Maftir for Shabbat \d{H}anukka corresponds to the day of \d{H}anukka. When Shabbat occurs on the first day, read from the beginning through the portion of "the first day". On subsequent days, read only the portion corresponding to that day. On the eighth day, begin with the portion for the eighth day and continue to the end.}
\setcounter{chap}{7}
\setcounter{verse}{1}
\twoverse{\textsf{יום א׳}}
{וַיְהִ֡י בְּיוֹם֩ כַּלּ֨וֹת מֹשֶׁ֜ה לְהָקִ֣ים אֶת\maqqaf הַמִּשְׁכָּ֗ן וַיִּמְשַׁ֨ח אֹת֜וֹ וַיְקַדֵּ֤שׁ אֹתוֹ֙ וְאֶת\maqqaf כׇּל\maqqaf כֵּלָ֔יו וְאֶת\maqqaf הַמִּזְבֵּ֖חַ וְאֶת\maqqaf כׇּל\maqqaf כֵּלָ֑יו וַיִּמְשָׁחֵ֖ם וַיְקַדֵּ֥שׁ אֹתָֽם׃}
{And it came to pass on the day that Moses had made an end of setting up the tabernacle, and had anointed it and sanctified it, and all the furniture thereof, and the altar and all the vessels thereof, and had anointed them and sanctified them;}
\twoverse{\arabic{verse}}
{וַיַּקְרִ֙יבוּ֙ נְשִׂיאֵ֣י יִשְׂרָאֵ֔ל רָאשֵׁ֖י בֵּ֣ית אֲבֹתָ֑ם הֵ֚ם נְשִׂיאֵ֣י הַמַּטֹּ֔ת הֵ֥ם הָעֹמְדִ֖ים עַל\maqqaf הַפְּקֻדִֽים׃}
{that the princes of Israel, the heads of their fathers’ houses, offered—these were the princes of the tribes, these are they that were over them that were numbered.}
\twoverse{\arabic{verse}}
{וַיָּבִ֨יאוּ אֶת\maqqaf קׇרְבָּנָ֜ם לִפְנֵ֣י יְהֹוָ֗ה שֵׁשׁ\maqqaf עֶגְלֹ֥ת צָב֙ וּשְׁנֵ֣י עָשָׂ֣ר בָּקָ֔ר עֲגָלָ֛ה עַל\maqqaf שְׁנֵ֥י הַנְּשִׂאִ֖ים וְשׁ֣וֹר לְאֶחָ֑ד וַיַּקְרִ֥יבוּ אוֹתָ֖ם לִפְנֵ֥י הַמִּשְׁכָּֽן׃}
{And they brought their offering before the \lord, six covered wagons, and twelve oxen: a wagon for every two of the princes, and for each one an ox; and they presented them before the tabernacle.}
\twoverse{\arabic{verse}}
{וַיֹּ֥אמֶר יְהֹוָ֖ה אֶל\maqqaf מֹשֶׁ֥ה לֵּאמֹֽר׃}
{And the \lord\space spoke unto Moses, saying:}
\twoverse{\arabic{verse}}
{קַ֚ח מֵֽאִתָּ֔ם וְהָי֕וּ לַעֲבֹ֕ד אֶת\maqqaf עֲבֹדַ֖ת אֹ֣הֶל מוֹעֵ֑ד וְנָתַתָּ֤ה אוֹתָם֙ אֶל\maqqaf הַלְוִיִּ֔ם אִ֖ישׁ כְּפִ֥י עֲבֹדָתֽוֹ׃}
{’Take it of them, that they may be to do the service of the tent of meeting; and thou shalt give them unto the Levites, to every man according to his service.’}
\twoverse{\arabic{verse}}
{וַיִּקַּ֣ח מֹשֶׁ֔ה אֶת\maqqaf הָעֲגָלֹ֖ת וְאֶת\maqqaf הַבָּקָ֑ר וַיִּתֵּ֥ן אוֹתָ֖ם אֶל\maqqaf הַלְוִיִּֽם׃}
{And Moses took the wagons and the oxen, and gave them unto the Levites.}
\twoverse{\arabic{verse}}
{אֵ֣ת \legarmeh  שְׁתֵּ֣י הָעֲגָל֗וֹת וְאֵת֙ אַרְבַּ֣עַת הַבָּקָ֔ר נָתַ֖ן לִבְנֵ֣י גֵרְשׁ֑וֹן כְּפִ֖י עֲבֹדָתָֽם׃}
{Two wagons and four oxen he gave unto the sons of Gershon, according to their service.}
\twoverse{\arabic{verse}}
{וְאֵ֣ת \legarmeh  אַרְבַּ֣ע הָעֲגָלֹ֗ת וְאֵת֙ שְׁמֹנַ֣ת הַבָּקָ֔ר נָתַ֖ן לִבְנֵ֣י מְרָרִ֑י כְּפִי֙ עֲבֹ֣דָתָ֔ם בְּיַד֙ אִֽיתָמָ֔ר בֶּֽן\maqqaf אַהֲרֹ֖ן הַכֹּהֵֽן׃}
{And four wagons and eight oxen he gave unto the sons of Merari, according unto their service, under the hand of Ithamar the son of Aaron the priest.}
\twoverse{\arabic{verse}}
{וְלִבְנֵ֥י קְהָ֖ת לֹ֣א נָתָ֑ן כִּֽי\maqqaf עֲבֹדַ֤ת הַקֹּ֙דֶשׁ֙ עֲלֵהֶ֔ם בַּכָּתֵ֖ף יִשָּֽׂאוּ׃}
{But unto the sons of Kohath he gave none, because the service of the holy things belonged unto them: they bore them upon their shoulders.}
\twoverse{\arabic{verse}}
{וַיַּקְרִ֣יבוּ הַנְּשִׂאִ֗ים אֵ֚ת חֲנֻכַּ֣ת הַמִּזְבֵּ֔חַ בְּי֖וֹם הִמָּשַׁ֣ח אֹת֑וֹ וַיַּקְרִ֧יבוּ הַנְּשִׂיאִ֛ם אֶת\maqqaf קׇרְבָּנָ֖ם לִפְנֵ֥י הַמִּזְבֵּֽחַ׃}
{And the princes brought the dedication-offering of the altar in the day that it was anointed, even the princes brought their offering before the altar.}
\twoverse{\arabic{verse}}
{וַיֹּ֥אמֶר יְהֹוָ֖ה אֶל\maqqaf מֹשֶׁ֑ה נָשִׂ֨יא אֶחָ֜ד לַיּ֗וֹם נָשִׂ֤יא אֶחָד֙ לַיּ֔וֹם יַקְרִ֙יבוּ֙ אֶת\maqqaf קׇרְבָּנָ֔ם לַחֲנֻכַּ֖ת הַמִּזְבֵּֽחַ׃ \setuma }
{And the \lord\space said unto Moses: ‘They shall present their offering each prince on his day, for the dedication of the altar.’}
\twoverse{\arabic{verse}}
{וַיְהִ֗י הַמַּקְרִ֛יב בַּיּ֥וֹם הָרִאשׁ֖וֹן אֶת\maqqaf קׇרְבָּנ֑וֹ נַחְשׁ֥וֹן בֶּן\maqqaf עַמִּינָדָ֖ב לְמַטֵּ֥ה יְהוּדָֽה׃}
{And he that presented his offering the first day was Nahshon the son of Amminadab, of the tribe of Judah;}
\twoverse{\arabic{verse}}
{וְקׇרְבָּנ֞וֹ קַֽעֲרַת\maqqaf כֶּ֣סֶף אַחַ֗ת שְׁלֹשִׁ֣ים וּמֵאָה֮ מִשְׁקָלָהּ֒ מִזְרָ֤ק אֶחָד֙ כֶּ֔סֶף שִׁבְעִ֥ים שֶׁ֖קֶל בְּשֶׁ֣קֶל הַקֹּ֑דֶשׁ שְׁנֵיהֶ֣ם \legarmeh  מְלֵאִ֗ים סֹ֛לֶת בְּלוּלָ֥ה בַשֶּׁ֖מֶן לְמִנְחָֽה׃}
{and his offering was one silver dish, the weight thereof was a hundred and thirty shekels, one silver basin of seventy shekels, after the shekel of the sanctuary; both of them full of fine flour mingled with oil for a meal-offering;}
\twoverse{\arabic{verse}}
{כַּ֥ף אַחַ֛ת עֲשָׂרָ֥ה זָהָ֖ב מְלֵאָ֥ה קְטֹֽרֶת׃}
{one golden pan of ten shekels, full of incense;}
\twoverse{\arabic{verse}}
{פַּ֣ר אֶחָ֞ד בֶּן\maqqaf בָּקָ֗ר אַ֧יִל אֶחָ֛ד כֶּֽבֶשׂ\maqqaf אֶחָ֥ד בֶּן\maqqaf שְׁנָת֖וֹ לְעֹלָֽה׃}
{one young bullock, one ram, one he-lamb of the first year, for a burnt-offering;}
\twoverse{\arabic{verse}}
{שְׂעִיר\maqqaf עִזִּ֥ים אֶחָ֖ד לְחַטָּֽאת׃}
{one male of the goats for a sin-offering;}
\twoverse{\arabic{verse}}
{וּלְזֶ֣בַח הַשְּׁלָמִים֮ בָּקָ֣ר שְׁנַ֒יִם֒ אֵילִ֤ם חֲמִשָּׁה֙ עַתּוּדִ֣ים חֲמִשָּׁ֔ה כְּבָשִׂ֥ים בְּנֵֽי\maqqaf שָׁנָ֖ה חֲמִשָּׁ֑ה זֶ֛ה קׇרְבַּ֥ן נַחְשׁ֖וֹן בֶּן\maqqaf עַמִּינָדָֽב׃ \petucha }
{and for the sacrifice of peace-offerings, two oxen, five rams, five he-goats, five he-lambs of the first year. This was the offering of Nahshon the son of Amminadab.}
\twoverse{\textsf{יום ב׳}}
{בַּיּוֹם֙ הַשֵּׁנִ֔י הִקְרִ֖יב נְתַנְאֵ֣ל בֶּן\maqqaf צוּעָ֑ר נְשִׂ֖יא יִשָּׂשכָֽר׃}
{On the second day Nethanel the son of Zuar, prince of Issachar, did offer:}
\twoverse{\arabic{verse}}
{הִקְרִ֨ב אֶת\maqqaf קׇרְבָּנ֜וֹ קַֽעֲרַת\maqqaf כֶּ֣סֶף אַחַ֗ת שְׁלֹשִׁ֣ים וּמֵאָה֮ מִשְׁקָלָהּ֒ מִזְרָ֤ק אֶחָד֙ כֶּ֔סֶף שִׁבְעִ֥ים שֶׁ֖קֶל בְּשֶׁ֣קֶל הַקֹּ֑דֶשׁ שְׁנֵיהֶ֣ם \legarmeh  מְלֵאִ֗ים סֹ֛לֶת בְּלוּלָ֥ה בַשֶּׁ֖מֶן לְמִנְחָֽה׃}
{he presented for his offering one silver dish, the weight thereof was a hundred and thirty shekels, one silver basin of seventy shekels, after the shekel of the sanctuary; both of them full of fine flour mingled with oil for a meal-offering;}
\twoverse{\arabic{verse}}
{כַּ֥ף אַחַ֛ת עֲשָׂרָ֥ה זָהָ֖ב מְלֵאָ֥ה קְטֹֽרֶת׃}
{one golden pan of ten shekels, full of incense;}
\twoverse{\arabic{verse}}
{פַּ֣ר אֶחָ֞ד בֶּן\maqqaf בָּקָ֗ר אַ֧יִל אֶחָ֛ד כֶּֽבֶשׂ\maqqaf אֶחָ֥ד בֶּן\maqqaf שְׁנָת֖וֹ לְעֹלָֽה׃}
{one young bullock, one ram, one he-lamb of the first year, for a burnt-offering;}
\twoverse{\arabic{verse}}
{שְׂעִיר\maqqaf עִזִּ֥ים אֶחָ֖ד לְחַטָּֽאת׃}
{one male of the goats for a sin-offering;}
\twoverse{\arabic{verse}}
{וּלְזֶ֣בַח הַשְּׁלָמִים֮ בָּקָ֣ר שְׁנַ֒יִם֒ אֵילִ֤ם חֲמִשָּׁה֙ עַתֻּדִ֣ים חֲמִשָּׁ֔ה כְּבָשִׂ֥ים בְּנֵי\maqqaf שָׁנָ֖ה חֲמִשָּׁ֑ה זֶ֛ה קׇרְבַּ֥ן נְתַנְאֵ֖ל בֶּן\maqqaf צוּעָֽר׃ \petucha }
{and for the sacrifice of peace-offerings, two oxen, five rams, five he-goats, five he-lambs of the first year. This was the offering of Nethanel the son of Zuar.}
\twoverse{\textsf{יום ג׳}}
{בַּיּוֹם֙ הַשְּׁלִישִׁ֔י נָשִׂ֖יא לִבְנֵ֣י זְבוּלֻ֑ן אֱלִיאָ֖ב בֶּן\maqqaf חֵלֹֽן׃}
{On the third day Eliab the son of Helon, prince of the children of Zebulun:}
\twoverse{\arabic{verse}}
{קׇרְבָּנ֞וֹ קַֽעֲרַת\maqqaf כֶּ֣סֶף אַחַ֗ת שְׁלֹשִׁ֣ים וּמֵאָה֮ מִשְׁקָלָהּ֒ מִזְרָ֤ק אֶחָד֙ כֶּ֔סֶף שִׁבְעִ֥ים שֶׁ֖קֶל בְּשֶׁ֣קֶל הַקֹּ֑דֶשׁ שְׁנֵיהֶ֣ם \legarmeh  מְלֵאִ֗ים סֹ֛לֶת בְּלוּלָ֥ה בַשֶּׁ֖מֶן לְמִנְחָֽה׃}
{his offering was one silver dish, the weight thereof was a hundred and thirty shekels, one silver basin of seventy shekels, after the shekel of the sanctuary; both of them full of fine flour mingled with oil for a meal-offering;}
\twoverse{\arabic{verse}}
{כַּ֥ף אַחַ֛ת עֲשָׂרָ֥ה זָהָ֖ב מְלֵאָ֥ה קְטֹֽרֶת׃}
{one golden pan of ten shekels, full of incense;}
\twoverse{\arabic{verse}}
{פַּ֣ר אֶחָ֞ד בֶּן\maqqaf בָּקָ֗ר אַ֧יִל אֶחָ֛ד כֶּֽבֶשׂ\maqqaf אֶחָ֥ד בֶּן\maqqaf שְׁנָת֖וֹ לְעֹלָֽה׃}
{one young bullock, one ram, one he-lamb of the first year, for a burnt-offering;}
\twoverse{\arabic{verse}}
{שְׂעִיר\maqqaf עִזִּ֥ים אֶחָ֖ד לְחַטָּֽאת׃}
{one male of the goats for a sin-offering;}
\twoverse{\arabic{verse}}
{וּלְזֶ֣בַח הַשְּׁלָמִים֮ בָּקָ֣ר שְׁנַ֒יִם֒ אֵילִ֤ם חֲמִשָּׁה֙ עַתֻּדִ֣ים חֲמִשָּׁ֔ה כְּבָשִׂ֥ים בְּנֵי\maqqaf שָׁנָ֖ה חֲמִשָּׁ֑ה זֶ֛ה קׇרְבַּ֥ן אֱלִיאָ֖ב בֶּן\maqqaf חֵלֹֽן׃ \petucha }
{and for the sacrifice of peace-offerings, two oxen, five rams, five he-goats, five he-lambs of the first year. This was the offering of Eliab the son of Helon.}
\twoverse{\textsf{יום ד׳}}
{בַּיּוֹם֙ הָרְבִיעִ֔י נָשִׂ֖יא לִבְנֵ֣י רְאוּבֵ֑ן אֱלִיצ֖וּר בֶּן\maqqaf שְׁדֵיאֽוּר׃}
{On the fourth day Elizur the son of Shedeur, prince of the children of Reuben:}
\twoverse{\arabic{verse}}
{קׇרְבָּנ֞וֹ קַֽעֲרַת\maqqaf כֶּ֣סֶף אַחַ֗ת שְׁלֹשִׁ֣ים וּמֵאָה֮ מִשְׁקָלָהּ֒ מִזְרָ֤ק אֶחָד֙ כֶּ֔סֶף שִׁבְעִ֥ים שֶׁ֖קֶל בְּשֶׁ֣קֶל הַקֹּ֑דֶשׁ שְׁנֵיהֶ֣ם \legarmeh  מְלֵאִ֗ים סֹ֛לֶת בְּלוּלָ֥ה בַשֶּׁ֖מֶן לְמִנְחָֽה׃}
{his offering was one silver dish, the weight thereof was a hundred and thirty shekels, one silver basin of seventy shekels, after the shekel of the sanctuary; both of them full of fine flour mingled with oil for a meal-offering;}
\twoverse{\arabic{verse}}
{כַּ֥ף אַחַ֛ת עֲשָׂרָ֥ה זָהָ֖ב מְלֵאָ֥ה קְטֹֽרֶת׃}
{one golden pan of ten shekels, full of incense;}
\twoverse{\arabic{verse}}
{פַּ֣ר אֶחָ֞ד בֶּן\maqqaf בָּקָ֗ר אַ֧יִל אֶחָ֛ד כֶּֽבֶשׂ\maqqaf אֶחָ֥ד בֶּן\maqqaf שְׁנָת֖וֹ לְעֹלָֽה׃}
{one young bullock, one ram, one he-lamb of the first year, for a burnt-offering;}
\twoverse{\arabic{verse}}
{שְׂעִיר\maqqaf עִזִּ֥ים אֶחָ֖ד לְחַטָּֽאת׃}
{one male of the goats for a sin-offering;}
\twoverse{\arabic{verse}}
{וּלְזֶ֣בַח הַשְּׁלָמִים֮ בָּקָ֣ר שְׁנַ֒יִם֒ אֵילִ֤ם חֲמִשָּׁה֙ עַתֻּדִ֣ים חֲמִשָּׁ֔ה כְּבָשִׂ֥ים בְּנֵֽי\maqqaf שָׁנָ֖ה חֲמִשָּׁ֑ה זֶ֛ה קׇרְבַּ֥ן אֱלִיצ֖וּר בֶּן\maqqaf שְׁדֵיאֽוּר׃ \petucha }
{and for the sacrifice of peace-offerings, two oxen, five rams, five he-goats, five he-lambs of the first year. This was the offering of Elizur the son of Shedeur.}
\twoverse{\textsf{יום ה׳}}
{בַּיּוֹם֙ הַחֲמִישִׁ֔י נָשִׂ֖יא לִבְנֵ֣י שִׁמְע֑וֹן שְׁלֻֽמִיאֵ֖ל בֶּן\maqqaf צוּרִֽישַׁדָּֽי׃}
{On the fifth day Shelumiel the son of Zurishaddai, prince of the children of Simeon:}
\twoverse{\arabic{verse}}
{קׇרְבָּנ֞וֹ קַֽעֲרַת\maqqaf כֶּ֣סֶף אַחַ֗ת שְׁלֹשִׁ֣ים וּמֵאָה֮ מִשְׁקָלָהּ֒ מִזְרָ֤ק אֶחָד֙ כֶּ֔סֶף שִׁבְעִ֥ים שֶׁ֖קֶל בְּשֶׁ֣קֶל הַקֹּ֑דֶשׁ שְׁנֵיהֶ֣ם \legarmeh  מְלֵאִ֗ים סֹ֛לֶת בְּלוּלָ֥ה בַשֶּׁ֖מֶן לְמִנְחָֽה׃}
{his offering was one silver dish, the weight thereof was a hundred and thirty shekels, one silver basin of seventy shekels, after the shekel of the sanctuary; both of them full of fine flour mingled with oil for a meal-offering;}
\twoverse{\arabic{verse}}
{כַּ֥ף אַחַ֛ת עֲשָׂרָ֥ה זָהָ֖ב מְלֵאָ֥ה קְטֹֽרֶת׃}
{one golden pan of ten shekels, full of incense;}
\twoverse{\arabic{verse}}
{פַּ֣ר אֶחָ֞ד בֶּן\maqqaf בָּקָ֗ר אַ֧יִל אֶחָ֛ד כֶּֽבֶשׂ\maqqaf אֶחָ֥ד בֶּן\maqqaf שְׁנָת֖וֹ לְעֹלָֽה׃}
{one young bullock, one ram, one he-lamb of the first year, for a burnt-offering;}
\twoverse{\arabic{verse}}
{שְׂעִיר\maqqaf עִזִּ֥ים אֶחָ֖ד לְחַטָּֽאת׃}
{one male of the goats for a sin-offering;}
\twoverse{\arabic{verse}}
{וּלְזֶ֣בַח הַשְּׁלָמִים֮ בָּקָ֣ר שְׁנַ֒יִם֒ אֵילִ֤ם חֲמִשָּׁה֙ עַתֻּדִ֣ים חֲמִשָּׁ֔ה כְּבָשִׂ֥ים בְּנֵי\maqqaf שָׁנָ֖ה חֲמִשָּׁ֑ה זֶ֛ה קׇרְבַּ֥ן שְׁלֻמִיאֵ֖ל בֶּן\maqqaf צוּרִֽישַׁדָּֽי׃ \petucha }
{and for the sacrifice of peace-offerings, two oxen, five rams, five he-goats, five he-lambs of the first year. This was the offering of Shelumiel the son of Zurishaddai.}
\twoverse{\textsf{יום ו׳}}
{בַּיּוֹם֙ הַשִּׁשִּׁ֔י נָשִׂ֖יא לִבְנֵ֣י גָ֑ד אֶלְיָסָ֖ף בֶּן\maqqaf דְּעוּאֵֽל׃}
{On the sixth day Eliasaph the son of Deuel, prince of the children of Gad:}
\twoverse{\arabic{verse}}
{קׇרְבָּנ֞וֹ קַֽעֲרַת\maqqaf כֶּ֣סֶף אַחַ֗ת שְׁלֹשִׁ֣ים וּמֵאָה֮ מִשְׁקָלָהּ֒ מִזְרָ֤ק אֶחָד֙ כֶּ֔סֶף שִׁבְעִ֥ים שֶׁ֖קֶל בְּשֶׁ֣קֶל הַקֹּ֑דֶשׁ שְׁנֵיהֶ֣ם \legarmeh  מְלֵאִ֗ים סֹ֛לֶת בְּלוּלָ֥ה בַשֶּׁ֖מֶן לְמִנְחָֽה׃}
{his offering was one silver dish, the weight thereof was a hundred and thirty shekels, one silver basin of seventy shekels, after the shekel of the sanctuary; both of them full of fine flour mingled with oil for a meal-offering;}
\twoverse{\arabic{verse}}
{כַּ֥ף אַחַ֛ת עֲשָׂרָ֥ה זָהָ֖ב מְלֵאָ֥ה קְטֹֽרֶת׃}
{one golden pan of ten shekels, full of incense;}
\twoverse{\arabic{verse}}
{פַּ֣ר אֶחָ֞ד בֶּן\maqqaf בָּקָ֗ר אַ֧יִל אֶחָ֛ד כֶּֽבֶשׂ\maqqaf אֶחָ֥ד בֶּן\maqqaf שְׁנָת֖וֹ לְעֹלָֽה׃}
{one young bullock, one ram, one he-lamb of the first year, for a burnt-offering;}
\twoverse{\arabic{verse}}
{שְׂעִיר\maqqaf עִזִּ֥ים אֶחָ֖ד לְחַטָּֽאת׃}
{one male of the goats for a sin-offering;}
\twoverse{\arabic{verse}}
{וּלְזֶ֣בַח הַשְּׁלָמִים֮ בָּקָ֣ר שְׁנַ֒יִם֒ אֵילִ֤ם חֲמִשָּׁה֙ עַתֻּדִ֣ים חֲמִשָּׁ֔ה כְּבָשִׂ֥ים בְּנֵי\maqqaf שָׁנָ֖ה חֲמִשָּׁ֑ה זֶ֛ה קׇרְבַּ֥ן אֶלְיָסָ֖ף בֶּן\maqqaf דְּעוּאֵֽל׃ \petucha }
{and for the sacrifice of peace-offerings, two oxen, five rams, five he-goats, five he-lambs of the first year. This was the offering of Eliasaph the son of Deuel.}
\twoverse{\textsf{יום ז׳}}
{בַּיּוֹם֙ הַשְּׁבִיעִ֔י נָשִׂ֖יא לִבְנֵ֣י אֶפְרָ֑יִם אֱלִֽישָׁמָ֖ע בֶּן\maqqaf עַמִּיהֽוּד׃}
{On the seventh day Elishama the son of Ammihud, prince of the children of Ephraim:}
\twoverse{\arabic{verse}}
{קׇרְבָּנ֞וֹ קַֽעֲרַת\maqqaf כֶּ֣סֶף אַחַ֗ת שְׁלֹשִׁ֣ים וּמֵאָה֮ מִשְׁקָלָהּ֒ מִזְרָ֤ק אֶחָד֙ כֶּ֔סֶף שִׁבְעִ֥ים שֶׁ֖קֶל בְּשֶׁ֣קֶל הַקֹּ֑דֶשׁ שְׁנֵיהֶ֣ם \legarmeh  מְלֵאִ֗ים סֹ֛לֶת בְּלוּלָ֥ה בַשֶּׁ֖מֶן לְמִנְחָֽה׃}
{his offering was one silver dish, the weight thereof was a hundred and thirty shekels, one silver basin of seventy shekels, after the shekel of the sanctuary; both of them full of fine flour mingled with oil for a meal-offering;}
\twoverse{\arabic{verse}}
{כַּ֥ף אַחַ֛ת עֲשָׂרָ֥ה זָהָ֖ב מְלֵאָ֥ה קְטֹֽרֶת׃}
{one golden pan of ten shekels, full of incense;}
\twoverse{\arabic{verse}}
{פַּ֣ר אֶחָ֞ד בֶּן\maqqaf בָּקָ֗ר אַ֧יִל אֶחָ֛ד כֶּֽבֶשׂ\maqqaf אֶחָ֥ד בֶּן\maqqaf שְׁנָת֖וֹ לְעֹלָֽה׃}
{one young bullock, one ram, one he-lamb of the first year, for a burnt-offering;}
\twoverse{\arabic{verse}}
{שְׂעִיר\maqqaf עִזִּ֥ים אֶחָ֖ד לְחַטָּֽאת׃}
{one male of the goats for a sin-offering;}
\twoverse{\arabic{verse}}
{וּלְזֶ֣בַח הַשְּׁלָמִים֮ בָּקָ֣ר שְׁנַ֒יִם֒ אֵילִ֤ם חֲמִשָּׁה֙ עַתֻּדִ֣ים חֲמִשָּׁ֔ה כְּבָשִׂ֥ים בְּנֵֽי\maqqaf שָׁנָ֖ה חֲמִשָּׁ֑ה זֶ֛ה קׇרְבַּ֥ן אֱלִישָׁמָ֖ע בֶּן\maqqaf עַמִּיהֽוּד׃ \petucha }
{and for the sacrifice of peace-offerings, two oxen, five rams, five he-goats, five he-lambs of the first year. This was the offering of Elishama the son of Ammihud.}
\twoverse{\textsf{יום ח׳}}
{בַּיּוֹם֙ הַשְּׁמִינִ֔י נָשִׂ֖יא לִבְנֵ֣י מְנַשֶּׁ֑ה גַּמְלִיאֵ֖ל בֶּן\maqqaf פְּדָהצֽוּר׃}
{On the eighth day Gamaliel the son of Pedahzur, prince of the children of Manasseh:}
\twoverse{\arabic{verse}}
{קׇרְבָּנ֞וֹ קַֽעֲרַת\maqqaf כֶּ֣סֶף אַחַ֗ת שְׁלֹשִׁ֣ים וּמֵאָה֮ מִשְׁקָלָהּ֒ מִזְרָ֤ק אֶחָד֙ כֶּ֔סֶף שִׁבְעִ֥ים שֶׁ֖קֶל בְּשֶׁ֣קֶל הַקֹּ֑דֶשׁ שְׁנֵיהֶ֣ם \legarmeh  מְלֵאִ֗ים סֹ֛לֶת בְּלוּלָ֥ה בַשֶּׁ֖מֶן לְמִנְחָֽה׃}
{his offering was one silver dish, the weight thereof was a hundred and thirty shekels, one silver basin of seventy shekels, after the shekel of the sanctuary; both of them full of fine flour mingled with oil for a meal-offering;}
\twoverse{\arabic{verse}}
{כַּ֥ף אַחַ֛ת עֲשָׂרָ֥ה זָהָ֖ב מְלֵאָ֥ה קְטֹֽרֶת׃}
{one golden pan of ten shekels, full of incense;}
\twoverse{\arabic{verse}}
{פַּ֣ר אֶחָ֞ד בֶּן\maqqaf בָּקָ֗ר אַ֧יִל אֶחָ֛ד כֶּֽבֶשׂ\maqqaf אֶחָ֥ד בֶּן\maqqaf שְׁנָת֖וֹ לְעֹלָֽה׃}
{one young bullock, one ram, one he-lamb of the first year, for a burnt-offering;}
\twoverse{\arabic{verse}}
{שְׂעִיר\maqqaf עִזִּ֥ים אֶחָ֖ד לְחַטָּֽאת׃}
{one male of the goats for a sin-offering;}
\twoverse{\arabic{verse}}
{וּלְזֶ֣בַח הַשְּׁלָמִים֮ בָּקָ֣ר שְׁנַ֒יִם֒ אֵילִ֤ם חֲמִשָּׁה֙ עַתֻּדִ֣ים חֲמִשָּׁ֔ה כְּבָשִׂ֥ים בְּנֵי\maqqaf שָׁנָ֖ה חֲמִשָּׁ֑ה זֶ֛ה קׇרְבַּ֥ן גַּמְלִיאֵ֖ל בֶּן\maqqaf פְּדָהצֽוּר׃ \petucha }
{and for the sacrifice of peace-offerings, two oxen, five rams, five he-goats, five he-lamb of the first year. This was the offering of Gamaliel the son of Pedahzur.}
\twoverse{\arabic{verse}}
{בַּיּוֹם֙ הַתְּשִׁיעִ֔י נָשִׂ֖יא לִבְנֵ֣י בִנְיָמִ֑ן אֲבִידָ֖ן בֶּן\maqqaf גִּדְעֹנִֽי׃}
{On the ninth day Abidan the son of Gideoni, prince of the children of Benjamin:}
\twoverse{\arabic{verse}}
{קׇרְבָּנ֞וֹ קַֽעֲרַת\maqqaf כֶּ֣סֶף אַחַ֗ת שְׁלֹשִׁ֣ים וּמֵאָה֮ מִשְׁקָלָהּ֒ מִזְרָ֤ק אֶחָד֙ כֶּ֔סֶף שִׁבְעִ֥ים שֶׁ֖קֶל בְּשֶׁ֣קֶל הַקֹּ֑דֶשׁ שְׁנֵיהֶ֣ם \legarmeh  מְלֵאִ֗ים סֹ֛לֶת בְּלוּלָ֥ה בַשֶּׁ֖מֶן לְמִנְחָֽה׃}
{his offering was one silver dish, the weight thereof was a hundred and thirty shekels, one silver basin of seventy shekels, after the shekel of the sanctuary; both of them full of fine flour mingled with oil for a meal-offering;}
\twoverse{\arabic{verse}}
{כַּ֥ף אַחַ֛ת עֲשָׂרָ֥ה זָהָ֖ב מְלֵאָ֥ה קְטֹֽרֶת׃}
{one golden pan of ten shekels, full of incense;}
\twoverse{\arabic{verse}}
{פַּ֣ר אֶחָ֞ד בֶּן\maqqaf בָּקָ֗ר אַ֧יִל אֶחָ֛ד כֶּֽבֶשׂ\maqqaf אֶחָ֥ד בֶּן\maqqaf שְׁנָת֖וֹ לְעֹלָֽה׃}
{one young bullock, one ram, one he-lamb of the first year, for a burnt-offering;}
\twoverse{\arabic{verse}}
{שְׂעִיר\maqqaf עִזִּ֥ים אֶחָ֖ד לְחַטָּֽאת׃}
{one male of the goats for a sin-offering;}
\twoverse{\arabic{verse}}
{וּלְזֶ֣בַח הַשְּׁלָמִים֮ בָּקָ֣ר שְׁנַ֒יִם֒ אֵילִ֤ם חֲמִשָּׁה֙ עַתֻּדִ֣ים חֲמִשָּׁ֔ה כְּבָשִׂ֥ים בְּנֵי\maqqaf שָׁנָ֖ה חֲמִשָּׁ֑ה זֶ֛ה קׇרְבַּ֥ן אֲבִידָ֖ן בֶּן\maqqaf גִּדְעֹנִֽי׃ \petucha }
{and for the sacrifice of peace-offerings, two oxen, five rams, five he-goats, five he-lambs of the first year. This was the offering of Abidan the son of Gideoni.}
\twoverse{\arabic{verse}}
{בַּיּוֹם֙ הָעֲשִׂירִ֔י נָשִׂ֖יא לִבְנֵ֣י דָ֑ן אֲחִיעֶ֖זֶר בֶּן\maqqaf עַמִּישַׁדָּֽי׃}
{On the tenth day Ahiezer the son of Ammishaddai, prince of the children of Dan:}
\twoverse{\arabic{verse}}
{קׇרְבָּנ֞וֹ קַֽעֲרַת\maqqaf כֶּ֣סֶף אַחַ֗ת שְׁלֹשִׁ֣ים וּמֵאָה֮ מִשְׁקָלָהּ֒ מִזְרָ֤ק אֶחָד֙ כֶּ֔סֶף שִׁבְעִ֥ים שֶׁ֖קֶל בְּשֶׁ֣קֶל הַקֹּ֑דֶשׁ שְׁנֵיהֶ֣ם \legarmeh  מְלֵאִ֗ים סֹ֛לֶת בְּלוּלָ֥ה בַשֶּׁ֖מֶן לְמִנְחָֽה׃}
{his offering was one silver dish, the weight thereof was a hundred and thirty shekels, one silver basin of seventy shekels, after the shekel of the sanctuary; both of them full of fine flour mingled with oil for a meal-offering;}
\twoverse{\arabic{verse}}
{כַּ֥ף אַחַ֛ת עֲשָׂרָ֥ה זָהָ֖ב מְלֵאָ֥ה קְטֹֽרֶת׃}
{one golden pan of ten shekels, full of incense;}
\twoverse{\arabic{verse}}
{פַּ֣ר אֶחָ֞ד בֶּן\maqqaf בָּקָ֗ר אַ֧יִל אֶחָ֛ד כֶּֽבֶשׂ\maqqaf אֶחָ֥ד בֶּן\maqqaf שְׁנָת֖וֹ לְעֹלָֽה׃}
{one young bullock, one ram, one he-lamb of the first year, for a burnt-offering;}
\twoverse{\arabic{verse}}
{שְׂעִיר\maqqaf עִזִּ֥ים אֶחָ֖ד לְחַטָּֽאת׃}
{one male of the goats for a sin-offering;}
\twoverse{\arabic{verse}}
{וּלְזֶ֣בַח הַשְּׁלָמִים֮ בָּקָ֣ר שְׁנַ֒יִם֒ אֵילִ֤ם חֲמִשָּׁה֙ עַתֻּדִ֣ים חֲמִשָּׁ֔ה כְּבָשִׂ֥ים בְּנֵֽי\maqqaf שָׁנָ֖ה חֲמִשָּׁ֑ה זֶ֛ה קׇרְבַּ֥ן אֲחִיעֶ֖זֶר בֶּן\maqqaf עַמִּישַׁדָּֽי׃ \petucha }
{and for the sacrifice of peace-offerings, two oxen, five rams, five he-goats, five he-lambs of the first year. This was the offering of Ahiezer the son of Ammishaddai.}
\twoverse{\arabic{verse}}
{בְּיוֹם֙ עַשְׁתֵּ֣י עָשָׂ֣ר י֔וֹם נָשִׂ֖יא לִבְנֵ֣י אָשֵׁ֑ר פַּגְעִיאֵ֖ל בֶּן\maqqaf עׇכְרָֽן׃}
{On the eleventh day Pagiel the son of Ochran, prince of the children of Asher:}
\twoverse{\arabic{verse}}
{קׇרְבָּנ֞וֹ קַֽעֲרַת\maqqaf כֶּ֣סֶף אַחַ֗ת שְׁלֹשִׁ֣ים וּמֵאָה֮ מִשְׁקָלָהּ֒ מִזְרָ֤ק אֶחָד֙ כֶּ֔סֶף שִׁבְעִ֥ים שֶׁ֖קֶל בְּשֶׁ֣קֶל הַקֹּ֑דֶשׁ שְׁנֵיהֶ֣ם \legarmeh  מְלֵאִ֗ים סֹ֛לֶת בְּלוּלָ֥ה בַשֶּׁ֖מֶן לְמִנְחָֽה׃}
{his offering was one silver dish, the weight thereof was a hundred and thirty shekels, one silver basin of seventy shekels, after the shekel of the sanctuary; both of them full of fine flour mingled with oil for a meal-offering;}
\twoverse{\arabic{verse}}
{כַּ֥ף אַחַ֛ת עֲשָׂרָ֥ה זָהָ֖ב מְלֵאָ֥ה קְטֹֽרֶת׃}
{one golden pan of ten shekels, full of incense;}
\twoverse{\arabic{verse}}
{פַּ֣ר אֶחָ֞ד בֶּן\maqqaf בָּקָ֗ר אַ֧יִל אֶחָ֛ד כֶּֽבֶשׂ\maqqaf אֶחָ֥ד בֶּן\maqqaf שְׁנָת֖וֹ לְעֹלָֽה׃}
{one young bullock, one ram, one he-lamb of the first year, for a burnt-offering;}
\twoverse{\arabic{verse}}
{שְׂעִיר\maqqaf עִזִּ֥ים אֶחָ֖ד לְחַטָּֽאת׃}
{one male of the goats for a sin-offering;}
\twoverse{\arabic{verse}}
{וּלְזֶ֣בַח הַשְּׁלָמִים֮ בָּקָ֣ר שְׁנַ֒יִם֒ אֵילִ֤ם חֲמִשָּׁה֙ עַתֻּדִ֣ים חֲמִשָּׁ֔ה כְּבָשִׂ֥ים בְּנֵֽי\maqqaf שָׁנָ֖ה חֲמִשָּׁ֑ה זֶ֛ה קׇרְבַּ֥ן פַּגְעִיאֵ֖ל בֶּן\maqqaf עׇכְרָֽן׃ \petucha }
{and for the sacrifice of peace-offerings, two oxen, five rams, five he-goats, five he-lambs of the first year. This was the offering of Pagiel the son of Ochran.}
\twoverse{\arabic{verse}}
{בְּיוֹם֙ שְׁנֵ֣ים עָשָׂ֣ר י֔וֹם נָשִׂ֖יא לִבְנֵ֣י נַפְתָּלִ֑י אֲחִירַ֖ע בֶּן\maqqaf עֵינָֽן׃}
{On the twelfth day Ahira the son of Enan, prince of the children of Naphtali:}
\twoverse{\arabic{verse}}
{קׇרְבָּנ֞וֹ קַֽעֲרַת\maqqaf כֶּ֣סֶף אַחַ֗ת שְׁלֹשִׁ֣ים וּמֵאָה֮ מִשְׁקָלָהּ֒ מִזְרָ֤ק אֶחָד֙ כֶּ֔סֶף שִׁבְעִ֥ים שֶׁ֖קֶל בְּשֶׁ֣קֶל הַקֹּ֑דֶשׁ שְׁנֵיהֶ֣ם \legarmeh  מְלֵאִ֗ים סֹ֛לֶת בְּלוּלָ֥ה בַשֶּׁ֖מֶן לְמִנְחָֽה׃}
{his offering was one silver dish, the weight thereof was a hundred and thirty shekels, one silver basin of seventy shekels, after the shekel of the sanctuary; both of them full of fine flour mingled with oil for a meal-offering;}
\twoverse{\arabic{verse}}
{כַּ֥ף אַחַ֛ת עֲשָׂרָ֥ה זָהָ֖ב מְלֵאָ֥ה קְטֹֽרֶת׃}
{one golden pan of ten shekels, full of incense;}
\twoverse{\arabic{verse}}
{פַּ֣ר אֶחָ֞ד בֶּן\maqqaf בָּקָ֗ר אַ֧יִל אֶחָ֛ד כֶּֽבֶשׂ\maqqaf אֶחָ֥ד בֶּן\maqqaf שְׁנָת֖וֹ לְעֹלָֽה׃}
{one young bullock, one ram, one he-lamb of the first year, for a burnt-offering;}
\twoverse{\arabic{verse}}
{שְׂעִיר\maqqaf עִזִּ֥ים אֶחָ֖ד לְחַטָּֽאת׃}
{one male of the goats for a sin-offering;}
\twoverse{\arabic{verse}}
{וּלְזֶ֣בַח הַשְּׁלָמִים֮ בָּקָ֣ר שְׁנַ֒יִם֒ אֵילִ֤ם חֲמִשָּׁה֙ עַתֻּדִ֣ים חֲמִשָּׁ֔ה כְּבָשִׂ֥ים בְּנֵֽי\maqqaf שָׁנָ֖ה חֲמִשָּׁ֑ה זֶ֛ה קׇרְבַּ֥ן אֲחִירַ֖ע בֶּן\maqqaf עֵינָֽן׃ \petucha }
{and for the sacrifice of peace-offerings, two oxen, five rams, five he-goats, five he-lambs of the first year. This was the offering of Ahira the son of Enan.}
\twoverse{\arabic{verse}}
{זֹ֣את \legarmeh  חֲנֻכַּ֣ת הַמִּזְבֵּ֗חַ בְּיוֹם֙ הִמָּשַׁ֣ח אֹת֔וֹ מֵאֵ֖ת נְשִׂיאֵ֣י יִשְׂרָאֵ֑ל קַעֲרֹ֨ת כֶּ֜סֶף שְׁתֵּ֣ים עֶשְׂרֵ֗ה מִֽזְרְקֵי\maqqaf כֶ֙סֶף֙ שְׁנֵ֣ים עָשָׂ֔ר כַּפּ֥וֹת זָהָ֖ב שְׁתֵּ֥ים עֶשְׂרֵֽה׃}
{This was the dedication-offering of the altar, in the day when it was anointed, at the hands of the princes of Israel: twelve silver dishes, twelve silver basins, twelve golden pans;}
\twoverse{\arabic{verse}}
{שְׁלֹשִׁ֣ים וּמֵאָ֗ה הַקְּעָרָ֤ה הָֽאַחַת֙ כֶּ֔סֶף וְשִׁבְעִ֖ים הַמִּזְרָ֣ק הָאֶחָ֑ד כֹּ֚ל כֶּ֣סֶף הַכֵּלִ֔ים אַלְפַּ֥יִם וְאַרְבַּע\maqqaf מֵא֖וֹת בְּשֶׁ֥קֶל הַקֹּֽדֶשׁ׃}
{each silver dish weighing a hundred and thirty shekels, and each basin seventy; all the silver of the vessels two thousand and four hundred shekels, after the shekel of the sanctuary;}
\twoverse{\arabic{verse}}
{כַּפּ֨וֹת זָהָ֤ב שְׁתֵּים\maqqaf עֶשְׂרֵה֙ מְלֵאֹ֣ת קְטֹ֔רֶת עֲשָׂרָ֧ה עֲשָׂרָ֛ה הַכַּ֖ף בְּשֶׁ֣קֶל הַקֹּ֑דֶשׁ כׇּל\maqqaf זְהַ֥ב הַכַּפּ֖וֹת עֶשְׂרִ֥ים וּמֵאָֽה׃}
{twelve golden pans, full of incense, weighing ten shekels apiece, after the shekel of the sanctuary; all the gold of the pans a hundred and twenty shekels;}
\twoverse{\arabic{verse}}
{כׇּל\maqqaf הַבָּקָ֨ר לָעֹלָ֜ה שְׁנֵ֧ים עָשָׂ֣ר פָּרִ֗ים אֵילִ֤ם שְׁנֵים\maqqaf עָשָׂר֙ כְּבָשִׂ֧ים בְּנֵֽי\maqqaf שָׁנָ֛ה שְׁנֵ֥ים עָשָׂ֖ר וּמִנְחָתָ֑ם וּשְׂעִירֵ֥י עִזִּ֛ים שְׁנֵ֥ים עָשָׂ֖ר לְחַטָּֽאת׃}
{all the oxen for the burnt-offering twelve bullocks, the rams twelve, the he-lambs of the first year twelve, and their meal-offering; and the males of the goats for a sin-offering twelve;}
\twoverse{\arabic{verse}}
{וְכֹ֞ל בְּקַ֣ר \legarmeh  זֶ֣בַח הַשְּׁלָמִ֗ים עֶשְׂרִ֣ים וְאַרְבָּעָה֮ פָּרִים֒ אֵילִ֤ם שִׁשִּׁים֙ עַתֻּדִ֣ים שִׁשִּׁ֔ים כְּבָשִׂ֥ים בְּנֵי\maqqaf שָׁנָ֖ה שִׁשִּׁ֑ים זֹ֚את חֲנֻכַּ֣ת הַמִּזְבֵּ֔חַ אַחֲרֵ֖י הִמָּשַׁ֥ח אֹתֽוֹ׃}
{and all the oxen for the sacrifice of peace-offerings twenty and four bullocks, the rams sixty, the he-goats sixty, the he-lambs of the first year sixty. This was the dedication-offering of the altar, after that it was anointed.}
\twoverse{\arabic{verse}}
{וּבְבֹ֨א מֹשֶׁ֜ה אֶל\maqqaf אֹ֣הֶל מוֹעֵד֮ לְדַבֵּ֣ר אִתּוֹ֒ וַיִּשְׁמַ֨ע אֶת\maqqaf הַקּ֜וֹל מִדַּבֵּ֣ר אֵלָ֗יו מֵעַ֤ל הַכַּפֹּ֙רֶת֙ אֲשֶׁר֙ עַל\maqqaf אֲרֹ֣ן הָעֵדֻ֔ת מִבֵּ֖ין שְׁנֵ֣י הַכְּרֻבִ֑ים וַיְדַבֵּ֖ר אֵלָֽיו׃ \petucha }
{And when Moses went into the tent of meeting that He might speak with him, then he heard the Voice speaking unto him from above the ark-cover that was upon the ark of the testimony, from between the two cherubim; and He spoke unto him.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{וַיְדַבֵּ֥ר יְהֹוָ֖ה אֶל\maqqaf מֹשֶׁ֥ה לֵּאמֹֽר׃}
{And the \lord\space spoke unto Moses, saying:}
\twoverse{\arabic{verse}}
{דַּבֵּר֙ אֶֽל\maqqaf אַהֲרֹ֔ן וְאָמַרְתָּ֖ אֵלָ֑יו בְּהַעֲלֹֽתְךָ֙ אֶת\maqqaf הַנֵּרֹ֔ת אֶל\maqqaf מוּל֙ פְּנֵ֣י הַמְּנוֹרָ֔ה יָאִ֖ירוּ שִׁבְעַ֥ת הַנֵּרֽוֹת׃}
{’Speak unto Aaron, and say unto him: When thou lightest the lamps, the seven lamps shall give light in front of the candlestick.’}
\twoverse{\arabic{verse}}
{וַיַּ֤עַשׂ כֵּן֙ אַהֲרֹ֔ן אֶל\maqqaf מוּל֙ פְּנֵ֣י הַמְּנוֹרָ֔ה הֶעֱלָ֖ה נֵרֹתֶ֑יהָ כַּֽאֲשֶׁ֛ר צִוָּ֥ה יְהֹוָ֖ה אֶת\maqqaf מֹשֶֽׁה׃}
{And Aaron did so: he lighted the lamps thereof so as to give light in front of the candlestick, as the \lord\space commanded Moses.}
\twoverse{\arabic{verse}}
{וְזֶ֨ה מַעֲשֵׂ֤ה הַמְּנֹרָה֙ מִקְשָׁ֣ה זָהָ֔ב עַד\maqqaf יְרֵכָ֥הּ עַד\maqqaf פִּרְחָ֖הּ מִקְשָׁ֣ה הִ֑וא כַּמַּרְאֶ֗ה אֲשֶׁ֨ר הֶרְאָ֤ה יְהֹוָה֙ אֶת\maqqaf מֹשֶׁ֔ה כֵּ֥ן עָשָׂ֖ה אֶת\maqqaf הַמְּנֹרָֽה׃ \petucha }
{And this was the work of the candlestick, beaten work of gold; unto the base thereof, and unto the flowers thereof, it was beaten work; according unto the pattern which the \lord\space had shown Moses, so he made the candlestick.}
\newhaftarah{חנכה (שבת ראשון)}
\label{haft_hanukka1}
\setcounter{chap}{2}
\setcounter{verse}{14}
\twoverse{\Roman{chap}:\arabic{verse}}
{רָנִּ֥י וְשִׂמְחִ֖י בַּת\maqqaf צִיּ֑וֹן כִּ֧י הִנְנִי\maqqaf בָ֛א וְשָׁכַנְתִּ֥י בְתוֹכֵ֖ךְ נְאֻם\maqqaf יְהֹוָֽה׃}
{’Sing and rejoice, O daughter of Zion; for, lo, I come, and I will dwell in the midst of thee, saith the \lord.}
\twoverse{\arabic{verse}}
{וְנִלְווּ֩ גוֹיִ֨ם רַבִּ֤ים אֶל\maqqaf יְהֹוָה֙ בַּיּ֣וֹם הַה֔וּא וְהָ֥יוּ לִ֖י לְעָ֑ם וְשָׁכַנְתִּ֣י בְתוֹכֵ֔ךְ וְיָדַ֕עַתְּ כִּי\maqqaf יְהֹוָ֥ה צְבָא֖וֹת שְׁלָחַ֥נִי אֵלָֽיִךְ׃}
{And many nations shall join themselves to the \lord\space in that day, and shall be My people, and I will dwell in the midst of thee’; and thou shalt know that the \lord\space of hosts hath sent me unto thee.}
\twoverse{\arabic{verse}}
{וְנָחַ֨ל יְהֹוָ֤ה אֶת\maqqaf יְהוּדָה֙ חֶלְק֔וֹ עַ֖ל אַדְמַ֣ת הַקֹּ֑דֶשׁ וּבָחַ֥ר ע֖וֹד בִּירוּשָׁלָֽ͏ִם׃}
{And the \lord\space shall inherit Judah as His portion in the holy land, and shall choose Jerusalem again.}
\twoverse{\arabic{verse}}
{הַ֥ס כׇּל\maqqaf בָּשָׂ֖ר מִפְּנֵ֣י יְהֹוָ֑ה כִּ֥י נֵע֖וֹר מִמְּע֥וֹן קׇדְשֽׁוֹ׃ \petucha }
{Be silent, all flesh, before the \lord; for He is aroused out of His holy habitation.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{וַיַּרְאֵ֗נִי אֶת\maqqaf יְהוֹשֻׁ֙עַ֙ הַכֹּהֵ֣ן הַגָּד֔וֹל עֹמֵ֕ד לִפְנֵ֖י מַלְאַ֣ךְ יְהֹוָ֑ה וְהַשָּׂטָ֛ן עֹמֵ֥ד עַל\maqqaf יְמִינ֖וֹ לְשִׂטְנֽוֹ׃}
{And he showed me Joshua the high priest standing before the angel of the \lord, and Satan standing at his right hand to accuse him.}
\twoverse{\arabic{verse}}
{וַיֹּ֨אמֶר יְהֹוָ֜ה אֶל\maqqaf הַשָּׂטָ֗ן יִגְעַ֨ר יְהֹוָ֤ה בְּךָ֙ הַשָּׂטָ֔ן וְיִגְעַ֤ר יְהֹוָה֙ בְּךָ֔ הַבֹּחֵ֖ר בִּירֽוּשָׁלָ֑͏ִם הֲל֧וֹא זֶ֦ה א֖וּד מֻצָּ֥ל מֵאֵֽשׁ׃}
{And the \lord\space said unto Satan: ‘The \lord\space rebuke thee, O Satan, yea, the \lord\space that hath chosen Jerusalem rebuke thee; is not this man a brand plucked out of the fire?’}
\twoverse{\arabic{verse}}
{וִיהוֹשֻׁ֕עַ הָיָ֥ה לָבֻ֖שׁ בְּגָדִ֣ים צוֹאִ֑ים וְעֹמֵ֖ד לִפְנֵ֥י הַמַּלְאָֽךְ׃}
{Now Joshua was clothed with filthy garments, and stood before the angel.}
\twoverse{\arabic{verse}}
{וַיַּ֣עַן וַיֹּ֗אמֶר אֶל\maqqaf הָעֹמְדִ֤ים לְפָנָיו֙ לֵאמֹ֔ר הָסִ֛ירוּ הַבְּגָדִ֥ים הַצֹּאִ֖ים מֵעָלָ֑יו וַיֹּ֣אמֶר אֵלָ֗יו רְאֵ֨ה הֶעֱבַ֤רְתִּי מֵעָלֶ֙יךָ֙ עֲוֺנֶ֔ךָ וְהַלְבֵּ֥שׁ אֹתְךָ֖ מַחֲלָצֽוֹת׃}
{And he answered and spoke unto those that stood before him, saying: ‘Take the filthy garments from off him.’ And unto him he said: ‘Behold, I cause thine iniquity to pass from thee, and I will clothe thee with robes.’}
\twoverse{\arabic{verse}}
{וָאֹמַ֕ר יָשִׂ֛ימוּ צָנִ֥יף טָה֖וֹר עַל\maqqaf רֹאשׁ֑וֹ וַיָּשִׂ֩ימוּ֩ הַצָּנִ֨יף הַטָּה֜וֹר עַל\maqqaf רֹאשׁ֗וֹ וַיַּלְבִּשֻׁ֙הוּ֙ בְּגָדִ֔ים וּמַלְאַ֥ךְ יְהֹוָ֖ה עֹמֵֽד׃}
{And I said: ‘Let them set a fair mitre upon his head.’ So they set a fair mitre upon his head, and clothed him with garments; and the angel of the \lord\space stood by.}
\twoverse{\arabic{verse}}
{וַיָּ֙עַד֙ מַלְאַ֣ךְ יְהֹוָ֔ה בִּיהוֹשֻׁ֖עַ לֵאמֹֽר׃}
{And the angel of the \lord\space forewarned Joshua, saying:}
\twoverse{\arabic{verse}}
{כֹּה\maqqaf אָמַ֞ר יְהֹוָ֣ה צְבָא֗וֹת אִם\maqqaf בִּדְרָכַ֤י תֵּלֵךְ֙ וְאִ֣ם אֶת\maqqaf מִשְׁמַרְתִּ֣י תִשְׁמֹ֔ר וְגַם\maqqaf אַתָּה֙ תָּדִ֣ין אֶת\maqqaf בֵּיתִ֔י וְגַ֖ם תִּשְׁמֹ֣ר אֶת\maqqaf חֲצֵרָ֑י וְנָתַתִּ֤י לְךָ֙ מַהְלְכִ֔ים בֵּ֥ין הָעֹמְדִ֖ים הָאֵֽלֶּה׃}
{’Thus saith the \lord\space of hosts: If thou wilt walk in My ways, and if thou wilt keep My charge, and wilt also judge My house, and wilt also keep My courts, then I will give thee free access among these that stand by.}
\twoverse{\arabic{verse}}
{שְֽׁמַֽע\maqqaf נָ֞א יְהוֹשֻׁ֣עַ \legarmeh  הַכֹּהֵ֣ן הַגָּד֗וֹל אַתָּה֙ וְרֵעֶ֙יךָ֙ הַיֹּשְׁבִ֣ים לְפָנֶ֔יךָ כִּֽי\maqqaf אַנְשֵׁ֥י מוֹפֵ֖ת הֵ֑מָּה כִּֽי\maqqaf הִנְנִ֥י מֵבִ֛יא אֶת\maqqaf עַבְדִּ֖י צֶֽמַח׃}
{Hear now, O Joshua the high priest, thou and thy fellows that sit before thee; for they are men that are a sign; for, behold, I will bring forth My servant the Shoot.}
\twoverse{\arabic{verse}}
{כִּ֣י \legarmeh  הִנֵּ֣ה הָאֶ֗בֶן אֲשֶׁ֤ר נָתַ֙תִּי֙ לִפְנֵ֣י יְהוֹשֻׁ֔עַ עַל\maqqaf אֶ֥בֶן אַחַ֖ת שִׁבְעָ֣ה עֵינָ֑יִם הִנְנִ֧י מְפַתֵּ֣חַ פִּתֻּחָ֗הּ נְאֻם֙ יְהֹוָ֣ה צְבָא֔וֹת וּמַשְׁתִּ֛י אֶת\maqqaf עֲוֺ֥ן הָאָֽרֶץ\maqqaf הַהִ֖יא בְּי֥וֹם אֶחָֽד׃}
{For behold the stone that I have laid before Joshua; upon one stone are seven facets; behold, I will engrave the graving thereof, saith the \lord\space of hosts: And I will remove the iniquity of that land in one day.}
\twoverse{\arabic{verse}}
{בַּיּ֣וֹם הַה֗וּא נְאֻם֙ יְהֹוָ֣ה צְבָא֔וֹת תִּקְרְא֖וּ אִ֣ישׁ לְרֵעֵ֑הוּ אֶל\maqqaf תַּ֥חַת גֶּ֖פֶן וְאֶל\maqqaf תַּ֥חַת תְּאֵנָֽה׃}
{In that day, saith the \lord\space of hosts, shall ye call every man his neighbour under the vine and under the fig-tree.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}
{וַיָּ֕שׇׁב הַמַּלְאָ֖ךְ הַדֹּבֵ֣ר בִּ֑י וַיְעִירֵ֕נִי כְּאִ֖ישׁ אֲשֶׁר\maqqaf יֵע֥וֹר מִשְּׁנָתֽוֹ׃}
{And the angel that spoke with me returned, and waked me, as a man that is wakened out of his sleep.}
\twoverse{\arabic{verse}}
{וַיֹּ֣אמֶר אֵלַ֔י מָ֥ה אַתָּ֖ה רֹאֶ֑ה \qk{וָאֹמַ֡ר}{ויאמר} רָאִ֣יתִי \pasek  וְהִנֵּ֣ה מְנוֹרַת֩ זָהָ֨ב כֻּלָּ֜הּ וְגֻלָּ֣הּ עַל\maqqaf רֹאשָׁ֗הּ וְשִׁבְעָ֤ה נֵרֹתֶ֙יהָ֙ עָלֶ֔יהָ שִׁבְעָ֤ה וְשִׁבְעָה֙ מֽוּצָק֔וֹת לַנֵּר֖וֹת אֲשֶׁ֥ר עַל\maqqaf רֹאשָֽׁהּ׃}
{And he said unto me: ‘What seest thou?’ And I said: ‘I have seen, and behold a candlestick all of gold, with a bowl upon the top of it, and its seven lamps thereon; there are seven pipes, yea, seven, to the lamps, which are upon the top thereof;}
\twoverse{\arabic{verse}}
{וּשְׁנַ֥יִם זֵיתִ֖ים עָלֶ֑יהָ אֶחָד֙ מִימִ֣ין הַגֻּלָּ֔ה וְאֶחָ֖ד עַל\maqqaf שְׂמֹאלָֽהּ׃}
{and two olive-trees by it, one upon the right side of the bowl, and the other upon the left side thereof.’}
\twoverse{\arabic{verse}}
{וָאַ֙עַן֙ וָֽאֹמַ֔ר אֶל\maqqaf הַמַּלְאָ֛ךְ הַדֹּבֵ֥ר בִּ֖י לֵאמֹ֑ר מָה\maqqaf אֵ֖לֶּה אֲדֹנִֽי׃}
{And I answered and spoke to the angel that spoke with me, saying: ‘What are these, my lord?’}
\twoverse{\arabic{verse}}
{וַ֠יַּ֠עַן הַמַּלְאָ֞ךְ הַדֹּבֵ֥ר בִּי֙ וַיֹּ֣אמֶר אֵלַ֔י הֲל֥וֹא יָדַ֖עְתָּ מָה\maqqaf הֵ֣מָּה אֵ֑לֶּה וָאֹמַ֖ר לֹ֥א אֲדֹנִֽי׃}
{Then the angel that spoke with me answered and said unto me: ‘Knowest thou not what these are?’ And I said: ‘No, my lord.’}
\twoverse{\arabic{verse}}
{וַיַּ֜עַן וַיֹּ֤אמֶר אֵלַי֙ לֵאמֹ֔ר זֶ֚ה דְּבַר\maqqaf יְהֹוָ֔ה אֶל\maqqaf זְרֻבָּבֶ֖ל לֵאמֹ֑ר לֹ֤א בְחַ֙יִל֙ וְלֹ֣א בְכֹ֔חַ כִּ֣י אִם\maqqaf בְּרוּחִ֔י אָמַ֖ר יְהֹוָ֥ה צְבָאֽוֹת׃}
{Then he answered and spoke unto me, saying: ‘This is the word of the \lord\space unto Zerubbabel, saying: Not by might, nor by power, but by My spirit, saith the \lord\space of hosts.}
\twoverse{\arabic{verse}}
{מִֽי\maqqaf אַתָּ֧ה הַֽר\maqqaf הַגָּד֛וֹל לִפְנֵ֥י זְרֻבָּבֶ֖ל לְמִישֹׁ֑ר וְהוֹצִיא֙ אֶת\maqqaf הָאֶ֣בֶן הָרֹאשָׁ֔ה תְּשֻׁא֕וֹת חֵ֥ן חֵ֖ן לָֽהּ׃ \petucha }
{Who art thou, O great mountain before Zerubbabel? thou shalt become a plain; and he shall bring forth the top stone with shoutings of Grace, grace, unto it.’}
\newperek
\newhaftarah{חנכה (שבת שני)}
\label{haft_hanukka2}
\setcounter{chap}{7}
\setcounter{verse}{40}
\twoverse{\Roman{chap}:\arabic{verse}}
{וַיַּ֣עַשׂ חִיר֔וֹם אֶ֨ת\maqqaf הַכִּיֹּר֔וֹת וְאֶת\maqqaf הַיָּעִ֖ים וְאֶת\maqqaf הַמִּזְרָק֑וֹת וַיְכַ֣ל חִירָ֗ם לַֽעֲשׂוֹת֙ אֶת\maqqaf כׇּל\maqqaf הַמְּלָאכָ֔ה אֲשֶׁ֥ר עָשָׂ֛ה לַמֶּ֥לֶךְ שְׁלֹמֹ֖ה בֵּ֥ית יְהֹוָֽה׃}
{And Hiram made the pots, and the shovels, and the basins. So Hiram made an end of doing all the work that he wrought for king Solomon in the house of the \lord:}
\twoverse{\arabic{verse}}
{עַמֻּדִ֣ים שְׁנַ֔יִם וְגֻלֹּ֧ת הַכֹּתָרֹ֛ת אֲשֶׁר\maqqaf עַל\maqqaf רֹ֥אשׁ הָעַמּוּדִ֖ים שְׁתָּ֑יִם וְהַשְּׂבָכ֣וֹת שְׁתַּ֔יִם לְכַסּ֗וֹת אֶת\maqqaf שְׁתֵּי֙ גֻּלּ֣וֹת הַכֹּתָרֹ֔ת אֲשֶׁ֖ר עַל\maqqaf רֹ֥אשׁ הָעַמּוּדִֽים׃}
{the two pillars, and the two bowls of the capitals that were on the top of the pillars; and the two networks to cover the two bowls of the capitals that were on the top of the pillars;}
\twoverse{\arabic{verse}}
{וְאֶת\maqqaf הָרִמֹּנִ֛ים אַרְבַּ֥ע מֵא֖וֹת לִשְׁתֵּ֣י הַשְּׂבָכ֑וֹת שְׁנֵֽי\maqqaf טוּרִ֤ים רִמֹּנִים֙ לַשְּׂבָכָ֣ה הָאֶחָ֔ת לְכַסּ֗וֹת אֶת\maqqaf שְׁתֵּי֙ גֻּלּ֣וֹת הַכֹּתָרֹ֔ת אֲשֶׁ֖ר עַל\maqqaf פְּנֵ֥י הָעַמּוּדִֽים׃}
{and the four hundred pomegranates for the two networks, two rows of pomegranates for each network, to cover the two bowls of the capitals that were upon the top of the pillars;}
\twoverse{\arabic{verse}}
{וְאֶת\maqqaf הַמְּכֹנ֖וֹת עָ֑שֶׂר וְאֶת\maqqaf הַכִּיֹּרֹ֥ת עֲשָׂרָ֖ה עַל\maqqaf הַמְּכֹנֽוֹת׃}
{and the ten bases, and the ten lavers on the bases;}
\twoverse{\arabic{verse}}
{וְאֶת\maqqaf הַיָּ֖ם הָאֶחָ֑ד וְאֶת\maqqaf הַבָּקָ֥ר שְׁנֵים\maqqaf עָשָׂ֖ר תַּ֥חַת הַיָּֽם׃}
{and the one sea, and the twelve oxen under the sea;}
\twoverse{\arabic{verse}}
{וְאֶת\maqqaf הַסִּיר֨וֹת וְאֶת\maqqaf הַיָּעִ֜ים וְאֶת\maqqaf הַמִּזְרָק֗וֹת וְאֵת֙ כׇּל\maqqaf הַכֵּלִ֣ים \qk{הָאֵ֔לֶּה}{האהל} אֲשֶׁ֨ר עָשָׂ֥ה חִירָ֛ם לַמֶּ֥לֶךְ שְׁלֹמֹ֖ה בֵּ֣ית יְהֹוָ֑ה נְחֹ֖שֶׁת מְמֹרָֽט׃}
{and the pots, and the shovels, and the basins; even all these vessels, which Hiram made for king Solomon, in the house of the \lord, were of burnished brass.}
\twoverse{\arabic{verse}}
{בְּכִכַּ֤ר הַיַּרְדֵּן֙ יְצָקָ֣ם הַמֶּ֔לֶךְ בְּמַעֲבֵ֖ה הָאֲדָמָ֑ה בֵּ֥ין סֻכּ֖וֹת וּבֵ֥ין צָרְתָֽן׃}
{In the plain of the Jordan did the king cast them, in the clay ground between Succoth and Zarethan.}
\twoverse{\arabic{verse}}
{וַיַּנַּ֤ח שְׁלֹמֹה֙ אֶת\maqqaf כׇּל\maqqaf הַכֵּלִ֔ים מֵרֹ֖ב מְאֹ֣ד מְאֹ֑ד לֹ֥א נֶחְקַ֖ר מִשְׁקַ֥ל הַנְּחֹֽשֶׁת׃}
{And Solomon left all the vessels unweighed, because they were exceeding many; the weight of the brass could not be found out.}
\twoverse{\arabic{verse}}
{וַיַּ֣עַשׂ שְׁלֹמֹ֔ה אֵ֚ת כׇּל\maqqaf הַכֵּלִ֔ים אֲשֶׁ֖ר בֵּ֣ית יְהֹוָ֑ה אֵ֚ת מִזְבַּ֣ח הַזָּהָ֔ב וְאֶת\maqqaf הַשֻּׁלְחָ֗ן אֲשֶׁ֥ר עָלָ֛יו לֶ֥חֶם הַפָּנִ֖ים זָהָֽב׃}
{And Solomon made all the vessels that were in the house of the \lord: the golden altar, and the table whereupon the showbread was, of gold;}
\twoverse{\arabic{verse}}
{וְאֶת\maqqaf הַ֠מְּנֹר֠וֹת חָמֵ֨שׁ מִיָּמִ֜ין וְחָמֵ֧שׁ מִשְּׂמֹ֛אול לִפְנֵ֥י הַדְּבִ֖יר זָהָ֣ב סָג֑וּר וְהַפֶּ֧רַח וְהַנֵּרֹ֛ת וְהַמֶּלְקַחַ֖יִם זָהָֽב׃}
{and the candlesticks, five on the right side, and five on the left, before the Sanctuary, of pure gold; and the flowers, and the lamps, and the tongs, of gold;}
\twoverse{\arabic{verse}}
{וְ֠הַסִּפּ֠וֹת וְהַֽמְזַמְּר֧וֹת וְהַמִּזְרָק֛וֹת וְהַכַּפּ֥וֹת וְהַמַּחְתּ֖וֹת זָהָ֣ב סָג֑וּר וְהַפֹּת֡וֹת לְדַלְתוֹת֩ הַבַּ֨יִת הַפְּנִימִ֜י לְקֹ֣דֶשׁ הַקֳּדָשִׁ֗ים לְדַלְתֵ֥י הַבַּ֛יִת לַהֵיכָ֖ל זָהָֽב׃ \petucha }
{and the cups, and the snuffers, and the basins, and the pans, and the fire-pans, of pure gold; and the hinges, both for the doors of the inner house, the most holy place, and for the doors of the house, that is, of the temple, of gold.}


\end{document}
