\documentclass[11pt, openany]{book}
\usepackage[
paperheight=9.68in,
paperwidth=7.44in,
top=0.625in,bottom=0.75in,
inner=1in,
outer=0.75in,
marginparsep=0.1in,
headsep=16pt
]{geometry}

\usepackage{chumash_format}

\newcommand{\texttitle}{ספר  ויקרא}
\newcommand{\engtitle}{\textsc{Leviticus}}

\fancyhead[C]{\texttitle \space\textendash\space \englishfont Introduction}

\begin{document}
	
\include{chumash_files/0introduction}
	
\clearpage
\mainmatter

\pagenumbering{arabic}
\setmainheader

\include{chumash_files/3Leviticus_chumash}
\addcontentsline{toc}{chapter}{פרשת במדבר למנחה בשבת בחקתי}
\fancyhead[C]{פרשת במדבר למנחה בשבת בחקתי}

\newparsha{במדבר}
\setcounter{chap}{1}
\setcounter{verse}{1}
\renewcommand{\aliya}[1]{\textsf{#1}}
\twoverse{\textsf{במדבר}}
{וַיְדַבֵּ֨ר יְהֹוָ֧ה אֶל־מֹשֶׁ֛ה בְּמִדְבַּ֥ר סִינַ֖י בְּאֹ֣הֶל מוֹעֵ֑ד בְּאֶחָד֩ לַחֹ֨דֶשׁ הַשֵּׁנִ֜י בַּשָּׁנָ֣ה הַשֵּׁנִ֗ית לְצֵאתָ֛ם מֵאֶ֥רֶץ מִצְרַ֖יִם לֵאמֹֽר׃}
{AND THE \lord\space spoke unto Moses in the wilderness of Sinai, in the tent of meeting, on the first day of the second month, in the second year after the were come out of the land of Egypt, saying:}
\twoverse{\arabic{verse}}
{שְׂא֗וּ אֶת־רֹאשׁ֙ כׇּל־עֲדַ֣ת בְּנֵֽי־יִשְׂרָאֵ֔ל לְמִשְׁפְּחֹתָ֖ם לְבֵ֣ית אֲבֹתָ֑ם בְּמִסְפַּ֣ר שֵׁמ֔וֹת כׇּל־זָכָ֖ר לְגֻלְגְּלֹתָֽם׃}
{‘Take ye the sum of all the congregation of the children of Israel, by their families, by their fathers’ houses, according to the number of names, every male, by their polls;}
\twoverse{\arabic{verse}}
{מִבֶּ֨ן עֶשְׂרִ֤ים שָׁנָה֙ וָמַ֔עְלָה כׇּל־יֹצֵ֥א צָבָ֖א בְּיִשְׂרָאֵ֑ל תִּפְקְד֥וּ אֹתָ֛ם לְצִבְאֹתָ֖ם אַתָּ֥ה וְאַהֲרֹֽן׃}
{from twenty years old and upward, all that are able to go forth to war in Israel: ye shall number them by their hosts, even thou and Aaron.}
\twoverse{\arabic{verse}}
{וְאִתְּכֶ֣ם יִהְי֔וּ אִ֥ישׁ אִ֖ישׁ לַמַּטֶּ֑ה אִ֛ישׁ רֹ֥אשׁ לְבֵית־אֲבֹתָ֖יו הֽוּא׃}
{And with you there shall be a man of every tribe, every one head of his fathers’ house.}
\twoverse{\textsf{לוי}}
{וְאֵ֙לֶּה֙ שְׁמ֣וֹת הָֽאֲנָשִׁ֔ים אֲשֶׁ֥ר יַֽעַמְד֖וּ אִתְּכֶ֑ם לִרְאוּבֵ֕ן אֱלִיצ֖וּר בֶּן־שְׁדֵיאֽוּר׃}
{And these are the names of the men that shall stand with you: of Reuben, Elizur the son of Shedeur.}
\twoverse{\arabic{verse}}
{לְשִׁמְע֕וֹן שְׁלֻמִיאֵ֖ל בֶּן־צוּרִֽישַׁדָּֽי׃}
{Of Simeon, Shelumiel the son of Zurishaddai.}
\twoverse{\arabic{verse}}
{לִֽיהוּדָ֕ה נַחְשׁ֖וֹן בֶּן־עַמִּינָדָֽב׃}
{Of Judah, Nahshon the son of Amminadab.}
\twoverse{\arabic{verse}}
{לְיִ֨שָּׂשכָ֔ר נְתַנְאֵ֖ל בֶּן־צוּעָֽר׃}
{Of Issachar, Nethanel the son of Zuar.}
\twoverse{\arabic{verse}}
{לִזְבוּלֻ֕ן אֱלִיאָ֖ב בֶּן־חֵלֹֽן׃}
{Of Zebulun, Eliab the son of Helon.}
\twoverse{\arabic{verse}}
{לִבְנֵ֣י יוֹסֵ֔ף לְאֶפְרַ֕יִם אֱלִישָׁמָ֖ע בֶּן־עַמִּיה֑וּד לִמְנַשֶּׁ֕ה גַּמְלִיאֵ֖ל בֶּן־פְּדָהצֽוּר׃}
{Of the children of Joseph: of Ephraim, Elishama the son of Ammihud; of Manasseh, Gamaliel the son of Pedahzur.}
\twoverse{\arabic{verse}}
{לְבִ֨נְיָמִ֔ן אֲבִידָ֖ן בֶּן־גִּדְעֹנִֽי׃}
{Of Benjamin, Abidan the son of Gideoni.}
\twoverse{\arabic{verse}}
{לְדָ֕ן אֲחִיעֶ֖זֶר בֶּן־עַמִּֽישַׁדָּֽי׃}
{Of Dan, Ahiezer the son of Ammishaddai.}
\twoverse{\arabic{verse}}
{לְאָשֵׁ֕ר פַּגְעִיאֵ֖ל בֶּן־עׇכְרָֽן׃}
{Of Asher, Pagiel the son of Ochran.}
\twoverse{\arabic{verse}}
{לְגָ֕ד אֶלְיָסָ֖ף בֶּן־דְּעוּאֵֽל׃}
{Of Gad, Eliasaph the son of Deuel.}
\twoverse{\arabic{verse}}
{לְנַ֨פְתָּלִ֔י אֲחִירַ֖ע בֶּן־עֵינָֽן׃}
{Of Naphtali, Ahira the son of Enan.’}
\twoverse{\arabic{verse}}
{אֵ֚לֶּה \qk{קְרוּאֵ֣י}{קריאי} הָעֵדָ֔ה נְשִׂיאֵ֖י מַטּ֣וֹת אֲבוֹתָ֑ם רָאשֵׁ֛י אַלְפֵ֥י יִשְׂרָאֵ֖ל הֵֽם׃}
{These were the elect of the congregation, the princes of the tribes of their fathers; they were the heads of the thousands of Israel.}
\twoverse{\textsf{ישראל}}
{וַיִּקַּ֥ח מֹשֶׁ֖ה וְאַהֲרֹ֑ן אֵ֚ת הָאֲנָשִׁ֣ים הָאֵ֔לֶּה אֲשֶׁ֥ר נִקְּב֖וּ בְּשֵׁמֹֽת׃\note{בספרי ספרד ואשכנז בְּשֵׁמֽוֹת}}
{And Moses and Aaron took these men that are pointed out by name.}
\twoverse{\arabic{verse}}
{וְאֵ֨ת כׇּל־הָעֵדָ֜ה הִקְהִ֗ילוּ בְּאֶחָד֙ לַחֹ֣דֶשׁ הַשֵּׁנִ֔י וַיִּתְיַֽלְד֥וּ עַל־מִשְׁפְּחֹתָ֖ם לְבֵ֣ית אֲבֹתָ֑ם בְּמִסְפַּ֣ר שֵׁמ֗וֹת מִבֶּ֨ן עֶשְׂרִ֥ים שָׁנָ֛ה וָמַ֖עְלָה לְגֻלְגְּלֹתָֽם׃}
{And they assembled all the congregation together on the first day of the second month, and they declared their pedigrees after their families, by their fathers’ houses, according to the number of names, from twenty years old and upward, by their polls.}
\twoverse{\arabic{verse}}
{כַּאֲשֶׁ֛ר צִוָּ֥ה יְהֹוָ֖ה אֶת־מֹשֶׁ֑ה וַֽיִּפְקְדֵ֖ם בְּמִדְבַּ֥ר סִינָֽי׃ \setuma }
{As the \lord\space commanded Moses, so did he number them in the wilderness of Sinai.}
\clearpage
\fancyhead[C]{\haftarahname}
\include{chumash_files/3leviticus_haftarot}

\vspace{12pt}
\section*{מפטיר לשבת ראש חודש}
\addcontentsline{toc}{section}{מפטיר לשבת ראש חודש}
\renewcommand{\haftarahname}{מפטיר לשבת ראש חודש}
\vspace{-12pt}
\nopagebreak
\label{maft_ShabRCh}
\haftnote{The Maftir for Shabbat Rosh \d{H}odesh is Numbers 28:9\verserangechar 15.}
\setcounter{chap}{28}
\setcounter{verse}{9}
\twoverse{\Roman{chap}:\arabic{verse}}{וּבְיוֹם֙ הַשַּׁבָּ֔ת שְׁנֵֽי־כְבָשִׂ֥ים בְּנֵֽי־שָׁנָ֖ה תְּמִימִ֑ם וּשְׁנֵ֣י עֶשְׂרֹנִ֗ים סֹ֧לֶת מִנְחָ֛ה בְּלוּלָ֥ה בַשֶּׁ֖מֶן וְנִסְכּֽוֹ׃}{And on the sabbath day two he-lambs of the first year without blemish, and two tenth parts of an ephah of fine flour for a meal-offering, mingled with oil, and the drink-offering thereof.}
\twoverse{\arabic{verse}}{עֹלַ֥ת שַׁבַּ֖ת בְּשַׁבַּתּ֑וֹ עַל־עֹלַ֥ת הַתָּמִ֖יד וְנִסְכָּֽהּ׃ \petucha}{This is the burnt-offering of every sabbath, beside the continual burnt-offering, and the drink-offering thereof.}
\twoverse{\arabic{verse}}{וּבְרָאשֵׁי֙ חׇדְשֵׁיכֶ֔ם תַּקְרִ֥יבוּ עֹלָ֖ה לַיהֹוָ֑ה פָּרִ֨ים בְּנֵֽי־בָקָ֤ר שְׁנַ֙יִם֙ וְאַ֣יִל אֶחָ֔ד כְּבָשִׂ֧ים בְּנֵי־שָׁנָ֛ה שִׁבְעָ֖ה תְּמִימִֽם׃}{And in your new moons ye shall present a burnt-offering unto the \lord: two young bullocks, and one ram, seven he-lambs of the first year without blemish;}
\twoverse{\arabic{verse}}{וּשְׁלֹשָׁ֣ה עֶשְׂרֹנִ֗ים סֹ֤לֶת מִנְחָה֙ בְּלוּלָ֣ה בַשֶּׁ֔מֶן לַפָּ֖ר הָאֶחָ֑ד וּשְׁנֵ֣י עֶשְׂרֹנִ֗ים סֹ֤לֶת מִנְחָה֙ בְּלוּלָ֣ה בַשֶּׁ֔מֶן לָאַ֖יִל הָֽאֶחָֽד׃}{and three tenth parts of an ephah of fine flour for a meal-offering, mingled with oil, for each bullock; and two tenth parts of fine flour for a meal-offering, mingled with oil, for the one ram;}
\twoverse{\arabic{verse}}{וְעִשָּׂרֹ֣ן עִשָּׂר֗וֹן סֹ֤לֶת מִנְחָה֙ בְּלוּלָ֣ה בַשֶּׁ֔מֶן לַכֶּ֖בֶשׂ הָאֶחָ֑ד עֹלָה֙ רֵ֣יחַ נִיחֹ֔חַ אִשֶּׁ֖ה לַיהֹוָֽה׃}{and a several tenth part of fine flour mingled with oil for a meal-offering unto every lamb; for a burnt-offering of a sweet savour, an offering made by fire unto the \lord.}
\twoverse{\arabic{verse}}{וְנִסְכֵּיהֶ֗ם חֲצִ֣י הַהִין֩ יִהְיֶ֨ה לַפָּ֜ר וּשְׁלִישִׁ֧ת הַהִ֣ין לָאַ֗יִל וּרְבִיעִ֥ת הַהִ֛ין לַכֶּ֖בֶשׂ יָ֑יִן זֹ֣את עֹלַ֥ת חֹ֙דֶשׁ֙ בְּחׇדְשׁ֔וֹ לְחׇדְשֵׁ֖י הַשָּׁנָֽה׃}{And their drink-offerings shall be half a hin of wine for a bullock, and the third part of a hin for the ram, and the fourth part of a hin for a lamb. This is the burnt-offering of every new moon throughout the months of the year.}
\twoverse{\arabic{verse}}{וּשְׂעִ֨יר עִזִּ֥ים אֶחָ֛ד לְחַטָּ֖את לַיהֹוָ֑ה עַל־עֹלַ֧ת הַתָּמִ֛יד יֵעָשֶׂ֖ה וְנִסְכּֽוֹ׃ \setuma }{And one he-goat for a sin-offering unto the \lord; it shall be offered beside the continual burnt-offering, and the drink-offering thereof.}
\newhaftarah{שבת ראש חודש}
\label{haft_ShabRCh}
\haftnote{The Haftarah for Shabbat Rosh \d{H}odesh is Isaiah 66:1\verserangechar 24.}
\setcounter{chap}{66}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}{כֹּ֚ה אָמַ֣ר יְהֹוָ֔ה הַשָּׁמַ֣יִם כִּסְאִ֔י וְהָאָ֖רֶץ הֲדֹ֣ם רַגְלָ֑י אֵי־זֶ֥ה בַ֙יִת֙ אֲשֶׁ֣ר תִּבְנוּ־לִ֔י וְאֵי־זֶ֥ה מָק֖וֹם מְנוּחָתִֽי׃}{Thus saith the \lord: The heaven is My throne, and the earth is My footstool; where is the house that ye may build unto Me? And where is the place that may be My resting-place?}
\twoverse{\arabic{verse}}{וְאֶת־כׇּל־אֵ֙לֶּה֙ יָדִ֣י עָשָׂ֔תָה וַיִּהְי֥וּ כׇל־אֵ֖לֶּה נְאֻם־יְהֹוָ֑ה וְאֶל־זֶ֣ה אַבִּ֔יט אֶל־עָנִי֙ וּנְכֵה־ר֔וּחַ וְחָרֵ֖ד עַל־דְּבָרִֽי׃}{For all these things hath My hand made, and so all these things came to be, saith the \lord; but on this man will I look, even on him that is poor and of a contrite spirit, and trembleth at My word.}
\twoverse{\arabic{verse}}{שׁוֹחֵ֨ט הַשּׁ֜וֹר מַכֵּה־אִ֗ישׁ זוֹבֵ֤חַ הַשֶּׂה֙ עֹ֣רֵֽף כֶּ֔לֶב מַעֲלֵ֤ה מִנְחָה֙ דַּם־חֲזִ֔יר מַזְכִּ֥יר לְבֹנָ֖ה מְבָ֣רֵֽךְ אָ֑וֶן גַּם־הֵ֗מָּה בָּֽחֲרוּ֙ בְּדַרְכֵיהֶ֔ם וּבְשִׁקּוּצֵיהֶ֖ם נַפְשָׁ֥ם חָפֵֽצָה׃}{He that killeth an ox is as if he slew a man; he that sacrificeth a lamb, as if he broke a dog’s neck; He that offereth a meal-offering, as if he offered swine’s blood; he that maketh a memorial-offering of frankincense, as if he blessed an idol; according as they have chosen their own ways, and their soul delighteth in their abominations;}
\twoverse{\arabic{verse}}{גַּם־אֲנִ֞י אֶבְחַ֣ר בְּתַעֲלֻלֵיהֶ֗ם וּמְגֽוּרֹתָם֙ אָבִ֣יא לָהֶ֔ם יַ֤עַן קָרָ֙אתִי֙ וְאֵ֣ין עוֹנֶ֔ה דִּבַּ֖רְתִּי וְלֹ֣א שָׁמֵ֑עוּ וַיַּעֲשׂ֤וּ הָרַע֙ בְּעֵינַ֔י וּבַאֲשֶׁ֥ר לֹא־חָפַ֖צְתִּי בָּחָֽרוּ׃ \setuma }{Even so I will choose their mockings, And will bring their fears upon them; Because when I called, none did answer; When I spoke, they did not hear, But they did that which was evil in Mine eyes, And chose that in which I delighted not.}
\twoverse{\arabic{verse}}{שִׁמְעוּ֙ דְּבַר־יְהֹוָ֔ה הַחֲרֵדִ֖ים אֶל־דְּבָר֑וֹ אָמְרוּ֩ אֲחֵיכֶ֨ם שֹׂנְאֵיכֶ֜ם מְנַדֵּיכֶ֗ם לְמַ֤עַן שְׁמִי֙ יִכְבַּ֣ד יְהֹוָ֔ה וְנִרְאֶ֥ה בְשִׂמְחַתְכֶ֖ם וְהֵ֥ם יֵבֹֽשׁוּ׃}{Hear the word of the \lord, Ye that tremble at His word: Your brethren that hate you, that cast you out for My name’s sake, have said: ‘Let the \lord\space be glorified, That we may gaze upon your joy’, But they shall be ashamed.}
\twoverse{\arabic{verse}}{ק֤וֹל שָׁאוֹן֙ מֵעִ֔יר ק֖וֹל מֵהֵיכָ֑ל ק֣וֹל יְהֹוָ֔ה מְשַׁלֵּ֥ם גְּמ֖וּל לְאֹיְבָֽיו׃}{Hark! an uproar from the city, Hark! it cometh from the temple, Hark! the \lord\space rendereth recompense to His enemies.}
\twoverse{\arabic{verse}}{בְּטֶ֥רֶם תָּחִ֖יל יָלָ֑דָה בְּטֶ֨רֶם יָב֥וֹא חֵ֛בֶל לָ֖הּ וְהִמְלִ֥יטָה זָכָֽר׃}{Before she travailed, she brought forth; Before her pain came, She was delivered of a man-child.}
\twoverse{\arabic{verse}}{מִֽי־שָׁמַ֣ע כָּזֹ֗את מִ֤י רָאָה֙ כָּאֵ֔לֶּה הֲי֤וּחַל אֶ֙רֶץ֙ בְּי֣וֹם אֶחָ֔ד אִם־יִוָּ֥לֵֽד גּ֖וֹי פַּ֣עַם אֶחָ֑ת כִּי־חָ֛לָה גַּם־יָלְדָ֥ה צִיּ֖וֹן אֶת־בָּנֶֽיהָ׃}{Who hath heard such a thing? Who hath seen such things? Is a land born in one day? Is a nation brought forth at once? For as soon as Zion travailed, She brought forth her children.}
\twoverse{\arabic{verse}}{הַאֲנִ֥י אַשְׁבִּ֛יר וְלֹ֥א אוֹלִ֖יד יֹאמַ֣ר יְהֹוָ֑ה אִם־אֲנִ֧י הַמּוֹלִ֛יד וְעָצַ֖רְתִּי אָמַ֥ר אֱלֹהָֽיִךְ׃ \setuma }{Shall I bring to the birth, and not cause to bring forth? Saith the \lord; Shall I that cause to bring forth shut the womb? Saith thy God.}
\twoverse{\arabic{verse}}{שִׂמְח֧וּ אֶת־יְרוּשָׁלַ֛͏ִם וְגִ֥ילוּ בָ֖הּ כׇּל־אֹהֲבֶ֑יהָ שִׂ֤ישׂוּ אִתָּהּ֙ מָשׂ֔וֹשׂ כׇּל־הַמִּֽתְאַבְּלִ֖ים עָלֶֽיהָ׃}{Rejoice ye with Jerusalem, And be glad with her, all ye that love her; Rejoice for joy with her, All ye that mourn for her;}
\twoverse{\arabic{verse}}{לְמַ֤עַן תִּֽינְקוּ֙ וּשְׂבַעְתֶּ֔ם מִשֹּׁ֖ד תַּנְחֻמֶ֑יהָ לְמַ֧עַן תָּמֹ֛צּוּ וְהִתְעַנַּגְתֶּ֖ם מִזִּ֥יז כְּבוֹדָֽהּ׃ \setuma }{That ye may suck, and be satisfied With the breast of her consolations; That ye may drink deeply with delight Of the abundance of her glory.}
\twoverse{\arabic{verse}}{כִּי־כֹ֣ה \legarmeh  אָמַ֣ר יְהֹוָ֗ה הִנְנִ֣י נֹטֶֽה־אֵ֠לֶ֠יהָ כְּנָהָ֨ר שָׁל֜וֹם וּכְנַ֧חַל שׁוֹטֵ֛ף כְּב֥וֹד גּוֹיִ֖ם וִֽינַקְתֶּ֑ם עַל־צַד֙ תִּנָּשֵׂ֔אוּ וְעַל־בִּרְכַּ֖יִם תְּשׇׁעֳשָֽׁעוּ׃}{For thus saith the \lord: Behold, I will extend peace to her like a river. And the wealth of the nations like an overflowing stream, and ye shall suck thereof: Ye shall be borne upon the side, and shall be dandled upon the knees.}
\twoverse{\arabic{verse}}{כְּאִ֕ישׁ אֲשֶׁ֥ר אִמּ֖וֹ תְּנַחֲמֶ֑נּוּ כֵּ֤ן אָֽנֹכִי֙ אֲנַ֣חֶמְכֶ֔ם וּבִירֽוּשָׁלַ֖͏ִם תְּנֻחָֽמוּ׃}{As one whom his mother comforteth, So will I comfort you; and ye shall be comforted in Jerusalem.}
\twoverse{\arabic{verse}}{וּרְאִיתֶם֙ וְשָׂ֣שׂ לִבְּכֶ֔ם וְעַצְמוֹתֵיכֶ֖ם כַּדֶּ֣שֶׁא תִפְרַ֑חְנָה וְנוֹדְעָ֤ה יַד־יְהֹוָה֙ אֶת־עֲבָדָ֔יו וְזָעַ֖ם אֶת־אֹיְבָֽיו׃ \setuma }{And when ye see this, your heart shall rejoice, and your bones shall flourish like young grass; and the hand of the \lord\space shall be known toward His servants, and He will have indignation against His enemies.}
\twoverse{\arabic{verse}}{כִּי־הִנֵּ֤ה יְהֹוָה֙ בָּאֵ֣שׁ יָב֔וֹא וְכַסּוּפָ֖ה מַרְכְּבֹתָ֑יו לְהָשִׁ֤יב בְּחֵמָה֙ אַפּ֔וֹ וְגַעֲרָת֖וֹ בְּלַהֲבֵי־אֵֽשׁ׃}{For, behold, the \lord\space will come in fire, And His chariots shall be like the whirlwind; to render His anger with fury, And His rebuke with flames of fire. }
\twoverse{\arabic{verse}}{כִּ֤י בָאֵשׁ֙ יְהֹוָ֣ה נִשְׁפָּ֔ט וּבְחַרְבּ֖וֹ אֶת־כׇּל־בָּשָׂ֑ר וְרַבּ֖וּ חַֽלְלֵ֥י יְהֹוָֽה׃}{For by fire will the \lord\space contend, And by His sword with all flesh; And the slain of the \lord\space shall be many.}
\twoverse{\arabic{verse}}{הַמִּתְקַדְּשִׁ֨ים וְהַמִּֽטַּהֲרִ֜ים אֶל־הַגַּנּ֗וֹת אַחַ֤ר \qk{אַחַת֙}{אחד} בַּתָּ֔וֶךְ אֹֽכְלֵי֙ בְּשַׂ֣ר הַחֲזִ֔יר וְהַשֶּׁ֖קֶץ וְהָעַכְבָּ֑ר יַחְדָּ֥ו יָסֻ֖פוּ נְאֻם־יְהֹוָֽה׃}{They that sanctify themselves and purify themselves to go unto the gardens, behind one in the midst, Eating swine’s flesh, and the detestable thing, and the mouse, Shall be consumed together, saith the \lord.}
\twoverse{\arabic{verse}}{וְאָנֹכִ֗י מַֽעֲשֵׂיהֶם֙ וּמַחְשְׁבֹ֣תֵיהֶ֔ם בָּאָ֕ה לְקַבֵּ֥ץ אֶת־כׇּל־הַגּוֹיִ֖ם וְהַלְּשֹׁנ֑וֹת וּבָ֖אוּ וְרָא֥וּ אֶת־כְּבוֹדִֽי׃}{For I [know] their works and their thoughts; [the time] cometh, that I will gather all nations and tongues; and they shall come, and shall see My glory.}
\twoverse{\arabic{verse}}{וְשַׂמְתִּ֨י בָהֶ֜ם א֗וֹת וְשִׁלַּחְתִּ֣י מֵהֶ֣ם \small ׀ \normalsize פְּ֠לֵיטִ֠ים אֶֽל־הַגּוֹיִ֞ם תַּרְשִׁ֨ישׁ פּ֥וּל וְל֛וּד מֹ֥שְׁכֵי קֶ֖שֶׁת תֻּבַ֣ל וְיָוָ֑ן הָאִיִּ֣ים הָרְחֹקִ֗ים אֲשֶׁ֨ר לֹֽא־שָׁמְע֤וּ אֶת־שִׁמְעִי֙ וְלֹא־רָא֣וּ אֶת־כְּבוֹדִ֔י וְהִגִּ֥ידוּ אֶת־כְּבוֹדִ֖י בַּגּוֹיִֽם׃}{And I will work a sign among them, and I will send such as escape of them unto the nations, to Tarshish, Pul and Lud, that draw the bow, to Tubal and Javan, to the isles afar off, that have not heard My fame, neither have seen My glory; and they shall declare My glory among the nations.}
\twoverse{\arabic{verse}}{וְהֵבִ֣יאוּ אֶת־כׇּל־אֲחֵיכֶ֣ם מִכׇּל־הַגּוֹיִ֣ם \small ׀ \normalsize מִנְחָ֣ה \small ׀ \normalsize לַיהֹוָ֡ה בַּסּוּסִ֡ים וּ֠בָרֶ֠כֶב וּבַצַּבִּ֨ים וּבַפְּרָדִ֜ים וּבַכִּרְכָּר֗וֹת עַ֣ל הַ֥ר קׇדְשִׁ֛י יְרוּשָׁלַ֖͏ִם אָמַ֣ר יְהֹוָ֑ה כַּאֲשֶׁ֣ר יָבִ֩יאוּ֩ בְנֵ֨י יִשְׂרָאֵ֧ל אֶת־הַמִּנְחָ֛ה בִּכְלִ֥י טָה֖וֹר בֵּ֥ית יְהֹוָֽה׃}{And they shall bring all your brethren out of all the nations for an offering unto the \lord, upon horses, and in chariots, and in fitters, and upon mules, and upon swift beasts, to My holy mountain Jerusalem, saith the \lord, as the children of Israel bring their offering in a clean vessel into the house of the \lord.}
\twoverse{\arabic{verse}}{וְגַם־מֵהֶ֥ם אֶקַּ֛ח לַכֹּהֲנִ֥ים לַלְוִיִּ֖ם אָמַ֥ר יְהֹוָֽה׃}{And of them also will I take for the priests and for the Levites, saith the \lord.}
\twoverse{\arabic{verse}}{כִּ֣י כַאֲשֶׁ֣ר הַשָּׁמַ֣יִם הַ֠חֲדָשִׁ֠ים וְהָאָ֨רֶץ הַחֲדָשָׁ֜ה אֲשֶׁ֨ר אֲנִ֥י עֹשֶׂ֛ה עֹמְדִ֥ים לְפָנַ֖י נְאֻם־יְהֹוָ֑ה כֵּ֛ן יַעֲמֹ֥ד זַרְעֲכֶ֖ם וְשִׁמְכֶֽם׃}{For as the new heavens and the new earth, which I will make, shall remain before Me, saith the \lord, so shall your seed and your name remain.}
\twoverse{\arabic{verse}}{וְהָיָ֗ה מִֽדֵּי־חֹ֙דֶשׁ֙ בְּחׇדְשׁ֔וֹ וּמִדֵּ֥י שַׁבָּ֖ת בְּשַׁבַּתּ֑וֹ יָב֧וֹא כׇל־בָּשָׂ֛ר לְהִשְׁתַּחֲוֺ֥ת לְפָנַ֖י אָמַ֥ר יְהֹוָֽה׃}{And it shall come to pass, that from one new moon to another, and from one sabbath to another, shall all flesh come to worship before Me, Saith the \lord.}
\twoverse{\arabic{verse}}{וְיָצְא֣וּ וְרָא֔וּ בְּפִגְרֵי֙ הָאֲנָשִׁ֔ים הַפֹּשְׁעִ֖ים בִּ֑י כִּ֣י תוֹלַעְתָּ֞ם לֹ֣א תָמ֗וּת וְאִשָּׁם֙ לֹ֣א תִכְבֶּ֔ה וְהָי֥וּ דֵרָא֖וֹן לְכׇל־בָּשָֽׂר׃ \small והיה מדי חדש בחדשו ומדי שבת בשבתו יבוא כל בשר להשתחות לפני אמר יהוה\small}{And they shall go forth, and look Upon the carcasses of the men that have rebelled against Me; For their worm shall not die, Neither shall their fire be quenched; And they shall be an abhorring unto all flesh. }
\newhaftarah{שבת מחר חודש}
\label{haft_macharchodesh}
\haftnote{The Haftarah for Shabbat Ma\d{h}ar \d{H}odesh is I Samuel 20:18\verserangechar 42.}
\setcounter{chap}{20}
\setcounter{verse}{18}
\twoverse{\Roman{chap}:\arabic{verse}}{וַיֹּאמֶר־ל֥וֹ יְהוֹנָתָ֖ן מָחָ֣ר חֹ֑דֶשׁ וְנִפְקַ֕דְתָּ כִּ֥י יִפָּקֵ֖ד מוֹשָׁבֶֽךָ׃}{And Jonathan said unto him: ‘To-morrow is the new moon; and thou wilt be missed, thy seat will be empty.}
\twoverse{\arabic{verse}}{וְשִׁלַּשְׁתָּ֙ תֵּרֵ֣ד מְאֹ֔ד וּבָאתָ֙ אֶל־הַמָּק֔וֹם אֲשֶׁר־נִסְתַּ֥רְתָּ שָּׁ֖ם בְּי֣וֹם הַֽמַּעֲשֶׂ֑ה וְיָ֣שַׁבְתָּ֔ אֵ֖צֶל הָאֶ֥בֶן הָאָֽזֶל׃}{And in the third day thou shalt hide thyself well, and come to the place where thou didst hide thyself in the day of work, and shalt remain by the stone Ezel.}
\twoverse{\arabic{verse}}{וַאֲנִ֕י שְׁלֹ֥שֶׁת הַחִצִּ֖ים צִדָּ֣ה אוֹרֶ֑ה לְשַֽׁלַּֽח־לִ֖י לְמַטָּרָֽה׃}{And I will shoot three arrows to the side-ward, as though I shot at a mark.}
\twoverse{\arabic{verse}}{וְהִנֵּה֙ אֶשְׁלַ֣ח אֶת־הַנַּ֔עַר לֵ֖ךְ מְצָ֣א אֶת־הַחִצִּ֑ים אִם־אָמֹר֩ אֹמַ֨ר לַנַּ֜עַר הִנֵּ֥ה הַחִצִּ֣ים \legarmeh  מִמְּךָ֣ וָהֵ֗נָּה קָחֶ֧נּוּ \small׀\small וָבֹ֛אָה כִּי־שָׁל֥וֹם לְךָ֛ וְאֵ֥ין דָּבָ֖ר חַי־יְהֹוָֽה׃}{And, behold, I will send the lad: Go, find the arrows. If I say unto the lad: Behold, the arrows are on this side of thee; take them, and come; for there is peace to thee and no hurt, as the \lord\space liveth.}
\twoverse{\arabic{verse}}{וְאִם־כֹּ֤ה אֹמַר֙ לָעֶ֔לֶם הִנֵּ֥ה הַחִצִּ֖ים מִמְּךָ֣ וָהָ֑לְאָה לֵ֕ךְ כִּ֥י שִֽׁלַּחֲךָ֖ יְהֹוָֽה׃}{But if I say thus unto the boy: Behold, the arrows are beyond thee; go thy way; for the \lord\space hath sent thee away.}
\twoverse{\arabic{verse}}{וְהַ֨דָּבָ֔ר אֲשֶׁ֥ר דִּבַּ֖רְנוּ אֲנִ֣י וָאָ֑תָּה הִנֵּ֧ה יְהֹוָ֛ה בֵּינִ֥י וּבֵינְךָ֖ עַד־עוֹלָֽם׃ \setuma }{And as touching the matter which I and thou have spoken of, behold, the \lord\space is between me and thee for ever.’}
\twoverse{\arabic{verse}}{וַיִּסָּתֵ֥ר דָּוִ֖ד בַּשָּׂדֶ֑ה וַיְהִ֣י הַחֹ֔דֶשׁ וַיֵּ֧שֶׁב הַמֶּ֛לֶךְ \qk{אֶל־}{על}הַלֶּ֖חֶם לֶאֱכֽוֹל׃}{So David hid himself in the field; and when the new moon was come, the king sat him down to the meal to eat.}
\twoverse{\arabic{verse}}{וַיֵּ֣שֶׁב הַ֠מֶּ֠לֶךְ עַל־מ֨וֹשָׁב֜וֹ כְּפַ֣עַם \legarmeh  בְּפַ֗עַם אֶל־מוֹשַׁב֙ הַקִּ֔יר וַיָּ֙קׇם֙ יְה֣וֹנָתָ֔ן וַיֵּ֥שֶׁב אַבְנֵ֖ר מִצַּ֣ד שָׁא֑וּל וַיִּפָּקֵ֖ד מְק֥וֹם דָּוִֽד׃}{And the king sat upon his seat, as at other times, even upon the seat by the wall; and Jonathan stood up, and Abner sat by Saul’s side; but David’s place was empty.}
\twoverse{\arabic{verse}}{וְלֹא־דִבֶּ֥ר שָׁא֛וּל מְא֖וּמָה בַּיּ֣וֹם הַה֑וּא כִּ֤י אָמַר֙ מִקְרֶ֣ה ה֔וּא בִּלְתִּ֥י טָה֛וֹר ה֖וּא כִּי־לֹ֥א טָהֽוֹר׃ \setuma }{Nevertheless Saul spoke not any thing that day; for he thought: ‘Something hath befallen him, he is unclean; surely he is not clean.’}
\twoverse{\arabic{verse}}{וַיְהִ֗י מִֽמׇּחֳרַ֤ת הַחֹ֙דֶשׁ֙ הַשֵּׁנִ֔י וַיִּפָּקֵ֖ד מְק֣וֹם דָּוִ֑ד \petucha וַיֹּ֤אמֶר שָׁאוּל֙ אֶל־יְהוֹנָתָ֣ן בְּנ֔וֹ מַדּ֜וּעַ לֹא־בָ֧א בֶן־יִשַׁ֛י גַּם־תְּמ֥וֹל גַּם־הַיּ֖וֹם אֶל־הַלָּֽחֶם׃}{And it came to pass on the morrow after the new moon, which was the second day, that David’s place was empty; and Saul said unto Jonathan his son: ‘Wherefore cometh not the son of Jesse to the meal, neither yesterday, nor to-day?’}
\twoverse{\arabic{verse}}{וַיַּ֥עַן יְהוֹנָתָ֖ן אֶת־שָׁא֑וּל נִשְׁאֹ֨ל נִשְׁאַ֥ל דָּוִ֛ד מֵעִמָּדִ֖י עַד־בֵּ֥ית לָֽחֶם׃}{And Jonathan answered Saul: ‘David earnestly asked leave of me to go to Beth-lehem;}
\twoverse{\arabic{verse}}{וַיֹּ֡אמֶר שַׁלְּחֵ֣נִי נָ֡א כִּ֣י זֶ֩בַח֩ מִשְׁפָּחָ֨ה לָ֜נוּ בָּעִ֗יר וְה֤וּא צִוָּה־לִי֙ אָחִ֔י וְעַתָּ֗ה אִם־מָצָ֤אתִי חֵן֙ בְּעֵינֶ֔יךָ אִמָּ֥לְטָה נָּ֖א וְאֶרְאֶ֣ה אֶת־אֶחָ֑י עַל־כֵּ֣ן לֹא־בָ֔א אֶל־שֻׁלְחַ֖ן הַמֶּֽלֶךְ׃ \setuma }{and he said: Let me go, I pray thee; for our family hath a sacrifice in the city; and my brother, he hath commanded me; and now, if I have found favour in thine eyes, let me get away, I pray thee, and see my brethren. Therefore he is not come unto the king’s table.’}
\twoverse{\arabic{verse}}{וַיִּֽחַר־אַ֤ף שָׁאוּל֙ בִּיה֣וֹנָתָ֔ן וַיֹּ֣אמֶר ל֔וֹ בֶּֽן־נַעֲוַ֖ת הַמַּרְדּ֑וּת הֲל֣וֹא יָדַ֗עְתִּי כִּֽי־בֹחֵ֤ר אַתָּה֙ לְבֶן־יִשַׁ֔י לְבׇ֨שְׁתְּךָ֔ וּלְבֹ֖שֶׁת עֶרְוַ֥ת אִמֶּֽךָ׃}{Then Saul’s anger was kindled against Jonathan, and he said unto him: ‘Thou son of perverse rebellion, do not I know that thou hast chosen the son of Jesse to thine own shame, and unto the shame of thy mother’s nakedness?}
\twoverse{\arabic{verse}}{כִּ֣י כׇל־הַיָּמִ֗ים אֲשֶׁ֤ר בֶּן־יִשַׁי֙ חַ֣י עַל־הָאֲדָמָ֔ה לֹ֥א תִכּ֖וֹן אַתָּ֣ה וּמַלְכוּתֶ֑ךָ וְעַתָּ֗ה שְׁלַ֨ח וְקַ֤ח אֹתוֹ֙ אֵלַ֔י כִּ֥י בֶן־מָ֖וֶת הֽוּא׃ \setuma }{For as long as the son of Jesse liveth upon the earth, thou shalt not be established, nor thy kingdom. Wherefore now send and fetch him unto me, for he deserveth to die.’}
\twoverse{\arabic{verse}}{וַיַּ֙עַן֙ יְה֣וֹנָתָ֔ן אֶת־שָׁא֖וּל אָבִ֑יו וַיֹּ֧אמֶר אֵלָ֛יו לָ֥מָּה יוּמַ֖ת מֶ֥ה עָשָֽׂה׃}{And Jonathan answered Saul his father, and said unto him: ‘Wherefore should he be put to death? what hath he done?’}
\twoverse{\arabic{verse}}{וַיָּ֨טֶל שָׁא֧וּל אֶֽת־הַחֲנִ֛ית עָלָ֖יו לְהַכֹּת֑וֹ וַיֵּ֙דַע֙ יְה֣וֹנָתָ֔ן כִּי־כָ֥לָה הִ֛יא מֵעִ֥ם אָבִ֖יו לְהָמִ֥ית אֶת־דָּוִֽד׃ \setuma }{And Saul cast his spear at him to smite him; whereby Jonathan knew that it was determined of his father to put David to death.}
\twoverse{\arabic{verse}}{וַיָּ֧קׇם יְהֽוֹנָתָ֛ן מֵעִ֥ם הַשֻּׁלְחָ֖ן בׇּחֳרִי־אָ֑ף וְלֹֽא־אָכַ֞ל בְּיוֹם־הַחֹ֤דֶשׁ הַשֵּׁנִי֙ לֶ֔חֶם כִּ֤י נֶעְצַב֙ אֶל־דָּוִ֔ד כִּ֥י הִכְלִמ֖וֹ אָבִֽיו׃ \setuma }{So Jonathan arose from the table in fierce anger, and did eat no food the second day of the month; for he was grieved for David, and because his father had put him to shame.}
\twoverse{\arabic{verse}}{וַיְהִ֣י בַבֹּ֔קֶר וַיֵּצֵ֧א יְהוֹנָתָ֛ן הַשָּׂדֶ֖ה לְמוֹעֵ֣ד דָּוִ֑ד וְנַ֥עַר קָטֹ֖ן עִמּֽוֹ׃}{And it came to pass in the morning, that Jonathan went out into the field at the time appointed with David, and a little lad with him.}
\twoverse{\arabic{verse}}{וַיֹּ֣אמֶר לְנַֽעֲר֔וֹ רֻ֗ץ מְצָ֥א נָא֙ אֶת־הַ֣חִצִּ֔ים אֲשֶׁ֥ר אָנֹכִ֖י מוֹרֶ֑ה הַנַּ֣עַר רָ֔ץ וְהוּא־יָרָ֥ה הַחֵ֖צִי לְהַעֲבִרֽוֹ׃}{And he said unto his lad: ‘Run, find now the arrows which I shoot.’ And as the lad ran, he shot an arrow beyond him.}
\twoverse{\arabic{verse}}{וַיָּבֹ֤א הַנַּ֙עַר֙ עַד־מְק֣וֹם הַחֵ֔צִי אֲשֶׁ֥ר יָרָ֖ה יְהוֹנָתָ֑ן וַיִּקְרָ֨א יְהוֹנָתָ֜ן אַחֲרֵ֤י הַנַּ֙עַר֙ וַיֹּ֔אמֶר הֲל֥וֹא הַחֵ֖צִי מִמְּךָ֥ וָהָֽלְאָה׃}{And when the lad was come to the place of the arrow which Jonathan had shot, Jonathan cried after the lad, and said: ‘Is not the arrow beyond thee?’}
\twoverse{\arabic{verse}}{וַיִּקְרָ֤א יְהֽוֹנָתָן֙ אַחֲרֵ֣י הַנַּ֔עַר מְהֵרָ֥ה ח֖וּשָׁה אַֽל־תַּעֲמֹ֑ד וַיְלַקֵּ֞ט נַ֤עַר יְהֽוֹנָתָן֙ אֶת־\qk{הַ֣חִצִּ֔ים}{החצי} וַיָּבֹ֖א אֶל־אֲדֹנָֽיו׃}{And Jonathan cried after the lad: ‘Make speed, hasten, stay not.’ And Jonathan’s lad gathered up the arrows, and came to his master.}
\twoverse{\arabic{verse}}{וְהַנַּ֖עַר לֹא־יָדַ֣ע מְא֑וּמָה אַ֤ךְ יְהֽוֹנָתָן֙ וְדָוִ֔ד יָדְע֖וּ אֶת־הַדָּבָֽר׃ \setuma }{But the lad knew not any thing; only Jonathan and David knew the matter.}
\twoverse{\arabic{verse}}{וַיִּתֵּ֤ן יְהֽוֹנָתָן֙ אֶת־כֵּלָ֔יו אֶל־הַנַּ֖עַר אֲשֶׁר־ל֑וֹ וַיֹּ֣אמֶר ל֔וֹ לֵ֖ךְ הָבֵ֥יא הָעִֽיר׃}{And Jonathan gave his weapons unto his lad, and said unto him: ‘Go, carry them to the city.’}
\twoverse{\arabic{verse}}{הַנַּ֘עַר֮ בָּא֒ וְדָוִ֗ד קָ֚ם מֵאֵ֣צֶל הַנֶּ֔גֶב וַיִּפֹּ֨ל לְאַפָּ֥יו אַ֛רְצָה וַיִּשְׁתַּ֖חוּ שָׁלֹ֣שׁ פְּעָמִ֑ים וַֽיִּשְּׁק֣וּ \legarmeh  אִ֣ישׁ אֶת־רֵעֵ֗הוּ וַיִּבְכּוּ֙ אִ֣ישׁ אֶת־רֵעֵ֔הוּ עַד־דָּוִ֖ד הִגְדִּֽיל׃}{And as soon as the lad was gone, David arose out of a place toward the South, and fell on his face to the ground, and bowed down three times; and they kissed one another, and wept one with another, until David exceeded.}
\twoverse{\arabic{verse}}{וַיֹּ֧אמֶר יְהוֹנָתָ֛ן לְדָוִ֖ד לֵ֣ךְ לְשָׁל֑וֹם אֲשֶׁר֩ נִשְׁבַּ֨עְנוּ שְׁנֵ֜ינוּ אֲנַ֗חְנוּ בְּשֵׁ֤ם יְהֹוָה֙ לֵאמֹ֔ר יְהֹוָ֞ה יִהְיֶ֣ה \legarmeh  בֵּינִ֣י וּבֵינֶ֗ךָ וּבֵ֥ין זַרְעִ֛י וּבֵ֥ין זַרְעֲךָ֖ עַד־עוֹלָֽם׃ \petucha}{And Jonathan said to David: ‘Go in peace, forasmuch as we have sworn both of us in the name of the \lord, saying: The \lord\space shall be between me and thee, and Between my seed and thy seed, for ever.’}
\vspace{12pt}
\section*{פרשת זכור}
\addcontentsline{toc}{section}{פרשת זכור}
\renewcommand{\haftarahname}{פרשת זכור}
\vspace{-12pt}
\nopagebreak
\label{maft_zachor}
\haftnote{The Maftir for Shabbat Zakhor is Deuteronomy 25:17\verserangechar 19.}
\setcounter{chap}{25}
\setcounter{verse}{17}
\twoverse{\Roman{chap}:\arabic{verse}}{זָכ֕וֹר אֵ֛ת אֲשֶׁר־עָשָׂ֥ה לְךָ֖ עֲמָלֵ֑ק בַּדֶּ֖רֶךְ בְּצֵאתְכֶ֥ם מִמִּצְרָֽיִם׃}{Remember what Amalek did unto thee by the way as ye came forth out of Egypt;}
\twoverse{\arabic{verse}}{אֲשֶׁ֨ר קָֽרְךָ֜ בַּדֶּ֗רֶךְ וַיְזַנֵּ֤ב בְּךָ֙ כׇּל־הַנֶּחֱשָׁלִ֣ים אַֽחֲרֶ֔יךָ וְאַתָּ֖ה עָיֵ֣ף וְיָגֵ֑עַ וְלֹ֥א יָרֵ֖א אֱלֹהִֽים׃}{how he met thee by the way, and smote the hindmost of thee, all that were enfeebled in thy rear, when thou wast faint and weary; and he feared not God.}
\twoverse{\arabic{verse}}{וְהָיָ֡ה בְּהָנִ֣יחַ יְהֹוָ֣ה אֱלֹהֶ֣יךָ \small׀\small לְ֠ךָ֠ מִכׇּל־אֹ֨יְבֶ֜יךָ מִסָּבִ֗יב בָּאָ֙רֶץ֙ אֲשֶׁ֣ר יְהֹוָה־אֱ֠לֹהֶ֠יךָ נֹתֵ֨ן לְךָ֤ נַחֲלָה֙ לְרִשְׁתָּ֔הּ תִּמְחֶה֙ אֶת־זֵ֣כֶר עֲמָלֵ֔ק מִתַּ֖חַת הַשָּׁמָ֑יִם לֹ֖א תִּשְׁכָּֽח׃ \petucha}{Therefore it shall be, when the \lord\space thy God hath given thee rest from all thine enemies round about, in the land which the \lord\space thy God giveth thee for an inheritance to possess it, that thou shalt blot out the remembrance of Amalek from under heaven; thou shalt not forget.}
\newhaftarah{פרשת זכור}
\label{haft_zachor}
\haftnote{The Haftarah for Shabbat Zakhor is I Samuel 15:2\verserangechar 34.}
\setcounter{chap}{15}
\setcounter{verse}{2}
\twoverse{\Roman{chap}:\arabic{verse}}{כֹּ֤ה אָמַר֙ יְהֹוָ֣ה צְבָא֔וֹת פָּקַ֕דְתִּי אֵ֛ת אֲשֶׁר־עָשָׂ֥ה עֲמָלֵ֖ק לְיִשְׂרָאֵ֑ל אֲשֶׁר־שָׂ֥ם לוֹ֙ בַּדֶּ֔רֶךְ בַּעֲלֹת֖וֹ מִמִּצְרָֽיִם׃}{Thus saith the \lord\space of hosts: I remember that which Amalek did to Israel, how he set himself against him in the way, when he came up out of Egypt.}
\twoverse{\arabic{verse}}{עַתָּה֩ לֵ֨ךְ וְהִכִּיתָ֜ה אֶת־עֲמָלֵ֗ק וְהַֽחֲרַמְתֶּם֙ אֶת־כׇּל־אֲשֶׁר־ל֔וֹ וְלֹ֥א תַחְמֹ֖ל עָלָ֑יו וְהֵמַתָּ֞ה מֵאִ֣ישׁ עַד־אִשָּׁ֗ה מֵֽעֹלֵל֙ וְעַד־יוֹנֵ֔ק מִשּׁ֣וֹר וְעַד־שֶׂ֔ה מִגָּמָ֖ל וְעַד־חֲמֽוֹר׃ \setuma }{Now go and smite Amalek, and utterly destroy all that they have, and spare them not; but slay both man and woman, infant and suckling, ox and sheep, camel and ass.’ .}
\twoverse{\arabic{verse}}{וַיְשַׁמַּ֤ע שָׁאוּל֙ אֶת־הָעָ֔ם וַֽיִּפְקְדֵם֙ בַּטְּלָאִ֔ים מָאתַ֥יִם אֶ֖לֶף רַגְלִ֑י וַעֲשֶׂ֥רֶת אֲלָפִ֖ים אֶת־אִ֥ישׁ יְהוּדָֽה׃}{And Saul summoned the people, and numbered them in Telaim, two hundred thousand footmen, and ten thousand men of Judah.}
\twoverse{\arabic{verse}}{וַיָּבֹ֥א שָׁא֖וּל עַד־עִ֣יר עֲמָלֵ֑ק וַיָּ֖רֶב בַּנָּֽחַל׃}{And Saul came to the city of Amalek, and lay in wait in the valley.}
\twoverse{\arabic{verse}}{וַיֹּ֣אמֶר שָׁא֣וּל אֶֽל־הַקֵּינִ֡י לְכוּ֩ סֻּ֨רוּ רְד֜וּ מִתּ֣וֹךְ עֲמָלֵקִ֗י פֶּן־אֹֽסִפְךָ֙ עִמּ֔וֹ וְאַתָּ֞ה עָשִׂ֤יתָֽה חֶ֙סֶד֙ עִם־כׇּל־בְּנֵ֣י יִשְׂרָאֵ֔ל בַּעֲלוֹתָ֖ם מִמִּצְרָ֑יִם וַיָּ֥סַר קֵינִ֖י מִתּ֥וֹךְ עֲמָלֵֽק׃}{And Saul said unto the Kenites: ‘Go, depart, get you down from among the Amalekites, lest I destroy you with them; for ye showed kindness to all the children of Israel, when they came up out of Egypt.’ So the Kenites departed from among the Amalekites.}
\twoverse{\arabic{verse}}{וַיַּ֥ךְ שָׁא֖וּל אֶת־עֲמָלֵ֑ק מֵֽחֲוִילָה֙ בּוֹאֲךָ֣ שׁ֔וּר אֲשֶׁ֖ר עַל־פְּנֵ֥י מִצְרָֽיִם׃}{And Saul smote the Amalekites, from Havilah as thou goest to Shur, that is in front of Egypt.}
\twoverse{\arabic{verse}}{וַיִּתְפֹּ֛שׂ אֶת־אֲגַ֥ג מֶלֶךְ־עֲמָלֵ֖ק חָ֑י וְאֶת־כׇּל־הָעָ֖ם הֶחֱרִ֥ים לְפִי־חָֽרֶב׃}{And he took Agag the king of the Amalekites alive, and utterly destroyed all the people with the edge of the sword.}
\twoverse{\arabic{verse}}{וַיַּחְמֹל֩ שָׁא֨וּל וְהָעָ֜ם עַל־אֲגָ֗ג וְעַל־מֵיטַ֣ב הַצֹּאן֩ וְהַבָּקָ֨ר וְהַמִּשְׁנִ֤ים וְעַל־הַכָּרִים֙ וְעַל־כׇּל־הַטּ֔וֹב וְלֹ֥א אָב֖וּ הַחֲרִימָ֑ם וְכׇל־הַמְּלָאכָ֛ה נְמִבְזָ֥ה וְנָמֵ֖ס אֹתָ֥הּ הֶחֱרִֽימוּ׃ \petucha}{But Saul and the people spared Agag, and the best of the sheep, and of the oxen, even the young of the second birth, and the lambs, and all that was good, and would not utterly destroy them; but every thing that was of no account and feeble, that they destroyed utterly.}
\twoverse{\arabic{verse}}{וַֽיְהִי֙ דְּבַר־יְהֹוָ֔ה אֶל־שְׁמוּאֵ֖ל לֵאמֹֽר׃}{Then came the word of the \lord\space unto Samuel, saying:}
\twoverse{\arabic{verse}}{נִחַ֗מְתִּי כִּֽי־הִמְלַ֤כְתִּי אֶת־שָׁאוּל֙ לְמֶ֔לֶךְ כִּי־שָׁב֙ מֵֽאַחֲרַ֔י וְאֶת־דְּבָרַ֖י לֹ֣א הֵקִ֑ים וַיִּ֙חַר֙ לִשְׁמוּאֵ֔ל וַיִּזְעַ֥ק אֶל־יְהֹוָ֖ה כׇּל־הַלָּֽיְלָה׃}{‘It repenteth Me that I have set up Saul to be king; for he is turned back from following Me, and hath not performed My commandments.’ And it grieved Samuel; and he cried unto the \lord\space all night.}
\twoverse{\arabic{verse}}{וַיַּשְׁכֵּ֧ם שְׁמוּאֵ֛ל לִקְרַ֥את שָׁא֖וּל בַּבֹּ֑קֶר וַיֻּגַּ֨ד לִשְׁמוּאֵ֜ל לֵאמֹ֗ר בָּֽא־שָׁא֤וּל הַכַּרְמֶ֙לָה֙ וְהִנֵּ֨ה מַצִּ֥יב לוֹ֙ יָ֔ד וַיִּסֹּב֙ וַֽיַּעֲבֹ֔ר וַיֵּ֖רֶד הַגִּלְגָּֽל׃}{And Samuel rose early to meet Saul in the morning; and it was told Samuel, saying: ‘Saul came to Carmel, and, behold, he is setting him up a monument, and is gone about, and passed on, and gone down to Gilgal.’}
\twoverse{\arabic{verse}}{וַיָּבֹ֥א שְׁמוּאֵ֖ל אֶל־שָׁא֑וּל וַיֹּ֧אמֶר ל֣וֹ שָׁא֗וּל בָּר֤וּךְ אַתָּה֙ לַֽיהֹוָ֔ה הֲקִימֹ֖תִי אֶת־דְּבַ֥ר יְהֹוָֽה׃}{And Samuel came to Saul; and Saul said unto him: ‘Blessed be thou of the \lord; I have performed the commandment of the \lord.’}
\twoverse{\arabic{verse}}{וַיֹּ֣אמֶר שְׁמוּאֵ֔ל וּמֶ֛ה קֽוֹל־הַצֹּ֥אן הַזֶּ֖ה בְּאׇזְנָ֑י וְק֣וֹל הַבָּקָ֔ר אֲשֶׁ֥ר אָנֹכִ֖י שֹׁמֵֽעַ׃}{And Samuel said: ‘What meaneth then this bleating of the sheep in mine ears, and the lowing of the oxen which I hear?’}
\twoverse{\arabic{verse}}{וַיֹּ֨אמֶר שָׁא֜וּל מֵעֲמָלֵקִ֣י הֱבִיא֗וּם אֲשֶׁ֨ר חָמַ֤ל הָעָם֙ עַל־מֵיטַ֤ב הַצֹּאן֙ וְהַבָּקָ֔ר לְמַ֥עַן זְבֹ֖חַ לַיהֹוָ֣ה אֱלֹהֶ֑יךָ וְאֶת־הַיּוֹתֵ֖ר הֶחֱרַֽמְנוּ׃ \petucha}{And Saul said: ‘They have brought them from the Amalekites; for the people spared the best of the sheep and of the oxen, to sacrifice unto the \lord\space thy God; and the rest we have utterly destroyed.’}
\twoverse{\arabic{verse}}{וַיֹּ֤אמֶר שְׁמוּאֵל֙ אֶל־שָׁא֔וּל הֶ֚רֶף וְאַגִּ֣ידָה לְּךָ֔ אֵת֩ אֲשֶׁ֨ר דִּבֶּ֧ר יְהֹוָ֛ה אֵלַ֖י הַלָּ֑יְלָה \qk{וַיֹּ֥אמֶר}{ויאמרו} ל֖וֹ דַּבֵּֽר׃ \setuma }{Then Samuel said unto Saul: ‘Stay, and I will tell thee what the \lord\space hath said to me this night.’ And he said unto him: ‘Say on.’}
\twoverse{\arabic{verse}}{וַיֹּ֣אמֶר שְׁמוּאֵ֔ל הֲל֗וֹא אִם־קָטֹ֤ן אַתָּה֙ בְּעֵינֶ֔יךָ רֹ֛אשׁ שִׁבְטֵ֥י יִשְׂרָאֵ֖ל אָ֑תָּה וַיִּמְשָׁחֲךָ֧ יְהֹוָ֛ה לְמֶ֖לֶךְ עַל־יִשְׂרָאֵֽל׃}{And Samuel said: ‘Though thou be little in thine own sight, art thou not head of the tribes of Israel? And the \lord\space anointed thee king over Israel;}
\twoverse{\arabic{verse}}{וַיִּשְׁלָחֲךָ֥ יְהֹוָ֖ה בְּדָ֑רֶךְ וַיֹּ֗אמֶר לֵ֣ךְ וְהַחֲרַמְתָּ֞ה אֶת־הַֽחַטָּאִים֙ אֶת־עֲמָלֵ֔ק וְנִלְחַמְתָּ֣ ב֔וֹ עַ֥ד כַּלּוֹתָ֖ם אֹתָֽם׃}{and the \lord\space sent thee on a journey, and said: Go and utterly destroy the sinners the Amalekites, and fight against them until they be consumed.}
\twoverse{\arabic{verse}}{וְלָ֥מָּה לֹֽא־שָׁמַ֖עְתָּ בְּק֣וֹל יְהֹוָ֑ה וַתַּ֙עַט֙ אֶל־הַשָּׁלָ֔ל וַתַּ֥עַשׂ הָרַ֖ע בְּעֵינֵ֥י יְהֹוָֽה׃ \setuma }{Wherefore then didst thou not hearken to the voice of the \lord, but didst fly upon the spoil, and didst that which was evil in the sight of the \lord?’}
\twoverse{\arabic{verse}}{וַיֹּ֨אמֶר שָׁא֜וּל אֶל־שְׁמוּאֵ֗ל אֲשֶׁ֤ר שָׁמַ֙עְתִּי֙ בְּק֣וֹל יְהֹוָ֔ה וָאֵלֵ֕ךְ בַּדֶּ֖רֶךְ אֲשֶׁר־שְׁלָחַ֣נִי יְהֹוָ֑ה וָאָבִ֗יא אֶת־אֲגַג֙ מֶ֣לֶךְ עֲמָלֵ֔ק וְאֶת־עֲמָלֵ֖ק הֶחֱרַֽמְתִּי׃}{And Saul said unto Samuel: ‘Yea, I have hearkened to the voice of the \lord, and have gone the way which the \lord\space sent me, and have brought Agag the king of Amalek, and have utterly destroyed the Amalekites.}
\twoverse{\arabic{verse}}{וַיִּקַּ֨ח הָעָ֧ם מֵהַשָּׁלָ֛ל צֹ֥אן וּבָקָ֖ר רֵאשִׁ֣ית הַחֵ֑רֶם לִזְבֹּ֛חַ לַיהֹוָ֥ה אֱלֹהֶ֖יךָ בַּגִּלְגָּֽל׃ \setuma }{But the people took of the spoil, sheep and oxen, the chief of the devoted things, to sacrifice unto the \lord\space thy God in Gilgal.’}
\twoverse{\arabic{verse}}{וַיֹּ֣אמֶר שְׁמוּאֵ֗ל הַחֵ֤פֶץ לַֽיהֹוָה֙ בְּעֹל֣וֹת וּזְבָחִ֔ים כִּשְׁמֹ֖עַ בְּק֣וֹל יְהֹוָ֑ה הִנֵּ֤ה שְׁמֹ֙עַ֙ מִזֶּ֣בַח ט֔וֹב לְהַקְשִׁ֖יב מֵחֵ֥לֶב אֵילִֽים׃}{And Samuel said: ‘Hath the \lord\space as great delight in burnt-offerings and sacrifices, As in hearkening to the voice of the \lord? Behold, to obey is better than sacrifice, And to hearken than the fat of rams.}
\twoverse{\arabic{verse}}{כִּ֤י חַטַּאת־קֶ֙סֶם֙ מֶ֔רִי וְאָ֥וֶן וּתְרָפִ֖ים הַפְצַ֑ר יַ֗עַן מָאַ֙סְתָּ֙ אֶת־דְּבַ֣ר יְהֹוָ֔ה וַיִּמְאָסְךָ֖ מִמֶּֽלֶךְ׃ \setuma }{For rebellion is as the sin of witchcraft, And stubbornness is as idolatry and teraphim. Because thou hast rejected the word of the \lord, He hath also rejected thee from being king.’}
\twoverse{\arabic{verse}}{וַיֹּ֨אמֶר שָׁא֤וּל אֶל־שְׁמוּאֵל֙ חָטָ֔אתִי כִּֽי־עָבַ֥רְתִּי אֶת־פִּֽי־יְהֹוָ֖ה וְאֶת־דְּבָרֶ֑יךָ כִּ֤י יָרֵ֙אתִי֙ אֶת־הָעָ֔ם וָאֶשְׁמַ֖ע בְּקוֹלָֽם׃}{And Saul said unto Samuel: ‘I have sinned; for I have transgressed the commandment of the \lord, and thy words; because I feared the people, and hearkened to their voice.}
\twoverse{\arabic{verse}}{וְעַתָּ֕ה שָׂ֥א נָ֖א אֶת־חַטָּאתִ֑י וְשׁ֣וּב עִמִּ֔י וְאֶֽשְׁתַּחֲוֶ֖ה לַיהֹוָֽה׃}{Now therefore, I pray thee, pardon my sin, and return with me, that I may worship the \lord.’}
\twoverse{\arabic{verse}}{וַיֹּ֤אמֶר שְׁמוּאֵל֙ אֶל־שָׁא֔וּל לֹ֥א אָשׁ֖וּב עִמָּ֑ךְ כִּ֤י מָאַ֙סְתָּה֙ אֶת־דְּבַ֣ר יְהוָ֔ה וַיִּמְאָסְךָ֣ יְהֹוָ֔ה מִֽהְי֥וֹת מֶ֖לֶךְ עַל־יִשְׂרָאֵֽל׃ \setuma }{And Samuel said unto Saul: ‘I will not return with thee; for thou hast rejected the word of the \lord, and the \lord\space hath rejected thee from being king over Israel.’}
\twoverse{\arabic{verse}}{וַיִּסֹּ֥ב שְׁמוּאֵ֖ל לָלֶ֑כֶת וַיַּחֲזֵ֥ק בִּכְנַף־מְעִיל֖וֹ וַיִּקָּרַֽע׃ \setuma }{And as Samuel turned about to go away, he laid hold upon the skirt of his robe, and it rent.}
\twoverse{\arabic{verse}}{וַיֹּ֤אמֶר אֵלָיו֙ שְׁמוּאֵ֔ל קָרַ֨ע יְהֹוָ֜ה אֶֽת־מַמְלְכ֧וּת יִשְׂרָאֵ֛ל מֵעָלֶ֖יךָ הַיּ֑וֹם וּנְתָנָ֕הּ לְרֵעֲךָ֖ הַטּ֥וֹב מִמֶּֽךָּ׃ \setuma }{And Samuel said unto him: ‘The \lord\space hath rent the kingdom of Israel from thee this day, and hath given it to a neighbour of thine, that is better than thou.}
\twoverse{\arabic{verse}}{וְגַם֙ נֵ֣צַח יִשְׂרָאֵ֔ל לֹ֥א יְשַׁקֵּ֖ר וְלֹ֣א יִנָּחֵ֑ם כִּ֣י לֹ֥א אָדָ֛ם ה֖וּא לְהִנָּחֵֽם׃}{And also the Glory of Israel will not lie nor repent; for He is not a man, that He should repent.’}
\twoverse{\arabic{verse}}{וַיֹּ֣אמֶר חָטָ֔אתִי עַתָּ֗ה כַּבְּדֵ֥נִי נָ֛א נֶ֥גֶד זִקְנֵֽי־עַמִּ֖י וְנֶ֣גֶד יִשְׂרָאֵ֑ל וְשׁ֣וּב עִמִּ֔י וְהִֽשְׁתַּחֲוֵ֖יתִי לַיהֹוָ֥ה אֱלֹהֶֽיךָ׃}{Then he said: ‘I have sinned; yet honour me now, I pray thee, before the elders of my people, and before Israel, and return with me, that I may worship the \lord\space thy God.’}
\twoverse{\arabic{verse}}{וַיָּ֥שׇׁב שְׁמוּאֵ֖ל אַחֲרֵ֣י שָׁא֑וּל וַיִּשְׁתַּ֥חוּ שָׁא֖וּל לַיהֹוָֽה׃ \setuma }{So Samuel returned after Saul; and Saul worshipped the \lord.}
\twoverse{\arabic{verse}}{וַיֹּ֣אמֶר שְׁמוּאֵ֗ל הַגִּ֤ישׁוּ אֵלַי֙ אֶת־אֲגַג֙ מֶ֣לֶךְ עֲמָלֵ֔ק וַיֵּ֣לֶךְ אֵלָ֔יו אֲגַ֖ג מַעֲדַנֹּ֑ת וַיֹּ֣אמֶר אֲגָ֔ג אָכֵ֖ן סָ֥ר מַר־הַמָּֽוֶת׃ \setuma }{Then said Samuel: ‘Bring ye hither to me Agag the king of the Amalekites.’ And Agag came unto him in chains. And Agag said: ‘Surely the bitterness of death is at hand.’}
\twoverse{\arabic{verse}}{וַיֹּ֣אמֶר שְׁמוּאֵ֔ל כַּאֲשֶׁ֨ר שִׁכְּלָ֤ה נָשִׁים֙ חַרְבֶּ֔ךָ כֵּן־תִּשְׁכַּ֥ל מִנָּשִׁ֖ים אִמֶּ֑ךָ וַיְשַׁסֵּ֨ף שְׁמוּאֵ֧ל אֶת־אֲגָ֛ג לִפְנֵ֥י יְהֹוָ֖ה בַּגִּלְגָּֽל׃ \setuma }{And Samuel said: As thy sword hath made women childless, So shall thy mother be childless among women. And Samuel hewed Agag in pieces before the \lord\space in Gilgal.}
\twoverse{\arabic{verse}}{וַיֵּ֥לֶךְ שְׁמוּאֵ֖ל הָרָמָ֑תָה וְשָׁא֛וּל עָלָ֥ה אֶל־בֵּית֖וֹ גִּבְעַ֥ת שָׁאֽוּל׃}{Then Samuel went to Ramah; and Saul went up to his house to Gibeath-shaul.}
\vspace{12pt}
\section*{פרשת פרה}
\addcontentsline{toc}{section}{פרשת פרה}
\renewcommand{\haftarahname}{פרשת פרה}
\vspace{-12pt}
\nopagebreak
\label{maft_parah}
\haftnote{The Maftir for Shabbat Parah is Numbers 19:1\verserangechar 22.}
\setcounter{chap}{19}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}{וַיְדַבֵּ֣ר יְהֹוָ֔ה אֶל־מֹשֶׁ֥ה וְאֶֽל־אַהֲרֹ֖ן לֵאמֹֽר׃}{And the \lord\space spoke unto Moses and unto Aaron, saying:}
\twoverse{\arabic{verse}}{זֹ֚את חֻקַּ֣ת הַתּוֹרָ֔ה אֲשֶׁר־צִוָּ֥ה יְהֹוָ֖ה לֵאמֹ֑ר דַּבֵּ֣ר \legarmeh  אֶל־בְּנֵ֣י יִשְׂרָאֵ֗ל וְיִקְח֣וּ אֵלֶ֩יךָ֩ פָרָ֨ה אֲדֻמָּ֜ה תְּמִימָ֗ה אֲשֶׁ֤ר אֵֽין־בָּהּ֙ מ֔וּם אֲשֶׁ֛ר לֹא־עָלָ֥ה עָלֶ֖יהָ עֹֽל׃}{This is the statute of the law which the \lord\space hath commanded, saying: Speak unto the children of Israel, that they bring thee a red heifer, faultless, wherein is no blemish, and upon which never came yoke.}
\twoverse{\arabic{verse}}{וּנְתַתֶּ֣ם אֹתָ֔הּ אֶל־אֶלְעָזָ֖ר הַכֹּהֵ֑ן וְהוֹצִ֤יא אֹתָהּ֙ אֶל־מִח֣וּץ לַֽמַּחֲנֶ֔ה וְשָׁחַ֥ט אֹתָ֖הּ לְפָנָֽיו׃}{And ye shall give her unto Eleazar the priest, and she shall be brought forth without the camp, and she shall be slain before his face.}
\twoverse{\arabic{verse}}{וְלָקַ֞ח אֶלְעָזָ֧ר הַכֹּהֵ֛ן מִדָּמָ֖הּ בְּאֶצְבָּע֑וֹ וְהִזָּ֞ה אֶל־נֹ֨כַח פְּנֵ֧י אֹֽהֶל־מוֹעֵ֛ד מִדָּמָ֖הּ שֶׁ֥בַע פְּעָמִֽים׃}{And Eleazar the priest shall take of her blood with his finger, and sprinkle of her blood toward the front of the tent of meeting seven times.}
\twoverse{\arabic{verse}}{וְשָׂרַ֥ף אֶת־הַפָּרָ֖ה לְעֵינָ֑יו אֶת־עֹרָ֤הּ וְאֶת־בְּשָׂרָהּ֙ וְאֶת־דָּמָ֔הּ עַל־פִּרְשָׁ֖הּ יִשְׂרֹֽף׃}{And the heifer shall be burnt in his sight; her skin, and her flesh, and her blood, with her dung, shall be burnt.}
\twoverse{\arabic{verse}}{וְלָקַ֣ח הַכֹּהֵ֗ן עֵ֥ץ אֶ֛רֶז וְאֵז֖וֹב וּשְׁנִ֣י תוֹלָ֑עַת וְהִשְׁלִ֕יךְ אֶל־תּ֖וֹךְ שְׂרֵפַ֥ת הַפָּרָֽה׃}{And the priest shall take cedar-wood, and hyssop, and scarlet, and cast it into the midst of the burning of the heifer.}
\twoverse{\arabic{verse}}{וְכִבֶּ֨ס בְּגָדָ֜יו הַכֹּהֵ֗ן וְרָחַ֤ץ בְּשָׂרוֹ֙ בַּמַּ֔יִם וְאַחַ֖ר יָבֹ֣א אֶל־הַֽמַּחֲנֶ֑ה וְטָמֵ֥א הַכֹּהֵ֖ן עַד־הָעָֽרֶב׃}{Then the priest shall wash his clothes, and he shall bathe his flesh in water, and afterward he may come into the camp, and the priest shall be unclean until the even.}
\twoverse{\arabic{verse}}{וְהַשֹּׂרֵ֣ף אֹתָ֔הּ יְכַבֵּ֤ס בְּגָדָיו֙ בַּמַּ֔יִם וְרָחַ֥ץ בְּשָׂר֖וֹ בַּמָּ֑יִם וְטָמֵ֖א עַד־הָעָֽרֶב׃}{And he that burneth her shall wash his clothes in water, and bathe his flesh in water, and shall be unclean until the even.}
\twoverse{\arabic{verse}}{וְאָסַ֣ף \legarmeh  אִ֣ישׁ טָה֗וֹר אֵ֚ת אֵ֣פֶר הַפָּרָ֔ה וְהִנִּ֛יחַ מִח֥וּץ לַֽמַּחֲנֶ֖ה בְּמָק֣וֹם טָה֑וֹר וְ֠הָיְתָ֠ה לַעֲדַ֨ת בְּנֵֽי־יִשְׂרָאֵ֧ל לְמִשְׁמֶ֛רֶת לְמֵ֥י נִדָּ֖ה חַטָּ֥את הִֽוא׃}{And a man that is clean shall gather up the ashes of the heifer, and lay them up without the camp in a clean place, and it shall be kept for the congregation of the children of Israel for a water of sprinkling; it is a purification from sin.}
\twoverse{\arabic{verse}}{וְ֠כִבֶּ֠ס הָאֹסֵ֨ף אֶת־אֵ֤פֶר הַפָּרָה֙ אֶת־בְּגָדָ֔יו וְטָמֵ֖א עַד־הָעָ֑רֶב וְֽהָיְתָ֞ה לִבְנֵ֣י יִשְׂרָאֵ֗ל וְלַגֵּ֛ר הַגָּ֥ר בְּתוֹכָ֖ם לְחֻקַּ֥ת עוֹלָֽם׃}{And he that gathereth the ashes of the heifer shall wash his clothes, and be unclean until the even; and it shall be unto the children of Israel, and unto the stranger that sojourneth among them, for a statute for ever.}
\twoverse{\arabic{verse}}{הַנֹּגֵ֥עַ בְּמֵ֖ת לְכׇל־נֶ֣פֶשׁ אָדָ֑ם וְטָמֵ֖א שִׁבְעַ֥ת יָמִֽים׃}{He that toucheth the dead, even any man’s dead body, shall be unclean seven days;}
\twoverse{\arabic{verse}}{ה֣וּא יִתְחַטָּא־ב֞וֹ בַּיּ֧וֹם הַשְּׁלִישִׁ֛י וּבַיּ֥וֹם הַשְּׁבִיעִ֖י יִטְהָ֑ר וְאִם־לֹ֨א יִתְחַטָּ֜א בַּיּ֧וֹם הַשְּׁלִישִׁ֛י וּבַיּ֥וֹם הַשְּׁבִיעִ֖י לֹ֥א יִטְהָֽר׃}{the same shall purify himself therewith on the third day and on the seventh day, and he shall be clean; but if he purify not himself the third day and the seventh day, he shall not be clean.}
\twoverse{\arabic{verse}}{כׇּֽל־הַנֹּגֵ֡עַ בְּמֵ֣ת בְּנֶ֩פֶשׁ֩ הָאָדָ֨ם אֲשֶׁר־יָמ֜וּת וְלֹ֣א יִתְחַטָּ֗א אֶת־מִשְׁכַּ֤ן יְהֹוָה֙ טִמֵּ֔א וְנִכְרְתָ֛ה הַנֶּ֥פֶשׁ הַהִ֖וא מִיִּשְׂרָאֵ֑ל כִּי֩ מֵ֨י נִדָּ֜ה לֹא־זֹרַ֤ק עָלָיו֙ טָמֵ֣א יִהְיֶ֔ה ע֖וֹד טֻמְאָת֥וֹ בֽוֹ׃}{Whosoever toucheth the dead, even the body of any man that is dead, and purifieth not himself—he hath defiled the tabernacle of the \lord—that soul shall be cut off from Israel; because the water of sprinkling was not dashed against him, he shall be unclean; his uncleanness is yet upon him.}
\twoverse{\arabic{verse}}{זֹ֚את הַתּוֹרָ֔ה אָדָ֖ם כִּֽי־יָמ֣וּת בְּאֹ֑הֶל כׇּל־הַבָּ֤א אֶל־הָאֹ֙הֶל֙ וְכׇל־אֲשֶׁ֣ר בָּאֹ֔הֶל יִטְמָ֖א שִׁבְעַ֥ת יָמִֽים׃}{This is the law: when a man dieth in a tent, every one that cometh into the tent, and every thing that is in the tent, shall be unclean seven days.}
\twoverse{\arabic{verse}}{וְכֹל֙ כְּלִ֣י פָת֔וּחַ אֲשֶׁ֛ר אֵין־צָמִ֥יד פָּתִ֖יל עָלָ֑יו טָמֵ֖א הֽוּא׃}{And every open vessel, which hath no covering close-bound upon it, is unclean.}
\twoverse{\arabic{verse}}{וְכֹ֨ל אֲשֶׁר־יִגַּ֜ע עַל־פְּנֵ֣י הַשָּׂדֶ֗ה בַּֽחֲלַל־חֶ֙רֶב֙ א֣וֹ בְמֵ֔ת אֽוֹ־בְעֶ֥צֶם אָדָ֖ם א֣וֹ בְקָ֑בֶר יִטְמָ֖א שִׁבְעַ֥ת יָמִֽים׃}{And whosoever in the open field toucheth one that is slain with a sword, or one that dieth of himself, or a bone of a man, or a grave, shall be unclean seven days.}
\twoverse{\arabic{verse}}{וְלָֽקְחוּ֙ לַטָּמֵ֔א מֵעֲפַ֖ר שְׂרֵפַ֣ת הַֽחַטָּ֑את וְנָתַ֥ן עָלָ֛יו מַ֥יִם חַיִּ֖ים אֶל־כֶּֽלִי׃}{And for the unclean they shall take of the ashes of the burning of the purification from sin, and running water shall be put thereto in a vessel.}
\twoverse{\arabic{verse}}{וְלָקַ֨ח אֵז֜וֹב וְטָבַ֣ל בַּמַּ֘יִם֮ אִ֣ישׁ טָהוֹר֒ וְהִזָּ֤ה עַל־הָאֹ֙הֶל֙ וְעַל־כׇּל־הַכֵּלִ֔ים וְעַל־הַנְּפָשׁ֖וֹת אֲשֶׁ֣ר הָֽיוּ־שָׁ֑ם וְעַל־הַנֹּגֵ֗עַ בַּעֶ֙צֶם֙ א֣וֹ בֶֽחָלָ֔ל א֥וֹ בַמֵּ֖ת א֥וֹ בַקָּֽבֶר׃}{And a clean person shall take hyssop, and dip it in the water, and sprinkle it upon the tent, and upon all the vessels, and upon the persons that were there, and upon him that touched the bone, or the slain, or the dead, or the grave.}
\twoverse{\arabic{verse}}{וְהִזָּ֤ה הַטָּהֹר֙ עַל־הַטָּמֵ֔א בַּיּ֥וֹם הַשְּׁלִישִׁ֖י וּבַיּ֣וֹם הַשְּׁבִיעִ֑י וְחִטְּאוֹ֙ בַּיּ֣וֹם הַשְּׁבִיעִ֔י וְכִבֶּ֧ס בְּגָדָ֛יו וְרָחַ֥ץ בַּמַּ֖יִם וְטָהֵ֥ר בָּעָֽרֶב׃}{And the clean person shall sprinkle upon the unclean on the third day, and on the seventh day; and on the seventh day he shall purify him; and he shall wash his clothes, and bathe himself in water, and shall be clean at even.}
\twoverse{\arabic{verse}}{וְאִ֤ישׁ אֲשֶׁר־יִטְמָא֙ וְלֹ֣א יִתְחַטָּ֔א וְנִכְרְתָ֛ה הַנֶּ֥פֶשׁ הַהִ֖וא מִתּ֣וֹךְ הַקָּהָ֑ל כִּי֩ אֶת־מִקְדַּ֨שׁ יְהֹוָ֜ה טִמֵּ֗א מֵ֥י נִדָּ֛ה לֹא־זֹרַ֥ק עָלָ֖יו טָמֵ֥א הֽוּא׃}{But the man that shall be unclean, and shall not purify himself, that soul shall be cut off from the midst of the assembly, because he hath defiled the sanctuary of the \lord; the water of sprinkling hath not been dashed against him: he is unclean.}
\twoverse{\arabic{verse}}{וְהָיְתָ֥ה לָהֶ֖ם לְחֻקַּ֣ת עוֹלָ֑ם וּמַזֵּ֤ה מֵֽי־הַנִּדָּה֙ יְכַבֵּ֣ס בְּגָדָ֔יו וְהַנֹּגֵ֙עַ֙ בְּמֵ֣י הַנִּדָּ֔ה יִטְמָ֖א עַד־הָעָֽרֶב׃}{And it shall be a perpetual statute unto them; and he that sprinkleth the water of sprinkling shall wash his clothes; and he that toucheth the water of sprinkling shall be unclean until even.}
\twoverse{\arabic{verse}}{וְכֹ֛ל אֲשֶׁר־יִגַּע־בּ֥וֹ הַטָּמֵ֖א יִטְמָ֑א וְהַנֶּ֥פֶשׁ הַנֹּגַ֖עַת תִּטְמָ֥א עַד־הָעָֽרֶב׃ \petucha}{And whatsoever the unclean person toucheth shall be unclean; and the soul that toucheth him shall be unclean until even.}
\newhaftarah{פרשת פרה}
\label{haft_parah}
\haftnote{The Haftarah for Shabbat Parah is Ezekiel 36:16\verserangechar 38.}
\setcounter{chap}{36}
\setcounter{verse}{16}
\twoverse{\Roman{chap}:\arabic{verse}}{וַיְהִ֥י דְבַר־יְהֹוָ֖ה אֵלַ֥י לֵאמֹֽר׃}{Moreover the word of the \lord\space came unto me, saying:}
\twoverse{\arabic{verse}}{בֶּן־אָדָ֗ם בֵּ֤ית יִשְׂרָאֵל֙ יֹשְׁבִ֣ים עַל־אַדְמָתָ֔ם וַיְטַמְּא֣וּ אוֹתָ֔הּ בְּדַרְכָּ֖ם וּבַעֲלִֽילוֹתָ֑ם כְּטֻמְאַת֙ הַנִּדָּ֔ה הָיְתָ֥ה דַרְכָּ֖ם לְפָנָֽי׃}{‘Son of man, when the house of Israel dwelt in their own land, they defiled it by their way and by their doings; their way before Me was as the uncleanness of a woman in her impurity.}
\twoverse{\arabic{verse}}{וָאֶשְׁפֹּ֤ךְ חֲמָתִי֙ עֲלֵיהֶ֔ם עַל־הַדָּ֖ם אֲשֶׁר־שָׁפְכ֣וּ עַל־הָאָ֑רֶץ וּבְגִלּוּלֵיהֶ֖ם טִמְּאֽוּהָ׃}{Wherefore I poured out My fury upon them for the blood which they had shed upon the land, and because they had defiled it with their idols;}
\twoverse{\arabic{verse}}{וָאָפִ֤יץ אֹתָם֙ בַּגּוֹיִ֔ם וַיִּזָּר֖וּ בָּאֲרָצ֑וֹת כְּדַרְכָּ֥ם וְכַעֲלִילוֹתָ֖ם שְׁפַטְתִּֽים׃}{and I scattered them among the nations, and they were dispersed through the countries; according to their way and according to their doings I judged them.}
\twoverse{\arabic{verse}}{וַיָּב֗וֹא אֶל־הַגּוֹיִם֙ אֲשֶׁר־בָּ֣אוּ שָׁ֔ם וַֽיְחַלְּל֖וּ אֶת־שֵׁ֣ם קׇדְשִׁ֑י בֶּאֱמֹ֤ר לָהֶם֙ עַם־יְהֹוָ֣ה אֵ֔לֶּה וּמֵאַרְצ֖וֹ יָצָֽאוּ׃}{And when they came unto the nations, whither they came, they profaned My holy name; in that men said of them: These are the people of the \lord, and are gone forth out of His land.}
\twoverse{\arabic{verse}}{וָאֶחְמֹ֖ל עַל־שֵׁ֣ם קׇדְשִׁ֑י אֲשֶׁ֤ר חִלְּלֻ֙הוּ֙ בֵּ֣ית יִשְׂרָאֵ֔ל בַּגּוֹיִ֖ם אֲשֶׁר־בָּ֥אוּ שָֽׁמָּה׃ \petucha}{But I had pity for My holy name, which the house of Israel had profaned among the nations, whither they came.}
\twoverse{\arabic{verse}}{לָכֵ֞ן אֱמֹ֣ר לְבֵֽית־יִשְׂרָאֵ֗ל כֹּ֤ה אָמַר֙ אֲדֹנָ֣י יֱהֹוִ֔ה לֹ֧א לְמַעַנְכֶ֛ם אֲנִ֥י עֹשֶׂ֖ה בֵּ֣ית יִשְׂרָאֵ֑ל כִּ֤י אִם־לְשֵׁם־קׇדְשִׁי֙ אֲשֶׁ֣ר חִלַּלְתֶּ֔ם בַּגּוֹיִ֖ם אֲשֶׁר־בָּ֥אתֶם שָֽׁם׃}{Therefore say unto the house of Israel: Thus saith the Lord \textsc{God}: I do not this for your sake, O house of Israel, but for My holy name, which ye have profaned among the nations, whither ye came.}
\twoverse{\arabic{verse}}{וְקִדַּשְׁתִּ֞י אֶת־שְׁמִ֣י הַגָּד֗וֹל הַֽמְחֻלָּל֙ בַּגּוֹיִ֔ם אֲשֶׁ֥ר חִלַּלְתֶּ֖ם בְּתוֹכָ֑ם וְיָדְע֨וּ הַגּוֹיִ֜ם כִּֽי־אֲנִ֣י יְהֹוָ֗ה נְאֻם֙ אֲדֹנָ֣י יֱהֹוִ֔ה בְּהִקָּדְשִׁ֥י בָכֶ֖ם לְעֵינֵיהֶֽם׃}{And I will sanctify My great name, which hath been profaned among the nations, which ye have profaned in the midst of them; and the nations shall know that I am the \lord, saith the Lord \textsc{God}, when I shall be sanctified in you before their eyes.}
\twoverse{\arabic{verse}}{וְלָקַחְתִּ֤י אֶתְכֶם֙ מִן־הַגּוֹיִ֔ם וְקִבַּצְתִּ֥י אֶתְכֶ֖ם מִכׇּל־הָאֲרָצ֑וֹת וְהֵבֵאתִ֥י אֶתְכֶ֖ם אֶל־אַדְמַתְכֶֽם׃}{For I will take you from among the nations, and gather you out of all the countries, and will bring you into your own land.}
\twoverse{\arabic{verse}}{וְזָרַקְתִּ֧י עֲלֵיכֶ֛ם מַ֥יִם טְהוֹרִ֖ים וּטְהַרְתֶּ֑ם מִכֹּ֧ל טֻמְאוֹתֵיכֶ֛ם וּמִכׇּל־גִּלּ֥וּלֵיכֶ֖ם אֲטַהֵ֥ר אֶתְכֶֽם׃}{And I will sprinkle clean water upon you, and ye shall be clean; from all your uncleannesses, and from all your idols, will I cleanse you.}
\twoverse{\arabic{verse}}{וְנָתַתִּ֤י לָכֶם֙ לֵ֣ב חָדָ֔שׁ וְר֥וּחַ חֲדָשָׁ֖ה אֶתֵּ֣ן בְּקִרְבְּכֶ֑ם וַהֲסִ֨רֹתִ֜י אֶת־לֵ֤ב הָאֶ֙בֶן֙ מִבְּשַׂרְכֶ֔ם וְנָתַתִּ֥י לָכֶ֖ם לֵ֥ב בָּשָֽׂר׃}{A new heart also will I give you, and a new spirit will I put within you; and I will take away the stony heart out of your flesh, and I will give you a heart of flesh.}
\twoverse{\arabic{verse}}{וְאֶת־רוּחִ֖י אֶתֵּ֣ן בְּקִרְבְּכֶ֑ם וְעָשִׂ֗יתִי אֵ֤ת אֲשֶׁר־בְּחֻקַּי֙ תֵּלֵ֔כוּ וּמִשְׁפָּטַ֥י תִּשְׁמְר֖וּ וַעֲשִׂיתֶֽם׃}{And I will put My spirit within you, and cause you to walk in My statutes, and ye shall keep Mine ordinances, and do them.}
\twoverse{\arabic{verse}}{וִישַׁבְתֶּ֣ם בָּאָ֔רֶץ אֲשֶׁ֥ר נָתַ֖תִּי לַאֲבֹֽתֵיכֶ֑ם וִהְיִ֤יתֶם לִי֙ לְעָ֔ם וְאָ֣נֹכִ֔י אֶהְיֶ֥ה לָכֶ֖ם לֵאלֹהִֽים׃}{And ye shall dwell in the land that I gave to your fathers; and ye shall be My people, and I will be your God.}
\twoverse{\arabic{verse}}{וְהוֹשַׁעְתִּ֣י אֶתְכֶ֔ם מִכֹּ֖ל טֻמְאֽוֹתֵיכֶ֑ם וְקָרָ֤אתִי אֶל־הַדָּגָן֙ וְהִרְבֵּיתִ֣י אֹת֔וֹ וְלֹא־אֶתֵּ֥ן עֲלֵיכֶ֖ם רָעָֽב׃}{And I will save you from all your uncleannesses; and I will call for the corn, and will increase it, and lay no famine upon you.}
\twoverse{\arabic{verse}}{וְהִרְבֵּיתִי֙ אֶת־פְּרִ֣י הָעֵ֔ץ וּתְנוּבַ֖ת הַשָּׂדֶ֑ה לְמַ֗עַן אֲ֠שֶׁ֠ר לֹ֣א תִקְח֥וּ ע֛וֹד חֶרְפַּ֥ת רָעָ֖ב בַּגּוֹיִֽם׃}{And I will multiply the fruit of the tree, and the increase of the field, that ye may receive no more the reproach of famine among the nations.}
\twoverse{\arabic{verse}}{וּזְכַרְתֶּם֙ אֶת־דַּרְכֵיכֶ֣ם הָרָעִ֔ים וּמַעַלְלֵיכֶ֖ם אֲשֶׁ֣ר לֹֽא־טוֹבִ֑ים וּנְקֹֽטֹתֶם֙ בִּפְנֵיכֶ֔ם עַ֚ל עֲוֺנֹ֣תֵיכֶ֔ם וְעַ֖ל תּוֹעֲבוֹתֵיכֶֽם׃}{Then shall ye remember your evil ways, and your doings that were not good; and ye shall loathe yourselves in your own sight for your iniquities and for your abominations.}
\twoverse{\arabic{verse}}{לֹ֧א לְמַעַנְכֶ֣ם אֲנִֽי־עֹשֶׂ֗ה נְאֻם֙ אֲדֹנָ֣י יֱהֹוִ֔ה יִוָּדַ֖ע לָכֶ֑ם בּ֧וֹשׁוּ וְהִכָּֽלְמ֛וּ מִדַּרְכֵיכֶ֖ם בֵּ֥ית יִשְׂרָאֵֽל׃ \setuma }{Not for your sake do I this, saith the Lord \textsc{God}, be it known unto you; be ashamed and confounded for your ways, O house of Israel.}
\twoverse{\arabic{verse}}{כֹּ֤ה אָמַר֙ אֲדֹנָ֣י יֱהֹוִ֔ה בְּיוֹם֙ טַהֲרִ֣י אֶתְכֶ֔ם מִכֹּ֖ל עֲוֺנֽוֹתֵיכֶ֑ם וְהֽוֹשַׁבְתִּי֙ אֶת־הֶ֣עָרִ֔ים וְנִבְנ֖וּ הֶחֳרָבֽוֹת׃}{Thus saith the Lord \textsc{God}: In the day that I cleanse you from all your iniquities, I will cause the cities to be inhabited, and the waste places shall be builded.}
\twoverse{\arabic{verse}}{וְהָאָ֥רֶץ הַנְּשַׁמָּ֖ה תֵּעָבֵ֑ד תַּ֚חַת אֲשֶׁ֣ר הָיְתָ֣ה שְׁמָמָ֔ה לְעֵינֵ֖י כׇּל־עוֹבֵֽר׃}{And the land that was desolate shall be tilled, whereas it was a desolation in the sight of all that passed by.}
\twoverse{\arabic{verse}}{וְאָמְר֗וּ הָאָ֤רֶץ הַלֵּ֙זוּ֙ הַנְּשַׁמָּ֔ה הָיְתָ֖ה כְּגַן־עֵ֑דֶן וְהֶעָרִ֧ים הֶחֳרֵב֛וֹת וְהַֽנְשַׁמּ֥וֹת וְהַנֶּהֱרָס֖וֹת בְּצוּר֥וֹת יָשָֽׁבוּ׃}{And they shall say: This land that was desolate is become like the garden of Eden; and the waste and desolate and ruined cities are fortified and inhabited.}
\twoverse{\arabic{verse}}{וְיָדְע֣וּ הַגּוֹיִ֗ם אֲשֶׁ֣ר יִֽשָּׁאֲרוּ֮ סְבִיבֽוֹתֵיכֶם֒ כִּ֣י \legarmeh  אֲנִ֣י יְהֹוָ֗ה בָּנִ֙יתִי֙ הַנֶּ֣הֱרָס֔וֹת נָטַ֖עְתִּי הַנְּשַׁמָּ֑ה אֲנִ֥י יְהֹוָ֖ה דִּבַּ֥רְתִּי וְעָשִֽׂיתִי׃ \setuma }{Then the nations that are left round about you shall know that I the \lord\space have builded the ruined places, and planted that which was desolate; I the \lord\space have spoken it, and I will do it.}
\twoverse{\arabic{verse}}{כֹּ֤ה אָמַר֙ אֲדֹנָ֣י יֱהֹוִ֔ה ע֗וֹד זֹ֛את אִדָּרֵ֥שׁ לְבֵֽית־יִשְׂרָאֵ֖ל לַעֲשׂ֣וֹת לָהֶ֑ם אַרְבֶּ֥ה אֹתָ֛ם כַּצֹּ֖אן אָדָֽם׃}{Thus saith the Lord \textsc{God}: I will yet for this be inquired of by the house of Israel, to do it for them; I will increase them with men like a flock.}
\twoverse{\arabic{verse}}{כְּצֹ֣אן קׇדָשִׁ֗ים כְּצֹ֤אן יְרוּשָׁלַ֙͏ִם֙ בְּמ֣וֹעֲדֶ֔יהָ כֵּ֤ן תִּֽהְיֶ֙ינָה֙ הֶעָרִ֣ים הֶחֳרֵב֔וֹת מְלֵא֖וֹת צֹ֣אן אָדָ֑ם וְיָדְע֖וּ כִּֽי־אֲנִ֥י יְהֹוָֽה׃ \petucha}{As the flock for sacrifice, as the flock of Jerusalem in her appointed seasons, so shall the waste cities be filled with flocks of men; and they shall know that I am the \lord.’}
\vspace{12pt}
\section*{פרשת החודש}
\addcontentsline{toc}{section}{מפטיר לפרשת החודש}
\renewcommand{\haftarahname}{מפטיר לפרשת החודש}
\haftnote{The Maftir for Shabbat Ha\d{H}odesh is Exodus 12:1\verserangechar 20.}
\label{maft_hachodesh}
\setcounter{chap}{12}
\setcounter{verse}{1}
\twoverse{\Roman{chap}:\arabic{verse}}
{וַיֹּ֤אמֶר יְהֹוָה֙ אֶל־מֹשֶׁ֣ה וְאֶֽל־אַהֲרֹ֔ן בְּאֶ֥רֶץ מִצְרַ֖יִם לֵאמֹֽר׃}
{And the \lord\space spoke unto Moses and Aaron in the land of Egypt, saying:}
\twoverse{\arabic{verse}}
{הַחֹ֧דֶשׁ הַזֶּ֛ה לָכֶ֖ם רֹ֣אשׁ חֳדָשִׁ֑ים רִאשׁ֥וֹן הוּא֙ לָכֶ֔ם לְחׇדְשֵׁ֖י הַשָּׁנָֽה׃}
{‘This month shall be unto you the beginning of months; it shall be the first month of the year to you.}
\twoverse{\arabic{verse}}
{דַּבְּר֗וּ אֶֽל־כׇּל־עֲדַ֤ת יִשְׂרָאֵל֙ לֵאמֹ֔ר בֶּעָשֹׂ֖ר לַחֹ֣דֶשׁ הַזֶּ֑ה וְיִקְח֣וּ לָהֶ֗ם אִ֛ישׁ שֶׂ֥ה לְבֵית־אָבֹ֖ת שֶׂ֥ה לַבָּֽיִת׃}
{Speak ye unto all the congregation of Israel, saying: In the tenth day of this month they shall take to them every man a lamb, according to their fathers’ houses, a lamb for a household;}
\twoverse{\arabic{verse}}
{וְאִם־יִמְעַ֣ט הַבַּ֘יִת֮ מִהְי֣וֹת מִשֶּׂה֒ וְלָקַ֣ח ה֗וּא וּשְׁכֵנ֛וֹ הַקָּרֹ֥ב אֶל־בֵּית֖וֹ בְּמִכְסַ֣ת נְפָשֹׁ֑ת אִ֚ישׁ לְפִ֣י אׇכְל֔וֹ תָּכֹ֖סּוּ עַל־הַשֶּֽׂה׃}
{and if the household be too little for a lamb, then shall he and his neighbour next unto his house take one according to the number of the souls; according to every man’s eating ye shall make your count for the lamb.}
\twoverse{\arabic{verse}}
{שֶׂ֥ה תָמִ֛ים זָכָ֥ר בֶּן־שָׁנָ֖ה יִהְיֶ֣ה לָכֶ֑ם מִן־הַכְּבָשִׂ֥ים וּמִן־הָעִזִּ֖ים תִּקָּֽחוּ׃}
{Your lamb shall be without blemish, a male of the first year; ye shall take it from the sheep, or from the goats;}
\twoverse{\arabic{verse}}
{וְהָיָ֤ה לָכֶם֙ לְמִשְׁמֶ֔רֶת עַ֣ד אַרְבָּעָ֥ה עָשָׂ֛ר י֖וֹם לַחֹ֣דֶשׁ הַזֶּ֑ה וְשָׁחֲט֣וּ אֹת֗וֹ כֹּ֛ל קְהַ֥ל עֲדַֽת־יִשְׂרָאֵ֖ל בֵּ֥ין הָעַרְבָּֽיִם׃}
{and ye shall keep it unto the fourteenth day of the same month; and the whole assembly of the congregation of Israel shall kill it at dusk.}
\twoverse{\arabic{verse}}
{וְלָֽקְחוּ֙ מִן־הַדָּ֔ם וְנָ֥תְנ֛וּ עַל־שְׁתֵּ֥י הַמְּזוּזֹ֖ת וְעַל־הַמַּשְׁק֑וֹף עַ֚ל הַבָּ֣תִּ֔ים אֲשֶׁר־יֹאכְל֥וּ אֹת֖וֹ בָּהֶֽם׃}
{And they shall take of the blood, and put it on the two side-posts and on the lintel, upon the houses wherein they shall eat it.}
\twoverse{\arabic{verse}}
{וְאָכְל֥וּ אֶת־הַבָּשָׂ֖ר בַּלַּ֣יְלָה הַזֶּ֑ה צְלִי־אֵ֣שׁ וּמַצּ֔וֹת עַל־מְרֹרִ֖ים יֹאכְלֻֽהוּ׃}
{And they shall eat the flesh in that night, roast with fire, and unleavened bread; with bitter herbs they shall eat it.}
\twoverse{\arabic{verse}}
{אַל־תֹּאכְל֤וּ מִמֶּ֙נּוּ֙ נָ֔א וּבָשֵׁ֥ל מְבֻשָּׁ֖ל בַּמָּ֑יִם כִּ֣י אִם־צְלִי־אֵ֔שׁ רֹאשׁ֥וֹ עַל־כְּרָעָ֖יו וְעַל־קִרְבּֽוֹ׃}
{Eat not of it raw, nor sodden at all with water, but roast with fire; its head with its legs and with the inwards thereof.}
\twoverse{\arabic{verse}}
{וְלֹא־תוֹתִ֥ירוּ מִמֶּ֖נּוּ עַד־בֹּ֑קֶר וְהַנֹּתָ֥ר מִמֶּ֛נּוּ עַד־בֹּ֖קֶר בָּאֵ֥שׁ תִּשְׂרֹֽפוּ׃}
{And ye shall let nothing of it remain until the morning; but that which remaineth of it until the morning ye shall burn with fire.}
\twoverse{\arabic{verse}}
{וְכָ֘כָה֮ תֹּאכְל֣וּ אֹתוֹ֒ מׇתְנֵיכֶ֣ם חֲגֻרִ֔ים נַֽעֲלֵיכֶם֙ בְּרַגְלֵיכֶ֔ם וּמַקֶּלְכֶ֖ם בְּיֶדְכֶ֑ם וַאֲכַלְתֶּ֤ם אֹתוֹ֙ בְּחִפָּז֔וֹן פֶּ֥סַח ה֖וּא לַיהֹוָֽה׃}
{And thus shall ye eat it: with your loins girded, your shoes on your feet, and your staff in your hand; and ye shall eat it in haste—it is the \lord’s passover.}
\twoverse{\arabic{verse}}
{וְעָבַרְתִּ֣י בְאֶֽרֶץ־מִצְרַ֘יִם֮ בַּלַּ֣יְלָה הַזֶּה֒ וְהִכֵּיתִ֤י כׇל־בְּכוֹר֙ בְּאֶ֣רֶץ מִצְרַ֔יִם מֵאָדָ֖ם וְעַד־בְּהֵמָ֑ה וּבְכׇל־אֱלֹהֵ֥י מִצְרַ֛יִם אֶֽעֱשֶׂ֥ה שְׁפָטִ֖ים אֲנִ֥י יְהֹוָֽה׃}
{For I will go through the land of Egypt in that night, and will smite all the first-born in the land of Egypt, both man and beast; and against all the gods of Egypt I will execute judgments: I am the \lord.}
\twoverse{\arabic{verse}}
{וְהָיָה֩ הַדָּ֨ם לָכֶ֜ם לְאֹ֗ת עַ֤ל הַבָּתִּים֙ אֲשֶׁ֣ר אַתֶּ֣ם שָׁ֔ם וְרָאִ֙יתִי֙ אֶת־הַדָּ֔ם וּפָסַחְתִּ֖י עֲלֵכֶ֑ם וְלֹֽא־יִֽהְיֶ֨ה בָכֶ֥ם נֶ֙גֶף֙ לְמַשְׁחִ֔ית בְּהַכֹּתִ֖י בְּאֶ֥רֶץ מִצְרָֽיִם׃}
{And the blood shall be to you for a token upon the houses where ye are; and when I see the blood, I will pass over you, and there shall no plague be upon you to destroy you, when I smite the land of Egypt.}
\twoverse{\arabic{verse}}
{וְהָיָה֩ הַיּ֨וֹם הַזֶּ֤ה לָכֶם֙ לְזִכָּר֔וֹן וְחַגֹּתֶ֥ם אֹת֖וֹ חַ֣ג לַֽיהֹוָ֑ה לְדֹרֹ֣תֵיכֶ֔ם חֻקַּ֥ת עוֹלָ֖ם תְּחׇגֻּֽהוּ׃}
{And this day shall be unto you for a memorial, and ye shall keep it a feast to the \lord; throughout your generations ye shall keep it a feast by an ordinance for ever.}
\twoverse{\arabic{verse}}
{שִׁבְעַ֤ת יָמִים֙ מַצּ֣וֹת תֹּאכֵ֔לוּ אַ֚ךְ בַּיּ֣וֹם הָרִאשׁ֔וֹן תַּשְׁבִּ֥יתוּ שְּׂאֹ֖ר מִבָּתֵּיכֶ֑ם כִּ֣י \legarmeh  כׇּל־אֹכֵ֣ל חָמֵ֗ץ וְנִכְרְתָ֞ה הַנֶּ֤פֶשׁ הַהִוא֙ מִיִּשְׂרָאֵ֔ל מִיּ֥וֹם הָרִאשֹׁ֖ן עַד־י֥וֹם הַשְּׁבִעִֽי׃}
{Seven days shall ye eat unleavened bread; howbeit the first day ye shall put away leaven out of your houses; for whosoever eateth leavened bread from the first day until the seventh day, that soul shall be cut off from Israel.}
\twoverse{\arabic{verse}}
{וּבַיּ֤וֹם הָרִאשׁוֹן֙ מִקְרָא־קֹ֔דֶשׁ וּבַיּוֹם֙ הַשְּׁבִיעִ֔י מִקְרָא־קֹ֖דֶשׁ יִהְיֶ֣ה לָכֶ֑ם כׇּל־מְלָאכָה֙ לֹא־יֵעָשֶׂ֣ה בָהֶ֔ם אַ֚ךְ אֲשֶׁ֣ר יֵאָכֵ֣ל לְכׇל־נֶ֔פֶשׁ ה֥וּא לְבַדּ֖וֹ יֵעָשֶׂ֥ה לָכֶֽם׃}
{And in the first day there shall be to you a holy convocation, and in the seventh day a holy convocation; no manner of work shall be done in them, save that which every man must eat, that only may be done by you.}
\twoverse{\arabic{verse}}
{וּשְׁמַרְתֶּם֮ אֶת־הַמַּצּוֹת֒ כִּ֗י בְּעֶ֙צֶם֙ הַיּ֣וֹם הַזֶּ֔ה הוֹצֵ֥אתִי אֶת־צִבְאוֹתֵיכֶ֖ם מֵאֶ֣רֶץ מִצְרָ֑יִם וּשְׁמַרְתֶּ֞ם אֶת־הַיּ֥וֹם הַזֶּ֛ה לְדֹרֹתֵיכֶ֖ם חֻקַּ֥ת עוֹלָֽם׃}
{And ye shall observe the feast of unleavened bread; for in this selfsame day have I brought your hosts out of the land of Egypt; therefore shall ye observe this day throughout your generations by an ordinance for ever.}
\twoverse{\arabic{verse}}
{בָּרִאשֹׁ֡ן בְּאַרְבָּעָה֩ עָשָׂ֨ר י֤וֹם לַחֹ֙דֶשׁ֙ בָּעֶ֔רֶב תֹּאכְל֖וּ מַצֹּ֑ת עַ֠ד י֣וֹם הָאֶחָ֧ד וְעֶשְׂרִ֛ים לַחֹ֖דֶשׁ בָּעָֽרֶב׃}
{In the first month, on the fourteenth day of the month at even, ye shall eat unleavened bread, until the one and twentieth day of the month at even.}
\twoverse{\arabic{verse}}
{שִׁבְעַ֣ת יָמִ֔ים שְׂאֹ֕ר לֹ֥א יִמָּצֵ֖א בְּבָתֵּיכֶ֑ם כִּ֣י \legarmeh  כׇּל־אֹכֵ֣ל מַחְמֶ֗צֶת וְנִכְרְתָ֞ה הַנֶּ֤פֶשׁ הַהִוא֙ מֵעֲדַ֣ת יִשְׂרָאֵ֔ל בַּגֵּ֖ר וּבְאֶזְרַ֥ח הָאָֽרֶץ׃}
{Seven days shall there be no leaven found in your houses; for whosoever eateth that which is leavened, that soul shall be cut off from the congregation of Israel, whether he be a sojourner, or one that is born in the land.}
\twoverse{\arabic{verse}}
{כׇּל־מַחְמֶ֖צֶת לֹ֣א תֹאכֵ֑לוּ בְּכֹל֙ מוֹשְׁבֹ֣תֵיכֶ֔ם תֹּאכְל֖וּ מַצּֽוֹת׃ \petucha }
{Ye shall eat nothing leavened; in all your habitations shall ye eat unleavened bread.’}
\newhaftarah{פרשת החודש}
\label{haft_hachodesh}
\haftnote{The Haftarah for Shabbat Ha\d{H}odesh is Ezekiel 45:16\verserangechar 46:18.}
\setcounter{chap}{45}
\setcounter{verse}{16}
\twoverse{\Roman{chap}:\arabic{verse}}{כֹּ֚ל הָעָ֣ם הָאָ֔רֶץ יִהְי֖וּ אֶל־הַתְּרוּמָ֣ה הַזֹּ֑את לַנָּשִׂ֖יא בְּיִשְׂרָאֵֽל׃}{All the people of the land shall give this offering for the prince in Israel.}
\twoverse{\arabic{verse}}{וְעַֽל־הַנָּשִׂ֣יא יִֽהְיֶ֗ה הָעוֹל֣וֹת וְהַמִּנְחָה֮ וְהַנֵּ֒סֶךְ֒ בַּחַגִּ֤ים וּבֶחֳדָשִׁים֙ וּבַשַּׁבָּת֔וֹת בְּכׇֽל־מוֹעֲדֵ֖י בֵּ֣ית יִשְׂרָאֵ֑ל הֽוּא־יַעֲשֶׂ֞ה אֶת־הַחַטָּ֣את וְאֶת־הַמִּנְחָ֗ה וְאֶת־הָעוֹלָה֙ וְאֶת־הַשְּׁלָמִ֔ים לְכַפֵּ֖ר בְּעַ֥ד בֵּֽית־יִשְׂרָאֵֽל׃ \setuma }{And it shall be the prince’s part to give the burnt-offerings, and the meal-offerings, and the drink-offerings, in the feasts, and in the new moons, and in the sabbaths, in all the appointed seasons of the house of Israel; he shall prepare the sin-offering, and the meal-offering, and the burnt-offering, and the peace-offerings, to make atonement for the house of Israel.}
\twoverse{\arabic{verse}}{כֹּה־אָמַר֮ אֲדֹנָ֣י יֱהֹוִה֒ בָּֽרִאשׁוֹן֙ בְּאֶחָ֣ד לַחֹ֔דֶשׁ תִּקַּ֥ח פַּר־בֶּן־בָּקָ֖ר תָּמִ֑ים וְחִטֵּאתָ֖ אֶת־הַמִּקְדָּֽשׁ׃}{Thus saith the Lord \textsc{God}: In the first month, in the first day of the month, thou shalt take a young bullock without blemish; and thou shalt purify the sanctuary.}
\twoverse{\arabic{verse}}{וְלָקַ֨ח הַכֹּהֵ֜ן מִדַּ֣ם הַחַטָּ֗את וְנָתַן֙ אֶל־מְזוּזַ֣ת הַבַּ֔יִת וְאֶל־אַרְבַּ֛ע פִּנּ֥וֹת הָעֲזָרָ֖ה לַמִּזְבֵּ֑חַ וְעַ֨ל־מְזוּזַ֔ת שַׁ֖עַר הֶחָצֵ֥ר הַפְּנִימִֽית׃}{And the priest shall take of the blood of the sin-offering, and put it upon the door-posts of the house, and upon the four corners of the settle of the altar, and upon the posts of the gate of the inner court.}
\twoverse{\arabic{verse}}{וְכֵ֤ן תַּֽעֲשֶׂה֙ בְּשִׁבְעָ֣ה בַחֹ֔דֶשׁ מֵאִ֥ישׁ שֹׁגֶ֖ה וּמִפֶּ֑תִי וְכִפַּרְתֶּ֖ם אֶת־הַבָּֽיִת׃}{And so thou shalt do on the seventh day of the month for every one that erreth, and for him that is simple; so shall ye make atonement for the house.}
\twoverse{\arabic{verse}}{בָּ֠רִאשׁ֠וֹן בְּאַרְבָּעָ֨ה עָשָׂ֥ר יוֹם֙ לַחֹ֔דֶשׁ יִהְיֶ֥ה לָכֶ֖ם הַפָּ֑סַח חָ֕ג שְׁבֻע֣וֹת יָמִ֔ים מַצּ֖וֹת יֵאָכֵֽל׃}{In the first month, in the fourteenth day of the month, ye shall have the passover; a feast of seven days; unleavened bread shall be eaten.}
\twoverse{\arabic{verse}}{וְעָשָׂ֤ה הַנָּשִׂיא֙ בַּיּ֣וֹם הַה֔וּא בַּעֲד֕וֹ וּבְעַ֖ד כׇּל־עַ֣ם הָאָ֑רֶץ פַּ֖ר חַטָּֽאת׃}{And upon that day shall the prince prepare for himself and for all the people of the land a bullock for a sin-offering.}
\twoverse{\arabic{verse}}{וְשִׁבְעַ֨ת יְמֵי־הֶחָ֜ג יַעֲשֶׂ֧ה עוֹלָ֣ה לַיהֹוָ֗ה שִׁבְעַ֣ת פָּ֠רִ֠ים וְשִׁבְעַ֨ת אֵילִ֤ים תְּמִימִם֙ לַיּ֔וֹם שִׁבְעַ֖ת הַיָּמִ֑ים וְחַטָּ֕את שְׂעִ֥יר עִזִּ֖ים לַיּֽוֹם׃}{And the seven days of the feast he shall prepare a burnt-offering to the \lord, seven bullocks and seven rams without blemish daily the seven days; and a he-goat daily for a sin-offering.}
\twoverse{\arabic{verse}}{וּמִנְחָ֗ה אֵיפָ֥ה לַפָּ֛ר וְאֵיפָ֥ה לָאַ֖יִל יַעֲשֶׂ֑ה וְשֶׁ֖מֶן הִ֥ין לָאֵיפָֽה׃}{And he shall prepare a meal-offering, an ephah for a bullock, and an ephah for a ram, and a hin of oil to an ephah.}
\twoverse{\arabic{verse}}{בַּשְּׁבִיעִ֡י בַּֽחֲמִשָּׁה֩ עָשָׂ֨ר י֤וֹם לַחֹ֙דֶשׁ֙ בֶּחָ֔ג יַעֲשֶׂ֥ה כָאֵ֖לֶּה שִׁבְעַ֣ת הַיָּמִ֑ים כַּֽחַטָּאת֙ כָּעֹלָ֔ה וְכַמִּנְחָ֖ה וְכַשָּֽׁמֶן׃ \setuma }{In the seventh month, in the fifteenth day of the month, in the feast, shall he do the like the seven days; to the sin-offering as well as the burnt-offering, and the meal-offering as well as the oil.}
\newperek
\twoverse{\Roman{chap}:\arabic{verse}}{כֹּה־אָמַר֮ אֲדֹנָ֣י יֱהֹוִה֒ שַׁ֜עַר הֶחָצֵ֤ר הַפְּנִימִית֙ הַפֹּנֶ֣ה קָדִ֔ים יִהְיֶ֣ה סָג֔וּר שֵׁ֖שֶׁת יְמֵ֣י הַֽמַּעֲשֶׂ֑ה וּבְי֤וֹם הַשַּׁבָּת֙ יִפָּתֵ֔חַ וּבְי֥וֹם הַחֹ֖דֶשׁ יִפָּתֵֽחַ׃}{Thus saith the Lord \textsc{God}: The gate of the inner court that looketh toward the east shall be shut the six working days; but on the sabbath day it shall be opened, and in the day of the new moon it shall be opened.}
\twoverse{\arabic{verse}}{וּבָ֣א הַנָּשִׂ֡יא דֶּ֩רֶךְ֩ אוּלָ֨ם הַשַּׁ֜עַר מִח֗וּץ וְעָמַד֙ עַל־מְזוּזַ֣ת הַשַּׁ֔עַר וְעָשׂ֣וּ הַכֹּהֲנִ֗ים אֶת־עֽוֹלָתוֹ֙ וְאֶת־שְׁלָמָ֔יו וְהִֽשְׁתַּחֲוָ֛ה עַל־מִפְתַּ֥ן הַשַּׁ֖עַר וְיָצָ֑א וְהַשַּׁ֥עַר לֹא־יִסָּגֵ֖ר עַד־הָעָֽרֶב׃}{And the prince shall enter by the way of the porch of the gate without, and shall stand by the post of the gate, and the priests shall prepare his burnt-offering and his peace-offerings, and he shall worship at the threshold of the gate; then he shall go forth; but the gate shall not be shut until the evening.}
\twoverse{\arabic{verse}}{וְהִשְׁתַּחֲו֣וּ עַם־הָאָ֗רֶץ פֶּ֚תַח הַשַּׁ֣עַר הַה֔וּא בַּשַּׁבָּת֖וֹת וּבֶחֳדָשִׁ֑ים לִפְנֵ֖י יְהֹוָֽה׃}{Likewise the people of the land shall worship at the door of that gate before the \lord\space in the sabbaths and in the new moons.}
\twoverse{\arabic{verse}}{וְהָ֣עֹלָ֔ה אֲשֶׁר־יַקְרִ֥ב הַנָּשִׂ֖יא לַיהֹוָ֑ה בְּי֣וֹם הַשַּׁבָּ֗ת שִׁשָּׁ֧ה כְבָשִׂ֛ים תְּמִימִ֖ם וְאַ֥יִל תָּמִֽים׃}{And the burnt-offering that the prince shall offer unto the \lord\space shall be in the sabbath day six lambs without blemish and a ram without blemish;}
\twoverse{\arabic{verse}}{וּמִנְחָה֙ אֵיפָ֣ה לָאַ֔יִל וְלַכְּבָשִׂ֥ים מִנְחָ֖ה מַתַּ֣ת יָד֑וֹ וְשֶׁ֖מֶן הִ֥ין לָאֵיפָֽה׃ \setuma }{and the meal-offering shall be an ephah for the ram, and the meal-offering for the lambs as he is able to give, and a hin of oil to an ephah.}
\twoverse{\arabic{verse}}{וּבְי֣וֹם הַחֹ֔דֶשׁ פַּ֥ר בֶּן־בָּקָ֖ר תְּמִימִ֑ם וְשֵׁ֧שֶׁת כְּבָשִׂ֛ים וָאַ֖יִל תְּמִימִ֥ם יִהְיֽוּ׃}{And in the day of the new moon it shall be a young bullock without blemish; and six lambs, and a ram; they shall be without blemish;}
\twoverse{\arabic{verse}}{וְאֵיפָ֨ה לַפָּ֜ר וְאֵיפָ֤ה לָאַ֙יִל֙ יַעֲשֶׂ֣ה מִנְחָ֔ה וְלַ֨כְּבָשִׂ֔ים כַּאֲשֶׁ֥ר תַּשִּׂ֖יג יָד֑וֹ וְשֶׁ֖מֶן הִ֥ין לָאֵיפָֽה׃}{and he shall prepare a meal-offering, an ephah for the bullock, and an ephah for the ram, and for the lambs according as his means suffice, and a hin of oil to an ephah.}
\twoverse{\arabic{verse}}{וּבְב֖וֹא הַנָּשִׂ֑יא דֶּ֣רֶךְ אוּלָ֤ם הַשַּׁ֙עַר֙ יָב֔וֹא וּבְדַרְכּ֖וֹ יֵצֵֽא׃}{And when the prince shall enter, he shall go in by the way of the porch of the gate, and he shall go forth by the way thereof.}
\twoverse{\arabic{verse}}{וּבְב֨וֹא עַם־הָאָ֜רֶץ לִפְנֵ֣י יְהֹוָה֮ בַּמּוֹעֲדִים֒ הַבָּ֡א דֶּרֶךְ־שַׁ֨עַר צָפ֜וֹן לְהִֽשְׁתַּחֲוֺ֗ת יֵצֵא֙ דֶּרֶךְ־שַׁ֣עַר נֶ֔גֶב וְהַבָּא֙ דֶּרֶךְ־שַׁ֣עַר נֶ֔גֶב יֵצֵ֖א דֶּרֶךְ־שַׁ֣עַר צָפ֑וֹנָה לֹ֣א יָשׁ֗וּב דֶּ֤רֶךְ הַשַּׁ֙עַר֙ אֲשֶׁר־בָּ֣א ב֔וֹ כִּ֥י נִכְח֖וֹ \qk{יֵצֵֽא}{יצאו}׃}{But when the people of the land shall come before the \lord\space in the appointed seasons, he that entereth by the way of the north gate to worship shall go forth by the way of the south gate; and he that entereth by the way of the south gate shall go forth by the way of the north gate; he shall not return by the way of the gate whereby he came in, but shall go forth straight before him.}
\twoverse{\arabic{verse}}{וְֽהַנָּשִׂ֑יא בְּתוֹכָ֤ם בְּבוֹאָם֙ יָב֔וֹא וּבְצֵאתָ֖ם יֵצֵֽאוּ׃}{And the prince, when they go in, shall go in in the midst of them; and when they go forth, they shall go forth together.}
\twoverse{\arabic{verse}}{וּבַחַגִּ֣ים וּבַמּוֹעֲדִ֗ים תִּהְיֶ֤ה הַמִּנְחָה֙ אֵיפָ֤ה לַפָּר֙ וְאֵיפָ֣ה לָאַ֔יִל וְלַכְּבָשִׂ֖ים מַתַּ֣ת יָד֑וֹ וְשֶׁ֖מֶן הִ֥ין לָאֵיפָֽה׃ \petucha}{And in the feasts and in the appointed seasons the meal-offering shall be an ephah for a bullock, and an ephah for a ram, and for the lambs as he is able to give, and a hin of oil to an ephah.}
\twoverse{\arabic{verse}}{וְכִֽי־יַעֲשֶׂה֩ הַנָּשִׂ֨יא נְדָבָ֜ה עוֹלָ֣ה אוֹ־שְׁלָמִים֮ נְדָבָ֣ה לַיהֹוָה֒ וּפָ֣תַֽח ל֗וֹ אֶת־הַשַּׁ֙עַר֙ הַפֹּנֶ֣ה קָדִ֔ים וְעָשָׂ֤ה אֶת־עֹֽלָתוֹ֙ וְאֶת־שְׁלָמָ֔יו כַּאֲשֶׁ֥ר יַעֲשֶׂ֖ה בְּי֣וֹם הַשַּׁבָּ֑ת וְיָצָ֛א וְסָגַ֥ר אֶת־הַשַּׁ֖עַר אַחֲרֵ֥י צֵאתֽוֹ׃}{And when the prince shall prepare a freewill-offering, a burnt-offering or peace-offerings as a freewill-offering unto the \lord, one shall open for him the gate that looketh toward the east, and he shall prepare his burnt-offering and his peace-offerings, as he doth on the sabbath day; then he shall go forth; and after his going forth one shall shut the gate.}
\twoverse{\arabic{verse}}{וְכֶ֨בֶשׂ בֶּן־שְׁנָת֜וֹ תָּמִ֗ים תַּעֲשֶׂ֥ה עוֹלָ֛ה לַיּ֖וֹם לַיהֹוָ֑ה בַּבֹּ֥קֶר בַּבֹּ֖קֶר תַּעֲשֶׂ֥ה אֹתֽוֹ׃}{And thou shalt prepare a lamb of the first year without blemish for a burnt-offering unto the \lord\space daily; morning by morning shalt thou prepare it.}
\twoverse{\arabic{verse}}{וּמִנְחָה֩ תַעֲשֶׂ֨ה עָלָ֜יו בַּבֹּ֤קֶר בַּבֹּ֙קֶר֙ שִׁשִּׁ֣ית הָאֵיפָ֔ה וְשֶׁ֛מֶן שְׁלִישִׁ֥ית הַהִ֖ין לָרֹ֣ס אֶת־הַסֹּ֑לֶת מִנְחָה֙ לַיהֹוָ֔ה חֻקּ֥וֹת עוֹלָ֖ם תָּמִֽיד׃}{And thou shalt prepare a meal-offering with it morning by morning, the sixth part of an ephah, and the third part of a hin of oil, to moisten the fine flour: a meal-offering unto the \lord\space continually by a perpetual ordinance.}
\twoverse{\arabic{verse}}{\qk{יַעֲשׂ֨וּ}{ועשו} אֶת־הַכֶּ֧בֶשׂ וְאֶת־הַמִּנְחָ֛ה וְאֶת־הַשֶּׁ֖מֶן בַּבֹּ֣קֶר בַּבֹּ֑קֶר עוֹלַ֖ת תָּמִֽיד׃ \setuma }{Thus shall they prepare the lamb, and the meal-offering, and the oil, morning by morning, for a continual burnt-offering.}
\twoverse{\arabic{verse}}{כֹּה־אָמַ֞ר אֲדֹנָ֣י יֱהֹוִ֗ה כִּֽי־יִתֵּ֨ן הַנָּשִׂ֤יא מַתָּנָה֙ לְאִ֣ישׁ מִבָּנָ֔יו נַחֲלָת֥וֹ הִ֖יא לְבָנָ֣יו תִּֽהְיֶ֑ה אֲחֻזָּתָ֥ם הִ֖יא בְּנַחֲלָֽה׃ \setuma }{Thus saith the Lord \textsc{God}: If the prince give a gift unto any of his sons, it is his inheritance, it shall belong to his sons; it is their possession by inheritance.}
\twoverse{\arabic{verse}}{וְכִֽי־יִתֵּ֨ן מַתָּנָ֜ה מִנַּחֲלָת֗וֹ לְאַחַד֙ מֵֽעֲבָדָ֔יו וְהָ֤יְתָה לּוֹ֙ עַד־שְׁנַ֣ת הַדְּר֔וֹר וְשָׁבַ֖ת לַנָּשִׂ֑יא אַ֚ךְ נַחֲלָת֔וֹ בָּנָ֖יו לָהֶ֥ם תִּהְיֶֽה׃}{But if he give of his inheritance a gift to one of his servants, it shall be his to the year of liberty; then it shall return to the prince; but as for his inheritance, it shall be for his sons.}
\twoverse{\arabic{verse}}{וְלֹא־יִקַּ֨ח הַנָּשִׂ֜יא מִנַּחֲלַ֣ת הָעָ֗ם לְהֽוֹנֹתָם֙ מֵאֲחֻזָּתָ֔ם מֵאֲחֻזָּת֖וֹ יַנְחִ֣ל אֶת־בָּנָ֑יו לְמַ֙עַן֙ אֲשֶׁ֣ר לֹא־יָפֻ֣צוּ עַמִּ֔י אִ֖ישׁ מֵאֲחֻזָּתֽוֹ׃}{Moreover the prince shall not take of the people’s inheritance, to thrust them wrongfully out of their possession; he shall give inheritance to his sons out of his own possession; that My people be not scattered every man from his possession.’}
\newhaftarah{שבת הגדול}
\label{haft_hagadol}
\haftnote{The Haftarah for Shabbat HaGadol is Malachi 3:4\verserangechar 24.}
\setcounter{chap}{3}
\setcounter{verse}{4}
\twoverse{\Roman{chap}:\arabic{verse}}
{וְעָֽרְבָה֙ לַֽיהֹוָ֔ה מִנְחַ֥ת יְהוּדָ֖ה וִירֽוּשָׁלָ֑͏ִם כִּימֵ֣י עוֹלָ֔ם וּכְשָׁנִ֖ים קַדְמֹנִיֹּֽת׃}
{Then shall the offering of Judah and Jerusalem be pleasant unto the \lord, as in the days of old, and as in ancient years.}
\twoverse{\arabic{verse}}
{וְקָרַבְתִּ֣י אֲלֵיכֶם֮ לַמִּשְׁפָּט֒ וְהָיִ֣יתִי \legarmeh  עֵ֣ד מְמַהֵ֗ר בַּֽמְכַשְּׁפִים֙ וּבַמְנָ֣אֲפִ֔ים וּבַנִּשְׁבָּעִ֖ים לַשָּׁ֑קֶר וּבְעֹשְׁקֵ֣י שְׂכַר־שָׂ֠כִ֠יר אַלְמָנָ֨ה וְיָת֤וֹם וּמַטֵּי־גֵר֙ וְלֹ֣א יְרֵא֔וּנִי אָמַ֖ר יְהֹוָ֥ה צְבָאֽוֹת׃}
{And I will come near to you to judgment; and I will be a swift witness against the sorcerers, and against the adulterers, and against false swearers; and against those that oppress the hireling in his wages, The widow, and the fatherless, and that turn aside the stranger from his right, and fear not Me, Saith the \lord\space of hosts.}
\twoverse{\arabic{verse}}
{כִּ֛י אֲנִ֥י יְהֹוָ֖ה לֹ֣א שָׁנִ֑יתִי וְאַתֶּ֥ם בְּנֵֽי־יַעֲקֹ֖ב לֹ֥א כְלִיתֶֽם׃}
{For I the \lord\space change not; and ye, O sons of Jacob, are not consumed.}
\twoverse{\arabic{verse}}
{לְמִימֵ֨י אֲבֹֽתֵיכֶ֜ם סַרְתֶּ֤ם מֵֽחֻקַּי֙ וְלֹ֣א שְׁמַרְתֶּ֔ם שׁ֤וּבוּ אֵלַי֙ וְאָשׁ֣וּבָה אֲלֵיכֶ֔ם אָמַ֖ר יְהֹוָ֣ה צְבָא֑וֹת וַאֲמַרְתֶּ֖ם בַּמֶּ֥ה נָשֽׁוּב׃}
{From the days of your fathers ye have turned aside from Mine ordinances, and have not kept them. Return unto Me, and I will return unto you, Saith the \lord\space of hosts. But ye say: ‘Wherein shall we return?’}
\twoverse{\arabic{verse}}
{הֲיִקְבַּ֨ע אָדָ֜ם אֱלֹהִ֗ים כִּ֤י אַתֶּם֙ קֹֽבְעִ֣ים אֹתִ֔י וַאֲמַרְתֶּ֖ם בַּמֶּ֣ה קְבַֽעֲנ֑וּךָ הַֽמַּעֲשֵׂ֖ר וְהַתְּרוּמָֽה׃}
{Will a man rob God? Yet ye rob Me. But ye say: ‘Wherein have we robbed Thee?’ In tithes and heave-offerings.}
\twoverse{\arabic{verse}}
{בַּמְּאֵרָה֙ אַתֶּ֣ם נֵֽאָרִ֔ים וְאֹתִ֖י אַתֶּ֣ם קֹבְעִ֑ים הַגּ֖וֹי כֻּלּֽוֹ׃}
{Ye are cursed with the curse, Yet ye rob Me, Even this whole nation.}
\twoverse{\arabic{verse}}
{הָבִ֨יאוּ אֶת־כׇּל־הַֽמַּעֲשֵׂ֜ר אֶל־בֵּ֣ית הָאוֹצָ֗ר וִיהִ֥י טֶ֙רֶף֙ בְּבֵיתִ֔י וּבְחָנ֤וּנִי נָא֙ בָּזֹ֔את אָמַ֖ר יְהֹוָ֣ה צְבָא֑וֹת אִם־לֹ֧א אֶפְתַּ֣ח לָכֶ֗ם אֵ֚ת אֲרֻבּ֣וֹת הַשָּׁמַ֔יִם וַהֲרִיקֹתִ֥י לָכֶ֛ם בְּרָכָ֖ה עַד־בְּלִי־דָֽי׃}
{Bring ye the whole tithe into the store-house, that there may be food in My house, and try Me now herewith, Saith the \lord\space of hosts, if I will not open you the windows of heaven, and pour you out a blessing, that there shall be more than sufficiency.}
\twoverse{\arabic{verse}}
{וְגָעַרְתִּ֤י לָכֶם֙ בָּֽאֹכֵ֔ל וְלֹא־יַשְׁחִ֥ת לָכֶ֖ם אֶת־פְּרִ֣י הָאֲדָמָ֑ה וְלֹא־תְשַׁכֵּ֨ל לָכֶ֤ם הַגֶּ֙פֶן֙ בַּשָּׂדֶ֔ה אָמַ֖ר יְהֹוָ֥ה צְבָאֽוֹת׃}
{And I will rebuke the devourer for your good, and he shall not destroy the fruits of your land; Neither shall your vine cast its fruit before the time in the field, Saith the \lord\space of hosts.}
\twoverse{\arabic{verse}}
{וְאִשְּׁר֥וּ אֶתְכֶ֖ם כׇּל־הַגּוֹיִ֑ם כִּֽי־תִהְי֤וּ אַתֶּם֙ אֶ֣רֶץ חֵ֔פֶץ אָמַ֖ר יְהֹוָ֥ה צְבָאֽוֹת׃ \petucha }
{And all nations shall call you happy; For ye shall be a delightsome land, Saith the \lord\space of hosts.}
\twoverse{\arabic{verse}}
{חָזְק֥וּ עָלַ֛י דִּבְרֵיכֶ֖ם אָמַ֣ר יְהֹוָ֑ה וַאֲמַרְתֶּ֕ם מַה־נִּדְבַּ֖רְנוּ עָלֶֽיךָ׃}
{Your words have been all too strong against Me, Saith the \lord. Yet ye say: ‘Wherein have we spoken against thee?’}
\twoverse{\arabic{verse}}
{אֲמַרְתֶּ֕ם שָׁ֖וְא עֲבֹ֣ד אֱלֹהִ֑ים וּמַה־בֶּ֗צַע כִּ֤י שָׁמַ֙רְנוּ֙ מִשְׁמַרְתּ֔וֹ וְכִ֤י הָלַ֙כְנוּ֙ קְדֹ֣רַנִּ֔ית מִפְּנֵ֖י יְהֹוָ֥ה צְבָאֽוֹת׃}
{Ye have said: ‘It is vain to serve God; and what profit is it that we have kept His charge, and that we have walked mournfully because of the \lord\space of hosts?}
\twoverse{\arabic{verse}}
{וְעַתָּ֕ה אֲנַ֖חְנוּ מְאַשְּׁרִ֣ים זֵדִ֑ים גַּם־נִבְנוּ֙ עֹשֵׂ֣י רִשְׁעָ֔ה גַּ֧ם בָּחֲנ֛וּ אֱלֹהִ֖ים וַיִּמָּלֵֽטוּ׃}
{And now we call the proud happy; yea, they that work wickedness are built up; yea, they try God, and are delivered.’}
\twoverse{\arabic{verse}}
{אָ֧ז נִדְבְּר֛וּ יִרְאֵ֥י יְהֹוָ֖ה אִ֣ישׁ אֶל־רֵעֵ֑הוּ וַיַּקְשֵׁ֤ב יְהֹוָה֙ וַיִּשְׁמָ֔ע וַ֠יִּכָּתֵ֠ב סֵ֣פֶר זִכָּר֤וֹן לְפָנָיו֙ לְיִרְאֵ֣י יְהֹוָ֔ה וּלְחֹשְׁבֵ֖י שְׁמֽוֹ׃}
{Then they that feared the \lord\space Spoke one with another; and the \lord\space hearkened, and heard, and a book of remembrance was written before Him, for them that feared the \lord, and that thought upon His name.}
\twoverse{\arabic{verse}}
{וְהָ֣יוּ לִ֗י אָמַר֙ יְהֹוָ֣ה צְבָא֔וֹת לַיּ֕וֹם אֲשֶׁ֥ר אֲנִ֖י עֹשֶׂ֣ה סְגֻלָּ֑ה וְחָמַלְתִּ֣י עֲלֵיהֶ֔ם כַּֽאֲשֶׁר֙ יַחְמֹ֣ל אִ֔ישׁ עַל־בְּנ֖וֹ הָעֹבֵ֥ד אֹתֽוֹ׃}
{And they shall be Mine, saith the \lord\space of hosts, in the day that I do make, even Mine own treasure; and I will spare them, as a man spareth His own son that serveth him.}
\twoverse{\arabic{verse}}
{וְשַׁבְתֶּם֙ וּרְאִיתֶ֔ם בֵּ֥ין צַדִּ֖יק לְרָשָׁ֑ע בֵּ֚ין עֹבֵ֣ד אֱלֹהִ֔ים לַאֲשֶׁ֖ר לֹ֥א עֲבָדֽוֹ׃ \petucha }
{Then shall ye again discern between the righteous and the wicked, between him that serveth God and him that serveth Him not.}
\twoverse{\arabic{verse}}
{כִּֽי־הִנֵּ֤ה הַיּוֹם֙ בָּ֔א בֹּעֵ֖ר כַּתַּנּ֑וּר וְהָי֨וּ כׇל־זֵדִ֜ים וְכׇל־עֹשֵׂ֤ה רִשְׁעָה֙ קַ֔שׁ וְלִהַ֨ט אֹתָ֜ם הַיּ֣וֹם הַבָּ֗א אָמַר֙ יְהֹוָ֣ה צְבָא֔וֹת אֲשֶׁ֛ר לֹא־יַעֲזֹ֥ב לָהֶ֖ם שֹׁ֥רֶשׁ וְעָנָֽף׃}
{For, behold, the day cometh, It burneth as a furnace; and all the proud, and all that work wickedness, shall be stubble; and the day that cometh shall set them ablaze, Saith the \lord\space of hosts, that it shall leave them neither root nor branch.}
\twoverse{\arabic{verse}}
{וְזָרְחָ֨ה לָכֶ֜ם יִרְאֵ֤י שְׁמִי֙ שֶׁ֣מֶשׁ צְדָקָ֔ה וּמַרְפֵּ֖א בִּכְנָפֶ֑יהָ וִיצָאתֶ֥ם וּפִשְׁתֶּ֖ם כְּעֶגְלֵ֥י מַרְבֵּֽק׃}
{But unto you that fear My name shall the sun of righteousness arise with healing in its wings; and ye shall go forth, and gambol as calves of the stall.}
\twoverse{\arabic{verse}}
{וְעַסּוֹתֶ֣ם רְשָׁעִ֔ים כִּֽי־יִהְי֣וּ אֵ֔פֶר תַּ֖חַת כַּפּ֣וֹת רַגְלֵיכֶ֑ם בַּיּוֹם֙ אֲשֶׁ֣ר אֲנִ֣י עֹשֶׂ֔ה אָמַ֖ר יְהֹוָ֥ה צְבָאֽוֹת׃ \petucha }
{And ye shall tread down the wicked; for they shall be ashes under the soles of your feet in the day that I do make, Saith the \lord\space of hosts.}
\twoverse{\arabic{verse}}
{\large זִ\normalsize כְר֕וּ תּוֹרַ֖ת מֹשֶׁ֣ה עַבְדִּ֑י אֲשֶׁר֩ צִוִּ֨יתִי אוֹת֤וֹ בְחֹרֵב֙ עַל־כׇּל־יִשְׂרָאֵ֔ל חֻקִּ֖ים וּמִשְׁפָּטִֽים׃}
{Remember ye the law of Moses My servant, which I commanded unto him in Horeb for all Israel, even statutes and ordinances.}
\twoverse{\arabic{verse}}
{הִנֵּ֤ה אָנֹכִי֙ שֹׁלֵ֣חַ לָכֶ֔ם אֵ֖ת אֵלִיָּ֣ה הַנָּבִ֑יא לִפְנֵ֗י בּ֚וֹא י֣וֹם יְהֹוָ֔ה הַגָּד֖וֹל וְהַנּוֹרָֽא׃}
{Behold, I will send you Elijah the prophet Before the coming Of the great and terrible day of the \lord.}
\twoverse{\arabic{verse}}
{וְהֵשִׁ֤יב לֵב־אָבוֹת֙ עַל־בָּנִ֔ים וְלֵ֥ב בָּנִ֖ים עַל־אֲבוֹתָ֑ם פֶּן־אָב֕וֹא וְהִכֵּיתִ֥י אֶת־הָאָ֖רֶץ חֵֽרֶם׃\small הנה אנכי שולח לכם את אליה הנביא לפני בוא יום יהוה הגדול והנורא\normalsize }
{And he shall turn the heart of the fathers to the children, and the heart of the children to their fathers; lest I come and smite the land with utter destruction.\small [Behold, I will send you Elijah the prophet before the coming of the great and terrible day of the \lord.]\normalsize}
\end{document}
